Homeostatic Synaptic Plasticity of Miniature Excitatory Postsynaptic Currents in Mouse Cortical Cultures Requires Neuronal Rab3A

Version 1

Abstract

Abstract

Following prolonged activity blockade, amplitudes of miniature excitatory postsynaptic currents (mEPSCs) increase, a form of homeostatic plasticity termed “synaptic scaling.” We previously showed that a presynaptic protein, the small GTPase Rab3A, is required for full expression of the increase in miniature endplate current amplitudes following prolonged blockade of action potential activity at the mouse neuromuscular junction in vivo (Wang et al., 2011), but it is unknown whether this form of Rab3A-dependent homeostatic plasticity shares any characteristics with central synapses. We show here that synaptic scaling of mEPSCs is impaired in mouse cortical neuron cultures prepared from Rab3A-/- and Rab3A Earlybird mutant mice. To determine if Rab3A is involved in the well-established homeostatic increase in postsynaptic AMPA-type receptors (AMPARs), we performed a series of experiments in which electrophysiological recordings of mEPSCs and confocal imaging of synaptic AMPAR immunofluorescence were assessed within the same cultures. We found that Rab3A is required for the increase in synaptic AMPARs following prolonged activity blockade, but the comparison of mEPSC amplitude and synaptic AMPARs in the same cultures revealed that mEPSC amplitude cannot solely be determined by postsynaptic AMPAR levels. Finally, we demonstrate that Rab3A is acting in neurons because selective loss of Rab3A in astrocytes did not disrupt homeostatic plasticity, whereas selective loss in neurons strongly reduced the homeostatic increase in mEPSC amplitudes. Taken together with the results at the neuromuscular junction, we propose that Rab3A is a presynaptic homeostatic regulator that controls quantal size on both sides of the synapse.

Introduction

One of the most studied phenomena triggered by prolonged activity blockade is the increase in amplitudes of miniature excitatory postsynaptic currents (mEPSCs), first demonstrated in cultures of dissociated cortical neurons (Turrigiano et al., 1998) and spinal cord neurons (O’Brien et al., 1998). This compensatory response is now termed homeostatic synaptic plasticity (Turrigiano and Nelson, 2004; Pozo and Goda, 2010) and has been shown to be dysregulated in several neurodevelopment, psychiatric, and neurodegenerative disorders. One of the initial studies of homeostatic synaptic plasticity demonstrated that AMPAergic mEPSCs were uniformly increased after prolonged activity blockade (Turrigiano et al., 1998), and the authors named this process synaptic scaling. An accompanying increase in AMPA-type glutamate receptors (AMPARs) was observed in both of the early studies of homeostatic synaptic plasticity (O’Brien et al., 1998; Turrigiano et al., 1998), and has been confirmed many times (a non-exhaustive list includes (Ju et al., 2004; Thiagarajan et al., 2005; Shepherd et al., 2006; Stellwagen and Malenka, 2006; Hou et al., 2008; Gainey et al., 2009; Soden and Chen, 2010; Correa et al., 2012; Altimimi and Stellwagen, 2013; Letellier et al., 2014; Xu and Pozzo- Miller, 2017).

Homeostatic synaptic plasticity is becoming increasingly implicated in both pathological brain conditions, for example, epilepsy, neuropsychiatric disorders, Huntington’s, and alcohol use disorder (Trasande and Ramirez, 2007; Fernandes and Carvalho, 2016; Wang et al., 2017; Lovinger and Abrahao, 2018; Smith-Dijak et al., 2019; Lignani et al., 2020; Suzuki et al., 2021; Kavalali and Monteggia, 2023), and essential normal functions such as sleep (Tononi and Cirelli, 2014; Diering et al., 2017; Torrado Pacheco et al., 2021), so a molecular understanding of the process is an extremely important next step. The homeostatic synaptic plasticity field has identified several proteins required for synaptic scaling of mEPSC amplitudes, the majority of which are involved in the regulation of AMPAR levels (Shepherd et al., 2006; Seeburg and Sheng, 2008; Gao et al., 2010; Anggono et al., 2011; Beique et al., 2011; Diering et al., 2014; Gainey et al., 2015; Tan et al., 2015; Pastuzyn and Shepherd, 2017; Sanderson et al., 2018).

In our previous work studying the increase in miniature endplate currents (mEPC) following prolonged in vivo activity-block of synaptic transmission at the mouse neuromuscular junction (NMJ), we were surprised to find no evidence of changes in acetylcholine receptor (AChR) levels (Wang et al., 2005). This result led us to search for presynaptic molecules that might homeostatically regulate mEPC amplitude, perhaps via the presynaptic quantum. In previous studies in chromaffin cells, we identified the small GTPase Rab3A, a synaptic vesicle protein, as a regulator of synaptic vesicle fusion pore opening (Wang et al., 2008), so we examined whether deletion of Rab3A (Rab3A-/-) might prevent homeostatic upregulation of mEPC amplitude. The results were clear: in the Rab3A-/- mouse, the homeostatic increase in mEPC amplitude was strongly reduced, and was completely abolished in mice expressing a point mutation in Rab3A, the Earlybird mutant (Rab3AEbd/Ebd) (Wang et al., 2011).

The Rab3A-/- mouse has minimal phenotypic abnormalities, with evoked synaptic transmission and mEPSCs essentially normal in hippocampal slices (Geppert et al., 1994). At the Rab3A-/-NMJ, reductions in evoked transmission were detected, but only under conditions of reduced extracellular calcium (Coleman et al., 2007). The most dramatic effect of loss of Rab3A is the disruption of a presynaptic form of long-term potentiation (LTP) at the mossy fiber-CA3 synapse (Weisskopf et al., 1994; Castillo et al., 1997). To our knowledge, there is no evidence of Rab3A involvement in the expression or trafficking of postsynaptic receptors. Our results at the mammalian NMJ suggest that in addition to its importance in mossy fiber LTP, Rab3A may be required for the homeostatic plasticity of mEPSC amplitude via a presynaptic mechanism.

In the current work, we explored whether the in vivo findings at the NMJ might also apply to the more typically studied homeostatic synaptic plasticity of mEPSC amplitude in dissociated cortical neuron cultures. We report that our findings in vivo at the NMJ were almost exactly recapitulated in cultures of dissociated cortical neurons: 1. strong reduction of the homeostatic increase in mEPSC amplitude in the absence of Rab3A; 2. complete abolishment of the homeostatic increase in mEPSC amplitude in the presence of the Rab3A Earlybird mutant; and 3. increased mEPSC amplitude in the Rab3A Earlybird mutant prior to activity blockade.

However, in contrast to the unchanged AChR levels at the mammalian NMJ, there was a modest increase in levels of the GluA2-type AMPARs at cortical synapses after activity- blockade, which appeared to be disrupted in cultures prepared from Rab3A-/-mice.

Importantly, when compared within the same cultures, GluA2 receptor levels did not always parallel mEPSC amplitudes. We also determined that Rab3A must be present in neurons, but not astrocytes, for full expression of homeostatic plasticity. The NMJ and cortical culture data taken together strongly suggest neuronal Rab3A is important for both postsynaptic receptor upregulation and a second mechanism affecting mEPSC amplitude, likely the amount of transmitter released by a single vesicle.

Materials and Methods

Animals

Rab3A+/- heterozygous mice were bred and genotyped as previously described (Kapfhamer et al., 2002; Wang et al., 2008). Rab3AEbd/Ebdmice were identified in an EU- mutagenesis screen of C57BL/6J mice, and after a cross to C3H/HeJ, were backcrossed for 3 generations to C57BL/6J (Kapfhamer et al., 2002). Rab3A+/Ebd heterozygous mice were bred at Wright State University and genotyped in a two-step procedure: 1. a PCR reaction with RabF1 and Dcaps3R as primers; and 2. a digestion with enzyme Bsp1286I (New England Biolabs) that distinguishes the Earlybird mutant by its different base-pair products. Rab3A+/- mice were backcrossed with Rab3A+/+ mice from the Earlybird heterozygous colony for 11 generations in an attempt to establish a single wild type strain, but differences in mEPSC amplitude and adrenal chromaffin cell calcium currents persisted, likely due to genes that are close to the Rab3A site, resulting in two wild type strains: 1. Rab3A+/+ from the Rab3A+/- colony, and 2. Rab3A+/+ from the Rab3A+/Ebd colony.

Primary Culture of Mouse Cortical Neurons

Primary dissociated cultures of mixed neuronal and astrocyte populations were prepared as previously described (Hanes et al., 2020). Briefly, postnatal day 0-2 (P0-P2) Rab3A+/+, Rab3A-/- or Rab3AEbd/Ebd neonates were euthanized by rapid decapitation, as approved by the Wright State University Institutional Animal Care and Use Committee, and brains were quickly removed. Each culture was prepared from the cortices harvested from two animals; neonates were not sexed. Cortices were collected in chilled Neurobasal-A media (Gibco) with osmolarity adjusted to 270 mOsm and supplemented with 40 U/ml DNAse I (ThermoFisher Scientific). The tissues were digested with papain (Worthington Biochemical) at 20 U/ml at 37°C for 20 minutes followed by trituration with a sterile, fire- polished Pasteur pipette, then filtered through a 100 μm cell strainer, and centrifuged at 1100 rpm for 2 minutes. After discarding the supernatant, the pellet was resuspended in room temperature Neurobasal-A media (270 mOsm), supplemented with 5% fetal bovine serum for astrocyte growth, and 2% B-27 supplement to promote neuronal growth (Gibco), L-glutamine, and gentamicin (ThermoFisher Scientific). Neurons were counted and plated at 0.15 * 106 cells/coverslip onto 12 mm coverslips pre-coated with poly-L-lysine (BioCoat, Corning). The culture media for the first day (0 DIV) was the same as the above Neurobasal-A media supplemented with FBS, B-27, L-glutamine, and gentamicin, and was switched after 24 hours (1 DIV) to media consisting of Neurobasal-A (270 mOsm), 2% B-27, and L-glutamine without FBS to avoid its toxic effects on neuronal viability and health (Stellwagen and Malenka, 2006). Half of the media was changed twice weekly and experiments were performed at 13-14 DIV. Two days prior to experiments, tetrodotoxin (TTX) (500 nM; Tocris), a potent Na+ channel blocker, was added to some cultures to chronically silence all network activity and induce homeostatic synaptic plasticity mechanisms, while untreated sister cultures served as controls. Cultures prepared from mutant mice were compared with cultures from wild-type mice from their respective colonies. Note that the cultures comprising the Rab3A+/+ data here are a subset of the data previously published in Hanes et al., 2020, and therefore the plots in Figure 1 are not identical to those in the previously published work. This smaller data set was restricted to the time period over which cultures were prepared from Rab3A-/- mice.

Block of activity with 48 hours TTX treatment increased mEPSC amplitudes in cortical cultures from wild type mice but not Rab3A-/- mice. A) 10 second current traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/- colony. B) Average traces for the recordings shown in (A). C) Box and whisker plots for average mEPSC amplitudes from 30 CON and 23 TTX cells (Rab3A+/+, Rab3A+/- colony; mean mEPSC amplitudes, CON, 13.9 ± 0.7 pA; TTX, 18.2 ± 0.9 pA, p = 4.58 * 10-4, Kruskal-Wallis test). D) Current traces recorded in pyramidal cortical neurons from a CON and TTX-treated neuron in cultures prepared from Rab3A-/- mice. E) Average traces for the recordings shown in (D). F) Box and whisker plots for average mEPSC amplitudes from 25 CON and 26 TTX cells, Rab3A-/- data (Rab3A-/-, mean mEPSC amplitudes, CON, 13.6 ± 0.1; TTX, 14.3 ± 0.6, p = 0.318, Kruskal-Wallis test). G) Analysis of pooled mEPSC amplitudes for cultures from Rab3A+/+ mice of the Rab3A+/- colony (i) Cumulative distribution functions (CDFs) of quantiles of mEPSC amplitudes pooled from CON (black curve) and TTX- treated cultures (red curve) (Kolmogorov Smirnov (KS) test, test statistic (D) = 0.149, p = 8.53 * 10-13. (ii) Quantiles pooled from CON and TTX-treated cells were ranked from smallest to largest, plotted against each other, and fit with linear regression functions. Gray line, identity; red dashed line, linear regression fit. Inset, full range of the data. (iii) The same ranked mEPSC quantiles were paired and the ratio of CON/TTX computed for each pair, then plotted as a function of the CON mEPSC for the pair. Dashed line at ratio = 1.0 (no plasticity). H) Analysis of pooled mEPSCs amplitudes for cultures from Rab3A-/- mice. (i) CDFs of quantiles pooled from CON (black curve) and TTX-treated cultures (red curve) (KS test, D = 0.070, p = 0.042). (ii) Quantiles pooled from CON and TTX-treated cells were ranked from smallest to largest, plotted against each other, and fit with a linear regression function. Gray lines, identity; red dashed lines, linear regression fits. Inset: full range of the data. (iii) The same ranked mEPSC quantiles were paired and the ratio of CON/TTX computed for each pair, then plotted as a function of the CON mEPSC for the pair. Dashed line at ratio = 1 (no plasticity).

Preparation of Astrocyte Feeder Layers

Astrocyte feeder layers were prepared from the cortices of P0-P2 Rab3A+/+ or Rab3A-/- mouse pups as described previously (Stellwagen and Malenka, 2006). Briefly, cortices were dissected and cells were dissociated as described above. Cell suspensions of mixed neuronal and astrocyte populations were plated onto glass coverslips pre-coated with poly-L-lysine in Dulbecco’s Modified Eagle Media (ThermoFisher Scientific) supplemented with 5% FBS (to promote astrocyte proliferation and to kill neurons), L- glutamine, and gentamicin, and maintained in an incubator at 37°C, 5% CO2; cultures were maintained in this manner for up to 1 month to generate purely astrocytic cultures (all neurons typically died off by 7 DIV). Culture media was replaced after 24 hours, and subsequent media changes were made twice weekly, replacing half of the culture media with fresh media. Feeder layers were not used for neuronal seeding until all native neurons were gone and astrocytes approached 100% confluency (visually inspected).

Plating of Neurons on Glial Feeder Layers

Cortical neurons were obtained as described above. The cell pellet obtained was resuspended in Neurobasal-A (osmolarity adjusted to 270 mOsm) containing B27 (2%, to promote neuronal growth), L-glutamine, and 5-fluorodeoxyuridine (FdU, a mitotic inhibitor; Sigma). Addition of FdU was used to prevent astrocyte proliferation and contamination of the feeder layer with new astrocytes, promoting only neuronal growth on the feeder layers (FdU-containing media was used for the maintenance of these cultures and all subsequent media changes). Astrocyte culture media was removed from the feeder layer cultures, and the neuronal cell suspension was plated onto the astrocyte feeder cultures. The culture strategy used to distinguish the relative roles of neuronal and astrocytic Rab3A is outlined in Figure 9. At 1 DIV, all of the culture media was removed and replaced with fresh Neurobasal-A media containing FdU described above, and half of the media was replaced twice per week for all subsequent media changes. Cultures were maintained in a 37°C, 5% CO2 incubator for 13-14 DIV.

Whole-Cell Voltage Clamp to Record mEPSCs

At 13-14 DIV, mEPSCs from TTX-treated and untreated sister cultures of Rab3A+/+ or Rab3A-/- neurons from the Rab3A+/- colony, or Rab3A+/+ or Rab3AEbd/Ebd neurons from the Rab3A+/Ebd colony, were recorded via whole-cell voltage clamp to assess the role of Rab3A in homeostatic synaptic plasticity. Recordings were taken from pyramidal neurons, which were identified visually by a prominent apical dendrite; images were taken of all cells recorded from. Cells were continuously perfused with a solution consisting of (in mM): NaCl (115), KCl (5), CaCl2 (2.5), MgCl2 (1.3), dextrose (23), sucrose (26), HEPES (4.2), pH = 7.2 (Stellwagen and Malenka, 2006). On the day of recording, the osmolarity of the media from the cultures was measured (normally 285 – 295 mOsm) and the perfusate osmolarity was adjusted to match the culture osmolarity, to protect against osmotic shock to the neurons. To isolate glutamatergic mEPSCs, TTX (500 nM) and picrotoxin (50 μM) were included in the perfusion solution to block action potentials and GABAergic currents, respectively. The NMDA receptor antagonist, APV, was not included in the perfusion solution because all mEPSCs were blocked by CNQX and picrotoxin, demonstrating no APV-sensitive mEPSCs were present (data not shown). Patch electrodes (3.5 – 5 MΩ) were filled with an internal solution containing (in mM): K-gluconate (128), NaCl (10), EGTA (1), CaCl2 (0.132), MgCl2 (2), HEPES (10), pH = 7.2.

Osmolarity was adjusted to 10 mOsm less than the perfusion solution osmolarity. Neurons were clamped at a voltage of -60 mV using an Axopatch 200B patch-clamp (Axon Instruments), recorded from for 2 – 5 minutes, and data were collected with Clampex 10.0/10.2 (Axon Instruments). The antagonist of Ca2+-permeable AMPA receptors (including GluA1 but not GluA2), N-naphthyl acetylspermine (NASPM, 20 μM; Tocris), was applied during recordings in a subset of experiments. Because NASPM is an open channel blocker, it was applied with a depolarizing high K+ solution (25 mM KCl, 95 mM NaCl). Baseline recordings were performed for 2 minutes in our standard perfusate, then were suspended while NASPM + High K+ solution was applied for 45 seconds, followed by a NASPM only solution for 5 minutes, after which recording was recommenced for 5 minutes (because we found in pilot experiments that frequency was reduced following NASPM application).

Data Analysis and Statistics of mEPSCs

Miniature excitatory postsynaptic currents were manually selected using Mini Analysis software (Synaptosoft) to identify mEPSCs. The program threshold was set at 3 pA but the smallest mEPSC selected was 4.01 pA. Records were filtered at 2 kHz using a low-pass Butterworth filter prior to selection. For computing mEPSC mean in each experimental condition, individual cell means were pooled across multiple cultures and compared with the non-parametric Kruskal-Wallis test, with n = the number of cells and the overall means presented as ± SEM. For cumulative probability distribution functions (CDFs) of mEPSC amplitude, 30 quantiles were computed for each cell and pooled across cultures (see (Hanes et al., 2020)), and a Kolmogorov-Smirnov test (KS test) was used to test for significant differences, with n = the number of mEPSC quantiles. The rank order plots were created by computing a matched number of quantiles from the two experimental conditions (usually control (CON) and TTX). We used an algorithm to identify the product closest to, but above, 24 quantiles for a data set. For example, if there were 13 CON cells and 12 TTX cells, we computed 24 quantiles for each of the 13 cells and 26 quantiles for each of the 12 cells, for an equal total number of 312 quantiles for both data sets. The quantiles were sorted from smallest to largest, TTX amplitudes plotted vs. CON amplitudes, and the relationship fit with a linear regression function with the intercept term allowed to vary. The ratio plots were created by taking the ratio of TTX/CON and plotting as a function of the CON amplitude quantiles.

Immunocytochemistry, microscopy, and data analysis

Primary cultures of mouse cortical neurons were grown for 13-14 DIV. Antibodies to GluA2 (mouse ab against N-terminal, EMD Millipore) were added directly to live cultures at 1:40 dilution, and incubated at 37 °C in a CO2 incubator for 45 minutes. Cultures were rinsed 3 times with PBS/5% donkey serum before being fixed with 4% paraformaldehyde. After 3 rinses in PBS/5% donkey serum, cultures were incubated in CY3-labeled donkey-anti-mouse secondary antibodies for 1 hour at room temperature, rinsed in PBS/5% donkey serum, permeabilized with 0.2% saponin, and incubated in chick anti-MAP2 (1:2500, AbCAM) and rabbit anti-VGLUT1 (1:4000, Synaptic Systems) for 1 hour at room temperature in PBS/5% donkey serum. After rinsing with PBS/5% donkey serum, coverslips were incubated with 488-anti chick and CY5-anti rabbit secondary antibodies for 1 hour at room temperature, rinsed, blotted to remove excess liquid, and inverted on a drop of Vectashield (Vector Labs). Coverslips were sealed with nail polish and stored at 4 °C for < 1 week before imaging. All secondary antibodies were from Jackson Immunoresearch and were used at 1:225 dilution.

Coverslips were viewed on a Fluoview FV1000 laser scanning confocal microscope with a 60x oil immersion, 1.35 NA objective. Once a pyramidal neuron was identified, Fluoview 2.1 software was used to zoom in on the primary dendrite (5X) and confocal sections were taken every 0.5 μm. Images were analyzed offline with ImagePro 6 (Cybernetics). The composite image was used to locate synaptic sites containing both VGLUT1 and GluA2 immunoreactivity in close apposition to each other and to the primary dendrite or a secondary branch. An area of interest (AOI) was manually drawn around the receptor cluster in the confocal section in which it was the brightest. The AOIs for a dendrite were saved in a single file; the AOI number and the confocal section it was associated with were noted for later retrieval. For quantification, AOIs were loaded, an individual AOI was called up on the appropriate section, and the count/measurement tool used to apply a threshold (400 or 450; identical for CON and TTX coverslips but different for different cultures); pixels within the cluster that were above the threshold were automatically outlined, and size, average intensity, and integral of the outlined region reported. For quantile sampling, only dendrites with a minimum of 6 synaptic sites were included. For creating CDFs, 30 quantiles were computed for each dendrite. For the rank order and ratio plots, we matched the total quantiles for CON and TTX GluA2 receptor characteristics as was done above for mEPSC amplitudes.

Results

We previously reported that mixed cultures of cortical neurons and astrocytes prepared from postnatal day 0-2 mouse pups responded to a block of action potential-mediated activity by a 48 hour TTX treatment with an increase in mEPSC amplitude (Hanes et al., 2020). Here, we asked the question whether cortical cultures prepared from mice lacking the small GTPase Rab3A, or, expressing a point mutation of Rab3A, Rab3A Earlybird, have an altered homeostatic plasticity response to a loss of network activity. To obtain Rab3A-/- and Rab3AEbd/Ebdhomozygotes, we established two mouse colonies of heterozygous breeders with cultures prepared from pups derived from a final breeding pair of homozygotes. Although we backcrossed Rab3A+/- with Rab3A+/Ebd for 11 generations, clear differences in mEPSC amplitudes in untreated cultures (see below) and in calcium current amplitudes in adrenal chromaffin cells (unpublished obs.) remained. Therefore, throughout this study we keep the two strains separate and there are two Rab3A+/+ or ‘wild type’ phenotypes.

Example current traces of spontaneously occurring mEPSCs recorded from pyramidal neurons in untreated (CON) 13-14 DIV cortical cultures and sister cultures treated with 500 nM TTX for 48 hours prepared from wild type animals in the Rab3A+/- colony are shown in Figure 1A. Average mEPSC waveforms from the same recordings are shown in Figure 1B. The mean mEPSC amplitudes for 30 control and 23 TTX-treated neurons are displayed in the box and whisker plot in Figure 1C; after activity blockade the average mEPSC amplitude increased from 13.9 ± 0.7 pA to 18.2 ± 0.9 pA (p = 4.58 * 10-4, Kruskal-Wallis test). Example current traces of mEPSCs and average mEPSC waveforms are shown in Figures 1D and E, respectively, for cortical cultures prepared from Rab3A-/- mice. In contrast to the behavior of neurons in cultures prepared from the wild type strain, the increase in mEPSC amplitudes in Rab3A-/- neurons after activity blockade was dramatically reduced, and the average mEPSC amplitude was not significantly increased, for 25 untreated cells and 26 TTX-treated cells (Figure 1F, 13.6 ± 0.1 vs. 14.3 ± 0.6, p = 0.318, Kruskal-Wallis test). To further examine homeostatic plasticity in the presence and absence of Rab3A, we computed 30 quantiles for the mEPSC amplitude distribution of each neuron (Hanes et al., 2020), pooled the quantiles, and plotted the data as cumulative distribution functions (CDFs) for CON and TTX. For mEPSC amplitudes from cultures prepared from Rab3A+/+ mice, the difference between CDFs was highly significant (Figure 1Gi, test statistic D = 0.172, p = 1.62 * 10-10, Kolmogorov Smirnov (KS) test). In contrast, the increase in mEPSC amplitude after activity blockade was dramatically reduced in cortical cultures from Rab3A-/- mice (Figure 1Hi). A KS test for the CDFs was significant, but the test statistic much smaller (0.070) and the p value much larger (0.042) than for the wild type strain.

It was originally proposed that loss of activity produced a uniform multiplicative increase in mEPSC amplitude across the entire distribution of mEPSCs (Turrigiano et al., 1998). This uniform scaling would be expected to preserve original differences in synaptic weights resulting from other forms of plasticity (Turrigiano, 1999; Turrigiano and Nelson, 2004). In addition, uniform scaling suggested a cell-wide mechanism (in fact, a culture-wide mechanism) to identically modify all synapses. However, we recently showed that in mouse cortical neurons, rat cortical neurons, and mouse hippocampal neurons, scaling is non-uniform, a phenomenon we called “divergent scaling;” the multiplicative factor is smallest (close to 1) for small mEPSC amplitudes and increases to ∼1.4 for larger mEPSC amplitudes (Hanes et al., 2020). In subsequent reviewing of the current literature (Koesters et al., 2022), we concluded that the majority of previous studies of homeostatic plasticity of mEPSC amplitude following activity blockade also show divergent scaling, most notably the study that originally defined synaptic scaling (Turrigiano et al., 1998). Divergent scaling is not obvious on the standard plot used to demonstrate synaptic scaling, the rank-order plot (Turrigiano et al., 1998), but is only apparent once the ratio of TTX mEPSC amplitude/CON mEPSC amplitude is computed and plotted as a function of CON mEPSC amplitude (Hanes et al., 2020). To compare between uniform and divergent scaling in the presence and absence of Rab3A, we created both rank order plots and ratio plots for Rab3A+/+ data. The rank order plot for Rab3A+/+ neurons (Figure 1Gii) shows that treatment with TTX appears to cause a uniform multiplication of control mEPSC amplitudes based on the linear fit to the data, which had a slope of 1.43 (R2 = 0.982). However, in the ratio plot for mEPSCs in Rab3A+/+ cultures, the scaling was clearly divergent with the smallest mEPSC amplitudes having the smallest ratios (at 1) and the ratio increasing with mEPSC amplitude to a maximum of 1.51 (Figure 1Giii). In the absence of Rab3A, the rank ordered mEPSC data follow the line of identity and the slope of the linear fit was 0.97 (R2 = 0.985), suggesting complete absence of homeostatic plasticity (Figure 1Hii). However, the more sensitive ratio plot reveals a residual divergent scaling to a peak ratio of 1.13 (Figure 1Hiii). We previously observed a residual homeostatic effect of activity blockade on mEPCs at the mouse NMJ in the absence of Rab3A (Wang et al., 2011) suggesting that there may be compensatory or redundant mechanisms after deletion of Rab3A.

We next examined whether expression of a single point mutant of Rab3A (Rab3AEbd/Ebd) would abolish synaptic scaling in the mouse cortical cultures as we previously showed for mEPCs at the mouse NMJ (Wang et al., 2011). Example current traces of mEPSCs, average mEPSC waveforms, and box and whisker plots are shown in Figures 2A and B, respectively, for cortical cultures prepared from wild type mice in the Rab3A+/Ebd colony. Treatment with TTX for 48 hours leads to a significant increase in the average mEPSC amplitude of 23 TTX-treated cells when compared to 20 untreated cells (Figure 2C, CON, 11.0 ± 0.6 pA; TTX, 15.0 ± 1.3 pA, p = 0.02, Kruskal-Wallis test). We note here that while the two strains respond very similarly to activity blockade, the mean mEPSC amplitude in untreated cultures was significantly different in the two wild type strains, 13.9 ± 0.7 pA, wild type from Rab3A+/- colony; 11.0 ± 0.6 pA, wild type from Rab3A+/Ebd colony (p = 0.004, Kruskal-Wallis test).

Block of activity with 48 hours TTX treatment increased mEPSC amplitudes in cortical cultures from wild type mice but not Rab3AEbd/Ebd mice, which had an increased mEPSC amplitude at baseline. A) 10 second current traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/Ebd colony. B) Average traces for the recordings shown in (A). C) Box and whisker plots for average mEPSC amplitudes from 20 CON and 23 TTX cells (Rab3A+/+, Rab3A+/Ebd colony, mean mEPSCs amplitudes, CON, 11.0 ± 0.6 pA; TTX, 15.0 ± 1.3 pA, p = 0.02, Kruskal-Wallis test). D) Current traces recorded in pyramidal-shaped cortical neurons from a CON and TTX-treated neuron in cultures prepared from Rab3AEbd/Ebd mice. E) Average traces for the recordings shown in (D). F) Box and whisker plots of average mEPSC amplitudes from 21 CON and 22 TTX cells (Rab3AEbd/Ebd, mean mEPSC amplitudes, CON, 15.1 ± 1.0 pA; TTX, 14.6 ± 1.1 pA, p = 0.81, Kruskal-Wallis test). G) Analysis of pooled mEPSC amplitudes for cultures from Rab3A+/+ mice of the Rab3A+/Ebd colony. (i) CDFs of quantiles pooled from CON (black curve) and TTX-treated cultures (red curve) (KS test, D = 0.177, p value = 2.88 * 10-9). (ii) Quantiles pooled from CON and TTX-treated cells were ranked from smallest to largest, plotted against each other, and fit with linear regression function. Gray line, identity; red dashed line, linear regression fit. Inset: full range of the data. (iii) The same ranked mEPSC quantiles were paired and the ratio of CON/TTX computed for each pair, then plotted as a function of the CON mEPSC for the pair. Dashed line at ratio = 1 (no plasticity). H) Analysis of pooled mEPSC amplitudes for cultures from Rab3AEbd/Ebdmice. (i) CDFs of quantiles pooled from CON (black curve) and TTX- treated cultures (red curve). (KS test, D = 0.067, p value = 0.097 (TTX mEPSCs were slightly reduced)). (ii) Quantiles pooled from CON and TTX-treated cells were ranked from smallest to largest, plotted against each other, and fit with a linear regression function. Gray line, identity; red dashed line, linear regression fit. Inset: full range of the data. (iii) The same ranked mEPSC quantiles were paired and the ratio of CON/TTX computed for each pair, then plotted as a function of the CON mEPSC for the pair. Dashed line at ratio = 1 (no plasticity).

We found a complete disruption of homeostatic plasticity in cortical cultures prepared from Rab3AEbd/Ebd mice, as can been seen in viewing example mEPSC traces and average mEPSC waveforms (Figures 2D and E, respectively). The lack of TTX effect was confirmed in a comparison of mEPSC amplitude means for 21 untreated and 22 TTX-treated cells (Figure 2F, CON, 15.1 ± 1.0 pA vs. TTX, 14.6 ± 1.1 pA, p = 0.81, Kruskal-Wallis test). For cultures prepared from the wild type strain of the Earlybird heterozygote colony, the CDFs of the pooled quantiles from control and TTX-treated cells were significantly different (Figure 2Gi, D = 0.177, p value = 2.88 * 10-9, KS test), but for cultures prepared from Rab3AEbd/Ebd mice, the CDF of mEPSC amplitudes from TTX treated cells was shifted to the left, indicating a slight reduction after activity blockade, which did not reach statistical significance (Figure 2Hi; D = 0.067, p = 0.097, KS test). The rank ordered data indicate that the increased slope after TTX observed in the wildtype data (Figure 2Gii, slope 1.73, R2 = 0.993) was completely abolished in the Earlybird data (Figure 2Hii, slope 1.03, R2 = 0.989). Finally, in the wild type data, the ratio of TTX mEPSC amplitude/CON mEPSC amplitude increases from 1.08 for the smallest amplitudes to over 1.60 for the largest amplitudes (Figure 2Giii), whereas for mEPSC amplitudes from Rab3AEbd/Ebdcultures, the ratio of TTX/CON actually falls below 1 (minimum 0.9, Figure 2 Hiii).

Our results show that the homeostatic increase of mEPSC amplitude after activity blockade is disrupted in both Rab3A-/- and the Rab3AEbd/Ebd cortical neurons, strongly supporting a crucial role of functioning Rab3A in the synaptic scaling process. However, it is important to note that the disruption differs for Rab3A-/- and Rab3AEbd/Ebd. In the Rab3A-/- data set, mEPSCs from untreated cultures were indistinguishable from mEPSCs from Rab3A+/+ untreated cultures, demonstrating loss of Rab3A has no impact on basal activity mEPSC amplitudes, but the increase in mEPSC amplitudes after activity blockade was strongly diminished in Rab3A-/-neurons. In the Rab3AEbd/Ebd data set, mEPSC amplitudes from untreated cultures were significantly larger than those of untreated cultures from wild type mice, as can be seen in Figure 3 for the CDFs (Figure 3A, D = 0.245; p = 1.52 * 10-16, KS test) and the box and whisker plot (Figure 3A Inset, Rab3A+/+, 11.0 ± 0.7 pA, vs. Rab3AEbd/Ebd, 15.1 ± 1.0 pA p = 0.0027). The linear fit of the rank ordered data of Rab3AEbd/Ebdvs. wildtype mEPSCs had a slope of 1.63 (Figure 3B). In the ratio plot (Figure 3C), scaling was more uniform than in any other data set we have examined so far, with values starting at 1.4, declining to 1.2, then rising again to 1.4 for the majority of the data, instead of beginning near 1 and rising to 1.5, as happened for homeostatic plasticity (Figure 1Giii). Given this distinct behavior, it cannot be stated that the presence of mutated Rab3A causes the identical effect on mEPSC amplitudes as that of activity blockade. It may be that the much longer time in the presence of the Rab3A Earlybird mutant compared to the 48 hour TTX treatment leads to a stable increase in the smallest mEPSCs.

The increase in mEPSC amplitudes in neurons from cultures prepared from Rab3AEbd/Ebd mice, relative to those from Rab3A+/+ mice in the same strain, showed less divergent scaling than that of the activity-block induced increase in mEPSC amplitudes typically observed in cultures from Rab3A+/+ mice. A) Cumulative distribution function of quantiles pooled from Rab3A+/+ (black curve) and Rab3AEbd/Ebd cultures (red curve), D = 0.245; p = 1.52 * 10-16, KS test. Inset: box and whisker plot of average mEPSC amplitudes from 20 Rab3A+/+ and 21 Rab3AEbd/Ebdcells, mean mEPSC amplitudes, Rab3A+/+, 11.0 ± 0.7 pA; Rab3AEbd/Ebd, 15.1 ± 1.0 pA, p = 0.0027, Kruskal-Wallis test. B) Quantiles pooled from Rab3A+/+ and Rab3AEbd/Ebd cells were ranked from smallest to largest, plotted against each other, and fit with linear regression functions. Gray line, identity; red dashed line, linear regression fit. C) The same ranked mEPSC quantiles were paired and the ratio of Rab3A+/+/Rab3AEbd/Ebd computed for each pair, then plotted as a function of the Rab3A+/+ mEPSC for the pair. Dashed line at ratio = 1.0 (no difference).

Alternatively, these are two distinct mechanisms of mEPSC amplitude augmentation that occlude each other. In any case, the increase in mEPSC amplitude in cultures from Rab3AEbd/Ebdmice is consistent with the increase in mEPC amplitude we observed at the Rab3AEbd/Ebd NMJ (Wang et al., 2011).

The small GTPase Rab3A is generally thought to function presynaptically to regulate synaptic vesicle trafficking, possibly in an activity-dependent manner (Castillo et al., 1997; Lonart et al., 1998; Leenders et al., 2001; Schluter et al., 2006; Coleman and Bykhovskaia, 2009; Tian et al., 2012). In contrast, homeostatic plasticity of mEPSC amplitude has been attributed to an increase in postsynaptic receptors on the surface of the dendrite (O’Brien et al., 1998; Turrigiano et al., 1998). Is Rab3A required for the increase in surface AMPA-type glutamate receptors that has been confirmed by multiple studies (see Introduction)? It should be noted that at the NMJ in vivo we could find no evidence for an increase in AChRs after TTX block of the sciatic nerve in vivo (Wang et al., 2005). The type of AMPA receptor that has been shown to be increased, calcium-permeable GluA1-containing AMPA receptor or calcium-impermeable GluA2-containing AMPA receptor, appears to depend on the experimental manipulation. For example, block of activity by APV (NMDA antagonist) combined with TTX (inhibitor of Na channels) induces an increase in mEPSC amplitude in dissociated hippocampal cultures that is completely reversed by acute application of the Ca-permeable receptor-specific inhibitor NASPM (Sutton et al., 2006), indicating the entirety of the homeostatic effect is due to those receptors, likely GluA1. In contrast, the increase in mEPSC amplitude in mouse hippocampal slice cultures induced by TTX alone was not affected by another GluA1-specific inhibitor, philanthotoxin, suggesting mediation by either Ca-impermeable receptors or a presynaptic effect on quantal size (Soden and Chen, 2010); see also, (Dubes et al., 2022). Because we used TTX alone to block network activity, we expected that NASPM would not reverse the TTX- induced increase in mEPSC amplitude in our mouse cortical cultures, and we found that this was indeed the case. Figure 4A shows that the TTX-induced increase in mean mEPSC amplitude was nearly identical in a set of 11 CON and 11 TTX-treated cells before and after NASPM treatment (before NASPM, CON 12.9 ± 3.5 pA; TTX, 17.5 ± 3.1 pA, p = 0.009; after NASPM, CON 11.9 ± 2.6 pA; TTX 16.1 ± 3.5 pA, p = 0.006, Kruskal-Wallis test). Application of NASPM caused a modest decrease in mEPSC amplitude in both untreated and TTX-treated cultures (Figure 4B, CON, before NASPM, 12.9 ± 3.5 pA; after, 11.9 ± 2.6 pA, p = 0.08; TTX, before NASPM, 17.5 ± 3.1 pA; after, 16.1 ± 3.5 pA, p = 0.08, paired t test), and also decreased mEPSC frequency (Figure 4C, CON, before NASPM, 1.84 sec-1; after NASPM, 1.56 sec-1; p = 0.003, paired t-test; TTX, before NASPM, 4.40 ± 3.51 mEPSCs sec-1; after NASPM, 2.68 ± 2.25 mEPSCs sec-1, p = 0.02, paired t-test), indicating that effective concentrations of NASPM were reached and mEPSCs due to AMPA receptors composed entirely of Ca-permeable subunits were abolished.

Ca2+-permeable, NASPM-sensitive glutamate receptors did not mediate the TTX- induced increase in mEPSC amplitude in cultures of dissociated mouse cortical neurons. A) mEPSC amplitudes for 11 mouse cortical neurons from untreated cultures (“CON”) and 11 neurons from TTX treated cultures (“TTX”) were recorded before and after application of 20 μM NASPM. mEPSC amplitudes were still significantly larger in TTX-treated cultures compared to untreated control cultures after NASPM treatment (TTX effect pre-NASPM, p = 0.009; TTX effect post-NASPM, p = 0.006, Kruskal-Wallis non-parametric test). B) Acute application of NASPM had a modest effect on mEPSC amplitude in cells from CON or TTX cultures that did not reach statistical significance (CON, pre-NASPM, 12.9 ± 1.1 pA, post-NASPM, 11.9 ± 0.8 pA, p = 0.08; TTX, pre-NASPM, 17.5 ± 0.9 pA, post-NASPM, 16.1 ± 1.0 pA, p = 0.08; student’s paired t-test). C) Acute application of NASPM consistently reduced mEPSC frequency in cells from untreated and cells from TTX-treated cultures (CON, pre-NASPM, 1.84 ± 1.82 mEPSCs sec-1, post-NASPM, 1.56 ± 1.74 mEPSCs sec-1; p = 0.003; TTX, pre-NASPM, 4.40 ± 3.51 mEPSCs sec-1, post-NASPM, 2.68 ± 2.25 mEPSCs sec-1, p = 0.02; student’s paired t-test).

Having established that GluA1 receptors are not contributing to the homeostatic increase in mEPSC amplitude, we turned to immunohistochemistry and confocal imaging to assess whether GluA2 receptor expression was increased in our wild type mouse cortical cultures following 48 hour treatment with TTX. Since mEPSCs necessarily report synaptic levels of receptors, we used VGLUT1-positivity to identify synapses on pyramidal primary apical dendrites labeled with MAP-2 immunofluorescence. Figure 5 shows 6 pairs of VGLUT1- and GluA2-immunofluorescent clusters (white frames) along each of two primary dendrites, one in an untreated cortical culture (CON, left), the other in a culture treated with TTX for 48 hours (TTX, right), both from Rab3A+/+ mice. A dendrite typically required ∼10 confocal sections to be fully captured, and the total number of synaptic pairs for all the sections imaged on a dendrite was usually < 20, so this is an atypically high number of pairs within a single section; these particular dendrites and sections were selected for illustration purposes. In addition to the synaptic pairs, we observed many GluA2-immunoreactive clusters not associated with VGLUT1 immunoreactivity; those along the dendrites may be extrasynaptic receptors, and those outside the dendrites may be on astrocytes (Fan et al., 1999). There are also GluA2 immunoreactive clusters that are close to VGLUT1 immunoreactivity but are not located along any apparent MAP2-positive neurite, suggesting axon-axonal contacts, although VGLUT1 has also been detected in astrocytes (Ormel et al., 2012). Only sites that contained both VGLUT1 and GluA2 immunoreactivity close to the primary MAP2-postive dendrite or a secondary branch were selected for analyses.

Confocal images of GluA2 and VGLUT1 immunofluorescence on MAP2-positive primary dendrites of pyramidal-shaped neurons in high density dissociated cortical cultures prepared from Rab3A+/+ mice. Shown are 5X confocal sections of two primary dendrites, one from an untreated culture (CON) and the other from a culture treated with TTX (TTX). Postsynaptic GluA2-positive receptor clusters were identified by their proximity to VGLUT1 immunoreactive presynaptic terminals within the same confocal section. GluA2-VGLUT1 synaptic pairs, which were of maximal brightness in these confocal slices, are indicated by white trapezoids. It can be seen that substantial numbers of GluA2-positive clusters located on the dendrites are unopposed by VGLUT1-positive terminals, and that some GluA2-positive and VGLUT1-positive clusters are found away from any MAP2 positivity, suggesting localization to axons or astrocytes. Scale bar = 10 μm.

Variability in the magnitude of the homeostatic response from culture to culture is averaged out in physiological experiments by the pooling of data from many cells across many cultures. To reduce the necessity for many cultures, we chose to pair experiments in the same cultures by recording mEPSCs from one set of coverslips, and processing another set of coverslips from the same culture for immunohistochemistry. We completed this matched paradigm of physiology and immunohistochemistry, which to our knowledge has never been done before, for 3 cultures prepared from Rab3A+/+ mice and 3 cultures prepared from Rab3A-/- mice. We present the results for mEPSC data and imaging data pooled across the 3 experiments in Figure 6 (Rab3A+/+), Figure 7 (Rab3A-/-), and Table 1. Levels of GluA2 immunoreactivity at synaptic sites were quantified by the size of the GluA2-positive receptor cluster and the average intensity value of the receptor cluster. We analyzed the imaging data the same way we analyzed mEPSC data, sampling 30 quantiles from each dendrite’s data set, and calculating the means, pooling quantiles across dendrites to create CDFs, and sorting quantiles from smallest to largest to produce rank order and ratio plots.

GluA2 receptor cluster size and average intensity increased following activity blockade in parallel with mEPSC amplitudes recorded in the same cortical cultures prepared from Rab3A+/+ mice, but the effects on receptor levels were less robust and differed from mEPSC amplitudes in characteristics of scaling. A) Cumulative distribution functions (CDFs) of mEPSC amplitudes recorded from untreated (“CON”) and TTX-treated cells (“TTX”) pooled from 3 cultures were significantly different (KS test, D = 0.162, p = 1.42 * 10-8, CON, n = 690, TTX, n = 720). Inset, mean mEPSC amplitudes for 23 CON cells and 24 TTX cells were significantly different (CON, 13.7 ± 4.5 pA; TTX, 16.4 ± 4.3 pA, p = 0.016, Kruskal-Wallis test). B) 24 mEPSC amplitude quantiles from each of 23 untreated cells were sorted from smallest to largest and plotted vs. 23 mEPSC amplitude quantiles from each of 24 TTX-treated cells, and the resulting relationship fit with a linear regression equation, Y = 1.106x + 1.22 (R2 = 0.984). Inset, full range of data used for fit. C) Sorted data from (B) were used to calculate the ratio of TTX mEPSC amplitude/CON mEPSC amplitude and the ratios plotted vs. CON mEPSC amplitude. D) CDFs of size of GluA2 receptor clusters imaged from primary dendrites in the same untreated cultures and TTX-treated cultures as in (A) – (C) were significantly different (KS test, D = 0.089, p = 0.002, CON, n = 870, TTX, n = 870). Inset, mean GluA2 cluster size for 29 CON dendrites and 29 TTX dendrites were not significantly different (CON, 0.97 ± 0.38 μm2; TTX, 1.15 ± 0.59 μm2, p = 0.44, Kruskal-Wallis test). E) 29 GluA2 cluster size quantiles from each of 29 untreated dendrites were sorted from smallest to largest and plotted vs. 29 cluster size quantiles from each of 29 TTX-treated dendrites, and the resulting relationship fit with a linear regression equation, Y = 1.328x -0.136 (R2 = 0.98). Inset, full range of data used for fit. F) Sorted data from (E) were used to calculate the ratio of TTX GluA2 cluster size/CON GluA2 cluster size and the ratios plotted vs. CON GluA2 cluster size. G) CDFs of intensity of GluA2 receptor clusters imaged from primary dendrites in untreated cultures and TTX-treated cultures were significantly different (KS test, D = 0.120, p = 6.73 * 10-6, CON, n = 870, TTX, n = 870). Inset, mean GluA2 cluster intensity for 29 CON dendrites and 29 TTX dendrites were not significantly different (CON, 673 ± 90 a.u.; TTX, 687 ± 72 a.u., p = 0.25, Kruskal-Wallis test). H) 29 GluA2 cluster intensity quantiles from 29 untreated dendrites were sorted from smallest to largest and plotted vs. 29 cluster intensity quantiles from 29 TTX-treated dendrites, and the resulting relationship fit with a linear regression equation, Y = 0.899x + 82 (R2 = 0.987). Inset, full range of data used for fit. I) Sorted data from (H) were used to calculate the ratio of TTX cluster intensity/CON cluster intensity and plotted vs. CON cluster intensity.

GluA2 receptor cluster size, average intensity, and mEPSC amplitudes recorded in the same cortical cultures prepared from Rab3A-/- mice were not increased following activity blockade. A) Cumulative distribution functions (CDFs) of mEPSC amplitudes recorded from untreated (“CON”) and TTX-treated cells (“TTX”) pooled from 3 Rab3A-/- cultures were not significantly different (KS test, D = 0.072, p = 0.085, CON, n = 630, TTX, n = 570). Inset, mean mEPSC amplitudes for 21 CON cells and 19 TTX cells were not significantly different (CON, 14.9 ± 3.8 pA; TTX, 14.0 ± 4.0 pA, p = 0.34, Kruskal-Wallis test). B) 19 mEPSC quantiles from each of 21 untreated cells were sorted from smallest to largest and plotted vs. 21 mEPSC quantiles from each of 19 TTX-treated cells, and the resulting relationship fit with a linear regression equation, Y = 0.93x + 0.13 (R2 = 0.997). Inset, full range of data used for fit. C) Sorted data from (B) were used to calculate the ratio of TTX mEPSC amplitude/CON mEPSC amplitude and plotted vs. CON mEPSC amplitude. D) CDFs of GluA2 receptor cluster size imaged from primary dendrites in the same untreated cultures and TTX-treated cultures as in (A) – (C) were not significantly different (KS test, D = 0.045, p = 0.31, CON, n = 900, TTX, n = 870). Inset, mean GluA2 cluster size for 30 CON dendrites and 29 TTX dendrites were not significantly different (CON, 0.93 ± 0.27 μm2; TTX, 0.91 ± 0.28 μm2, p = 0.74, Kruskal-Wallis test). E) 29 GluA2 cluster size quantiles from each of 30 untreated dendrites were sorted from smallest to largest and plotted vs. 30 cluster size quantiles from each of 29 TTX-treated dendrites, and the resulting relationship fit with a linear regression equation, Y = 1.01x -0.03 (R2 = 0.993). Inset, full range of data used for fit. F) Sorted data from (E) were used to calculate the ratio of TTX GluA2 cluster size/CON GluA2 cluster size and the ratios plotted vs. CON GluA2 cluster size. G) CDFs of GluA2 receptor cluster average intensity imaged from primary dendrites in untreated cultures and TTX-treated cultures were not significantly different (KS test, D = 0.0.060, p = 0.08, CON, n = 900, TTX, n = 870). Inset, mean GluA2 cluster intensity for 30 CON dendrites and 29 TTX dendrites were not significantly different (CON, 766 ± 68 a.u.; TTX, 776 ± 79 a.u., p = 0.47, Kruskal-Wallis test). H) 29 GluA2 cluster intensity quantiles from 30 untreated dendrites were sorted from smallest to largest and plotted vs. 30 cluster intensity quantiles from 29 TTX-treated dendrites, and the resulting relationship fit with a linear regression equation, Y = 1.09x - 60 (R2 = 0.993). Inset, full range of data used for fit. I) Sorted data from (H) were used to calculate the ratio of TTX cluster intensity/CON cluster intensity and the ratios plotted vs. CON cluster intensity.

In the data pooled from the 3 matched experiments, neurons from Rab3A+/+ cultures showed a significant increase in mean mEPSC amplitudes following activity blockade (Figure 6A inset, CON, 13.7 ± 4.5 pA, n = 23; TTX, 16.4 ± 4.3 pA, n = 24; p = 0.016, Kruskal-Wallis test); a significant shift of the TTX CDF to larger mEPSC amplitude values (Figure 6A, D = 0.162, p = 1.42 * 10-8, KS test); a slope of 1.11 in the rank order plot (Figure 6B, R2 = 0.98); and a mean ratio of the 50th to 75th percentile of 1.24 in the ratio plot (Figure 6C). These data indicate that the homeostatic response averaged across the 3 cultures was very similar to, but slightly smaller than, that of the previous data set presented in Figure 1. The means for size and intensity of GluA2 receptor clusters, while showing trends to higher values after activity blockade, were not significantly different (Figure 6D inset, size, CON, 0.97 ± 0.38 μm2; TTX, 1.15 ± 0.59 μm2, p = 0.44; Figure 6G inset, intensity, CON, 673 ± 90, TTX, 687 ± 72, p = 0.25, Kruskal-Wallis test; Table 1). However, when comparing the CDFs, where the sample size is much greater (equal to usually close to 30 quantiles * number of cells), the shifts to larger values for the TTX CDF did reach statistical significance for GluA2 receptor cluster size (Figure 6D, D = 0.089, p = 0.002, KS test) and intensity (Figure 6G, D = 0.120, p = 6.73 * 10-6, KS test). In the rank order plot for GluA2 receptor cluster sizes (Figure 6E), the slope was actually larger than that for mEPSC amplitude (1.33), but for intensity (Figure 6H), the slope value was below 1 (0.90). The weaker effects on synaptic GluA2 receptor levels relative to mEPSC amplitude suggest the possibility that GluA2 receptors are not the sole determining factor in the increased mEPSC amplitude following activity blockade. In addition, the plot of the ratio of TTX GluA2 receptor cluster size/control size as a function of ranked control values showed a surprising deviation from the divergent scaling we have observed for mEPSC amplitude (compare Figure 6C to Figure 6F). The smallest GluA2 receptor clusters showed a TTX/CON ratio as high as 3, but the mean ratio across the 50th to 75th percentile settled down to 1.11 within < 20 samples. This finding of larger ratios for the smallest GluA2 receptor clusters is supported by a previous study that followed the same fluorescently labeled postsynaptic sites over time following activity blockade. Wang and colleagues found that the biggest increases after activity blockade occurred at the smallest/dimmest synaptic sites (Wang et al., 2019). A similar inverse relationship was reported for changes in individual fluorescently labeled postsynaptic sites followed over time under normal activity conditions (Minerbi et al., 2009; Statman et al., 2014). The mismatch in ratio values between the smallest mEPSC amplitudes and smallest GluA2 receptor cluster sizes may be due to the inability of physiological assays to detect the mEPSCs coming from the smallest synapses.

We proceeded to determine the homeostatic responses of mEPSC amplitude and GluA2 receptor levels in 3 cultures prepared from Rab3A-/- mice. As shown for the data set in Figure 1, mean mEPSC amplitude was not increased following activity blockade in the data pooled from this new set of 3 Rab3A-/- cultures (Figure 7A inset, CON 14.9 ± 3.8 pA, n = 21; TTX, 14.0 ± 4.0 pA, n = 19; p = 0.34, Kruskal-Wallis test). Notably, the TTX CDF was shifted to smaller mEPSC amplitude values (Figure 7A, D = 0.072, p = 0.085, KS test), the slope of the rank order plot was below 1 (Figure 7B, slope 0.93, R2 = 0.997), and the mean ratio across the 50th to 75th percentile was below 1 (Figure 7C, 0.96). Thus, the disruption of homeostatic plasticity of mEPSC amplitude shown with the previous data sets was recapitulated in this data set. For immunohistochemistry results from the same Rab3A-/- cultures, we found that mean GluA2 receptor cluster size was unchanged following activity blockade (Figure 7D inset, CON, 0.93 ± 0.27 μm2, TTX, 0.91 ± 0.28 μm2, p = 0.74, Kruskal-Wallis test), the TTX CDF was not significantly shifted (Figure 7D, D = 0.045, p = 0.31, KS test), the slope on the rank order plot was 1.01 (Figure 7E), and the ratio hovered at or below 1 except for a small group of ratios at the high end of the control values (Figure 7F). For intensity, mean values were not increased after activity blockade (Figure 7G inset, CON, 766 ± 68, TTX, 776 ± 79, p = 0.47, Kruskal-Wallis test), the CDFs were not significantly different (Figure 7G, D = 0.060, p = 0.080), the slope on the rank order plot was 1.09 (Figure 7H), and the mean ratio value across the 50th to 75th percentile was 1.02 (Figure 7I). Taken together, these results indicate that the modest increases in GluA2 receptor cluster size and intensity following activity blockade observed in wild-type cultures do not occur in the absence of Rab3A.

As noted above, the magnitude of the homeostatic effect on mEPSC amplitude appeared to be more robust than that on receptor levels in cultures from Rab3A+/+ mice. Further evidence that there is not a one-to-one correspondence in the homeostatic response of mEPSC amplitudes and GluA2 receptor levels was apparent when the individual experiments’ data were compared. The electrophysiology experiments were technically difficult because in order to evaluate whether a particular culture displayed homeostatic plasticity of mEPSC amplitude, we set a minimum requirement of recording from 6 cells per condition, meaning a successful experiment required at least 12 (6 CON and 6 TTX) usable recordings (i.e. low holding current, stable baseline, low noise, etc.) in one day. Ultimately, our sample size ranged from 6 to 10 cells per condition per culture. Figures 8A-C show the CDFs and ratio plots for 3 individual Rab3A+/+ cultures. Rank order plots are not included here because the ratio plot is more sensitive than the rank order plot when determining effect magnitude. We found that the mEPSC amplitude effects differed from the GluA2 cluster size effects at both the broad level (increases vs decreases), and in the specific ways in which the effect was non-uniform. Broadly, although two of the three Rab3A+/+ experiments show a homeostatic increase in the TTX CDF and a ratio of TTX/CON > 1 for both mEPSC amplitudes (Figures 8Ai,iii and 8Bi,iii) and synaptic GluA2 receptor cluster sizes (Figures 8Aii,iv and 8Bii,iv), Culture #3 shows an increase in mEPSC amplitude but a decrease in the GluA2 receptor cluster size, based on both the shifts in CDFs (mEPSC, Figure 8Ci, vs GluA2 receptor cluster size, Figure 8Cii) and the majority of ratio values (mEPSC, Figure 8Ciii, vs. GluA2 receptor cluster size, Figure 8Civ). Similarly, we found an obvious mismatch between the mEPSC amplitude data and the GluA2 cluster size data for one of the three Rab3A-/- experiments. For Culture #3, the mEPSC TTX CDF shows a shift to the left (Figure 8Fi), and the ratios were below 1 for control mEPSC amplitudes < 20 pA and above 1 for control mEPSC amplitudes > 20 pA (Figure 8Fiii). In the same culture, GluA2 receptor cluster size CDF shows a shift to the right after activity blockade (Figure 8Fii), and in the ratio plot, values were at or above 1 throughout the range of control sizes (Figure 8Fiv).

mEPSC amplitudes and GluA2 receptor cluster sizes did not always change in parallel after treatment with TTX in cultures from Rab3A+/+ and Rab3A-/- mice. A) Rab3A+/+ Culture #1: mEPSC amplitudes and GluA2 receptor cluster sizes increased in parallel after treatment in TTX. i) CDFs of mEPSC amplitudes of cells from untreated and TTX treated dishes were significantly different (KS test, D = 0.16, p = 0.010; CON, n = 180, TTX, n = 210). ii) CDFs of GluA2 receptor cluster sizes of dendrites from untreated and TTX treated dishes were significantly different (KS test, D = 0.17, p = 3.79 * 10-4; CON, n = 300, TTX, n = 300). iii) 35 mEPSC amplitude quantiles from each of 6 untreated cells and 30 quantiles from each of 7 TTX-treated cells were sorted from smallest to largest and used to calculate the ratio of TTX mEPSC amplitude/CON mEPSC amplitude, which was plotted vs. CON mEPSC amplitude. iv) 30 GluA2 receptor cluster size quantiles from each of 10 untreated dendrites and 30 GluA2 receptor cluster size quantiles from each of 10 TTX-treated dendrites were used to calculate the ratio of TTX GluA2 cluster size/CON GluA2 cluster size, and plotted vs. CON GluA2 cluster size. Inset, full range of data. B) Rab3A+/+ Culture #2: mEPSC amplitudes and GluA2 receptor cluster sizes increased in parallel after treatment in TTX. i) CDFs of mEPSC amplitudes of cells from untreated and TTX treated dishes were significantly different (KS test, D = 0.21, p = 1.06 * 10-4; CON, n = 210, TTX, n = 240). ii) CDFs of GluA2 receptor cluster sizes of dendrites from untreated and TTX treated dishes were significantly different (KS test, D = 0.21, p = 5.39 * 10-6; CON, n = 270, TTX, n = 300). iii) 32 mEPSC amplitude quantiles from each of 7 untreated cells, and 28 mEPSC amplitude quantiles from each of 8 TTX treated cells were sorted from smallest to largest and were used to calculate the ratio of TTX mEPSC amplitude/CON mEPSC amplitude, and plotted vs. CON mEPSC amplitude. iv) 30 GluA2 receptor cluster size quantiles from each of 9 untreated dendrites and 27 GluA2 receptor cluster size quantiles from each of 10 TTX-treated dendrites were sorted from smallest to largest and used to calculate the ratio of TTX GluA2 cluster/CON GluA2 cluster size, and plotted vs. CON GluA2 cluster size. Inset, full range of data. C) Rab3A+/+ Culture #3: mEPSC amplitudes increased, but GluA2 receptor cluster sizes decreased, after treatment in TTX. i) CDFs of mEPSC amplitudes of cells from untreated and TTX treated dishes were significantly different (KS test, D = 0.15, p = 0.002; CON, n = 300, TTX, n = 270). ii) CDFs of GluA2 receptor cluster sizes of dendrites from untreated and TTX treated dishes were significantly different, but the TTX CDF is shifted to the left (KS test, D = 0.16, p = 0.001; CON, n = 300, TTX, n = 270) iii) 27 mEPSC amplitude quantiles from each of 10 untreated cells, and 30 mEPSC amplitude quantiles from each of 9 TTX-treated cells, were sorted from smallest to largest, used to calculate the ratio of TTX mEPSC amplitude/ CON mEPSC amplitude, and plotted vs. CON mEPSC amplitude. iv) 27 GluA2 receptor cluster size quantiles from each of 10 untreated dendrites and 30 GluA2 receptor cluster sizes from each of 9 TTX-treated dendrites were sorted from smallest to largest, used to calculate the ratio of TTX GluA2 receptor cluster size/CON GluA2 receptor cluster size, and plotted vs. CON GluA2 receptor cluster size. D) Rab3A-/- Culture #1 (note, this culture is distinct from that of Rab3A+/+ Culture #1). Neither mEPSC amplitudes, nor GluA2 receptor cluster size, increased following TTX treatment. i) CDFs of mEPSC amplitudes of cells from untreated and TTX-treated cultures were not significantly different (KS test, D = 0.05, p = 0.90; CON, n = 210, TTX, n = 180). ii) CDFs of GluA2 receptor cluster sizes of dendrites from untreated and TTX (Figure 8 continued) treated dishes were not significantly different, although the TTX CDF is shifted slightly to the left (KS test, D = 0.10, p = 0.08; CON, n = 300, TTX, n = 300). iii) 30 mEPSC amplitude quantiles from each of 7 untreated cells and 35 mEPSC quantiles from each of 6 TTX-treated cells were sorted from smallest to largest, used to calculate the ratio TTX mEPSC amplitude/CON mEPSC amplitude, and plotted vs. CON mEPSC amplitude. iv) 30 GluA2 receptor cluster size quantiles from each of 10 untreated dendrites and 10 TTX-treated dendrites were sorted from smallest to largest, used to calculate the ratio TTX cluster size/CON cluster size, and plotted vs. CON cluster size. E) Rab3A-/- Culture #2 (distinct from Rab3A+/+ Culture #2) mEPSC amplitudes and GluA2 receptor cluster sizes decreased after TTX treatment. i) TTX mEPSC amplitudes CDF is shifted slightly to the left compared to the CON mEPSC amplitudes CDF but the difference is not significant (KS test, D = 0.08, p = 0.28, CON, n = 240, TTX, n = 240). ii) TTX GluA2 receptor cluster sizes CDF is significantly shifted to the left compared to CON GluA2 receptor cluster sizes CDF (KS test, D = 0.13, p = 0.01; CON, n = 300, TTX, n = 300). iii) 24 mEPSC amplitude quantiles from each of 8 untreated cells and 8 TTX- treated cells were sorted from smallest to largest, used to calculate the ratio TTX mEPSC amplitude/CON mEPSC amplitude, and plotted vs. CON mEPSC amplitude. iv) 27 GluA2 receptor cluster size quantiles from each of 10 untreated dendrites, and 30 GluA2 receptor cluster size quantiles from each of 9 TTX treated dendrites, were sorted from smallest to largest, used to calculate the ratio TTX GluA2 cluster size/CON cluster size, and plotted vs. CON cluster size. Both mEPSC amplitude and GluA2 receptor cluster size ratios were below 1 across the majority of data. F) Rab3A-/- Culture #3 (distinct from Rab3A+/+ Culture #3). mEPSC amplitudes were decreased following TTX treatment, but GluA2 receptor cluster sizes were increased. i) TTX CDF of mEPSC amplitudes was significantly shifted to the left of CON CDF (KS test, D = 0.21, p = 4.85 * 10-4, CON, n = 180, TTX, n = 180). ii) TTX CDF of GluA2 receptor cluster sizes was significantly shifted to the right of CON CDF (KS test, D = 0.15, p = 0.003, CON, n = 300, TTX, n = 300). iii) 30 mEPSC amplitude quantiles from each of 6 untreated and 6 TTX-treated cells were sorted from smallest to largest, used to calculate the ratio TTX mEPSC amplitude/CON mEPSC amplitude, and plotted vs. CON mEPSC amplitude. iv) 30 GluA2 receptor cluster size quantiles from each of 10 untreated dendrites and 10 TTX-treated dendrites were sorted from smallest to largest, used to calculate the ratio TTX GluA2 cluster size/CON GluA2 cluster size, and plotted vs. CON CluA2 cluster size.

Perhaps more striking than the broad mismatches, which could be attributed to the small sample sizes, is the complete lack of correspondence of TTX/CON ratios in the mEPSC amplitudes compared to GluA2 cluster sizes in the same cultures. In cultures from Rab3A+/+ mice, mEPSC amplitudes show divergent scaling similar to what we have previously reported, with the smallest CON amplitudes showing the smallest ratios, and the ratio increasing monotonically to a plateau, or, a peak followed by a gradual decline, as CON mEPSC amplitude becomes very large. In contrast, the very smallest GluA2 cluster sizes in Rab3A+/+ cultures have the largest increase, with ratios as high as 3 in the two cultures that showed a broad overall increase. The ratios decline dramatically to values around 1.5, but then remain substantially above 1 throughout the entire range of data. Interestingly, the dramatic increase in the size of the smallest GluA2 clusters is absent in the two cultures from Rab3A-/- mice that broadly showed no increase, suggesting this aspect is also dependent on Rab3A. It is not even possible to cut off the ratios at a certain point, to address a threshold effect such as GluA2 cluster sizes below a certain value not being measurable electrophysiologically, and get the ratios for the remaining data to match up. Taken together, these data indicate that variation in homeostatic effects on GluA2 cluster size is not responsible for the variation in homeostatic effects on mEPSC amplitude.

In summary, our matched mEPSC and receptor cluster results indicate that: 1. activity blockade resulted in an increase in mEPSC amplitudes and synaptic GluA2 receptors; 2. loss of Rab3A disrupted both the increase in mEPSC amplitudes and the increase in GluA2 receptor levels; and 3. The homeostatic effects on mEPSC amplitudes and the GluA2 cluster sizes within the same cultures differed at both the broad level, and in the specific way the ratios were non- uniform, making it difficult to conclude modulation of GluA2 cluster size is the sole determinant of homeostatic increases in mEPSC amplitude. Although it remains a possibility that the disparities in homeostatic effects on mEPSC amplitudes and GluA2 cluster size TTX/CON ratios might arise due to a sampling error, given the small number of cells sampled in each individual experiment (≤ 10), it seems unlikely that differences arising from sampling error alone would be so consistent. However, only when we are able to more easily sample larger numbers of cells, or it becomes possible to measure mEPSCs and receptor levels in the same cells, will we be able to resolve this question.

We have demonstrated here for the first time that the synaptic vesicle protein Rab3A influences the homeostatic regulation of postsynaptic GluA2 receptors. What are possible ways that Rab3A could exert a postsynaptic action? It has previously been shown that exogenous addition of TNFα to hippocampal cultures causes an increase in surface expression of GluA1 receptors (although not GluA2 receptors), and that the homeostatic increase in mEPSC amplitudes is abolished in cultures prepared from the TNFα deletion mouse (Stellwagen et al., 2005; Stellwagen and Malenka, 2006). Furthermore, neurons from TNFα+/+ mice plated on astrocytic feeder layers derived from TNFα-/- mice fail to show the increase in mEPSC amplitude after TTX treatment, indicating that the TNFα inducing the receptor increases following activity blockade comes from the astrocytes (Stellwagen and Malenka, 2006). Rab3A has been detected in astrocytes (Maienschein et al., 1999; Hong et al., 2016), so to determine whether Rab3A is acting via regulating TNFα release from astrocytes, we performed experiments similar to those of Stellwagen and Malenka (2006) (schema illustrated in Figure 9, left side). We compared the effect of activity blockade on mEPSC amplitudes recorded from cortical neurons from Rab3A+/+ mice plated on Rab3A+/+ astrocytic feeder layers (Figure 9A); neurons from Rab3A+/+ mice plated on Rab3A-/- astrocytes (Figure 9B), and neurons from Rab3A-/- mice plated on Rab3A+/+ astrocytes (Figure 9C). If Rab3A is required for TNFα release from astrocytes, then Rab3A+/+ neurons plated on Rab3A-/- astrocytes should not show a homeostatic increase in mEPSC amplitude after treatment with TTX, and any cultures with Rab3A+/+ astrocytes should have a normal homeostatic response. We found the opposite result: mEPSC amplitudes increased dramatically in cultures where Rab3A was present in neurons (Figures 9A and B), but much more modestly increased in cultures where Rab3A was present only in astrocytes (Figure 9C).

Rab3A in neurons, not astrocytes, was required for full homeostatic plasticity after activity blockade. A-C, Data from dissociated cortical neurons plated on an astrocyte feeder layer, each prepared separately from the type of mice depicted in the schema at left: A) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A+/+ mice; i) CDFs for mEPSC amplitudes from untreated (CON, black curve) and TTX-treated (TTX, red curve) cultures (KS test, test statistic, D = 0.141, p = 2.74 * 10-5, n = 510 CON, 600 TTX samples; inset, means, CON, 13.3 ± 0.5 pA, n = 17 cells, TTX, 16.7 ± 1.2 pA, n = 20 cells, p = 0.03, Kruskal Wallis test; ii) rank order plot and linear regression fit, slope = 1.89, intercept = -8.51, R2 = 0.97; iii) ratio plot of TTX mEPSC amplitude/CON amplitude vs. CON amplitude; B) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A-/- mice; i) CDFs of mEPSC amplitudes, CON, black curve; TTX, red curve (KS test, test statistic 0.273, p = 2.98 * 10-11), n = 330 CON, 330 TTX samples; inset, means, CON, 13.3 ± 1.0 pA, n = 11 cells, TTX, 18.8 ± 1.4 pA, n = 11 cells, p = 0.005, Kruskal-Wallis test; ii) rank order plot and linear regression fit, slope = 1.76, intercept = -4.64, R2 = 0.98; iii) ratio plot of TTX mEPSC amplitude/TTX mEPSC amplitude vs. CON mEPSC amplitude; C) Neurons from Rab3A-/- neurons plated on astrocytes from Rab3A+/+ mice; i) CDFs of mEPSC amplitudes, CON, black curve, TTX, red curve (KS test, test statistic 0.126, p = 0.005), n = 420 CON, 330 TTX samples; inset, means, CON, 15.2 ± 1.1 pA, n = 14 cells, TTX, 16.9 ± 0.7 pA, n = 11 cells, p = 0.125, Kruskal-Wallis test;; ii) rank order plot and linear regression fit; slope = 1.01, intercept = 1.53, R2 = 0.97; ratio plot, TTX mEPSC amplitude/CON mEPSC amplitude vs. CON mEPSC amplitude. Gray line, identity; red dashed line, linear regression fit. Insets in rank order plots depict full range of data.

Interestingly, the homeostatic effects appear to be larger in the feeder layer cultures, compared to the effects observed in our previous neuronal/astrocyte mixed cultures prepared directly onto poly-L-lysine coated coverslips, but the ratio plots clearly show scaling was divergent under both plating conditions (compare Figures 1Giii with Figures 9Aiii and Biii). We noticed that in the astrocyte feeder layer cultures, and in the matched mEPSC amplitude and GluA2 receptor measurement experiments, the mean mEPSC amplitude in the untreated cultures prepared from Rab3A-/- mice was slightly, but not significantly, larger, compared to the mean mEPSC amplitude for untreated cultures prepared from Rab3A+/+ mice in the same experiments (astrocyte feeder layer, Rab3A+/+ neurons on Rab2A+/+ astrocytes, 13.3 ± 0.5 pA, Rab3A-/- on Rab3A+/+ astrocytes, 15.2 ± 1.1 pA; matched mEPSC and receptor experiments, Rab3A+/+ 13.7 ± 4.5 pA, Rab3A-/- 14.9 ± 3.8 pA. It is therefore possible that loss of Rab3A, like expression of the Rab3A Earlybird mutant, affects basal mEPSC amplitude, albeit to a lesser extent. However, it could also be that these differences reflect random variation in mEPSC amplitude from culture to culture.

In summary, in the absence of Rab3A from astrocytes, the divergent scaling of mEPSC amplitude following activity blockade was completely normal, whereas in the absence of Rab3A in neurons, scaling was greatly diminished. This result makes it highly unlikely that Rab3A is required for the release of TNFα, or another factor from astrocytes, that induces a homeostatic upregulation of postsynaptic receptors and thereby increases mEPSC amplitude following TTX treatment. Neuronal Rab3A appears to mediate the homeostatic increase in mEPSC amplitude following activity blockade.

Discussion

We found that homeostatic synaptic plasticity of mEPSC amplitude in dissociated mixed cultures of mouse cortical neurons and astrocytes behaved remarkably similar to the mouse NMJ in response to loss of Rab3A function: scaling up of mEPSC amplitude following prolonged network silencing by TTX was strongly diminished in cultures from Rab3A-/-, and in cultures from Rab3AEbd/Ebd mice, basal mEPSC amplitude was increased compared to that of wild-type cultures and was not further modified following 48-hour treatment with TTX. These results suggest that normal function of the presynaptic vesicle protein Rab3A is required for the homeostatic scaling up of mEPSC amplitude in cortical cultures and at the NMJ in vivo.

Rab3A not likely to regulate receptor trafficking

We demonstrated that an increase in synaptic GluA2 receptors accompanies the increase in mEPSC amplitude in cultures of dissociated mouse cortical neurons after 48-hour TTX treatment. By examining the same cultures with both mEPSC and immunofluorescence measurements, we found: 1. the effect on GluA2 receptor levels, whether measured as the size of a synaptic cluster or the density of receptors in a cluster, was less robust than the effect on mEPSC amplitudes; 2. the homeostatic effect on mEPSC amplitudes and GluA2 receptors substantially differed from each other both broadly (mEPSC amplitudes could increase while GluA2 cluster sizes decreased, and vice versa), and in the details of non-uniform scaling, one example being that the smallest mEPSC amplitudes showed the least effects but the smallest GluA2 clusters showed the greatest effects. Interestingly, there are other examples where the effect of activity blockade by either TTX alone (Hu et al., 2010), TTX and AP5 (Letellier et al., 2014), or DNQX (Blackman et al., 2012) on AMPAR levels did not match that of mEPSC amplitudes, although in these cases, the increases in receptors were larger than the increases of mEPSC amplitudes. The striking differences between TTX effects on AMPARs and mEPSC amplitudes indicates that there may be another factor contributing to mEPSC amplitude besides levels of AMPAR expression. One possibility is alterations in the type of AMPA receptor expressed, since GluA1 homomers have much greater conductance than GluA2-containing receptors (Oh and Derkach, 2005; Benke and Traynelis, 2019). Increases in mEPSC amplitudes not accompanied by increases in receptor numbers have been attributed to such a switch (Hou et al., 2015; Silva et al., 2019; Dubes et al., 2022).

In addition to finding a disparity between synaptic GluA2 receptor and mEPSC responses to prolonged activity blockade, there are two other reasons we do not think Rab3A impacts homeostatic synaptic plasticity solely through postsynaptic regulation of GluA2 receptor trafficking to the plasma membrane. First, prolonged activity blockade with TTX at the NMJ led to an increase in mEPC amplitude that was not accompanied by an increase in AChR levels, yet it was dependent on Rab3A (Wang et al., 2011). Second, for Rab3A to modulate AMPAR trafficking, it must be located in the postsynaptic dendrite. Multiple presynaptic molecules, such as SNARE proteins, synaptotagmins, and NSF have been identified in the postsynaptic compartment (Lledo et al., 1998; Nishimune et al., 1998; Osten et al., 1998; Song et al., 1998; Araki et al., 2010; Kennedy et al., 2010; Suh et al., 2010; Jurado et al., 2013; Hussain and Davanger, 2015; Gu et al., 2016; Wu et al., 2017), but to our knowledge, Rab3A has not.

Rab3A may regulate quantal size presynaptically

Based on our findings, the lack of postsynaptic Rab3A expression and the known presynaptic expression of Rab3A, Rab3A may regulate mEPSC amplitude presynaptically by modulating: a) the amount of transmitter transported into the vesicle, b) the size of the vesicle, or c) fusion pore characteristics. mRNA for the glutamate transporter, VGLUT1, was increased in rat cortical cultures after 48-hr treatment with TTX (De Gois et al., 2005), and VGLUT1 relative to synapsin I in rat hippocampal cultures was increased after 48-hr treatment with AP-5 (Wilson et al., 2005). This activity-dependent regulation of VGLUT1 was recently corroborated in a proteome study (Dorrbaum et al., 2020), and it has also been shown that mEPSC amplitudes are increased after overexpression of VGLUT at the Drosophila NMJ (Daniels et al., 2004) and in rat hippocampal cultures (Wilson et al., 2005). These data suggest that increases in VGLUT1 after activity-blockade could contribute to the homeostatic increase in mEPSC amplitude. Although there is no direct evidence for an interaction between VGLUT1 and Rab3A, Rab3A levels are reduced 50% in hippocampal extracts from mice lacking VGLUT1 (Fremeau et al., 2004). A Rab3A-dependent modulation of presynaptic quantal size would explain the homeostatic effects on mEPSC amplitudes that are not matched by changes in receptor levels.

Interestingly, Rab3A plays a role in regulating vesicle size. In adrenal chromaffin cells from mice lacking all 4 Rab3 isoforms (A, B, C and D) or mice heterozygous for Rab3A and lacking the B, C, and D isoforms (Rab3A+/-BCD-/-) with only 1 copy of Rab3A, there was a 20% increase in large dense core vesicle (DCV) diameter that did not reach significance (p = 0.11) (Schonn et al., 2010). Similarly, in the exocrine pancreas and parotid gland of the Rab3D-/-, large DCV diameter was increased by 22% (Riedel et al., 2002). This magnitude of change in diameter would produce a greater than 80% increase in volume. Also, an increase in volume has been shown to increase miniature endplate junctional current amplitude at the Drosophila NMJ (Karunanithi et al., 2002). There have not been any reports that small synaptic vesicle size is increased in Rab3-/-mice, but small synaptic vesicle size is regulated by activity: vesicle size increased in retinotectal terminals within the optic tectum following enucleation in pigeon (Cuenod et al., 1970), with similar results in cat and monkey (Lloret and Saavedra, 1975), in rat caudate nucleus following cortical ablation (Kawana et al., 1971), and in rat ventral spinal cord following immobilization (Cheresharov et al., 1978).

Finally, we previously showed that at the mouse NMJ, both loss of Rab3A and activity blockade by TTX caused an increase in the frequency of abnormally slow rising and/or prolonged duration mEPCs (Wang et al., 2011). We further demonstrated that the loss of Rab3A increased the frequency of very small amplitude fusion pore feet in mouse chromaffin cells (Wang et al., 2008). Taken together, these data suggested that Rab3A acts to prevent abnormal fusion pore release events. However, at the NMJ and in chromaffin cells, the kinetics (rise time, decay, half width) of the vast majority of release events were unchanged in the absence of Rab3A, making it unlikely that the shift in the distribution of mEPSC amplitudes is mediated by a Rab3A-dependent change in fusion pore kinetics.

Neuronal, not astrocytic, Rab3A may be required for the homeostatic release of a signaling molecule

Our results support the surprising possibility that presynaptic Rab3A regulates postsynaptic AMPARs, rather than occurring through a direct effect on the trafficking of AMPARs by Rab3A in the dendrite. Astrocytic release of TNFα was shown to mediate the increase in AMPARs in prolonged activity-blocked hippocampal cultures (Stellwagen and Malenka, 2006), but we found that loss of Rab3A in astrocytes does not disrupt the increase in mEPSC after activity blockade. Further, we previously showed that the homeostatic increase in NMJ mEPC amplitude was completely normal in the absence of TNFα (Wang et al., 2011).

Another possible way Rab3A might mediate regulation of postsynaptic AMPARs is through the presynaptic release of a signaling molecule, such as brain-derived neurotrophic factor (BDNF), that acts anterogradely to alter postsynaptic AMPAR levels. Addition of exogenous BDNF to neuronal cultures prevents the increase in mEPSC amplitude following activity blockade (Rutherford et al., 1998; Benevento et al., 2016) but see (Smith-Dijak et al., 2019), and BDNF mRNA is reduced after activity blockade in vitro (Benevento et al., 2016; Miyasaka and Yamamoto, 2021) and in vivo (Castren et al., 1992). Also, after reduction of secreted BDNF in culture media via the BDNF scavengers TrkB-FC or TrkB-IgG, mEPSC amplitudes are increased (Rutherford et al., 1998; Benevento et al., 2016; Smith-Dijak et al., 2019). Rab3B, a Rab3 family member with strong homology to Rab3A but expressed more highly in inhibitory nerve terminals (Tsetsenis et al., 2011), is modified epigenetically following TTX treatment by the same DNA methylation pathway as that for BDNF (Benevento et al., 2016). In addition, reduction of Rab3A in astrocytes decreases astrocytic BDNF release (Hong et al., 2016), and treatment of cultures with BDNF can increase Rab3A levels (Takei et al., 1997; Thakker-Varia et al., 2001) but see (Shinoda et al., 2014). Taken together, these results suggest an interplay between BDNF and Rab3A in the activity-dependent regulation of AMPA receptors.

Rab3A is required for divergent scaling

We have established that homeostatic synaptic plasticity in mouse cortical cultures, rat cortical cultures, and mouse hippocampal cultures demonstrates divergent, rather than uniform, scaling (Hanes et al., 2020; Koesters et al., 2022). The deviation from uniformity suggests that individual synapses may be independently regulated, and do not require a global, cell-wide signal. Our conclusion is supported by multiple studies demonstrating local activity- dependent regulation of mEPSC amplitudes at subsets of synapses (Ju et al., 2004; Sutton et al., 2006; Hou et al., 2008; Beique et al., 2011; Hou et al., 2011; Letellier et al., 2014). It is further supported by the recent finding that in hippocampal cultures treated with TTX, AMPA receptors increase selectively at synaptopodin-positive sites, which are already larger and contain higher amounts of AMPA receptors (Dubes et al., 2022). The authors provide evidence that the process depends on the microRNA miR-124: miR-124 is reduced after activity blockade; and overexpression of miR-124 blocks the TTX-induced increase in AMPA receptors and mEPSC amplitudes. Increases in miR-124 have also been linked to decreases in mEPC amplitudes at the NMJs of the slow channel syndrome mouse (mSCS) in which prolongation of ACh currents causes excess excitation (Zhu et al., 2013). Zhu and colleagues provide evidence for a postsynaptic calpain-Cdk5-nNOS pathway that is activated by increased calcium levels, with the NO feeding back retrogradely to regulate miR-124, which in turn directly interacts with Rab3A to affect quantal content (Zhu et al., 2013), but it is also possible that Rab3A is contributing to the mEPSC amplitude changes in mSCS. In sum, these studies suggest miR-124 could tie together presynaptic changes (Rab3A, transmitter released from a single vesicle) and postsynaptic changes (synaptopodin, AMPA receptors) in divergent homeostatic synaptic plasticity.

A model for the role of Rab3A in homeostatic plasticity of mEPSC amplitude

We propose a model for Rab3A function in homeostatic synaptic plasticity of mEPSC amplitude in which Rab3A contributes to presynaptic quantal size through the amount of transmitter released by fusion of a single vesicle, and to postsynaptic quantal size through the number of postsynaptic receptors. Normally, Rab3A cycles between GTP and GDP-bound forms (“State 1,” indicated by red tag in Figure 10A) and promotes synaptic vesicle mobilization and fusion, at the same time maintaining vesicle size/transmitter content within a narrow range. In this normally cycling Rab3A state, an anterograde signal dependent on Rab3A maintains GluA2 receptors at their normal, restricted level. Activity blockade leads to a buildup of an alternate form of Rab3A, possibly one that is unable to cycle through its GTP and GDP-bound forms (“State 2,” indicated by gray tag in Figure 10B). Without the quality control mechanism conferred by State 1 Rab3A, vesicle size/transmitter content skews to higher values, and the anterograde signal is abnormal, leading to GluA2 receptors levels rising outside normal limits. Not shown here, the Earlybird point mutation may permanently be in “State 2,” as the mutation is in the guanine nucleotide binding region (Kapfhamer), which would lead to increased transmitter released from single vesicles and increased receptor levels.

Rab3A is required for increased mEPSC amplitude after activity blockade via 2 mechanisms: 1. increased amount of transmitter released from a single vesicle and 2. increased postsynaptic receptors. Left: in the presence of normal activity, Rab3A is in its control state (“State 1,” red), maintaining mEPSC amplitude within a limited range by regulating either vesicle size or transmitter content, or both, and by regulating the number of postsynaptic receptors, likely through release of an anterograde signal. Right: after activity is blocked, Rab3A accumulates in an alternate functional state (“State 2,” gray) resulting in quantal size expanding to larger ranges because of increased vesicle size or transmitter content and because of increased numbers of receptors no longer being under the normal control of the anterograde signal.

We have demonstrated an essential role of the presynaptic vesicle protein Rab3A in the homeostatic increase in mEPSC amplitude and GluA2 AMPA receptor levels at synaptic sites in mouse dissociated cortical cultures. To our knowledge, this is the first evidence of a presynaptic protein being implicated in the homeostatic regulation of mEPSC amplitude and AMPARs in neuronal cultures. The current results extend our previous findings that showed the homeostatic increase in mEPC amplitudes at the mouse NMJ in vivo depends on Rab3A. Under conditions where postsynaptic receptors are close to being saturated by the amount of transmitter in a single vesicle (possibly in cortical cultures but not at the NMJ), it may be necessary to increase both the amount of transmitter released and the number or function of postsynaptic receptors in order to observe a homeostatic increase in the physiological signal, mEPSC amplitude. Our results suggest that in cortical cultures, neuronal Rab3A is a key player regulating the homeostatic increase in synaptic strength on both sides of the synapse.

eLife Assessment

This study presents valuable findings on the role of the small GTPase Rab3A in homeostatic plasticity. While the study demonstrates that Rab3A is required for homeostatic scaling, the evidence supporting the model put forward by the authors is incomplete. The work will be of interest to researchers in the field of synaptic transmission and plasticity.

Reviewer #1 (Public Review):

Koesters and colleagues investigated the role of the presynaptic small GTPase Rab3A in homeostatic scaling of miniature synaptic transmission in primary mouse cortical cultures using electrophysiology and immunohistochemistry. The major finding is that TTX incubation for 48 hours does not induce an increase in the amplitude of excitatory synaptic miniature events in neuronal cultures derived from Rab3A KO and Rab3A Earlybird mutant mice. NASPM application had comparable effects on mEPSC amplitude in control and after TTX, implying that Ca2+-permeable glutamate receptors are unlikely modulated during synaptic scaling. Immunohistochemical analysis revealed an increase in GluA2 puncta size and intensity in wild type, but not Rab3A KO cultures. Finally, they provide evidence that loss of Rab3A in neurons, but not astrocytes, blocks homeostatic scaling. Based on these data, the authors propose a model in which presynaptic Rab3A is required for homeostatic scaling of synaptic transmission through GluA2-dependent and independent mechanisms.

While the title of the manuscript is mostly supported by data of solid quality, many conclusions, as well as the final model, cannot be derived from the results presented. Importantly, the results do not indicate that Rab3A modulates quantal size on both sides of the synapse. Moreover, several analysis approaches seem inappropriate.

The following points should be addressed:

1. The model shown in Figure 10 is not supported by the data. The authors neither provide evidence for two different functional states of Rab3A being involved in mEPSC amplitude modulation, nor for a change in glutamate content of vesicles. Furthermore, the data do not fully support the conclusion of a presynaptic role for Rab3A in homeostatic scaling.

2. The analysis of mEPSC data using quantile sampling followed by ratio calculation is not meaningful under the tested experimental conditions because of the following reasons: (i) The analysis implicitly assumes that all events have been detected. The prominent mEPSC frequency increase after TTX suggests that this is not the case, i.e., many (small) mEPSCs are likely missed under control conditions. (ii) The analysis is used to conclude how events of a certain size are altered by TTX treatment. However, this analysis compares the smallest mEPSCs of the TTX condition with the smallest control mEPSCs, but this is not a pre-post experimental design. Variation between cells and between coverslips will markedly affect the results and lead to misleading interpretations. (iii) The ratio (TTX/control) vs. control plots seem to suffer from a division by small value artifact (see Figure 6F). Correspondingly, ratio-analysis differs considerably for different control conditions (Fig. 1Giii, Fig. 2Giii, Fig. 6C, Fig. 9A).

3. As noted by the authors in a previous publication (Hanes et al. 2020), statistical analysis of CDFs suffers from n-inflation. In addition, the quantile sampling method chosen violates an important assumption of the K-S test. Indeed, p-values for these comparisons are typically several orders of magnitude smaller. Given that the statistical N most likely corresponds to the number of cultures (see, e.g., https://doi.org/10.1371/journal.pbio.2005282), CDF comparisons are not informative and should thus not be used to draw conclusions from the data. The plots can be informative, though.

4. How does recoding noise and the mEPSC amplitude threshold affect "divergent scaling"?

5. What is the justification for the line fits of the ratio data/how was the fit range chosen?

6. TTX application induces a significant increase in mEPSC amplitude in Rab3A-/- mice in two out of three data sets (Figs. 1 and 9). Hence, the major conclusion that Rab3A is required for homeostatic scaling is only partially supported by the data.

7. Line 289: A comparison of p-values between conditions does not allow any meaningful conclusions.

8. There is a significant increase in baseline mEPSC amplitude in Rab3AEbd/Ebd (15 pA) vs. Rab3Aebd/+ (11 pA) cultures, but not in Rab3A-/- (13.6 pA) vs. Rab3A+/- (13.9 pA). Although the nature of scaling was different between Rab3AEbd/Ebd vs. Rab3AEbd/+, and Rab3AEbd/Ebd with vs. without TTX, the question arises whether the increase in mEPSC amplitude in Rab3AEbd/Ebd is Rab3A dependent. Could a Rab3A independent mechanism occlude scaling?

9. Figure 4: NASPM appears to have a stronger effect on mEPSC frequency in the TTX condition vs. control (-40% vs. -15%). A larger sample size might be necessary to draw definitive conclusions on the contribution of Ca2+-permeable AMPARs.

10. The authors discuss previous papers showing changes in VGLUT1 intensity. Was VGLUT intensity altered in the stainings presented in the manuscript?

11. The change in GluA2 area or fluorescence intensity upon TTX treatment in controls is modest. How does the GluA2 integral change?

12. The quantitative comparison between physiology and microscopy data is problematic. The authors report a mismatch in ratio values between the smallest mEPSC amplitudes and smallest GluA2 receptor cluster sizes (l. 464; Figure 8). Is this comparison affected by the fluorescence intensity threshold? What was the rationale for a threshold of 400 a.u. or 450 a.u.? How does this threshold compare to the mEPSC threshold of 3 pA? The conclusion that an increase in AMPAR levels is not fully responsible for the observed mEPSC increase is mainly based on the rank-order analysis of GluA2 intensity, yielding a slope of ~0.9. There are several points to consider here: (i) GluA2 fluorescence intensity did increase on average, as did GluA2 cluster size. (ii) The increase in GluA2 cluster size is very similar to the increase in mEPSC amplitude (each approx. 18-20%). (iii) Are there any reports that fluorescence intensity values are linearly reporting mEPSC amplitudes (in this system)? Antibody labelling efficiency, and false negatives of mEPSC recordings may influence the results. The latter was already noted by the authors. (iv) It is not entirely clear if their imaging experiments will sample from all synapses. Other AMPAR subtypes than GluA2 could contribute, as could kainate or NMDA receptors.

Furthermore, the statement "complete lack of correspondence of TTX/CON ratios" is not supported by the data presented (l. 515ff). First, under the assumption that no scaling occurs in Rab3A-/- , the TTX/CON ratios show a 20-30% change, which indicates the variation of this readout. Second, the two examples shown in Figure 8 for Rab3A+/+ are actually quite similar (culture #1 and #2), particularly when ignoring the leftmost section of the data, which is heavily affected by the raw values approaching zero.

13. Figure 7A: TTX CDF was shifted to smaller mEPSC amplitude values in Rab3A-/- cultures. How can this be explained?

Reviewer #2 (Public Review):

In this study, Koesters et al. investigated whether Rab3A, a small GTPase that regulates synaptic vesicle fusion pore opening, is required for excitatory synaptic scaling in response to TTX-induced activity suppression in dissociated mouse cortical neuronal culture. They first show that, while pyramidal neurons from wild-type (WT) littermates show normal synaptic scaling in response to 48h of TTX treatment (~30% increase in the mean mEPSC amplitude), those from two different mouse lines with either deletion (Rab3A-/-) or loss-of-function mutation of Rab3A (Rab3AEbd/Ebd) fail to engage this homeostatic compensation. They perform cumulative distribution analysis to show that the mEPSC population has gone through divergent scaling in WT neurons. Similarly, this phenomenon is absent in neurons from the two Rab3A mouse lines. They further demonstrate that GluA2-containing AMPARs likely account for the increase in mEPSC amplitudes by comparing measurements before and after washing in blockers specific for GluA2-lacking AMPARs. Subsequently, they perform electrophysiology and immunohistochemistry side by side for WT neurons from the same culture following TTX treatment, and find that both mEPSC amplitudes and GluA2 cluster sizes have shifted towards higher values, while GluA2 cluster intensity remains unchanged. Importantly, all these homeostatic compensations are absent in Rab3A-/- neurons. Finally, they mix neurons and astrocyte feeders either from WT or Rab3A-/- mice, which reveals that neuronal but not astrocytic Rab3A knockout leads to impaired scaling up of mEPSCs. They conclude that Rab3A is required for homeostatic scaling up of mEPSC amplitude in cortical neurons, most likely from the presynaptic side.

Although the authors have raised an interesting question, their conclusion is not well supported by the data presented. I list my technical and conceptual concerns below.

Technical concerns:

1. The culture condition is questionable. The authors saw no NMDAR current present during spontaneous recordings, which is worrisome since NMDARs should be active in cultures with normal network activity (Watt et al., 2000; Sutton et al., 2006). It is important to ensure there is enough spiking activity before doing any activity manipulation. Similarly, it is also unknown whether spiking activity is normal in Rab3A KO/Ebd neurons.

2. Selection of mEPSC events is not conducted in an unbiased manner. Manually selecting events is insufficient for cumulative distribution analysis, where small biases could skew the entire distribution. Since the authors claim their ratio plot is a better method to detect the uniformity of scaling than the well-established rank-order plot, it is important to use an unbiased population to substantiate this claim.

3. Immunohistochemistry data analysis is problematic. The authors only labeled dendrites without doing cell-fills to look at morphology, so it is questionable how they differentiate branches from pyramidal neurons and interneurons. Since glutamatergic synapses on these two types of neuron scale in the opposite directions, it is crucial to show that only pyramidal neurons are included for analysis.

Conceptual concerns:

The only novel finding here is the implicated role for Rab3A in synaptic scaling, but insights into mechanisms behind this observation are lacking. The author claims that Rab3A likely regulates scaling from the presynaptic side, yet there is no direct evidence from data presented. In its current form, this study's contribution to the field is very limited.

1. Their major argument for this is that homeostatic effects on mEPSC amplitudes and GluA2 cluster sizes do not match. This is inconsistent with reports from multiple labs showing that upscaling of mEPSC amplitude and GluA2 accumulation occur side by side during scaling (Ibata et al., 2008; Pozo et al., 2012; Tan et al., 2015; Silva et al., 2019). Further, because the acquisition and quantification methods for mEPSC recordings and immunohistochemistry imaging are entirely different (each with its own limitations in signal detection), it is not convincing that the lack of proportional changes must signify a presynaptic component.

2. The authors also speculate in the discussion that presynaptic Rab3A could be interacting with retrograde BDNF signaling to regulate postsynaptic AMPARs. Without data showing Rab3A-dependent presynaptic changes after TTX treatment, this argument is not compelling. In this retrograde pathway, BDNF is synthesized in and released from dendrites (Jakawich et al., 2010; Thapliyal et al., 2022), and it is entirely possible for postsynaptic Rab3A to interfere with this process cell-autonomously.

3. The authors propose that a change in AMPAR subunit composition from GluA2-containing ones to GluA1 homomers may account for the distinct changes in mEPSC amplitudes and GluA2 clusters. However, their data from the Naspm wash-in experiments clearly show that GluA1 homomer contributions have not changed before and after TTX treatment.

Ibata K, Sun Q, Turrigiano GG (2008) Rapid synaptic scaling induced by changes in postsynaptic firing. Neuron 57:819-826.

Jakawich SK, Nasser HB, Strong MJ, McCartney AJ, Perez AS, Rakesh N, Carruthers CJL, Sutton MA (2010) Local Presynaptic Activity Gates Homeostatic Changes in Presynaptic Function Driven by Dendritic BDNF Synthesis. Neuron 68:1143-1158.

Pozo K, Cingolani LA, Bassani S, Laurent F, Passafaro M, Goda Y (2012) β3 integrin interacts directly with GluA2 AMPA receptor subunit and regulates AMPA receptor expression in hippocampal neurons. Proceedings of the National Academy of Sciences 109:1323-1328.

Silva MM, Rodrigues B, Fernandes J, Santos SD, Carreto L, Santos MAS, Pinheiro P, Carvalho AL (2019) MicroRNA-186-5p controls GluA2 surface expression and synaptic scaling in hippocampal neurons. Proceedings of the National Academy of Sciences 116:5727-5736.

Sutton MA, Ito HT, Cressy P, Kempf C, Woo JC, Schuman EM (2006) Miniature Neurotransmission Stabilizes Synaptic Function via Tonic Suppression of Local Dendritic Protein Synthesis. Cell 125:785-799.

Tan HL, Queenan BN, Huganir RL (2015) GRIP1 is required for homeostatic regulation of AMPAR trafficking. Proceedings of the National Academy of Sciences 112:10026-10031.

Thapliyal S, Arendt KL, Lau AG, Chen L (2022) Retinoic acid-gated BDNF synthesis in neuronal dendrites drives presynaptic homeostatic plasticity. eLife 11:e79863.

Watt AJ, Rossum MCW van, MacLeod KM, Nelson SB, Turrigiano GG (2000) Activity Coregulates Quantal AMPA and NMDA Currents at Neocortical Synapses. Neuron 26:659-670.

Reviewer #3 (Public Review):

Summary: The authors clearly demonstrate the Rab3A plays a role in HSP at excitatory synapses, with substantially less plasticity occurring in the Rab3A KO neurons. There is also no apparent HSP in the Earlybird Rab3A mutation, although baseline synaptic strength seems already elevated. In this context, it is unclear if the plasticity is absent or just occluded by a ceiling effect due the synapses already being strengthened. The authors do appropriately discuss both options. There are also differences in genetic background between the Rab3A KO and Earlybird mutants that could also impact the results, which are also noted. The authors have solid data showing that Rab3A is unlikely to be active in astrocytes, Finally, they attempt to study the linkage between synaptic strength during HSP and AMPA receptor trafficking, and conclude that trafficking is largely not responsible for the changes in synaptic strength.

Strengths: This work adds another player into the mechanisms underlying an important form of synaptic plasticity. The plasticity is only reduced, suggesting Rab3A is only partially required and perhaps multiple mechanisms contribute. The authors speculate about some possible novel mechanisms.

Weaknesses: However, the rather strong conclusions on the dissociation of AMPAR trafficking and synaptic response are made from somewhat weaker data. The key issue is the GluA2 immunostaining in comparison with the mESPC recordings. Their imaging method involves only assessing puncta clearly associated with a MAP2 labeled dendrite. This is a small subset of synapses, judging from the sample micrographs (Fig 5). To my knowledge, this is a new and unvalidated approach that could represent a particular subset of synapses not representative of the synapses contributing to the mEPSC change (they are also sampling different neurons for the two measurements; an additional unknown detail is how far from the cell body were the analyzed dendrites for immunostaining). While the authors acknowledge that a sampling issue could explain the data, they still use this data to draw strong conclusions about the lack of AMPAR trafficking contribution to the mEPSC amplitude change. This apparent difference may be a methodological issue rather than a biological one, and at this point it is impossible to differentiate these. It will unfortunately be difficult to validate their approach. Perhaps if they were to drive NMDA-dependent LTD or chemLTP, and show alignment of the imaging and ephys, that would help. More helpful would be recordings and imaging from the same neurons but this is challenging. Sampling from identified synapses would of course be ideal, perhaps from 2P uncaging combined with SEP-labeled AMPARs, but this is more challenging still. But without data to validate the method, it seems unwarranted to make such strong conclusions such as that AMPAR trafficking does not underlie the increase in mEPSC amplitude, given the previous data supporting such a model.

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a frequency effect that is quite unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. It is also unclear why the authors argue this proves that the NASPM was at an effective concentration (lines 399-400). Further, the amplitude data show a strong trend towards smaller amplitude. The p value for both control and TTX neurons was 0.08 - it is very difficult to argue that there is no effect. And the decrease is larger in the TTX neurons. Considering the strong claims for a pre-synaptic locus and the use of this data to justify only looking at GluA2 by immunostaining, these data do not offer much support of the conclusions. Between the sampling issues and perhaps looking at the wrong GluA subunit, it seems premature to argue that trafficking is not a contributor to the mEPSC amplitude change, especially given the substantial support for that hypothesis. Further, even if trafficking is not the major contributor, there could be shifts in conductance (perhaps due to regulation of auxiliary subunits) that does not necessitate a pre-synaptic locus. While the authors are free to hypothesize such a mechanism, it would be prudent to acknowledge other options and explanations.

The frequency data are missing from the paper, with the exception of the NASPM dataset. The mEPSC frequencies should be reported for all experiments, particularly given that Rab3A is generally viewed as a pre-synaptic protein regulating release. Also, in the NASPM experiments, the average frequency is much higher in the TTX treated cultures. Is this statistically above control values?

Unaddressed issues that would greatly increase the impact of the paper:

1. Is Rab3A acting pre-synaptically, post-synaptically or both? The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where it is acting (pre or post) would aid substantially in understanding its role (and particularly the hypothesized and somewhat novel idea that the amount of glutamate released per vesicle is altered in HSP). They could use sparse knock-down of Rab3A, or simply mix cultures from KO and WT mice (with appropriate tags/labels). The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. The more support for their suggestion of a pre-synaptic site of control, the better.

2. Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs and/or a decrease of GABA-packaging in vesicles (ie the opposite of whatever is happening at excitatory synapses). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling; an effect only at excitatory synapses would argue for a more specific role just at these synapses.

Author response:

The following is the authors’ response to the original reviews.

The detailed, thorough critique provided by the three reviewers is very much appreciated. We believe the manuscript is greatly improved by the changes we have made based on those reviews. The major changes are described below, followed by a point by point response.

Major Changes:

(1) We revised our model (old Fig. 10; new Fig. 9) to keep the explanation focused on the data shown in the current study. Specifically, references to GTP/GDP states of Rab3A and changes in the presynaptic quantum have been removed and the mechanisms depicted are confined to pre- or post-synaptic Rab3A participating in either controlling release of a trophic factor that regulates surface GluA2 receptors (pre- or postsynaptic) or directly affecting fusion of GluA2-receptor containing vesicles (postsynaptic).

(2) We replaced all cumulative density function plots and ratio plots, based on multiple quantile samples per cell, with box plots of cell means. This affects new Figures 1, 2, 3, 5, 6, 7 and 8. All references to “scaling,” “divergent scaling,” or “uniform scaling,” have been removed. New p values for comparison of means are provided above every box plot in Figures 1, 2, 3, 5, 6, 7 and 8. The number of cultures is provided in the figure legends.

(3) We have added frequency to Figures 1, 2 and 8. Frequency values overall are more variable, and the effect of activity blockade less robust, than for mEPSC amplitudes. We have added text indicating that the increase in frequency after activity blockade was significant in neurons from cultures prepared from WT in the Rab3A+/- colony but not cultures prepared from KO mice (Results, lines 143 to 147, new Fig. 1G. H). The TTX-induced increase in frequency was significant in the NASPM experiments before NASPM, but not after NASPM (Results, lines 231 to 233, new Fig. 3, also cultures from WT in Rab3A+/- colony). The homeostatic plasticity effect on frequency did not reach significance in WT on WT glia cultures or

WT on KO glia cultures, possibly due to the variability of frequency, combined with smaller sample sizes (Results, lines 400 to 403, new Fig. 8). In the cultures prepared from WT mice in the Rab3A+/Ebd colony, there was a trend towards higher frequency after TTX that did not reach statistical significance, and in cultures prepared from mutant mice, the p value was large, suggesting disruption of the effect, which appears to be due to an increase in frequency in untreated cultures, similar to the behavior of mEPSC amplitudes in neurons from mutant mice (Results, lines 161-167). In sum, the effect of activity on frequency requires Rab3A and Ca2+-permeable receptors, and is mimicked by the presence of the Rab3A Earlybird mutant. We have also added a discussion of these results (Discussion, lines 427-435).

(4) In the revised manuscript we have added analysis of VGLUT1 levels for the same synaptic sites that we previously analyzed GluA2 levels, and these data are described in Results, lines 344 to 371, and appear in new Table 2. In contrast to previous studies, we did not find any evidence for an increase in VGLUT1 levels after activity blockade. We reviewed those studies to determine whether there might be differences in the experimental details that could explain the lack of effect we observed. In (De Gois et al., 2005), the authors measured mRNA and performed western blots to show increases in VGLUT1 after TTX treatment in older rat cortical cultures (DIV 19). The study performs immunofluorescence imaging of VGLUT1 but only after bicuculline treatment (it decreases), not after TTX treatment. In (Wilson et al.,

2005), the hippocampal cultures are treated with AP5, not TTX, and the VGLUT1 levels in immunofluorescence images are reported relative to synapsin I. That the type of activity blockade matters is illustrated by the failure of Wilson and colleagues to observe a consistent increase in VGLUT1/Synapsin ratio in cultures treated with AMPA receptor blockade (NBQX; supplementary information). These points have been added to the Discussion, lines 436 to 447.)

Reviewer #1:

(1) (model…is not supported by the data), (2) (The analysis of mEPSC data using quantile sampling…), (3) (…statistical analysis of CDFs suffers from n-inflation…), (4) (How does recording noise and the mEPSC amplitude threshold affect “divergent scaling?”) (5) (…justification for the line fits of the ratio data…), (7) (A comparison of p-values between conditions….) and (10) (Was VGLUT intensity altered in the stainings presented in the manuscript?)

The major changes we made, described above, address Reviewer #1’s points. The remaining points are addressed below.

(6) TTX application induces a significant increase in mEPSC amplitude in Rab3A-/- mice in two out of three data sets (Figs. 1 and 9). Hence, the major conclusion that Rab3A is required for homeostatic scaling is only partially supported by the data.

The p values based on CDF comparisons were problematic, but the point we were making is that they were much larger for amplitudes measured in cultures prepared from Rab3A-/- mice (Fig. 1, p = 0.04) compared to those from cultures prepared from Rab3A+/+ mice (Fig. 1, p = 4.6 * 10-4). Now that we are comparing means, there are no significant TTX-induced effects on mEPSC amplitudes for Rab3A-/- data. However, acknowledging that some increase after activity blockade remains, we describe homeostatic plasticity as being impaired or not significant, rather than abolished, by loss of Rab3A, (Abstract, lines 37 to 39; Results, lines 141 to 143; Discussion, lines 415 to 418).

(8) There is a significant increase in baseline mEPSC amplitude in Rab3AEbd/Ebd (15 pA) vs. Rab3AEbd/+ (11 pA) cultures, but not in Rab3A-/- (13.6 pA) vs. Rab3A+/- (13.9 pA). Although the nature of scaling was different between Rab3AEbd/Ebd vs. Rab3AEbd/+ and Rab3AEbd/Ebd with vs. without TTX, the question arises whether the increase in mEPSC amplitude in Rab3AEbd/Ebd is Rab3A dependent. Could a Rab3A independent mechanism occlude scaling?

The Reviewer is concerned that the increase in mEPSC amplitude in the presence of the Rab3A point mutant may be through a ‘non-Rab3A’ mechanism (a concern raised by the lack of such effect in cultures from the Rab3A-/- mice), and secondly, that the already large mEPSC cannot be further increased by the homeostatic plasticity mechanism. It must always be considered that a mutant with an altered genetic sequence may bind to novel partners, causing activities that would not be either facilitated or inhibited by the original molecule. We have added this caveat to Results, lines 180 to 186 We added that a number of other manipulations, implicating individual molecules in the homeostatic mechanism, have caused an increase in mEPSC amplitude at baseline, potentially nonspecifically occluding the ability of activity blockade to induce a further increase (Results lines 186 to 189). Still, it is a strong coincidence that the novel activity of the mutant Rab3A would affect mEPSC amplitude, the same characteristic that is affected by activity blockade in a Rab3A dependent manner, a point which we added to Results, lines 189 to 191.

(9) Figure 4: NASPM appears to have a stronger effect on mEPSC frequency in the TTX condition vs. control (-40% vs -15%). A larger sample size might be necessary to draw definitive conclusions on the contribution of Ca2+-permeable AMPARs.

Our results, even with the modest sample size of 11 cells, are clear: NASPM does not disrupt the effect of TTX treatment on mEPSC amplitude (new Fig. 3A). It also looks like there is a greater magnitude effect of NAPSM on frequency in TTX-treated cells; we note this, but point out that nevertheless, these mEPSCs are not contributing to the increase in mEPSC amplitude (Results, lines 238-241).

(11) The change in GluA2 area or fluorescence intensity upon TTX treatment in controls is modest. How does the GluA2 integral change?

We had reported that GluA2 area showed the most prominent increase following activity blockade, with intensity changing very little. When we examined the integral, it closely matched the change in area. We have added the values for integral to new Fig. 5 D, H; new Fig. 6 A-C; new Fig. 7 A-C and new Table 1 (for GluA2) and new Table 2 (for VGLUT1). These results are described in the text in the following places: Results, lines 289-292; 298-299; 311-319; 328-324). For VGLUT1, both area and intensity changed modestly, and the integral appeared to be a combination of the two, being higher in magnitude and resulting in smaller p values than either area or intensity (Results, lines 344-348; 353-359; new Table 2).

(12) The quantitative comparison between physiology and microscopy data is problematic. The authors report a mismatch in ratio values between the smallest mEPSC amplitudes and the smallest GluA2 receptor cluster sizes (l. 464; Figure 8). Is this comparison affected by the fluorescence intensity threshold? What was the rationale for a threshold of 400 a.u. or 450 a.u.? How does this threshold compare to the mEPSC threshold of 3 pA.

This concern is partially addressed by no longer comparing the rank ordered mEPSC amplitudes with the rank ordered GluA2 receptor characteristics. We had used multiple thresholds in the event that an experiment was not analyzable with the chosen threshold (this in fact happened for VGLUT1, see end of this paragraph). We created box plots of the mean GluA2 receptor cluster size, intensity and integral, for experiments in which we used all three thresholds, to determine if the effect of activity blockade was different depending on which threshold was applied, and found that there was no obvious difference in the results (Author response image 1). Nevertheless, since there is no need to use a different threshold for any of the 6 experiments (3 WT and 3KO), for new Figures 5, 6 and 7 we used the same threshold for all data, 450; described in Methods, lines 746 to 749. For VGLUT1 levels, it was necessary to use a different threshold for Rab3A+/+ Culture #1 (400), but a threshold of 200 for the other five experiments (Methods, lines 751-757). The VGLUT1 immunofluorescent sites in Culture #1 had higher levels overall, and the low threshold caused the entire AOI to be counted as the synapse, which clearly included background levels outside of the synaptic site. Conversely, to use a threshold of 400 on the other experiments meant that the synaptic site found by the automated measurement tool was much smaller that what was visible by eye. In our judgement it would have been meaningless to adhere to a single threshold for VGLUT1 data.

Using different thresholds does not substantially alter GluA2 receptor cluster size data.

A) Rab3A+/+ Culture #1, size data for three different thresholds, depicted above each graph. B) Rab3A+/+ Culture #2, size data for three different thresholds, depicted above each graph. Note scale bar in A is different from B, to highlight differences for different thresholds. (Culture #3 was only analyzed with 450 threshold).

The conclusion that an increase in AMPAR levels is not fully responsible for the observed mEPSC increase is mainly based on the rank-order analysis of GluA2 intensity, yielding a slope of ~0.9. There are several points to consider here: (i) GluA2 fluorescence intensity did increase on average, as did GluA2 cluster size.

(ii) The increase in GluA2 cluster size is very similar to the increase in mEPSC amplitude (each approx. 1820%). (iii) Are there any reports that fluorescence intensity values are linearly reporting mEPSC amplitudes (in this system)? Antibody labelling efficiency, and false negatives of mEPSC recordings may influence the results. The latter was already noted by the authors.

Our comparison between mEPSC amplitude and GluA2 receptor cluster characteristics has been reexamined in the revised version using means rather than rank-ordered data in rank-order plots or ratio plots. Importantly, all of these methods revealed that in one out of three WT cultures (Culture #3) GluA2 receptor cluster size (old Fig. 8, old Table 1; new Fig. 6, new Table 1), intensity and integral (new Fig. 6, new Table 1) values decreased following activity blockade while in the same culture, mEPSC amplitudes increased. It is based on this lack of correspondence that we conclude that increases in mEPSC amplitude are not fully explained by increases in GluA2 receptors, and suggest there may be other contributors. These points are made in the Abstract (lines 108-110); Results (lines 319 to 326; 330337; 341-343) and the Discussion (lines 472 to 474). To our knowledge, there are not any reports that quantitatively compare receptor levels (area, intensity or integrals) to mEPSC amplitudes in the same cultures. We examined the comparisons very closely for 5 studies that used TTX to block activity and examined receptor levels using confocal imaging at identified synapses (Hou et al., 2008; Ibata et al., 2008; Jakawich et al., 2010a; Xu and Pozzo-Miller, 2017; Dubes et al., 2022). We were specifically looking for whether the receptor data were more variable than the mEPSC amplitude data, as we found. However, for 4 of the studies, sample sizes were very different so that we cannot simply compare the p values. Below is a table of the comparisons.

In Xu 2017 the sample sizes are close enough that we feel comfortable concluding that the receptor data were slightly more variable (p < 0.05) than mEPSC data (p<0.01) but recognize that it is speculative to say our finding has been confirmed. A discussion of these articles is in Discussion, lines 456-474.

(iv) It is not entirely clear if their imaging experiments will sample from all synapses. Other AMPAR subtypes than GluA2 could contribute, as could kainite or NMDA receptors.

While our imaging data only examined GluA2, we used the application of NASPM to demonstrate Ca2+permeable receptors did not contribute quantitatively to the increase in mEPSC amplitude following TTX treatment. Since GluA3 and GluA4 are also Ca2+-permeable, the findings in new Figure 3 (old Fig. 4) likely rule out these receptors as well. There are also reports that Kainate receptors are Ca2+-permeable and blocked by NASPM (Koike et al., 1997; Sun et al., 2009), suggesting the NASPM experiment also rules out the contribution of Kainate receptors. Finally, given our recording conditions, which included normal magnesium levels in the extracellular solution as well as TTX to block action-potential evoked synaptic transmission, NMDA receptors would not be available to contribute currents to our recordings due to block by magnesium ions at resting Vm. These points have been added to the Methods section, lines 617 to 677 (NMDA); 687-694 (Ca2+-permeable AMPA receptors and Kainate receptors).

Furthermore, the statement “complete lack of correspondence of TTX/CON ratios” is not supported by the data presented (l. 515ff). First, under the assumption that no scaling occurs in Rab3A-/-, the TTX/CON ratios show a 20-30% change, which indicates the variation of this readout. Second, the two examples shown in Figure 8 for Rab3A+/+ are actually quite similar (culture #1 and #2, particularly when ignoring the leftmost section of the data, which is heavily affected by the raw values approaching zero.

We are no longer presenting ratio plots in the revised manuscript, so we do not base our conclusion that mEPSC amplitude data is not always corresponding to GluA2 receptor data on the difference in behavior of TTX/CON ratio values, but only on the difference in direction of the TTX effect in one out of three cultures. We agree with the reviewer that the ratio plots are much more sensitive to differences between control and treated values than the rank order plot, and we feel these differences are important, for example, there is still a homeostatic increase in the Rab3A-/- cultures, and the effect is still divergent rather than uniform. But the comparison of ratio data will be presented elsewhere.

(13) Figure 7A: TTX CDF was shifted to smaller mEPSC amplitude values in Rab3A-/- cultures. How can this be explained?

While this result is most obvious in CDF plots, we still observe a trend towards smaller mEPSC amplitudes after TTX treatment in two of three individual cultures prepared from Rab3A-/- mice when comparing means (new Fig. 7, Table 1) which did not reach statistical significance for the pooled data (new Fig. 5, new Table 1). There was not any evidence of this decrease in the larger data set (new Fig. 1) nor for Rab3A-/- neurons on Rab3A+/+ glia (new Fig. 8). Given that this effect is not consistent, we did not comment on it in the revised manuscript. It may be that there is a non-Rab3A-dependent mechanism that results in a decrease in mEPSC amplitude after activity blockade, which normally pulls down the magnitude of the activity-dependent increase typically observed. But studying this second component would be difficult given its magnitude and inconsistent presentation.

Reviewer #1 (Recommendations For the Authors):

(1) Abstract, last sentence: The conclusion of the present manuscript should be primarily based on the results presented. At present, it is mainly based on a previous publication by the authors.

We have revised the last sentence to reflect actual findings of the current study (Abstract, lines 47 to 49).

(2) Line 55: “neurodevelopmental”

This phrase has been removed.

(3) Line 56: “AMPAergic” should be replaced by AMPAR-mediated

This sentence was removed when all references to “scaling” were removed; no other instances of “AMPAergic” are present.

(4) Figure 9: The use of BioRender should be disclosed in the Figure Legend.

We used BioRender in new Figures 3, 7 and 8, and now acknowledge BioRender in those figure legends.

(5) Figure legends and results: The number of cultures should be indicated for each comparison.

Number of cultures has been added to the figure legends.

(6) Line 289: A comparison of p-values between conditions does not allow any meaningful conclusions.

Agreed, therefore we have removed CDFs and the KS test comparison p values. All comparisons in the revised manuscript are for cell means.

(7) Line 623ff: The argument referring to NMJ data is weak, given that different types of receptors are involved.

We still think it is valid to point out that Rab3A is required for the increase in mEPC at the NMJ but that ACh receptors do not increase (Discussion, lines 522 to 525). We are not saying that postsynaptic receptors do not contribute in cortical cultures, only that there could be another Rab3A-dependent mechanism that also affects mEPSC amplitude.

(8) Plotting data points outside of the ranges should be avoided (e.g., Fig. 2Giii, 7F).

These two figures are no longer present in the revised manuscript. In revising figures, we made sure no other plots have data points outside of the ranges.

(9) The rationale for investigating Rab3AEbd/Ebd remains elusive and should be described.

A rationale for investigating Rab3AEbd/Ebd is that if the results are similar to the KO, it strengthens the evidence for Rab3A being involved in homeostatic synaptic plasticity. In addition, since its phenotype of early awakening was stronger than that demonstrated in Rab3A KO mice (Kapfhamer et al., 2002), it was possible we would see a more robust effect. These points have been added to the Results, lines 118 to 126.

(10) Figures 3 and 4, as well as Figure 5 and 6 could be merged.

In the revised version, Figure 3 has been eliminated since its main point was a difference in scaling behavior. Figure 4 has been expanded to include a model of how NASPM could reduce frequency (new Fig. 3.) Images of the pyramidal cell body have been added to Figure 5 (new Fig. 4), and Figure 6 has been completely revised and now includes pooled data for both Rab3A+/+ and Rab3A-/- cultures, for mEPSC amplitude, GluA2 receptor cluster size, intensity and integral.

(11) Figure 5: The legend refers to MAP2, but this is not indicated in the figure.

MAP2 has now been added to the labels for each image and described in the figure legend (new Fig. 4).

Reviewer #2:

Technical concerns:

(1) The culture condition is questionable. The authors saw no NMDAR current present during spontaneous recordings, which is worrisome since NMDARs should be active in cultures with normal network activity (Watt et al., 2000; Sutton et al., 2006). It is important to ensure there is enough spiking activity before doing any activity manipulation. Similarly it is also unknown whether spiking activity is normal in Rab3AKO/Ebd neurons.

In the studies cited by the reviewer, NMDA currents were detected under experimental conditions in which magnesium was removed. In our recordings, we have normal magnesium (1.3 mM) and also TTX, which prevents the necessary depolarization to allow inward current through NMDA receptors. This point has been added to our Methods, lines 674 to 677. We acknowledge we do not know the level of spiking in cultures prepared from Rab3A+/+, Rab3A-/- or Rab3A_Ebd/Ebd_ mice. Given the similar mEPSC amplitude for untreated cultures from WT and KO studies, we think it unlikely that activity was low in the latter, but it remains a possibility for untreated cultures from Rab3A_Ebd/Ebd_ mice, where mEPSC amplitude was increased. These points are added to the Methods, lines 615 to 622.

(2) Selection of mEPSC events is not conducted in an unbiased manner. Manually selecting events is insufficient for cumulative distribution analysis, where small biases could skew the entire distribution. Since the authors claim their ratio plot is a better method to detect the uniformity of scaling than the well-established rank-order plot, it is important to use an unbiased population to substantiate this claim.

We no longer include any cumulative distributions or ratio plot analysis in the revised version. We have added the following text to Methods, lines 703 to 720:

“MiniAnalysis selects many false positives with the automated feature when a small threshold amplitude value is employed, due to random fluctuations in noise, so manual re-evaluation of the automated process is necessary to eliminate false positives. If the threshold value is set high, there are few false positives but small amplitude events that visually are clearly mEPSCs are missed, and manual re-evaluation is necessary to add back false negatives or the population ends up biased towards large mEPSC amplitudes. As soon as there is a manual step, bias is introduced. Interestingly, a manual reevaluation step was applied in a recent study that describes their process as ‘unbiased (Wu et al., 2020). In sum, we do not believe it is currently possible to perform a completely unbiased detection process. A fully manual detection process means that the same criterion (“does this look like an mEPSC?”) is applied to all events, not just the false positives, or the false negatives, which prevents the bias from being primarily at one end or the other of the range of mEPSC amplitudes. It is important to note that when performing the MiniAnalysis process, the researcher did not know whether a record was from an untreated cell or a TTX-treated cell.”

(3) Immunohistochemistry data analysis is problematic. The authors only labeled dendrites without doing cell-fills to look at morphology, so it is questionable how they differentiate branches from pyramidal neurons and interneurons. Since glutamatergic synapse on these two types of neuron scale in the opposite directions, it is crucial to show that only pyramidal neurons are included for analysis.

We identified neurons with a pyramidal shape and a prominent primary dendrite at 60x magnification without the zoom feature. This should have been made clear in the description of imaging. We have added an image of the two selected cells to our figure of dendrites (old Fig. 5, new Fig. 4), and described this process in the Methods, lines 736 to 739, and Results, lines 246 to 253. Given the morphology of the neurons selected it is highly unlikely that the dendrites we analyzed came from interneurons.

Conceptual Concerns

The only novel finding here is the implicated role for Rab3A in synaptic scaling, but insights into mechanisms behind this observation are lacking. The authors claim that Rab3A likely regulates scaling from the presynaptic side, yet there is no direct evidence from data presented. In its current form, this study’s contribution to the field is very limited.

We have demonstrated that loss of Rab3A and expression of a Rab3A point mutant disrupt homeostatic plasticity of mEPSC amplitudes, and that in the absence of Rab3A, the increase in GluA2 receptors at synaptic sites is abolished. Further, we show that this effect cannot be through release of a factor, like TNFα, from astrocytes. In the new version, we add the finding that VGLUT1 is not increased after activity blockade, ruling out this presynaptic factor as a contributor to homeostatic increases in mEPSC amplitude. We show for the first time by examining mEPSC amplitudes and GluA2 receptors in the same cultures that the increases in GluA2 receptors are not as consistent as the increases in mEPSC amplitude, suggesting the possibility of another contributor to homeostatic increases in mEPSC amplitude. We first proposed this idea in our previous study of Rab3A-dependent homeostatic increases in mEPC amplitudes at the mouse neuromuscular junction. In sum, we dispute that there is only one novel finding and that we have no insights into mechanism. We acknowledge that we have no direct evidence for regulation from the presynaptic side, and have removed this claim from the revised manuscript. We have retained the Discussion of potential mechanisms affecting the presynaptic quantum and evidence that Rab3A is implicated in these mechanisms (vesicle size, fusion pore kinetics; Discussion, lines 537 to 563). One way to directly show that the amount of transmitter released for an mEPSC has been modified after activity blockade is to demonstrate that a fast off-rate antagonist has become less effective at inhibiting mEPSCs (because the increased glutamate released out competes it; see (Liu et al., 1999) and (Wilson et al., 2005) for example experiments). This set of experiments is underway but will take more time than originally expected, because we are finding surprisingly large decreases in frequency, possibly the result of mEPSCs with very low glutamate concentration that are completely inhibited by the dose used. Once mEPSCs are lost, it is difficult to compare the mEPSC amplitude before and after application of the antagonist. Therefore we intend to include this experiment in a future report, once we determine the reason for the frequency reduction, or, can find a dose where this does not occur.

(1) Their major argument for this is that homeostatic effects on mEPSC amplitudes and GluA2 cluster sizes do not match. This is inconsistent with reports from multiple labs showing that upscaling of mEPSC amplitude and GluA2 accumulation occur side by side during scaling (Ibata et al., 2008; Pozo et al., 2012; Tan et al., 2015; Silva et al., 2019). Further, because the acquisition and quantification methods for mEPSC recordings and immunohistochemistry imaging are entirely different (each with its own limitations in signal detection), it is not convincing that the lack of proportional changes must signify a presynaptic component.

Within the analyses in the revised manuscript, which are now based only on comparison of cell/dendrite means, we find a very good match in the magnitude of increase for the pooled data of mEPSC amplitudes and GluA2 receptor cluster sizes (+19.7% and +20.0% respectively; new Table 1). However, when looking at individual cultures, we had one of three WT cultures in which mEPSC amplitude increased 17.2% but GluA2 cluster size decreased 9.5%. This result suggests that while activity blockade does lead to an increase in GluA2 receptors after activity blockade, the effect is more variable than that for mEPSC amplitude. We went back to published studies to see if this has been previously observed, but found that it was difficult to compare because the sample sizes were different for the two characteristics (see Author response table 1). We included these particular 5 studies because they use the same treatment (TTX), examine receptors using imaging of identified synaptic sites, and record mEPSCs in their cultures (although the authors do not indicate that imaging and recordings are done simultaneously on the same cultures.) Only one of the studies listed by the Reviewer is in our group (Ibata et al., 2008). The study by (Tan et al., 2015) uses western blots to measure receptors; the study by (Silva et al., 2019) blocks activity using a combination of AMPA and NMDA receptor blockers; the study by (Pozo et al., 2012) correlates mEPSC amplitude changes with imaging but not in response to activity blockade, instead for changing the expression of GluA2. While it may seem like splitting hairs to reject studies that use other treatment protocols, there is ample evidence that the mechanisms of homeostatic plasticity depend on how activity was altered, see the following studies for several examples of this (Sutton et al., 2006; Soden and Chen, 2010; Fong et al., 2015). A discussion of the 5 articles we selected is in the revised manuscript, Discussion, lines 456 to 474. In sum, we provide evidence that activity blockade is associated with an overall increase in GluA2 receptors; what we propose is that this increase, being more variable, does not fully explain the increase in mEPSC amplitude. However, we acknowledge that the disparity could be explained by the differences in limitations of the two methods (Discussion, lines 469-472).

(2) The authors also speculate in the discussion that presynaptic Rab3A could be interacting with retrograde BDNF signaling to regulate postsynaptic AMPARs. Without data showing Rab3A-dependent presynaptic changes after TTX treatment, this argument is not compelling. In this retrograde pathway, BDNF is synthesized in and released from dendrites (Jakawich et al., 2010b; Thapliyal et al., 2022), and it is entirely possible for postsynaptic Rab3A to interfere with this process cell-autonomously.

We have added the information that Rab3A could control BDNF from the postsynaptic cell and included the two references provided by the reviewer, Discussion, lines 517 to 518. We have added new evidence, recently published, that the Rab3 family has been shown to regulate targeting of EGF receptors to rafts (among other plasma membrane molecules), with Rab3A itself clearly present in nonneuronal cells (Diaz-Rohrer et al., 2023) (added to Discussion, lines 509 to 515).

(3) The authors propose that a change in AMPAR subunit composition from GluA2-containing ones to GluA1 homomers may account for the distinct changes in mEPSC amplitudes and GluA2 clusters. However, their data from the NASPM wash-in experiments clearly show that the GluA1 homomer contributions have not changed before and after TTX treatment.

We have revised this section in the Discussion, lines 534 to 536, to clarify that any change due to GluA1 homomers should have been detectable by a greater ability of NASPM to reverse the TTX-induced increase.

Reviewer #2 (Recommendations for the Authors):

For authors to have more convincing arguments in general, they will need to clarify/improve certain details in their data collection by addressing the above technical concerns. Additionally, the authors should design experiments to test whether Rab3A regulates scaling from pre- or post-synaptic site. For example, they could sparsely knock out Rab3A in WT neurons to test the postsynaptic possibility. On the other hand, their argument for a presynaptic role would be much more compelling if they could show whether there are clear functional changes such as in vesicle sizes and release probability in the presynaptic terminal of Rab3AKO neurons.

An important next step is to identify whether Rab3A is acting pre- or post-synaptically (Discussion, lines 572 to 573), but these experiments will be undertaken in the future. It would not add much to simply show vesicle size is altered in the KO (and we do not necessarily expect this since mEPSC amplitude is normal in the KO). It will be very difficult to establish that vesicle size is changing with activity blockade and that this change is prevented in the Rab3A KO, because we are looking for a ~25% increase in vesicle volume, which would correspond to a ~7.5% increase in diameter. Finally, we do not believe demonstrating changes in release probability tell us anything about a presynaptic role for Rab3A in regulating the size of the presynaptic quantum.

Reviewer #3 (Public Review)

Weaknesses: However, the rather strong conclusions on the dissociation of AMPAR trafficking and synaptic response are made from somewhat weaker data. The key issue is the GluA2 immunostaining in comparison with the mEPSC recordings. Their imaging method involves only assessing puncta clearly associated with a MAP2 labeled dendrite. This is a small subset of synapses, judging from the sample micrographs (Fig. 5). To my knowledge, this is a new and unvalidated approach that could represent a particular subset of synapses not representative of the synapses contributing to the mEPSC change (they are also sampling different neurons for the two measurements; an additional unknown detail is how far from the cell body were the analyzed dendrites for immunostaining.) While the authors acknowledge that a sampling issue could explain the data, they still use this data to draw strong conclusions about the lack of AMPAR trafficking contribution to the mEPSC amplitude change. This apparent difference may be a methodological issue rather than a biological one, and at this point it is impossible to differentiate these. It will unfortunately be difficult to validate their approach. Perhaps if they were to drive NMDAdependent LTD or chemLTP, and show alignment of the imaging and ephys, that would help. More helpful would be recordings and imaging from the same neurons but this is challenging. Sampling from identified synapses would of course be ideal, perhaps from 2P uncaging combined with SEP-labeled AMPARs, but this is more challenging still. But without data to validate the method, it seems unwarranted to make such strong conclusions such as that AMPAR trafficking does not underlie the increase in mEPSC amplitude, given the previous data supporting such a model.

In the new version, we soften our conclusion regarding the mismatch between GluA2 receptor levels and mEPSC amplitudes, now only stating that receptors may not be the sole contributor to the TTX effect on mEPSC amplitude (Discussion, lines 472 to 474). With our analysis in the new version focusing on comparisons of cell means, the GluA2 receptor cluster size and the mEPSC amplitude data match well in magnitude for the data pooled across the 3 matched cultures (20.0% and 19.7%, respectively, see new Table 1). However, in one of the three cultures the direction of change for GluA2 receptors is opposite that of mEPSC amplitudes (Table 1, Culture #3, -9.5% vs +17.2%, respectively).

It is unlikely that the lack of matching of homeostatic plasticity in one culture, but very good matching in two other cultures, can be explained by an unvalidated focus on puncta associated with MAP2 positive dendrites. We chose to restrict analysis of synaptic GluA2 receptors to the primary dendrite in order to reduce variability, reasoning that we are always measuring synapses for an excitatory pyramidal neuron, synapses that are relatively close to the cell body, on the consistently identifiable primary dendrite. We measured how far this was for the two cells depicted in old Figure 5 (new Fig. 4). Because we always used the 5X zoom window which is a set length, and positioned it within ~10 microns of the cell body, these cells give a ball park estimate for the usual distances. For the untreated cell, the average distance from the cell body was 38.5 ± 2.8 µm; for the TTX-treated cell, it was 42.4 ± 3.2 µm (p = 0.35, KruskalWallis test). We have added these values to the Results, lines 270 to 274.

We did not mean to propose that AMPA receptor levels do not contribute at all to mEPSC amplitude, and we acknowledge there are clear cases where the two characteristics change in parallel (for example, in the study cited by Reviewer #2, (Pozo et al., 2012), increases in GluA2 receptors due to exogenous expression are closely matched by increases in mEPSC amplitudes.) What our matched culture experiments demonstrate is that in the case of TTX treatment, both GluA2 receptors and mEPSC amplitudes increase on average, but sometimes mEPSC amplitudes can increase in the absence of an increase in GluA2 receptors (Culture #3, Rab3A+/+ cultures), and sometimes mEPSC amplitudes do not increase even though GluA2 receptor levels do increase (Culture #3, Rab3A-/- cultures). Therefore, it would not add anything to our argument to examine receptors and mEPSCs in NMDA-dependent LTP, a different plasticity paradigm in which changes in receptors and mEPSCs may more closely align. It has been demonstrated that mEPSCs of widely varying amplitude can be recorded from a single synaptic site (Liu and Tsien, 1995), so we would need to measure a large sample of individual synapse recordings to detect a modest shift in average values due to activity blockade. In addition, it would be essential to express fluorescent AMPA receptors in order to correlate receptor levels in the same cells we record from (or at the same synapses). And yet, even after these heroics, one is still left with the issue that the two methods, electrophysiology and fluorescent imaging, have distinct limitations and sources of variability that may obscure any true quantitative correlation.

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a frequency effect that is quite unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. It is also unclear why the authors argue this proves that NASPM was at an effective concentration (lines 399-400). Further, the amplitude data show a strong trend towards smaller amplitude. The p value for both control and TTX neurons was 0.08 – it is very difficult to argue that there is no effect. And the decrease is larger in the TTX neurons. Considering the strong claims for a presynaptic locus and the use of this data to justify only looking at GluA2 by immunostaining, these data do not offer much support of the conclusions. Between the sampling issues and perhaps looking at the wrong GluA subunit, it seems premature to argue that trafficking is not a contributor to the mEPSC amplitude change, especially given the substantial support for that hypothesis. Further, even if trafficking is not the major contributor, there could be shifts in conductance (perhaps due to regulation of auxiliary subunits) that does not necessitate a pre-synaptic locus. While the authors are free to hypothesize such a mechanism, it would be prudent to acknowledge other options and explanations.

We have created a model cartoon to explain how NASPM could reduce mEPSC frequency (new Fig. 3D). mEPSCs that arise from a synaptic site that has only Ca2+-permeable AMPA receptors will be completely blocked by NASPM, if the NASPM concentration is maximal. The reason we conclude that we have sufficient NASPM reaching the cells is that the frequency is decreased, as expected if there are synaptic sites with only Ca2+-permeable AMPA receptors. We previously were not clear that there is an effect of NASPM on mEPSC amplitude, although it did not reach statistical significance (new Fig. 3B). Where there is no effect is on the TTX-induced increase in mEPSC amplitude, which remains after the acute NASPM application (new Fig. 3A). We have revised the description of these findings in Results, lines 220 to 241. In reviewing the literature further, we could find no previous studies demonstrating an increase in conductance in GluA2 or Ca2+-impermeable receptors, only in GluA1 homomers. In other words, any conductance change would have been due to a change in GluA1 homomers, and should have been visible as a disruption of the homeostatic plasticity by NASPM application. We have added text to Results, lines 211 to 217; 236-241; Discussion, lines 420 to 422; 526-536 and Methods, lines 685 to 695 regarding this point.

The frequency data are missing from the paper, with the exception of the NASPM dataset. The mEPSC frequencies should be reported for all experiments, particularly given that Rab3A is generally viewed as a pre-synaptic protein regulating release. Also, in the NASPM experiments, the average frequency is much higher in the TTX treated cultures. Is this statistically above control values?

This comment is addressed by the major change #3, above.

Unaddressed issues that would greatly increase the impact of the paper:

(1) Is Rab3A activity pre-synaptically, post-synaptically or both. The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where is it acting (pre or post) would aid substantially in understanding its role (and particularly the hypothesized and somewhat novel idea that the amount of glutamate released per vesicle is altered in HSP). They could use sparse knockdown of Rab3A, or simply mix cultures from KO and WT mice (with appropriate tags/labels). The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. The more support for their suggestion of a pre-synaptic site of control, the better.

This is similar to the request of Reviewer #2, Recommendations to the Authors. An important next step is to identify whether Rab3A is working pre- or postsynaptically. However, it is possible that it is acting pre-synaptically to anterogradely regulate trafficking of AMPAR, as we have depicted in our model, new Fig. 9. To demonstrate that the presynaptic quantum is being altered, we would need to show that vesicle size is increased, or the amount of transmitter being released during an mEPSC is increased after activity blockade. To that end, we are currently performing experiments using a fast off-rate antagonist. As described above in response to Reviewer #2’s Conceptual Concerns, we find dramatic decreases in frequency not explained by the 30-60% inhibition observed for the largest amplitude mEPSCs, which suggests the possibility that small mEPSCs are more sensitive than large mEPSCs and therefore may have less transmitter. Due to these complexities and the delay while we test other antagonists to see if the effect is specific to fast-off rate antagonists, we are not including these results here.

(2) Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs and/or a decrease of GABA-packaging in vesicles (ie the opposite of whatever is happening at excitatory synapses.). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling, an effect only at excitatory synapses would argue for a more specific role just at these synapses.

It will be important to determine if homeostatic synaptic plasticity at inhibitory synapses on excitatory neurons is sensitive to Rab3A deletion, especially in light of the fact that unlike many of the other molecules implicated in homeostatic increases in mEPSCS, Rab3A is not a molecule known to be selective for glutamate receptor trafficking (in contrast to Arc/Arg3.1 or GRIP1, for example). Such a study would warrant its own publication.

Reviewer #3 (Recommendations for the Authors):

There are a number of minor points or suggestions for the authors:

Is RIM1 part of this pathway (or expected to be)? Some discussion of this would be nice.

RIM, Rab3-interacting molecule, has been implicated at the drosophila neuromuscular junction in a presynaptic form of homeostatic synaptic plasticity in which evoked release is increased after block of postsynaptic receptors (Muller et al., 2012), a plasticity that also requires Rab3-GAP (Muller et al., 2011). To our knowledge there is no evidence that RIM is involved in the homeostatic plasticity of mEPSC amplitude after activity blockade by TTX. The Rim1a KO does not have a change in mEPSC amplitude relative to WT (Calakos et al., 2004), but that is not unexpected given the normal mEPSC amplitude in neurons from cultures prepared from Rab3A-/- mice in the current study. It would be interesting to look at homeostatic plasticity in cortical cultures prepared from Rim1a or other RIM deletion mice, but we have not added these points to the revised manuscript since there are a number of directions one could go in attempting to define the molecular pathway and we feel it is more important to discuss the potential location of action and physiological mechanisms.

Is the Earlybird mutation a GOF? More information about this mutation would help.

We have added a description of how the Earlybird mutation was identified, in a screen for rest:activity mutants (Results, lines 118 to 123). Rab3A Earlybird mice have a shortened circadian period, shifting their wake cycle earlier and earlier. When Rab3A deletion mice were tested in the same activity raster plot measurements, the shift was smaller than that for the Earlybird mutant, suggesting the possibility that it is a dominant negative mutation.

The high K used in the NASPM experiments seems a bit unusual. Have the authors done high K/no drug controls to see if this affects the synapses in any way?

We used the high K based on previous studies that indicated the blocking effect of the Ca2+-permeable receptor blockers was use dependent (Herlitze et al., 1993; Iino et al., 1996; Koike et al., 1997). We reasoned that a modest depolarization would increase the frequency of AMPA receptor mEPSCs and allow access of the NASPM. We have added this point to the Methods, lines 695 to 708.

The NASPM experiments do not show that GluA1 does not contribute (line 401), only that GluA1 homomers are not contributing (much – see above). GluA1/A2 heteromers are quite likely involved. Also, the SEM is missing from the WT pre/post NASPM data.

Imaging of GluA2-positive sites will not distinguish between GluA2 homomers and GluA2-GluA1 heteromers, so we have added this clarification to Results, lines 242 to 246. We have remade the NASPM pre-post line plots so that the mean values and error bars are more visible (new Fig. 3B, C).

It seems odd to speculate based on non-significant findings (line 650-1), with lower significance (p = 0.11) than findings being dismissed in the paper (NASPM on mEPSC amplitude; p = 0.08).

We did not mean to dismiss the effect of NASPM on mEPSC amplitude (new Fig. 3B), rather, we dismiss the effect of NASPM on the homeostatic increase in mEPSC amplitude caused by TTX treatment (new Fig. 3A). We have emphasized this distinction in Results, lines 223 to 225, and Discussion, lines 420 to 422, as well as adding that the stronger effect of NASPM on frequency after TTX treatment suggests an activity-dependent increase in the number of synapses expressing only Ca2+ permeable homomers (Results, lines 236 to 241; Discussion, lines 431 to 435).

Fig. 4 could be labeled better (to make it clear that B is amplitude and C is freq from the same cells).

Fig. 4 has been revised—now the amplitude and frequency plots from the same condition (new Fig. 3, B, C; CON or TTX) are in a vertical line and the figure legend states that the frequency data are from the same cells as in Fig. 3A.

The raw amplitude data seems a bit hidden in the inset panels – I would suggest these data are at least as important as the cumulative distributions in the main panel. Maybe re-organizing the figures would help.

We have removed all cumulative distributions, rank order plots, and ratio plots. The box plots are now full size in new Figures 1, 2, 5, 6, 7 and 8.

I’m not sure I would argue in the paper that 12 cells a day is a limiting issue for experiments. It doesn’t add anything and doesn’t seem like that high a barrier. It is fine to just say it is difficult and therefore there is a limited amount of data meeting the criteria.

We have removed the comment regarding difficulty.

Calakos N, Schoch S, Sudhof TC, Malenka RC (2004) Multiple roles for the active zone protein RIM1alpha in late stages of neurotransmitter release. Neuron 42:889-896.

De Gois S, Schafer MK, Defamie N, Chen C, Ricci A, Weihe E, Varoqui H, Erickson JD (2005) Homeostatic scaling of vesicular glutamate and GABA transporter expression in rat neocortical circuits. J Neurosci 25:7121-7133.

Diaz-Rohrer B, Castello-Serrano I, Chan SH, Wang HY, Shurer CR, Levental KR, Levental I (2023) Rab3 mediates a pathway for endocytic sorting and plasma membrane recycling of ordered microdomains. Proc Natl Acad Sci U S A 120:e2207461120.

Dubes S, Soula A, Benquet S, Tessier B, Poujol C, Favereaux A, Thoumine O, Letellier M (2022) miR-124dependent tagging of synapses by synaptopodin enables input-specific homeostatic plasticity. EMBO J 41:e109012.

Fong MF, Newman JP, Potter SM, Wenner P (2015) Upward synaptic scaling is dependent on neurotransmission rather than spiking. Nat Commun 6:6339.

Herlitze S, Raditsch M, Ruppersberg JP, Jahn W, Monyer H, Schoepfer R, Witzemann V (1993) Argiotoxin detects molecular differences in AMPA receptor channels. Neuron 10:1131-1140.

Hou Q, Zhang D, Jarzylo L, Huganir RL, Man HY (2008) Homeostatic regulation of AMPA receptor expression at single hippocampal synapses. Proc Natl Acad Sci U S A 105:775-780.

Ibata K, Sun Q, Turrigiano GG (2008) Rapid synaptic scaling induced by changes in postsynaptic firing. Neuron 57:819-826.

Iino M, Koike M, Isa T, Ozawa S (1996) Voltage-dependent blockage of Ca(2+)-permeable AMPA receptors by joro spider toxin in cultured rat hippocampal neurones. J Physiol 496 ( Pt 2):431437.

Jakawich SK, Neely RM, Djakovic SN, Patrick GN, Sutton MA (2010a) An essential postsynaptic role for the ubiquitin proteasome system in slow homeostatic synaptic plasticity in cultured hippocampal neurons. Neuroscience 171:1016-1031.

Jakawich SK, Nasser HB, Strong MJ, McCartney AJ, Perez AS, Rakesh N, Carruthers CJ, Sutton MA (2010b) Local presynaptic activity gates homeostatic changes in presynaptic function driven by dendritic BDNF synthesis. Neuron 68:1143-1158.

Kapfhamer D, Valladares O, Sun Y, Nolan PM, Rux JJ, Arnold SE, Veasey SC, Bucan M (2002) Mutations in Rab3a alter circadian period and homeostatic response to sleep loss in the mouse. Nat Genet 32:290-295.

Koike M, Iino M, Ozawa S (1997) Blocking effect of 1-naphthyl acetyl spermine on Ca(2+)-permeable AMPA receptors in cultured rat hippocampal neurons. Neurosci Res 29:27-36.

Liu G, Tsien RW (1995) Properties of synaptic transmission at single hippocampal synaptic boutons. Nature 375:404-408.

Liu G, Choi S, Tsien RW (1999) Variability of neurotransmitter concentration and nonsaturation of postsynaptic AMPA receptors at synapses in hippocampal cultures and slices. Neuron 22:395409.

Muller M, Pym EC, Tong A, Davis GW (2011) Rab3-GAP controls the progression of synaptic homeostasis at a late stage of vesicle release. Neuron 69:749-762.

Muller M, Liu KS, Sigrist SJ, Davis GW (2012) RIM controls homeostatic plasticity through modulation of the readily-releasable vesicle pool. J Neurosci 32:16574-16585.

Pozo K, Cingolani LA, Bassani S, Laurent F, Passafaro M, Goda Y (2012) beta3 integrin interacts directly with GluA2 AMPA receptor subunit and regulates AMPA receptor expression in hippocampal neurons. Proc Natl Acad Sci U S A 109:1323-1328.

Silva MM, Rodrigues B, Fernandes J, Santos SD, Carreto L, Santos MAS, Pinheiro P, Carvalho AL (2019) MicroRNA-186-5p controls GluA2 surface expression and synaptic scaling in hippocampal neurons. Proc Natl Acad Sci U S A 116:5727-5736.

Soden ME, Chen L (2010) Fragile X protein FMRP is required for homeostatic plasticity and regulation of synaptic strength by retinoic acid. J Neurosci 30:16910-16921.

Sun HY, Bartley AF, Dobrunz LE (2009) Calcium-permeable presynaptic kainate receptors involved in excitatory short-term facilitation onto somatostatin interneurons during natural stimulus patterns. J Neurophysiol 101:1043-1055.

Sutton MA, Ito HT, Cressy P, Kempf C, Woo JC, Schuman EM (2006) Miniature neurotransmission stabilizes synaptic function via tonic suppression of local dendritic protein synthesis. Cell 125:785-799.

Tan HL, Queenan BN, Huganir RL (2015) GRIP1 is required for homeostatic regulation of AMPAR trafficking. Proc Natl Acad Sci U S A 112:10026-10031.

Thapliyal S, Arendt KL, Lau AG, Chen L (2022) Retinoic acid-gated BDNF synthesis in neuronal dendrites drives presynaptic homeostatic plasticity. Elife 11.

Wilson NR, Kang J, Hueske EV, Leung T, Varoqui H, Murnick JG, Erickson JD, Liu G (2005) Presynaptic regulation of quantal size by the vesicular glutamate transporter VGLUT1. J Neurosci 25:62216234.

Wu YK, Hengen KB, Turrigiano GG, Gjorgjieva J (2020) Homeostatic mechanisms regulate distinct aspects of cortical circuit dynamics. Proc Natl Acad Sci U S A 117:24514-24525.

Xu X, Pozzo-Miller L (2017) EEA1 restores homeostatic synaptic plasticity in hippocampal neurons from Rett syndrome mice. J Physiol 595:5699-5712.

Version 2

Abstract

Abstract

Following prolonged activity blockade, amplitudes of miniature excitatory postsynaptic currents (mEPSCs) increase, a form of plasticity termed “homeostatic synaptic plasticity.” We previously showed that a presynaptic protein, the small GTPase Rab3A, is required for full expression of the increase in miniature endplate current amplitudes following prolonged blockade of action potential activity at the mouse neuromuscular junction in vivo (Wang et al., 2011), but it is unknown whether this form of Rab3A-dependent homeostatic plasticity shares any characteristics with central synapses. We show here that homeostatic synaptic plasticity of mEPSCs is impaired in mouse cortical neuron cultures prepared from Rab3A−/− and mutant mice expressing a single point mutation of Rab3A, Rab3A Earlybird mice. To determine if Rab3A is involved in the well-established homeostatic increase in postsynaptic AMPA-type receptors (AMPARs), we performed a series of experiments in which electrophysiological recordings of mEPSCs and confocal imaging of synaptic AMPAR immunofluorescence were assessed within the same cultures. We found that Rab3A was required for the increase in synaptic AMPARs following prolonged activity blockade, but the increase in mEPSC amplitudes was not always accompanied by an increase in postsynaptic AMPAR levels, suggesting other factors may contribute. Finally, we demonstrate that Rab3A is acting in neurons because only selective loss of Rab3A in neurons, not glia, disrupted the homeostatic increase in mEPSC amplitudes. This is the first demonstration that neuronal Rab3A is required for homeostatic synaptic plasticity and that it does so partially through regulation of the surface expression of AMPA receptors.

Introduction

One of the most studied phenomena triggered by prolonged activity blockade is the increase in amplitudes of miniature excitatory postsynaptic currents (mEPSCs) in neurons. First demonstrated in cultures of dissociated cortical neurons (Turrigiano et al., 1998) and spinal cord neurons (O’Brien et al., 1998), the compensatory response to prolonged loss of activity has been dubbed “homeostatic synaptic plasticity” (Turrigiano and Nelson, 2004; Pozo and Goda, 2010). These first two studies provided evidence that glutamate receptors were also increased after inactivity, via an increased response to exogenously applied glutamate (Turrigiano et al., 1998) and increased immunofluorescent labeling of GluA1 receptors at synaptic sites (O’Brien et al., 1998). The accompanying increase in synaptic AMPA receptors has been confirmed with immunofluorescence in multiple studies of hippocampal and cortical cultures treated with TTX (Wierenga et al., 2005; Hou et al., 2008; Ibata et al., 2008; Jakawich et al., 2010a; Tatavarty et al., 2013; Xu and Pozzo-Miller, 2017; Dubes et al., 2022). It is now well accepted that the homeostatic increase in mEPSC amplitudes in neurons is due to an increase in postsynaptic AMPA receptors.

In our previous work, we found that a TTX cuff applied for 48 hr around the sciatic nerve in mice led to an increase in the amplitude of the miniature endplate currents (mEPCs) recorded in tibialis anterior muscles. Surprisingly, and unlike central synapses, we could find no evidence that the acetylcholine receptors levels at the neuromuscular junction (NMJ) were increased in the TTX-cuff treated mice (Wang et al., 2005). This result led us to search for presynaptic molecules that might homeostatically regulate the presynaptic quantum. In previous studies in chromaffin cells, we identified the small GTPase Rab3A, a presynaptic vesicle protein, as a regulator of synaptic vesicle fusion pore opening (Wang et al., 2008), so we examined whether deletion of Rab3A (Rab3A−/−) might prevent homeostatic upregulation of mEPC amplitude at the NMJ. The results were clear: the homeostatic increase in mEPC amplitude induced by the TTX cuff was strongly reduced in the Rab3A−/− mouse, and was completely abolished in the Earlybird mutant (Rab3AEbd/Ebd), which has a single point mutation in Rab3A that causes a shift towards early awakening that is more dramatic than that of the Rab3A deletion mouse (Kapfhamer et al., 2002; Wang et al., 2011). These results led us to conclude that the homeostatic increase in mEPC amplitude at the NMJ is a presynaptic phenomenon.

It has remained somewhat of a mystery what the role of Rab3A is in synaptic transmission. The Rab3A−/− mouse has minimal phenotypic abnormalities, with evoked synaptic transmission and mEPSCs essentially normal in hippocampal slices (Geppert et al., 1994). At the Rab3A−/− NMJ, reductions in evoked transmission were detected, but only under conditions of reduced extracellular calcium (Coleman et al., 2007). There are some modest changes in short term plasticity during repetitive stimulus trains, with increased depression observed in response to moderate frequencies in hippocampal slices (Geppert et al., 1994) and at the mammalian NMJ (Coleman and Bykhovskaia, 2009), but increased facilitation in response to high frequency stimulation at the mammalian NMJ (Coleman and Bykhovskaia, 2010). The most dramatic effect of loss of Rab3A is the abolishment of a presynaptic form of long-term potentiation (LTP) at the mossy fiber-CA3 synapse that is induced by 25 Hz stimulation in the presence of NMDA blockers (Weisskopf et al., 1994; Castillo et al., 1997). Our result that prolonged inactivity-induced plasticity of mEPC amplitude at the mammalian NMJ is disrupted with loss of function of Rab3A further cements a role for Rab3A in activity-dependent plasticity of synaptic strength. However, a crucial question that remains is whether the TTX-cuff induced effect on the amplitude of the spontaneous synaptic current at the NMJ shares underlying mechanisms with central synapses. We set out to answer this question for the more well-studied phenomenon of homeostatic plasticity of mEPSCs in dissociated mouse cortical neurons prepared from wild type, Rab3A−/− mice and Rab3aEbd/Ebd mice treated with TTX for 48 hr. We show here that homeostatic synaptic plasticity of mEPSC amplitude was strongly reduced in cultures from Rab3A−/− mice, and completely abolished in cultures from Rab3AEbd/Ebd mice, supporting a shared mechanism with the neuromuscular junction. Unexpectedly, the modest increase in GluA2 immunofluorescence we observed in wild type cultures was not present in Rab3A−/− cultures. Loss of Rab3A in neurons, but not in glia, disrupted the increase in mEPSC amplitudes after TTX treatment. These results support the idea that the increase in mEPSC amplitude is due to more postsynaptic receptors, and that neuronal Rab3A is necessary for this regulation. However, we found that GluA2 receptors and mEPSC amplitudes did not always change in parallel. This mismatch suggests there may be contributors other than postsynaptic receptor numbers to the homeostatic plasticity of mEPSC amplitude.

Results

We previously reported that mixed cultures of cortical neurons and glia prepared from postnatal day 0-2 mouse pups responded to a block of action potential-mediated activity by a 48 hr TTX treatment with an increase in mEPSC amplitude (Hanes et al., 2020). Here, we asked the question whether cortical cultures prepared from mice lacking the small GTPase Rab3A, or, expressing a point mutation of Rab3A, Rab3A Earlybird, have an altered homeostatic plasticity response to a loss of network activity. The Rab3A Earlybird mutation was discovered in a mutagenesis screen in mice for genes involved in the rest-activity cycle (Kapfhamer et al., 2002). The Earlybird mutant was named for its phenotype, a significantly shorter circadian period which resulted in early awakening. In that study, the authors concluded that the Earlybird mutation was likely dominant-negative, as the shift in circadian period was substantially more dramatic than that observed in the Rab3A−/− mice. As described in the Introduction, the effects of loss of Rab3A have been modest, so we included experiments using the Rab3A Earlybird mutant mice to see if there would be a more robust phenotype, and in doing so, strengthen conclusions based on the Rab3A deletion mice. To obtain Rab3A−/− and Rab3AEbd/Ebd homozygotes, we established two mouse colonies of heterozygous breeders with cultures prepared from pups derived from a final breeding pair of homozygotes. Although we backcrossed Rab3A+/− with Rab3A+/Ebd for 11 generations, clear differences in mEPSC amplitudes in untreated cultures (see below) and in calcium current amplitudes in adrenal chromaffin cells (unpublished obs.) remained. Therefore, throughout this study we keep the two strains separate and there are two Rab3A+/+ or ‘wild type’ phenotypes.

We found that loss of Rab3A resulted in a greatly impaired homeostatic increase after activity blockade. Example current traces of spontaneously occurring mEPSCs recorded from pyramidal neurons in untreated (CON) 13-14 DIV cortical cultures and sister cultures treated with 500 nM TTX for 48 hr prepared from wild type and Rab3A−/− mice in the Rab3A+/− colony are shown in Figures 1A and B, respectively. Average mEPSC waveforms from the same recordings are shown in Figures 1C and D. The mean mEPSC amplitudes for 30 control and 23 TTX-treated neurons from Rab3A+/+ cultures are displayed in the box and whisker plot in Figure 1E; after activity blockade the average mEPSC amplitude increased from 13.9 ± 0.7 pA to 18.2 ± 0.9 pA. In contrast, in cultures prepared from Rab3A−/− mice, the average mEPSC amplitude was not significantly increased, for 25 untreated cells and 26 TTX-treated cells (Figure 1F, 13.6 ± 0.1 pA vs. 14.3 ± 0.6 pA). We found that TTX treatment also resulted in an increase in mEPSC frequency in cultures prepared from Rab3A+/+ mice, as shown in Figure 1G (CON, 2.26 ± 0.37 sec−1; TTX, 4.62 ± 0.74 sec−1). This TTX-induced frequency increase was strongly reduced in cultures from Rab3A−/− mice, but there were a few outliers with very high frequencies still present after TTX treatment (Figure 1H, CON, 2.74 ± 0.49 Hz; TTX, 3.23 ± 0.93 Hz).

Loss of Rab3A prevented the TTX-induced increases in amplitudes and frequency of mEPSCs recorded in cultured mouse cortical neurons.

(A) Ten second example traces recorded at –60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/− colony. (B) Ten second example traces recorded at –60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A−/− mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A+/+ mice (Rab3A+/− colony; CON, N = 30 cells, 13.9 ± 0.7 pA; TTX, N = 23 cells, 18.2 ± 0.9 pA; from 11 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A−/− mice (CON, N = 25 cells, 13.6 ± 0.7 pA; TTX, N = 26 cells, 14.3 ± 0.6 pA); from 11 cultures. (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 2.26 ± 0.37 sec−1; TTX, 4.62 ± 0.74 sec−1). (H) Box plots for average mEPSC frequency for same Rab3A−/−cells as in (F) (CON, 2.74 ± 0.49 sec−1; TTX, 3.23 ± 0.93 sec−1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. For all p-values, * with underline indicates significance with p < 0.05.

We next examined homeostatic plasticity in cultures prepared from wild type mice in the Rab3A+/Ebd colony. Example current traces of mEPSCs, and average mEPSC waveforms, are shown in Figures 2A and C, respectively, for cortical cultures prepared from wild type mice in the Rab3A+/Ebd colony. Treatment with TTX for 48 hr led to a significant increase in the average mEPSC amplitude of 23 TTX-treated cells when compared to 20 untreated cells (Figure 2E, CON, 11.0 ± 0.6 pA; TTX, 15.0 ± 1.3 pA). We note here that while the two wild type strains respond very similarly to activity blockade, the mean mEPSC amplitude in untreated cultures was significantly different in the two wild type strains, 13.9 ± 0.7 pA, wild type from Rab3A+/− colony; 11.0 ± 0.6 pA, wild type from Rab3A+/Ebd colony (p = 0.004, Kruskal-Wallis test).

Normally functioning Rab3A was required for TTX-induced homeostatic plasticity of mEPSC amplitudes in cultured mouse cortical neurons.

(A) Ten second example traces recorded at –60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/Ebd colony. (B) Ten second example traces recorded at –60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3AEbd/Ebd mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3A+/+ mice (Rab3A+/Ebd colony; CON, N = 20 cells, 11.0 ± 0.6 pA; TTX, N = 23 cells, 15.1 ± 1.2 pA; from 6 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3AEbd/Ebd mice (CON, N = 21 cells, 15.1 ± 1.0 pA; TTX, N = 22 cells, 14.6 ± 1.1 pA; from 7 cultures.) (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 1.15 ± 0.19 sec−1; TTX, 2.54 ± 0.55 sec−1). (H) Box plots for average mEPSC frequency for same Rab3AEbd/Ebd cells as in (F) (CON, 1.71 ± 0.41 sec−1; TTX, 3.05 ± 0.80 sec−1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. For all p-values, * with underline indicates significance with p < 0.05.

We found a complete disruption of homeostatic plasticity in cortical cultures prepared from Rab3AEbd/Ebd mice, as can been seen in viewing example mEPSC traces and average mEPSC waveforms (Figures 2B and D, respectively). The lack of TTX effect was confirmed in a comparison of mEPSC amplitude means for 21 untreated and 22 TTX-treated cells (Figure 2F, CON, 15.1 ± 1.0 pA vs. TTX, 14.6 ± 1.1 pA). There was a trend towards increased mEPSC frequency in TTX-treated cultures from this wild type strain, but the difference was not as robust as for the wild type cultures from Rab3A+/− colony, and did not reach statistical significance (Figure 2G, CON, 1.15 ± 0.19 sec−1; TTX, 2.54 ± 0.55 sec−1). This trend was not observed in neurons from cultures prepared from Rab3AEbd/Ebd mice, likely due to an increase in frequency in untreated cells—frequencies remained high in TTX-treated cells (Figure 2H, CON, 1.71 ± 0.41 sec−1; TTX, 3.05 ± 0.80 sec−1).

Our results show that the homeostatic increase in mEPSC amplitude after activity blockade is disrupted in both Rab3A−/− and the Rab3AEbd/Ebd cortical neurons, strongly supporting a crucial role of functioning Rab3A in this process. However, it is important to note that the disruption differs for Rab3A−/− and Rab3AEbd/Ebd. In the Rab3A−/− data set, mEPSCs from untreated cultures were indistinguishable from mEPSCs from Rab3A+/+ untreated cultures, demonstrating loss of Rab3A has no impact on basal activity mEPSC amplitudes, but the increase in mEPSC amplitudes after activity blockade was strongly diminished in Rab3A−/− neurons. In the Rab3AEbd/Ebddata set, mEPSC amplitudes from untreated cultures were significantly larger than those of untreated cultures from wild type mice, as can be seen comparing the CON values in Figures 2E and F (CON, Rab3A+/+, 11.0 ± 0.7 pA, vs. CON, Rab3AEbd/Ebd, 15.1 ± 1.0 pA, p = 0.0027, Kruskal-Wallis test). The increase in mEPSC amplitude in cultures from Rab3AEbd/Ebd mice is consistent with the increase in mEPC amplitude we observed at the Rab3AEbd/EbdNMJ (Wang et al., 2011). Similar increases in mEPSC amplitude at baseline, combined with a failure to increase further after activity blockade, were noted in cultures prepared from mECP2, AKAP, Homer1a, and Arc/Arg3.1 deletion mice (Shepherd et al., 2006; Hu et al., 2010; Xu and Pozzo-Miller, 2017; Sanderson et al., 2018) as well as in cultures treated with ubiquitin proteasome inhibitors (Jakawich et al., 2010a) or microRNA 186-5p (Silva et al., 2019), suggesting the possibility of a generalized phenomenon in which increased mEPSC amplitude is induced when homeostatic regulation is disrupted. Since we did not observe such an increase in mEPSC amplitude at baseline in cultures from Rab3A−/− mice, it remains a possibility that the point mutant may bind to novel partners, causing activities that would not be either facilitated or inhibited by Rab3A. Still, it is a strong coincidence that the novel activity of the mutant Earlybird Rab3A would affect mEPSC amplitude, the same characteristic that is modulated by activity blockade in a Rab3A dependent manner. Finally, although we cannot rule out that once mEPSC amplitude is increased at baseline, it is not possible for homeostatic plasticity to increase mEPSC amplitude further due to a non-specific ceiling effect, the simplest interpretation is that the presence of the mutant protein mimics the condition of activity blockade.

The small GTPase Rab3A is generally thought to function presynaptically to regulate synaptic vesicle trafficking, possibly in an activity-dependent manner (Castillo et al., 1997; Lonart et al., 1998; Leenders et al., 2001; Schluter et al., 2006; Coleman and Bykhovskaia, 2009; Tian et al., 2012). In contrast, homeostatic plasticity of mEPSC amplitude has been attributed to an increase in postsynaptic receptors on the surface of the dendrite (O’Brien et al., 1998; Turrigiano et al., 1998). Is Rab3A required for the increase in surface AMPA-type glutamate receptors that has been confirmed by multiple studies (see Introduction)? At the NMJ in vivo we could find no evidence for an increase in AChRs after TTX block of the sciatic nerve in vivo (Wang et al., 2005). The type of AMPA receptor that has been shown to be increased, calcium-permeable GluA1-containing AMPA receptor or calcium-impermeable GluA2-containing AMPA receptor, appears to depend on the experimental manipulation. For example, block of activity by APV (NMDA antagonist) combined with TTX (inhibitor of Na+ channels) induces an increase in mEPSC amplitude in dissociated hippocampal cultures that is completely reversed by acute application of the Ca2+-permeable receptor-specific inhibitor NASPM (Sutton et al., 2006), indicating the entirety of the homeostatic effect is due to an increase in homomeric Ca2+-permeable AMPA receptors, likely GluA1, although GluA3 and GluA4 are also Ca2+-permeable (Hollmann et al., 1991; Burnashev et al., 1992). Heteromeric receptors of any of these subunits in combination with GluA2 would have rendered the receptors Ca2+-impermeable and no longer sensitive to NASPM (Blaschke et al., 1993; Bochet et al., 1994; Jonas and Burnashev, 1995). In contrast, the increase in mEPSC amplitude in mouse hippocampal slice cultures induced by TTX alone was not affected by another Ca2+-permeable AMPA receptor inhibitor, philanthotoxin, suggesting mediation by either GluA2-containing AMPA receptors or a presynaptic effect on quantal size (Soden and Chen, 2010); see also, (Dubes et al., 2022).

Because we used TTX alone to block network activity, we expected that NASPM would not reverse the TTX-induced increase in mEPSC amplitude in our mouse cortical cultures, and we found that this was indeed the case. Figure 3A shows that the TTX-induced increase in mean mEPSC amplitude was nearly identical in a set of 11 CON and 11 TTX-treated cells before and after NASPM treatment (before NASPM, CON 12.9 ± 3.5 pA; TTX, 17.5 ± 3.1 pA; after NASPM, CON 11.9 ± 2.6 pA; TTX 16.1 ± 3.5 pA). This result indicates that the TTX-induced increase in mEPSC amplitude does not depend on Ca2+-permeable receptors, since the effect of TTX clearly remained when their presence was removed by acute NASPM application. We do not think that there was any technical issue with the NASPM application, because overall, mEPSC amplitudes were reduced modestly in both untreated and TTX-treated cultures (Figure 3B, CON, before NASPM, 12.9 ± 3.5 pA; after, 11.9 ± 2.6 pA; TTX, before NASPM, 17.5 ± 3.1 pA; after, 16.1 ± 3.5 pA). Furthermore, NASPM consistently reduced mEPSC frequency (Figure 3C, CON, before NASPM, 1.84 ± 0.55 sec−1; after NASPM, 1.56 ± 0.53 sec−1; TTX, before NASPM, 4.40 ± 3.51 sec−1; after NASPM, 2.68 ± 2.25 sec−1). Similar to the data in Figure 1, frequency is significantly increased after TTX treatment (p = 0.04), but this difference is no longer significant after NASPM application (p = 0.20). A loss in frequency after acute application of NASPM suggests there are synaptic sites that express only homomers of Ca2+-permeable receptors; mEPSCs from these sites would be expected to be completely blocked by NASPM (see cartoon description in Figure 3D, Left). The magnitude of the frequency reduction following acute NASPM appears to be greater after TTX treatment, suggesting that loss of activity promotes the establishment of synaptic sites that contain only Ca2+-permeable receptors. However, these new sites do not appear to contribute to the increase in mEPSC amplitude (Figure 3A). The lack of effect of NASPM also rules out the contribution of Ca2+-permeable forms of Kainate receptors GluA5 and GluA6 (Koike et al., 1997; Sun et al., 2009).

Homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures treated with TTX for 48 hr was not affected by acute inhibition of Ca2+ permeable AMPA receptors by NASPM (20 μm).

(A) Box plot comparison of the TTX effect on mEPSC amplitudes in the same pyramidal neurons before and after application of 20 μM NASPM. (N = 11 cells from 3 cultures; Pre-NASPM, CON: 12.9 ± 1.1 pA; TTX: 17.5 ± 0.9 pA; post-NASPM, CON: 11.9 ± 0.8 pA; TTX: 16.1 ± 1.0 pA). (B) Line series plot of mEPSC amplitudes before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NASPM: 12.9 ± 1.1 pA; post-NASPM: 11.9 ± 0.8 pA; TTX, pre-NASPM: 17.5 ± 0.9 pA; post-NASPM: 16.1 ± 0.1 pA. (C) Line series plot of mEPSC frequency before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NAPSM: 1.84 ± 0.55 sec−1; post-NASPM; 1.56 ± 0.53 sec−1; TTX, pre-NASPM: 4.40 ± 1.06 sec−1; post-NASPM, 2.68 ± 0.68 sec−1. (D) A proposed mechanism for why NASPM had a robust effect on frequency without greatly affecting amplitude. A dendrite with spines (Top) is expanded on three postsynaptic sites (Middle) to show possible types of AMPA receptor distributions: Left, at a site comprised only of Ca2+-permeable AMPA receptors (NASPM-sensitive, GluA2-lacking receptors (black)), NASPM would completely inhibit the mEPSC, causing a decrease in frequency to be measured in the overall population; Right, at a site comprised only of Ca2+-impermeable AMPA receptors (NASPM-insensitive, GluA2-containing receptors (green)), NASPM would have no effect on mEPSC amplitude; Middle, at a site comprised of a mix of Ca2+-permeable and Ca2+-impermeable AMPA receptors, NASPM would partially inhibit the mEPSC. Diagram produced in BioRender.com (2024). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. Line series plot p values are from student’s paired t test. For all p-values, * with underline indicates significance with p < 0.05.

Having established that homomeric Ca2+-permeable receptors are not contributing to the homeostatic increase in mEPSC amplitude, we turned to immunohistochemistry and confocal imaging to assess whether GluA2 receptor expression, which will identify either GluA2 homomers or GluA1-GluA2 heteromers, was increased in our wild type mouse cortical cultures following 48 hr treatment with TTX. Since mEPSCs necessarily report synaptic levels of receptors, we used VGLUT1-immunofluorescence to identify synapses on pyramidal primary apical dendrites labeled with MAP-2 immunofluorescence. We focused on the primary dendrite of pyramidal neurons as a way to reduce variability that might arise from being at widely ranging distances from the cell body, or, from inadvertently sampling dendritic regions arising from other neurons, possibly inhibitory neurons. Figure 4 (Top) shows two pyramidal neurons, one from an untreated culture on the left, and one from a TTX-treated culture on the right, prepared from Rab3A+/+ mice. To measure characteristics of synaptically located GluA2 receptor clusters, we zoomed in on the primary dendrite (region in white rectangle). The zoomed regions for single confocal sections of the selected area are shown below, and contain pairs of VGLUT1– and GluA2-immunofluorescent clusters indicated by white trapezoidal frames. A dendrite typically required ∼10 confocal sections to be fully captured, and the total number of synaptic pairs for all the sections imaged on a dendrite was usually < 20, so this is an atypically high number of pairs within a single section; these particular dendrites and sections were selected for illustration purposes. In addition to the synaptic pairs, we observed many GluA2-immunofluorescent clusters not associated with VGLUT1 immunofluorescence (a few are indicated with white arrows), most probably extrasynaptic receptors. There were also GluA2-positive clusters present outside of MAP2-positive dendrites, which may be located on astrocytes (Fan et al., 1999). Finally, GluA2 immunofluorescent clusters close to VGLUT1 immunofluorescence but not located along any apparent MAP2-positive neurites suggests the presence of axon-axonal contacts, although VGLUT1 has also been detected in astrocytes (Ormel et al., 2012). Only sites that contained both VGLUT1 and GluA2 immunofluorescence close to the primary MAP2-postive dendrite or a secondary branch along this primary dendrite were selected for analyses. In the experiment from which these images were collected, we analyzed the distance from the cell body for each synapse. The average distance from the cell body, for dendrites from the untreated cell, was 38.5 ± 2.8 μm; for the TTX-treated cell, it was 42.4 ± 3.2 μm (p = 0.35, Kruskal-Wallis test). Since the greatest distance from the cell body is set by the outer limit of the zoom window, and that window was placed adjacent to the cell body, distances in other experiments would be in this same range.

Identification of synaptic GluA2 receptor immunofluorescence on primary dendrites of pyramidal neurons in high density mouse cortical co-cultures prepared from Rab3A+/+ mice.

(Top) Non-zoomed single confocal sections collected with a 60X oil immersion objective of recognizably pyramidal-shaped neurons, presumed to be excitatory neurons, selected for synaptic GluA2 analysis. A neuron was selected from an untreated coverslip (CON, Left) and a TTX-treated coverslip (TTX, Right) from the same culture prepared from Rab3A+/+ animals (= Rab3A+/+ Culture #2 in Figure 6). The white rectangular boxes indicate 5X zoomed areas shown in the images below. Note that the depth of the single confocal section of the non-zoomed neuron image is not at the same depth as the confocal section in zoomed dendritic image, so some features are not visible in both images. Scale bar, 20 μm. (Bottom) 5X zoom single confocal sections selected for demonstration purposes because they had an unusually high number of identified synaptic pairs along the primary dendrite contained within a single confocal section. Synaptic pairs, highlighted with white trapezoids, were identified based on close proximity of GluA2 (red) and VGLUT1 (white) immunofluorescence, apposed to the MAP2 immunofluorescent primary dendrite (green). Some apparent synaptic pairs are not highlighted with a white trapezoid because there was a different confocal section in which the two immunofluorescent sites were maximally bright. There were GluA2 positive clusters located on the primary dendrites that were not apposed to VGLUT1-positive terminals (4 of these non-synaptic GluA2 clusters are highlighted with white arrows in the MAP2-GluA2 and MAP2-GluA2-VGLUT1 panels). Scale bar, 10 μm.

Variability in the magnitude of the homeostatic response from culture to culture is averaged out in physiological experiments by the pooling of data from many cells across many cultures. To reduce the necessity for many cultures, we chose to pair experiments in the same cultures by recording mEPSCs from one set of coverslips, and processing another set of coverslips from the same culture for immunofluorescence. We completed this matched paradigm of physiology and immunofluorescence, which to our knowledge has never before been done, for 3 cultures prepared from Rab3A+/+ mice and 3 cultures prepared from Rab3A−/− mice. We present the results for mEPSC data and imaging data pooled across the 3 experiments in Figure 5, with the data summarized in Table 1. Levels of GluA2 immunofluorescence at synaptic sites were quantified by the size of the GluA2-immunofluorescent receptor cluster, the average intensity value, and the integral of intensity values across the area of the cluster. In the data pooled from the 3 matched experiments, neurons from Rab3A+/+ cultures showed a significant increase in mean mEPSC amplitudes following activity blockade (Figure 5A; Table 1). This result indicates that the homeostatic response averaged across the 3 cultures was very similar to, but slightly smaller than, that of the data set presented in Figure 1. The means for size, intensity and integral of GluA2 immunofluorescent receptor clusters showed trends to higher values after activity blockade, but the differences did not reach statistical significance (Figure 5B size; Figure 5C intensity; Figure 5D integral; Table 1). These results support previous studies demonstrating an increase in AMPA receptors following activity blockade, with the magnitude of the changes in GluA2 receptor cluster size and integral matching that of mEPSC amplitude, at around 20% (Table 1), but not achieving the same level of statistical significance.

Comparison of mEPSC amplitude and GluA2 receptor cluster data, pooled from 3 cortical co-cultures prepared from Rab3A+/+ mice and 3 co-cultures prepared from Rab3A−/− mice.

(A) Mean mEPSC amplitudes recorded in pyramidal neurons from untreated (CON) and TTX-treated coverslips (TTX) of co-cultures prepared from Rab3A+/+ mice (CON, N = 23 cells, 13.7 ± 0.9 pA; TTX, N = 24 cells, 16.4 ± 0.9 pA; from 3 cultures). (B) Mean sizes of synaptic GluA2 receptor clusters, identified by proximity to VGLUT1-positive terminals, on primary dendrites of pyramidal neurons of untreated and TTX-treated coverslips from the same cultures as in (A) (CON, N = 29 dendrites, 0.90 ± 0.06 μm2; TTX, N = 28 dendrites, 1.08 ± 0.10 μm2). (C) Mean intensities of the same GluA2 receptor clusters as in (B) (CON, 684 ± 15 A.U.; TTX, 699 ± 12 A.U.). (D) Mean integrals of the same GluA2 receptor clusters as in (B). (CON, 317,898 ± 26,726 A.U.; TTX, 389,487 ± 41,465 A.U.). (E) Mean mEPSC amplitudes in pyramidal neurons from untreated and TTX-treated coverslips of mouse cortical co-cultures from Rab3A−/− mice (CON, N = 21 cells, 14.9 ± 0.8 pA; TTX, N = 21 cells, 13.5 ± 0.9 pA; from 3 cultures). (F) Mean sizes of GluA2 receptor clusters, identified by proximity to VGLUT1-positive terminals, on primary dendrites of pyramidal neurons of untreated and TTX-treated coverslips from the same cultures as in (E). (CON, N = 30 dendrites, 0.93 ± 0.05 μm2; TTX, N = 30 dendrites, 0.92 ± 0.05 μm2). (G) Mean intensities of the same GluA2 receptor clusters as in (F). (CON, 766 ± 12 A.U.; TTX, 776 ± 78 A.U.). (H) Mean integrals of the same GluA2 receptor clusters as in (F). (CON, 369,436 ± 23,439 A.U.; TTX, 366,389 ± 25,049 A.U.). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10 and 90 percentiles; open circles indicate means of individual cells; line, median; dot, mean. p values (shown on the graphs) were determined with Kruskal-Wallis test. For all p-values, * with underline indicates significance with p < 0.05.

As expected, mean mEPSC amplitude was not increased following activity blockade in the data pooled from the new set of 3 Rab3A−/− cultures (Figure 5E). For immunofluorescence results from the same Rab3A−/− cultures, no trend towards higher values was apparent for GluA2 receptor cluster characteristics (Figure 5F size; Figure 5G intensity; Figure 5H integral; Table 1). Taken together, these results indicate that the modest increases in GluA2 receptor cluster size and integral following activity blockade observed in wild-type cultures do not occur in the absence of Rab3A. These results suggest the surprising finding that the presynaptic vesicle protein Rab3A is involved in the regulation of postsynaptic AMPA receptors during homeostatic plasticity.

As noted above, the homeostatic increase in mEPSC amplitude in cultures from Rab3A+/+ mice reached significance whereas that of receptor levels did not. To determine where this disparity was coming from, we compared mEPSC amplitude and GluA2 receptor data within individual experiments. In order to evaluate whether a particular culture displayed homeostatic plasticity of mEPSC amplitude, we set a minimum requirement of recording from 6 cells per condition, meaning a successful experiment required at least 12 usable recordings in one day (6 CON and 6 TTX; i.e. low holding current, stable baseline, low noise, etc.). Ultimately, our sample size ranged from 6 to 10 cells per condition per culture. Figures 6A-C show the box plots for mEPSC amplitudes, and GluA2 receptor cluster size, intensity and integrals, for 3 individual Rab3A+/+ cultures. With these small sample sizes, the differences between untreated and TTX-treated cells were not statistically significant for mEPSC amplitudes in any of the 3 cultures, but an upward trend occurred in every case, ranging from 12 to 28% (Table 1). Second, in two of the three Rab3A+/+ experiments (Cultures #1 and #2) GluA2 receptor cluster size increased 22 and 56%, respectively, significantly so for Culture #2 (Table 1). However, in one experiment (Culture #3), GluA2 receptor cluster size showed a downward trend of 9.5%. Because we have 10 CON mean mEPSC amplitudes and 9 TTX mean mEPSC amplitudes that indicate this particular culture very likely has a normal homeostatic increase in mEPSC amplitude induced by activity blockade, the lack of increase in GluA2 receptor cluster characteristics suggests that mEPSC amplitudes can increase in the absence of an increase in receptors. Furthermore, Culture #3’s opposite results from Cultures #1 and #2, explain why in the pooled data set, the increase in GluA2 receptor characteristics after TTX treatment do not reach statistical significance.

In cultures prepared from Rab3A+/+ mice, increases in mEPSC amplitudes following TTX treatment were not always matched by increases in GluA2 receptor cluster characteristics.

(A) Culture #1: mean mEPSC amplitudes for pyramidal neurons from untreated coverslips (CON) and TTX-treated coverslips (TTX) (CON, N = 6, 14.2 ± 2.2 pA; TTX, N = 6, 15.9 ± 1.9 pA), compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken of dendrites from other coverslips in the same culture (CON, N = 10, TTX, N = 9; size, CON, 0.73 ± 0.09 μm2; TTX, 0.89 ± 0.09 μm2; intensity, CON, 682 ± 15 A.U.; TTX, 704 ± 11 A.U.; integral, CON, 256,364 ± 35,242 A.U.; TTX, 316,647 ± 35,077 A.U.). (B) Culture #2: mean mEPSC amplitudes for pyramidal neurons from untreated and TTX-treated coverslips (CON, N = 7, 13.8 ± 2.4 pA; TTX, N = 8, 17.7 ± 1.8 pA), compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken of dendrites from other coverslips from the same culture (CON, N = 9, TTX, N = 9; size, CON, 0.91 ± 0.12 μm2; TTX, 1.42 ± 0.24 μm2; intensity, CON, 621 ± 14 A.U.; TTX, 679 ± 26 A.U.; integral, CON, 296,300 ± 44,108 A.U.; TTX, 512,336 ± 96,892 A.U.). (C) Culture #3: mean mEPSC amplitudes for pyramidal neurons from untreated coverslips and TTX-treated coverslips (CON, N = 10, 13.4 ± 0.8 pA; TTX, N = 9, 15.7 ± 1.0 pA,) compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken from other coverslips from the same culture. (CON, N = 10, TTX, N = 9; size, CON, 1.05 ± 0.11 μm2; TTX, 0.95 ± 0.15 μm2; intensity, CON, 744 ± 28 A.U.; TTX, 714 ± 25 A.U.; integral, CON, 398,870 ± 49,500 A.U.; TTX, 347,571 ± 65,454 A.U.). Cultures were fixed and processed for immunofluorescence on the same day as mEPSC recordings, and imaged within 1 week. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. p-values denoted with * and underline indicates significance of p < 0.05.

Figures 7A-C show the box plots for mEPSC amplitudes, and GluA2 receptor cluster size, intensity and integrals, for 3 individual Rab3A−/− cultures. Here, mEPSC amplitudes showed a downward trend in all 3 cultures, as did GluA2 receptor characteristics for two of the three cultures. But we found another obvious mismatch between the mEPSC amplitude data and the GluA2 receptor cluster data for one of the Rab3A−/− cultures. For Culture #3 (Figure 7C), the box plot of mEPSC mean amplitudes showed a downward trend after TTX treatment, decreasing by 7% (Table 1), whereas GluA2 receptor cluster size, intensity and integral all showed upward trends, the increase ranging from 6 to 23% (Table 1). This result may suggest that the disruption of the homeostatic increase in receptors in the absence of Rab3A is not absolute. It also indicates that synaptic GluA2 receptor levels can increase without an accompanying increase in mEPSC amplitudes.

In cultures prepared from Rab3A−/− mice, a lack of effect of TTX treatment in mEPSC amplitudes was not always matched by a lack of effect in synaptic GluA2 receptor cluster characteristics.

(A) Culture #1: mean mEPSC amplitudes for pyramidal neurons from untreated coverslips (CON) and TTX-treated coverslips (TTX) (CON, N = 7, 15.4 ± 1.2 pA; TTX, N = 6, 14.4 ± 0.8 pA), compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken of dendrites from other coverslips in the same culture (CON, N = 10, TTX, N = 10; size, CON, 0.87 ± 0.07 μm2; TTX, 0.80 ± 0.12 μm2; intensity, CON, 735 ± 10 A.U.; TTX, 731 ± 25 A.U.; integral, CON, 328,477 ± 28,250; TTX, 302,594 ± 53,052 A.U.). (B) Culture #2: mean mEPSC amplitudes for pyramidal neurons from untreated and TTX-treated coverslips (CON, N = 8, 15.2 ± 1.8 pA; TTX, N = 9, 13.5 ± 1.8 pA), compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken of dendrites from other coverslips in the same culture (CON, N = 10, TTX, N = 10; size, CON, 1.05 ± 0.10 μm2; TTX, 0.93 ± 0.05 μm2; intensity, CON, 800 ± 26 A.U.; TTX, 781 ± 21 A.U.; integral, CON, 425,552 ± 49,486 A.U.; TTX, 361,740 ± 21,963 A.U.). (C) Culture #3: mean mEPSC amplitudes for pyramidal neurons from untreated and TTX-treated coverslips (CON, N = 6, 13.9 ± 1.2 pA; TTX, N = 6, 12.9 ± 1.5 pA) compared to mean sizes, intensities, and integrals of synaptic GluA2 receptor clusters in images taken of dendrites from other coverslips in the same culture (CON, N = 10, TTX, N = 10; size, CON, 0.89 ± 0.08 μm2; TTX, 1.03 ± 0.08 μm2; intensity, CON, 765 ± 22 A.U.; TTX, 814 ± 23 A.U.; integral, CON, 354,279 ± 38,755 A.U.; TTX, 434,833 ± 42,350 A.U.). Cultures were fixed and processed for immunofluorescence on the same day as mEPSC recordings, and imaged within 1 week. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test.

In summary, our matched mEPSC and receptor cluster results indicate that: 1. activity blockade resulted in an increase in mEPSC amplitudes usually accompanied by an increase in synaptic GluA2 receptors; 2. loss of Rab3A disrupted both the increase in mEPSC amplitudes and the increase in GluA2 receptor levels; and 3. The differences between homeostatic responses of mEPSC amplitudes and the GluA2 cluster sizes within the same cultures suggest there may be other factors that contribute to mEPSC amplitude.

Having used VGLUT1 immunofluorescence to mark synaptic sites, we could also ask whether there was a presynaptic effect of activity blockade on the size, intensity or integral of the VGLUT1 signal at sites apposed to the previously examined GluA2 postsynaptic sites. In the pooled data from cultures prepared from Rab3A+/+ mice, we found no change in the size of the VGLUT1-positive regions, and trends towards reduced intensity and integral (Table 2). These data do not support the idea that activity blockade caused an increase in the amount of VGLUT1 per vesicle, although we cannot rule out that a simultaneous reduction in the number of vesicles obscured the effect, since we did not have an independent label for synaptic vesicles. In cultures prepared from Rab3A−/− mice, the size, intensity and integral of the VGLUT1-positive regions showed upward trends (Table 2), suggesting the possibility that the modest decreases observed for cultures prepared from wild type mice depend on Rab3A, but it is difficult to make any conclusions when the p values are far from reaching significance. Similarly to the GluA2 receptor cluster data, the likely reason that the pooled data effects did not reach significance is because the effects were not consistent across all three cultures, with one Rab3A+/+ culture showing an upward trend instead of a downward trend, and one Rab3A−/− culture showing a downward trend instead of an upward trend (Table 2). It should be noted that there was no consistent inverse pattern for GluA2 vs. VGLUT1. While in cultures prepared from Rab3A−/− mice, when GluA2 characteristics trended downward, VGLUT1 characteristics trended upward, this pattern did not hold for cultures prepared from Rab3A+/+ mice. In Rab3A+/+ Culture #2, GluA2 showed significant increases, and VGLUT1 showed essentially no change. In Rab3A+/+ Culture #3, both GluA2 and VGLUT1 trended downwards. Taken together, the GluA2 and VGLUT1 immunofluorescence data indicate that GluA2 receptors likely contribute to the increase in mEPSC amplitudes after activity blockade in Rab3A+/+ cultures, and the disruption of that increase in Rab3A−/− cultures, but VGLUT1 levels change in the opposite direction from mEPSC amplitudes. Interestingly, VGLUT1 levels are strongly affected by loss of Rab3A, with a highly significant decrease in integral in untreated cultures from Rab3A−/− mice, compared to untreated cultures from Rab3A+/+ mice (size, p = 0.019; intensity, p = 0.004; integral, p = 4.6 * 10−4; Table 2).

We have demonstrated here for the first time that the synaptic vesicle protein Rab3A influences the homeostatic regulation of postsynaptic GluA2 receptors. What are possible ways that Rab3A could exert a postsynaptic action? It has previously been shown that exogenous addition of TNFα to hippocampal cultures causes an increase in surface expression of GluA1 receptors (although not GluA2 receptors), and that the homeostatic increase in mEPSC amplitudes is abolished in cultures prepared from the TNFα deletion mouse (Stellwagen et al., 2005; Stellwagen and Malenka, 2006). Furthermore, neurons from TNFα+/+ mice plated on glial feeder layers derived from TNFα−/− mice fail to show the increase in mEPSC amplitude after TTX treatment, indicating that the TNFα inducing the receptor increases following activity blockade comes from the glial cells (Stellwagen and Malenka, 2006). It has recently been shown that the source of TNFα is astrocytes rather than microglia (Heir et al., 2024). Rab3A has been detected in astrocytes (Maienschein et al., 1999; Hong et al., 2016), so to determine whether Rab3A is acting via regulating TNFα release from astrocytes, we performed experiments similar to those of Stellwagen and Malenka (2006) (schema illustrated in Figure 8, Left). It should be noted that under our culturing conditions, the glial feeder layers would be expected to be composed mainly of astrocytes (Heir et al., 2024). We compared the effect of activity blockade on mEPSC amplitudes recorded from cortical neurons from Rab3A+/+ mice plated on Rab3A+/+ glial feeder layers (Figure 8A); neurons from Rab3A+/+ mice plated on Rab3A−/− glia (Figure 8B), and neurons from Rab3A−/− mice plated on Rab3A+/+ glia (Figure 8C). If Rab3A is required for TNFα release from glia, then Rab3A+/+ neurons plated on Rab3A−/− glia should not show a homeostatic increase in mEPSC amplitude after treatment with TTX, and any cultures with Rab3A+/+ glia should have a normal homeostatic response. We found the opposite result: mEPSC amplitudes increased dramatically in cultures where Rab3A was present in neurons (WT on WT, CON, 13.3 ± 0.5 pA, TTX, 16.7 ± 1.2 pA; WT on KO, CON, 13.3 ± 1.0 pA, TTX, 18.8 ± 1.4 pA), but the increase was greatly diminished in cultures where Rab3A was present only in glia (KO on WT, CON, 15.2 ± 1.1 pA, TTX, 16.9 ± 0.7 pA). It is interesting to note that in the glial feeder layer culture, the loss of Rab3A in the neurons may have caused an increase in mEPSC amplitude at baseline, similar to the Earlybird mutant in dissociated cultures, but it is also possible this can be attributed to culture to culture variability. Regarding mEPSC frequency, in cultures with Rab3A+/+ neurons, frequency trended upward after activity blockade, whereas in the culture with Rab3A−/− neurons, frequency was already increased, and trended downward after activity blockade.

Rab3A in neurons, not astrocytes, was required for full TTX-induced homeostatic plasticity.

(A)-(C), mEPSC amplitude (Middle) and frequency (Right) data from dissociated cortical neurons plated on an astrocyte feeder layer, each prepared separately from the type of mice depicted in the schema (Left): (A) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A+/+ mice.

Box plots for mEPSC amplitudes (CON, N = 17 cells, 13.3 ± 0.5 pA; TTX, N = 20 cells, 16.7 ± 1.2 pA) and mEPSC frequency in the same recordings (mean, CON, 2.54 ± 0.57 sec−1; TTX, 3.48 ± 0.64 sec−1); from 4 cultures. (B) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A−/− mice. Box plots for average mEPSC amplitude (CON, N = 11 cells, 13.3 ± 1.0 pA; TTX, N = 11 cells, 18.8 ± 1.4 pA) and average mEPSC frequency in the same recordings (CON, 2.01 ± 0.41 sec−1; TTX, 4.47 ± 1.53 sec−1); from 2 cultures. (C) Neurons from Rab3A−/− neurons plated on astrocytes from Rab3A+/+ mice. Box plots for average mEPSC amplitude (CON, N = 14 cells, 15.2 ± 1.1 pA; TTX, N = 11 cells, 16.9 ± 0.7 pA) and mEPSC frequency in the same recordings (CON, 4.47 ± 1.21 sec−1; TTX, 3.02 ± 0.70 sec−1); from 3 cultures. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. p-values denoted with * and underline indicates significance of p < 0.05.

In summary, in the absence of Rab3A from glia, the homeostatic plasticity of mEPSC amplitude following activity blockade was completely normal, whereas in the absence of Rab3A in neurons, homeostatic plasticity was greatly diminished. This result makes it highly unlikely that Rab3A is required for the release of TNFα, or another factor from glia, that induces a homeostatic upregulation of postsynaptic receptors and thereby increases mEPSC amplitude following TTX treatment. Neuronal Rab3A appears to mediate the homeostatic increase in mEPSC amplitude following activity blockade.

Discussion

We found that homeostatic synaptic plasticity of mEPSC amplitude in dissociated mixed cultures of mouse cortical neurons and glia behaved remarkably similar to the mouse NMJ in response to loss of Rab3A function: in cultures from Rab3A−/− mice, the increase in mEPSC amplitudes following prolonged network silencing by TTX was strongly diminished, and in cultures from Rab3AEbd/Ebd mice, basal mEPSC amplitude was increased compared to that of wild-type cultures and was not further modified by TTX treatment. These results suggest that normal function of the synaptic vesicle protein Rab3A is required for the homeostatic increase of mEPSC amplitude in cortical cultures. The effect of TTX on mEPSC amplitude was not due to increases in Ca2+-permeable homomers because acute NASPM application was unable to reverse the increase in amplitude after TTX treatment. We used imaging of immunofluorescence to demonstrate a strong trend towards increased GluA2 levels at synapses in activity-blocked cultures from Rab3A+/+ mice that was no longer apparent in the absence of Rab3A, suggesting the surprising idea that this protein, previously implicated primarily in the regulation of calcium-dependent vesicle fusion, is also important for the regulation of surface expression of postsynaptic receptors. We found a significant increase in mEPSC frequency after activity blockade that was no longer significant in the absence of Rab3A, which was less surprising as frequency is known to depend on presynaptic factors. At the neuromuscular junction, higher mEPC frequencies that occur in 0.5 mM extracellular calcium and above were significantly reduced in the Rab3A deletion mouse (Coleman et al., 2007). On the other hand, the increase in frequency after TTX treatment is also strongly diminished after acute application of NASPM, suggesting there may be an increase in the number of synapses with only Ca2+-permeable homomers. It remains to be determined if this effect is Rab3A-dependent or if there are two distinct mechanisms contributing to homeostatic plasticity of frequency.

The levels of the glutamate transporter, VGLUT1, were not increased after activity blockade under our experimental conditions. A similar lack of increase in VGLUT1 levels, relative to synapsin in imaging of immunofluorescence, was previously observed in hippocampal cultures treated with NBQX, an AMPA receptor blocker (see supplemental data, (Wilson et al., 2005)). On the other hand, treatment with AP5, an NMDA receptor blocker, was able to induce an increase in VGLUT1 levels relative to synapsin (Wilson et al., 2005), suggesting that this presynaptic response depends on the type of activity manipulation. Another study showed increased VGLUT1 mRNA and protein in western blots in cortical cultures treated with TTX, but only examined VGLUT1 at synapses using imaging following treatment with bicuculline (De Gois et al., 2005). Taken together with our findings, it appears that there is not strong evidence that the amount of VGLUT1 at synapses is a contributing factor to the increase in mEPSC amplitude after activity blockade with TTX.

Homeostatic plasticity of postsynaptic receptors was less consistent than that of mEPSC amplitude

While we found that the effect of TTX treatment on mEPSC amplitude and GluA2 receptor levels were of very similar magnitude, the effect on GluA2 receptor levels was not statistically significant for the pooled data sets acquired in the same cultures. We determined that this discrepancy was due to one of the three cultures showing a downward trend in GluA2 receptor levels after activity blockade. In light of the extensive literature documenting increases in GluA1 or GluA2 receptors, or both, after activity blockade, the less than convincing increase in GluA2 receptor levels in the current work came as a surprise. Therefore we reviewed what has previously been shown, and were able to identify only 5 studies that examined the effect of activity blockade using TTX alone, reported levels of AMPA receptors specifically localized to synapses, and recorded mEPSCs in the same study, although not in the same cultures (Hou et al., 2008; Ibata et al., 2008; Jakawich et al., 2010a; Xu and Pozzo-Miller, 2017; Dubes et al., 2022). Since our issue is not with the magnitude of the changes being different for receptors and mEPSC amplitudes, but with statistical significance (or consistency) being different between the two methods, we wanted to determine whether the previous studies showed any evidence of greater variability in receptor measurements. Only a single study had comparable sample sizes for receptors and mEPSC amplitudes (either because number of receptor sites is used instead of cells, or because the number of cells differs for the two types of experiments), and the p value was < 0.01 for mEPSC amplitudes (n = 9 and 10 cells for CON and TTX, respectively), but < 0.05 for receptor data (n = 13 and 14 cells for CON and TTX) (Xu and Pozzo-Miller, 2017). This single example is consistent with the idea that the receptor data are more variable than mEPSC data. Even if a difference in consistency could be established in multiple studies, it remains a possibility that the disparity is attributable to the unique limitations of the two types of experiments. Nevertheless, the finding that mEPSC amplitude can increase despite a decrease in receptors suggests receptors may not be the only contributor to the TTX effect on mEPSC amplitude.

Neuronal, not glial, Rab3A may be required for the homeostatic release of a signaling molecule

Our results support the surprising possibility that the presynaptic vesicle protein Rab3A regulates postsynaptic AMPARs. Astrocytic release of TNFα was shown to mediate the increase in mEPSC amplitudes in prolonged TTX-treated cultures of dissociated hippocampal neurons and cultures of hippocampal slices (Stellwagen and Malenka, 2006; Heir et al., 2024), but we found that loss of Rab3A in glia did not disrupt the increase in mEPSC amplitude after activity blockade. Further, we previously showed that the homeostatic increase in NMJ mEPC amplitude was completely normal in the absence of TNFα (Wang et al., 2011). This leaves neuronal Rab3A as the potential controller of a soluble factor regulating surface expression of receptors.

One possible signaling molecule being controlled by Rab3A is brain-derived neurotrophic factor (BDNF). Addition of exogenous BDNF to neuronal cultures prevents the increase in mEPSC amplitude following activity blockade (Rutherford et al., 1998; Benevento et al., 2016) but see (Smith-Dijak et al., 2019), and BDNF mRNA is reduced after activity blockade in vitro (Benevento et al., 2016; Miyasaka and Yamamoto, 2021) and in vivo (Castren et al., 1992). Also, after reduction of secreted BDNF in culture media via the BDNF scavengers TrkB-FC or TrkB-IgG, mEPSC amplitudes are increased (Rutherford et al., 1998; Benevento et al., 2016; Smith-Dijak et al., 2019). Reduction of Rab3A in astrocytes decreases astrocytic BDNF release (Hong et al., 2016). Taken together, these studies suggest that activity blockade could cause a Rab3A-dependent inhibition of BDNF release, leading to an increase in postsynaptic receptors. However, we have no evidence that Rab3A is required for the regulation of BDNF release from neurons, so it is just as likely that after activity blockade, Rab3A acts in a positive way to promote release of a trophic molecule that stimulates the insertion of postsynaptic receptors. It is unlikely that Rab3A is required when activity is normal because receptor levels do not change in the same way with loss of Rab3A as they do with loss of activity. Rather, our data fit with a model in which Rab3A mainly becomes engaged in response to activity blockade (Figure 9).

Model depicting the requirement for neuronal Rab3A in the homeostatic increase in synaptic GluA2 receptor levels after treatment with TTX.

Under normal activity levels (Left), neither presynaptic or postsynaptic Rab3A is engaged in the regulation of GluA2 receptor levels. After prolonged activity blockade (Right), Rab3A located in the presynaptic terminal promotes the release of a trophic factor (orange dots) that diffuses across the synapse and facilitates transport of postsynaptic vesicles containing GluA2 AMPA receptors (dark orange channels) to the surface membrane. Rab3A located in the postsynaptic dendrite directly promotes fusion of vesicles containing GluA2 AMPA receptors. Fusion of presynaptic vesicles containing glutamate (black dots) causes mEPSCs when glutamate binds to AMPA receptors. Adapted from “Synaptic Cleft” template, by BioRender.com (2024).

Is Rab3A acting in the postsynaptic cell to cause homeostatic increase in mEPSC amplitude?

To our knowledge, there is no evidence localizing Rab3A to postsynaptic sites. Other presynaptic molecules, such as SNARE proteins, synaptotagmins, and NSF, have been identified in the postsynaptic compartment (Lledo et al., 1998; Nishimune et al., 1998; Osten et al., 1998; Song et al., 1998; Araki et al., 2010; Kennedy et al., 2010; Suh et al., 2010; Jurado et al., 2013; Hussain and Davanger, 2015; Gu et al., 2016; Wu et al., 2017) but Rab3A localization was not examined in those studies. On the other hand, a recent report implicates Rab3A in the localization of plasma membrane proteins in rafts in HEK cells and Jurkat-T cells, indicating Rab3A can be found in locations other than the presynaptic nerve terminal (Diaz-Rohrer et al., 2023). In this study a reduction in all Rab3 isoforms led to an impairment of plasma membrane localization of epidermal growth factor receptor (EGFR), so it is possible that Rab3A is acting similarly in postsynaptic dendrites to promote transport of GluA2 receptors to the surface after activity blockade (Figure 9). Another possibility, not depicted in Figure 9, is that Rab3A acts postsynaptically to regulate the release of a trophic factor that in turn regulates surface expression of GluA2 receptors. This idea is consistent with evidence that BDNF is synthesized and released locally from dendrites (Jakawich et al., 2010b; Thapliyal et al., 2022).

Alternate contributors to homeostatic increase in mEPSC amplitude

We do not think Rab3A impacts homeostatic synaptic plasticity solely through pre-or postsynaptic regulation of GluA2 receptor trafficking to the plasma membrane. First, prolonged activity blockade with TTX at the NMJ led to an increase in mEPC amplitude that was not accompanied by any increase in AChR levels, yet it was dependent on Rab3A (Wang et al., 2011). Second, mEPSC amplitudes increased in three cortical cultures treated with TTX, but GluA2 receptor levels in one of the cultures actually went down. An increase in conductance could explain the increase in mEPSC amplitudes in the absence of receptor level increases, but all evidence to date suggests that the only way to increase conductance is to either switch from GluA2-containing heteromeric receptors to GluA1 homomers, which have much greater conductance (Oh and Derkach, 2005; Soares et al., 2013; Hou et al., 2015; Xu and Pozzo-Miller, 2017; Benke and Traynelis, 2019; Silva et al., 2019; Dubes et al., 2022), or, to post-translationally modify the receptor. Such conductance-altering modifications have been demonstrated only for GluA1 (Derkach et al., 1999; Kristensen et al., 2011; Jenkins and Traynelis, 2012). Importantly, any change in conductance due to an increase in GluA1 homomers or modification of GluA1 should have been detected by NASPM being able to disrupt the homeostatic increase in mEPSC amplitude, and this was not what we observed.

The other contributor to mEPSC amplitude besides receptors is the size of the presynaptic quantum, or the amount of transmitter released during a single fusion event. As stated above, we found a downward trend in VGLUT1 levels after activity blockade. This downward trend was not apparent in cultures prepared from Rab3A−/− mice, suggesting a possible Rab3A-dependent effect. Interestingly, Rab3A levels are reduced 50% in hippocampal extracts from mice lacking VGLUT1 (Fremeau et al., 2004). Coupled with our finding that loss of Rab3A was associated with a highly significant decrease in VGLUT1 levels in untreated cultures, these results suggest an interdependence between Rab3A and VGLUT1. More importantly, our results rule out VGLUT1 levels as the source for increased mEPSC amplitude.

The amount of transmitter released during vesicle fusion can be affected by the kinetics of the fusion pore opening (Chang et al., 2017). We previously used amperometric measurements in mouse adrenal chromaffin cells to show loss of Rab3A increased the occurrence of very small amplitude fusion pore feet (Wang et al., 2008). Paired with our finding that there is an increase in the occurrence of very slow-rising, abnormally shaped mEPCs at the neuromuscular junction of Rab3A−/− mice, our data suggest that small synaptic vesicles may have a fusion pore step under certain circumstances (see also (Chiang et al., 2018)). If activity blockade causes an increase in mEPSC amplitude due to a more rapid opening of a fusion pore, or a larger fusion pore conductance, it is possible that Rab3A is required for this modulation.

Finally, a simple way to increase the amount of transmitter released by a vesicle is to enlarge vesicle size. Larger vesicle diameter has been observed following activity disruption in multiple paradigms, including ocular deprivation, cortical ablation, and immobilization (Kawana et al., 1971; Lloret and Saavedra, 1975; Cheresharov et al., 1978). Loss of Rab3 family members is associated with increased vesicle size (Riedel et al., 2002; Schonn et al., 2010). Finally, an increase in volume has been shown to increase miniature endplate junctional current amplitude at the Drosophila NMJ (Karunanithi et al., 2002). Rab3A could be required for an activity-dependent increase in vesicle size, but the increase in diameter needed for a 25% increase in transmitter, and therefore mEPSC amplitude, would be difficult to detect.

To our knowledge, this is the first evidence of a presynaptic protein being implicated in the homeostatic regulation of mEPSC amplitude and AMPARs in neuronal cultures. Homeostatic synaptic plasticity is becoming increasingly implicated in essential normal functions such as sleep (Tononi and Cirelli, 2014; Diering et al., 2017; Torrado Pacheco et al., 2021), and pathological conditions such as epilepsy, neuropsychiatric disorders, Huntington’s, and alcohol use disorder (Trasande and Ramirez, 2007; Fernandes and Carvalho, 2016; Wang et al., 2017; Lovinger and Abrahao, 2018; Smith-Dijak et al., 2019; Lignani et al., 2020; Suzuki et al., 2021; Kavalali and Monteggia, 2023). Therefore our identification of a novel regulatory pathway involving Rab3A may have important therapeutic implications. In future work it will be important to determine whether Rab3A is acting presynaptically or postsynaptically to regulate GluA2 receptors, and what other contributor to mEPSC amplitude depends on Rab3A, possibly fusion pore characteristics or vesicle volume.

Materials and methods

Animals. To determine the role of Rab3A in homeostatic plasticity in mouse cortical cultures, we employed two distinct genetic mouse strains with altered Rab3A function. Rab3A deletion mice were generated as follows: Rab3A+/− heterozygous mice were bred and genotyped as previously described (Kapfhamer et al., 2002; Wang et al., 2008), and maintained as a colony using multiple heterozygous breeding pairs. Homozygous Rab3A+/+ or Rab3A−/− mouse pups were obtained by breeding Rab3A+/+ pairs and Rab3A−/− pairs, respectively, so that the pups were the first generation progeny of a homozygous mating. This protocol minimizes the tendency of multiple generations of Rab3A+/+ and Rab3A−/− homozygous breedings to produce progeny that are more and more genetically distinct. Rab3AEbd/Ebd mice were identified in an EU-mutagenesis screen of C57BL/6J mice, and after a cross to C3H/HeJ, were backcrossed for 3 generations to C57BL/6J (Kapfhamer et al., 2002). Rab3A+/Ebd heterozygous mice were bred at Wright State University and genotyped in a two-step procedure: 1. a PCR reaction (forward primer: TGA CTC CTT CAC TCC AGC CT; reverse primer: TGC ACT GCA TTA AAT GAC TCC T) followed by 2. a restriction digest with enzyme Bsp1286I (New England Biolabs) that distinguishes the Earlybird mutant by its different base-pair products. Rab3A+/− mice were backcrossed with Rab3A+/+ mice from the Earlybird heterozygous colony for 11 generations in an attempt to establish a single wild type strain, but differences in mEPSC amplitude and adrenal chromaffin cell calcium currents persisted, likely due to genes that are close to the Rab3A site, resulting in two wild type strains: 1. Rab3A+/+ from the Rab3A+/− colony, and 2. Rab3A+/+ from the Rab3A+/Ebd colony.

Primary Culture of Mouse Cortical Neurons. Primary dissociated cultures of mixed neuronal and glia populations were prepared as previously described (Hanes et al., 2020). Briefly, postnatal day 0-2 (P0-P2) Rab3A+/+, Rab3A−/− or Rab3AEbd/Ebd neonates were euthanized by rapid decapitation, as approved by the Wright State University Institutional Animal Care and Use Committee, and brains were quickly removed. Each culture was prepared from the cortices harvested from two animals; neonates were not sexed. Cortices were collected in chilled Neurobasal-A media (Gibco) with osmolarity adjusted to 270 mOsm and supplemented with 40 U/ml DNAse I (ThermoFisher Scientific). The tissues were digested with papain (Worthington Biochemical) at 20 U/ml at 37°C for 20 minutes followed by trituration with a sterile, fire-polished Pasteur pipette, then filtered through a 100 μm cell strainer, and centrifuged at 1100 rpm for 2 minutes. After discarding the supernatant, the pellet was resuspended in room temperature Neurobasal-A media (270 mOsm), supplemented with 5% fetal bovine serum for glia growth, and 2% B-27 supplement to promote neuronal growth (Gibco), L-glutamine, and gentamicin (ThermoFisher Scientific). Neurons were counted and plated at 0.15 * 106 cells/coverslip onto 12 mm coverslips pre-coated with poly-L-lysine (BioCoat, Corning). This plating density produces a “high density” culture characterized by a complex mesh of neuronal processes criss-crossing the field of view, completely filling the space between cell bodies (see Figure 4). We found in preliminary experiments that an increase in mEPSC amplitude in mouse cortical cultures was inconsistent when cell density was low, likely due to lower levels of baseline activity, although it is a limitation of this study that we did not directly measure activity levels in cultures prepared from Rab3A+/+, Rab3A−/− or Rab3AEbd/Ebd mice, and therefore do not know if differences in results are due to differences in activity levels. While unlikely for Rab3A+/+ and Rab3A−/− studies, given the mEPSC amplitude in untreated cultures that was relatively small, it remains possible that in Rab3AEbd/Ebd studies, the larger mEPSC amplitude in untreated cultures was due to a loss of activity in these cultures. The culture media for the first day (0 DIV) was the same as the above Neurobasal-A media supplemented with FBS, B-27, L-glutamine, and gentamicin, and was switched after 24 hours (1 DIV) to media consisting of Neurobasal-A (270 mOsm), 2% B-27, and L-glutamine without FBS to avoid its toxic effects on neuronal viability and health (Stellwagen and Malenka, 2006). Half of the media was changed twice weekly and experiments were performed at 13-14 DIV. Two days prior to experiments, tetrodotoxin (TTX) (500 nM; Tocris), a potent Na+ channel blocker, was added to some cultures to chronically silence all network activity and induce homeostatic synaptic plasticity mechanisms, while untreated sister cultures served as controls. Cultures prepared from mutant mice were compared with cultures from wild-type mice from their respective colonies. Note that the cultures comprising the Rab3A+/+ data here are a subset of the data previously published in Hanes et al., 2020. This smaller data set was restricted to the time period over which cultures were also prepared from Rab3A−/− mice.

Preparation of Glial Feeder Layers. Glial feeder layers were prepared from the cortices of P0-P2 Rab3A+/+ or Rab3A−/− mouse pups as described previously (Stellwagen and Malenka, 2006). Briefly, cortices were dissected and cells were dissociated as described above. Cell suspensions of mixed neuronal and glial populations were plated onto glass coverslips pre-coated with poly-L-lysine in Dulbecco’s Modified Eagle Media (ThermoFisher Scientific) supplemented with 5% FBS (to promote glial proliferation and to kill neurons), L-glutamine, and gentamicin, and maintained in an incubator at 37°C, 5% CO2; cultures were maintained in this manner for up to 1 month to generate purely glial cultures (all neurons typically died off by 7 DIV). Culture media was replaced after 24 hours, and subsequent media changes were made twice weekly, replacing half of the culture media with fresh media. Feeder layers were not used for neuronal seeding until all native neurons were gone and glia approached 100% confluency (visually inspected).

Plating of Neurons on Glial Feeder Layers. Cortical neurons were obtained as described above. The cell pellet obtained was resuspended in Neurobasal-A (osmolarity adjusted to 270 mOsm) containing B27 (2%, to promote neuronal growth), L-glutamine, and 5-fluorodeoxyuridine (FdU, a mitotic inhibitor; Sigma). Addition of FdU was used to prevent glial proliferation and contamination of the feeder layer with new glia, promoting only neuronal growth on the feeder layers (FdU-containing media was used for the maintenance of these cultures and all subsequent media changes). Glial culture media was removed from the feeder layer cultures, and the neuronal cell suspension was plated onto the glial feeder cultures. The culture strategy used to distinguish the relative roles of neuronal and glial Rab3A is outlined in Figure 8, Left. At 1 DIV, all of the culture media was removed and replaced with fresh Neurobasal-A media containing FdU described above, and half of the media was replaced twice per week for all subsequent media changes. Cultures were maintained in a 37°C, 5% CO2 incubator for 13-14 DIV.

Whole-Cell Voltage Clamp to Record mEPSCs. At 13-14 DIV, mEPSCs from TTX-treated and untreated sister cultures of Rab3A+/+ or Rab3A−/− neurons from the Rab3A+/− colony, or Rab3A+/+ or Rab3AEbd/Ebd neurons from the Rab3A+/Ebd colony, were recorded via whole-cell voltage clamp to assess the role of Rab3A in homeostatic synaptic plasticity. Recordings were taken from pyramidal neurons, which were identified visually by a prominent apical dendrite; images were taken of all cells recorded from. Cells were continuously perfused with a solution consisting of (in mM): NaCl (115), KCl (5), CaCl2 (2.5), MgCl2 (1.3), dextrose (23), sucrose (26), HEPES (4.2), pH = 7.2 (Stellwagen and Malenka, 2006). On the day of recording, the osmolarity of the media from the cultures was measured (normally 285 – 295 mOsm) and the perfusate osmolarity was adjusted to match the culture osmolarity, to protect against osmotic shock to the neurons. To isolate glutamatergic mEPSCs, TTX (500 nM) and picrotoxin (50 μM) were included in the perfusion solution to block action potentials and GABAergic currents, respectively. The NMDA receptor antagonist, APV, was not included in the perfusion solution because in pilot studies, all mEPSCs were blocked by 10 µM CNQX and 50 μm picrotoxin, demonstrating no APV-sensitive mEPSCs were present (data not shown). Previous studies have described NMDA mEPSCs in neuronal cultures, but recordings were performed in 0 Mg2+ (Watt et al., 2000; Sutton et al., 2006). In our recordings, the presence of extracellular Mg2+ (1.3 mM) and TTX would mean that the majority of NMDA receptors are blocked by extracellular Mg2+ at resting Vm. Patch electrodes (3.5 – 5 MΩ) were filled with an internal solution containing (in mM): K-gluconate (130), NaCl (10), EGTA (1), CaCl2 (0.132), MgCl2 (2), HEPES (10), pH = 7.2. Osmolarity was adjusted to 10 mOsm less than the perfusion solution osmolarity, which we have found improves success rate of achieving low access resistance recordings. Neurons were clamped at a voltage of –60 mV using an Axopatch 200B patch-clamp (Axon Instruments), recorded from for 2 – 5 minutes, and data were collected with Clampex 10.0/10.2 (Axon Instruments).

NASPM application. The antagonist of Ca2+-permeable AMPA receptors, N-naphthyl acetylspermine (NASPM, 20 μM; Tocris), was applied during recordings in a subset of experiments. NASPM is a synthetic analog of Joro Spider Toxin (JSTX) (Koike et al., 1997), and chemicals in this family block Ca2+ permeable glutamate receptors (Blaschke et al., 1993; Herlitze et al., 1993; Iino et al., 1996). The presence of the edited GluA2 subunit disrupts both the Ca2+ permeability and the sensitivity to JSTX and related substances (Blaschke et al., 1993; Bochet et al., 1994; Jonas and Burnashev, 1995). Therefore, 20 μM NASPM is expected to completely block AMPA receptors containing GluA1, GluA3 or GluA4 subunits (Hollmann et al., 1991; Burnashev et al., 1992), or the Kainate receptors containing GluA5 or GluA6 (Koike et al., 1997; Sun et al., 2009), but be ineffective against any heteromer or homomer containing GluA2 subunits. Because the effect of the spider toxins and their analogues are use-dependent (Herlitze et al., 1993; Iino et al., 1996; Koike et al., 1997), NASPM was applied with a mildly depolarizing high K+ solution (25 mM KCl, 95 mM NaCl) which we expected to trigger release of glutamate and opening of glutamate-activated receptors, allowing entry of the pore blocker. Baseline recordings were performed for 2 minutes in our standard perfusate, then were suspended while NASPM + High K+ solution was applied for 45 seconds, followed by a NASPM only solution for 5 minutes, after which recording was recommenced for 5 minutes (because we found in pilot experiments that frequency was reduced following NASPM application). Analysis of mEPSCs. Miniature excitatory postsynaptic currents (mEPSCs) were manually selected using Mini Analysis software (Synaptosoft), a standard program used for mEPSC event detection and analysis. Records were filtered at 2 kHz using a low-pass Butterworth filter prior to selection. The program threshold was set at 3 pA but the smallest mEPSC selected was 4.01 pA. MiniAnalysis selects many false positives with the automated feature when a small threshold amplitude value is employed, due to random fluctuations in noise, so manual re-evaluation of the automated process is necessary to eliminate false positives. If the threshold value is set high, there are few false positives but small amplitude events that visually are clearly mEPSCs are missed, and manual re-evaluation is necessary to add back false negatives or the population ends up biased towards large mEPSC amplitudes. As soon as there is a manual step, bias is introduced. Interestingly, a manual reevaluation step was applied in a recent study that describes their process as ‘unbiased’ (Wu et al., 2020). In sum, we do not believe it is currently possible to perform a completely unbiased detection process. A fully manual detection process means that the same criterion (“does this look like an mEPSC?”) is applied to all events, not just the false positives, or the false negatives, which prevents the bias from being primarily at one end or the other of the range of mEPSC amplitudes. It is important to note that when performing the MiniAnalysis process, the researcher did not know whether a record was from an untreated cell or a TTX-treated cell.

Immunofluorescence, microscopy, and data analysis. Primary cultures of mouse cortical neurons were grown for 13-14 DIV. Antibodies to GluA2 (mouse ab against N-terminal, EMD Millipore) were added directly to live cultures at 1:40 dilution, and incubated at 37 °C in a CO2 incubator for 45 minutes. Cultures were rinsed 3 times with PBS/5% donkey serum before being fixed with 4% paraformaldehyde. After 3 rinses in PBS/5% donkey serum, cultures were incubated in CY3-labeled donkey-anti-mouse secondary antibodies for 1 hour at room temperature, rinsed in PBS/5% donkey serum, permeabilized with 0.2% saponin, and incubated in chick anti-MAP2 (1:2500, AbCAM) and rabbit anti-VGLUT1 (1:4000, Synaptic Systems) for 1 hour at room temperature in PBS/5% donkey serum. After rinsing with PBS/5% donkey serum, coverslips were incubated with 488-anti chick and CY5-anti rabbit secondary antibodies for 1 hour at room temperature, rinsed, blotted to remove excess liquid, and inverted on a drop of Vectashield (Vector Labs). Coverslips were sealed with nail polish and stored at 4 °C for < 1 week before imaging. All secondary antibodies were from Jackson Immunoresearch and were used at 1:225 dilution.

Coverslips were viewed on a Fluoview FV1000 laser scanning confocal microscope with a 60x oil immersion, 1.35 NA objective. Once a pyramidal neuron was identified and a single confocal section of the cell acquired (see Top, Figure 4), Fluoview 2.1 software was used to zoom in on the primary dendrite (5X) and a series of confocal sections were taken every 0.5 μm. Images were analyzed offline with ImagePro 6 (Cybernetics). The composite image was used to locate synaptic sites containing both VGLUT1 and GluA2 immunoreactivity in close apposition to each other and to the primary dendrite or a secondary branch identified with MAP2 immunoreactivity. An area of interest (AOI) was manually drawn around the GluA2 cluster in the confocal section in which it was the brightest. The AOIs for a dendrite were saved in a single file; the AOI number and the confocal section it was associated with were noted for later retrieval. For quantification, AOIs were loaded, an individual AOI was called up on the appropriate section, and the count/measurement tool used to apply a threshold. Multiple thresholds were used, 400-550, but we found that within a given experiment, the magnitude of the TTX effect was not altered by different thresholds, therefore, all data presented here use a threshold of 450. Pixels within the cluster that were above the threshold were automatically outlined, and size, average intensity, and integral of the outlined region saved to Origin2020 (OriginLab) file for calculating mean values. For analyzing VGLUT1 sites, previous AOIs were examined, and if not directly over the VGLUT1 site, moved to encompass it (this happened when GluA2 and VGLUT1 were side by side rather than overlapping). The threshold was 200 for all VGLUT1 data except Rab3A+/+ Culture #1, where threshold was 400. There was no threshold that worked for both that culture and the other cultures, with too much of the background regions being included for Culture #1 with a threshold of 200, and too little of the synaptic site area being included for the other Cultures with a threshold of 400.

Statistics. In our previous publication (Hanes et al., 2020), we described the characteristics of homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures using multiple mEPSC amplitude quantiles from each cell to create data sets with thousands of samples. We compared the cumulative distribution functions for mEPSC amplitudes in untreated vs. TTX-treated cultures, as well as sorting the values from smallest to largest, and plotted TTX values vs. control values in the rank order plot first described in (Turrigiano et al., 1998) to demonstrate synaptic scaling. Finally, we plotted the ratio of TTX values/control values vs. control values to show that the ratio, which represents the effect of TTX, increased with increasing control values, which we termed “divergent scaling.” Here, we wanted to avoid the inflation of sample size caused by pooling multiple mEPSCs per cell in order to statistically compare the effects of TTX in cultures of wild type mice, Rab3A−/− mice and Rab3AEbd/Ebd mutant mice. Therefore, we computed the mean mEPSC amplitude for each cell and compared means from untreated and TTX-treated cultures with the non-parametric Kruskal-Wallis test, with N = the number of cells. The overall means are presented ± SEM in the text, and p values are displayed above each plot. The same process was carried out for the characteristics of synaptic GluA2 receptor clusters and VGLUT1 sites in imaging experiments, with N = the number of dendrites (one dendrite was sampled for each cell). Although we quantified three characteristics of the GluA2 clusters and VGLUT1 sites, size, average intensity, and integral, we have not corrected p values for multiple comparisons. Means are presented as box plots + data, with the box at 25 and 75 percentiles, whiskers at 10 and 90 percentiles, a dot indicating the mean, and a line indicating the median. In Figure 3, mEPSC amplitudes were measured within the same cell before and after NASPM application, so a paired T test was used to compare the pre– and post-NASPM mEPSC amplitudes. Statistics were computed and plots were created using Origin2020 (OriginLab).

eLife Assessment

This study presents useful findings on the role of the small GTPase Rab3A in homeostatic synaptic plasticity following activity suppression. While the study demonstrates that Rab3A is required for homeostatic scaling, the evidence supporting the model put forward by the authors is incomplete. The work will be of interest to the field of synaptic transmission and synaptic plasticity.

Reviewer #1 (Public review):

Koesters and colleagues investigated the role of the small GTPase Rab3A in homeostatic scaling of miniature synaptic transmission in primary mouse cortical cultures using electrophysiology and immunohistochemistry. The major finding is that TTX incubation for 48 hours does not induce an increase in the amplitude of excitatory synaptic miniature events in neuronal cultures derived from Rab3A KO and Rab3A Earlybird mutant mice. NASPM application had comparable effects on mEPSC amplitude in control and after TTX, implying that Ca2+-permeable glutamate receptors are unlikely modulated during synaptic scaling. Immunohistochemical analysis revealed no significant changes in GluA2 puncta size, intensity, and integral in control and Rab3A KO cultures. Finally, they provide evidence that loss of Rab3A in neurons, but not astrocytes, blocks homeostatic scaling. Based on these data, the authors propose a model in which neuronal Rab3A is required for homeostatic scaling of synaptic transmission through GluA2-dependent and independent mechanisms.

While the title of the manuscript is mostly supported by data of solid quality, many conclusions, as well as the final model, cannot be derived from the results presented. Importantly, the data do not support that GluA2 levels change upon TTX treatment in control cultures, rendering conclusions regarding Rab3A's role in TTX-dependent GluA2 modulation spurious. Other aspects of the model, such as a Rab3A-dependent release of a tropic factor, cannot be derived from the data.

The following points should be addressed:

(1) There is no (significant) increase in GluA2 levels (intensity, area, or integral) upon TTX treatment in controls (Fig. 5). Conclusions regarding Rab3As role in TTX-dependent GluA2 modulation should be revised accordingly. Hence, the data shown in Fig. 4 - 7 do not allow drawing conclusions in the context of Rab3A-dependent GluA2 modulation and scaling.

(2) The effects of Rab3A on TTX-induced mini frequency modulation remains unclear, because TTX does not induce a change in mini frequency in the Rab3A+/Ebd control (Fig. 2). The respective conclusions should be revised accordingly (l. 427).

(3) The model is still not supported by the data. In particular, data supporting a negative regulation of Rab3A by APs, Rab3A-dependent release of a tropic factor, or a Rab3A-dependent increase in GluA2 abundance are not presented.

(4) Data points are not overlapping and appear "quantal" in most box plots. How were the data rounded?

Reviewer #2 (Public review):

In the revised manuscript, the authors investigated the role of a presynaptic protein, Rab3A, in the homeostatic synaptic plasticity in cultured cortical neurons. The study was motivated by their previous findings that Rab3A is required for expression of similar homeostatic mechanisms at the neuromuscular junction. The authors first show that untreated WT neurons express homeostatic synaptic plasticity in response to 48h of TTX treatment (upregulation of both mEPSC amplitude and frequency), whereas neurons lacking Rab3A or carrying a dominant negative mutated Rab3A (earlybird) do not. They also demonstrate that only neuronal, but not glial Rab3A is responsible for this defect. Furthermore, they confirm the increased mEPSC amplitudes in WT neurons reflect the addition of GluA2-containing AMPA receptors rather than calcium-permeable ones, as previously reported by multiple labs. However, the increase in mEPSC amplitude is not accompanied by a corresponding upregulation of GluA2 synaptic clusters according to their IHC data (cluster size and intensity trend slightly upwards but not reaching significance). They further show that this modest upward trend is absent in Rab3A KO neurons, and conclude that Rab3A is involved in postsynaptic GluA2 upregulation during homeostatic synaptic plasticity.

When compared to the original version, the authors have done an admirable job in switching to more established ways to assess homeostatic synaptic plasticity by comparing the mean mEPSC amplitude and frequency, which has greatly improved the legibility of the manuscript to the public. Their data in Figures 1,2, and 8 clearly demonstrate that functional Rab3A in cortical neurons is required for the homeostatic regulation of mEPSCs.

However, the authors still have not provided further investigation of the mechanisms behind the role of Rab3A in this form of plasticity, and the revision therefore has added little to the significance of the study. Moreover, the experimental design for the investigation of the mismatch between mEPSC amplitude and GluA2 cluster fluorescence remains questionable, making it difficult to draw any credible conclusions from groups of data that not only look similar to the eye but also show no significance statistically.

A major claim the authors want to make is that Rab3A, although a presynaptic protein, regulates postsynaptic GluA2, and they do this by showing in Figure 5 that the upward trend of GluA2 cluster size and intensity is absent in Rab3A KO neurons. First, it is difficult to convince readers that this upward trend is real in Figures 5B-D without getting more samples. Second, the authors pick GluA2 clusters on the primary dendrites, whereas mEPSC events come from a much larger synapse population (e.g., more distal), therefore it makes sense that these two forms of measurement do not show matching changes, and this caveat could be addressed by sampling more diverse dendritic locations. Without a convincing phenotype in WT neurons, the support for this claim is weak.

Another claim of the authors is that this mismatch between mEPSC amplitude and GluA2 cluster sizes with the same culture suggests there are other factors contributing to the mEPSC amplitude. They do this by comparing results from individual culture dissociations, which greatly suffer from undersampling (Figure 6). In particular, they point out that 2 out of 3 dissociations show "matching" upward trends in mEPSC and GluA2 cluster (figure 6A and 6B) while the third one shows opposite trends, and use this to support their claim. Anyone who has done culture preparation would know the high variability between dissociations, which is why culture data are always pooled for assessment of any population trend. Anything could have happened to this particular dissociation (culture #3, figure 6C), and drawing conclusion from this single incident does little to support this claim. At least, they should double the dissociation numbers, and their claim would be much more convincing if a similar phenomenon occurs again. Besides, as mentioned above, all these mismatching trends could just be due to sampling differences.

In summary, this study establishes that neuronal Rab3A plays a role in homeostatic synaptic plasticity, but so do a number of other molecules that have been implicated in homeostatic synaptic plasticity in the past two decades (only will grow with the new techniques such as RNAseq). Without going beyond this finding and demonstrating how exactly Rab3A participates in the induction and/or expression of this form of plasticity, or maybe the potential Rab3A-mediated functional and behavioral defects in vivo, the contribution of the current study to the field is limited. However, given the presynaptic location of Rab3A, this finding could serve as a starting point for researchers interested in pre-postsynaptic cross-talk during homeostatic plasticity in general.

Reviewer #3 (Public review):

This manuscript presents a number of interesting findings that have the potential to increase our understanding of the mechanism underlying homeostatic synaptic plasticity (HSP). The data broadly support that Rab3A plays a role in HSP, although the site and mechanism of action remain uncertain.

The authors clearly demonstrate the Rab3A plays a role in HSP at excitatory synapses, with substantially less plasticity occurring in the Rab3A KO neurons. There is also no apparent HSP in the Earlybird Rab3A mutation, although baseline synaptic strength seems already elevated. In this context, it is unclear if the plasticity is absent or just occluded by a ceiling effect due the synapses already being strengthened. Occlusion may also occur in the mixed cultures, with Rab3A missing from neurons but not astrocytes. The authors do appropriately discuss both options. There are also differences in genetic background between the Rab3A KO and Earlybird mutants that could also impact the results, which are also noted. The authors have solid data showing that Rab3A is unlikely to be active in astrocytes, Finally, they attempt to study the linkage between synaptic strength during HSP and AMPA receptor trafficking and conclude that trafficking may not be solely responsible for the changes in synaptic strength.

Strengths:

This work adds another player into the mechanisms underlying an important form of synaptic plasticity. The plasticity is likely only reduced, suggesting Rab3A is only partially required and perhaps multiple mechanisms contribute. The authors speculate about some possible novel mechanisms.

However, the conclusions on the partial dissociation of AMPAR trafficking and synaptic response are made from somewhat weaker data. On average, across 3 culture sets, they saw similar magnitude of change in mEPSC amplitude and GluA2 cluster area and integral, but the GluA2 data was not significant. This is likely due to the nature of the datasets. Their imaging method involves only assessing puncta pairs (GluA2/VGlut1) clearly associated with a MAP2 labeled dendrite. This is a small subset of synapses, with usually less than 20 synapses per neuron analyzed (as stated by the authors). The mEPSC recordings will be averaging across several hundred events, which likely represent a hundred or more synapses given reasonable expectations on release probability. It has been reported, in work from this lab as well as by direct monitoring of tagged AMPARs during HSP (Wang, et al., 2019), that individual synapses are quite variable in their response. So there will almost necessarily be higher variability in the imaging data due to the smaller number of synapses sampled. The overall trends, though, are in alignment with previous data implicating receptor trafficking as the mechanism for HSP. However, the authors go on to evaluate each of the individual cultures, where 2 show similar changes between the mEPSC data and GluA2 clusters, and 1 culture showing little/no change in GluA2 clusters. The n's are very low here, and none of the datasets are significant. They want to conclude for this culture, there was a change in mEPSC amplitude that was not accompanied by a change in GluA2 at synaptic sites. But these data are collected from different coverslips, and due to the low n's, the potential under-sampling of the GluA2 clusters, and neuron-to-neuron variability, it is very hard to distinguish if this apparent difference is a methodological issue rather than a biological one. Much stronger data would be necessary to conclude that additional factors beyond receptor trafficking are required for HSP.

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a frequency effect that is unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. However the change in frequency seems to argue (as the authors do) that some synapses only have CP-AMPARs, while the rest of the synapses have few or none. Another possibility is that there are pre-synaptic NASPM-sensitive receptors that influence release probability. Further, the amplitude data show a strong trend towards smaller amplitude following NASPM treatment (Fig 3B). The p value for both control and TTX neurons was 0.08 - it is very difficult to argue that there is no effect. The decrease on average is larger in the TTX neurons, and some cells show a strong effect. It is possible there is some heterogeneity between neurons on whether GluA1/A2 heteromers or GluA1 homomers are added during HSP. This would impact the weakly supported conclusions about the GluA2 imaging vs mEPSC amplitude data.

Unaddressed issues that would greatly increase the impact of the paper:

(1) Is Rab3A acting pre-synaptically, post-synaptically or both? The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where it is acting (pre or post) would aid substantially in understanding its role. They could use sparse knock-down of Rab3A, or simply mix cultures from KO and WT mice (with appropriate tags/labels). The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. The more support for their suggestion of a pre-synaptic site of control, the better.

(2) Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs (ie the opposite of whatever is happening at excitatory synapses). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling; an effect only at excitatory synapses would argue for a more specific role just at these synapses.

Author response:

The following is the authors’ response to the previous reviews.

Since multiple Reviewers requested that the results describing effects of TTX treatment on GluA2 receptor levels detected by immunofluorescence and confocal imaging be revised, we have made substantial changes, which are described below. We believe the changes have greatly improved the manuscript and thank the reviewers for their comments.

Lack of significant increase in GluA2 receptor data is due to too few cultures sampled; anything could have happened [in one] particular dissociation. A concern that the TTX effect might vary greatly from culture to culture was why we felt it was important to match the receptor measurements on the same cultures that we recorded mEPSCs. We now present the culture means in Figure 5A (mEPSCs) and 5B (GluA2 receptor cluster size). These plots make it clear that the variability in the GluA2 receptor cluster size effect is not attributable to a failure of that culture to show a homeostatic effect. That is, the variability in GluA2 receptor effect is independent of the variability in mEPSC effect. To increase sample size, we examined 2 additional cultures for synaptic GluA2 receptor levels in control vs. TTX treatment. These cultures showed very modest increases (Figure 5C). When cell means from these experiments were pooled with those from the 3 matched cultures, the TTX effect was still not statistically significant (Figure 5G).

Lack of significant increase in GluA2 receptor data is due to the choice to restrict our analysis to the primary dendrite, close to the cell body. We restricted our analysis to the primary dendrite because Figure 3 in Turrigiano et al, 1998, shows the increased response to exogenously applied glutamate after TTX treatment is greatest close to the cell body and wanes as the glutamate is applied further away (added to Results, new lines 388-389).

Variability in GluA2 receptor data is due to the much smaller number of synapses sampled, compared to mEPSCs. We matched the sampling for mEPSC amplitude data to that of imaging data by taking only 20 samples from each electrophysiological recording. Each mEPSC represents one synapse; in a set of 20 mEPSCs some might come from the same synapse, so that we are sampling from £ 20 synapses. The effect of TTX on mEPSC amplitudes remained significant despite the reduced samples per cell (Figure 5A).

Why do we fail to show a significant increase in receptors when this has been shown in many studies?

We have added to our discussion the point that several studies, including Wang et al. 2019, use the number of puncta, rather than the number of cells, as the sample number. We ran an analysis of GluA2 receptor cluster size where we sampled multiple synapses per cell, and used the number of clusters as the sample n. We found that even with as few as 6 synapses randomly selected from each cell, the effect of TTX on GluA2 receptor cluster size became highly significant (p = 0.001 for data from 3 cultures and p = 0.005 for data from 5 cultures) (see new lines 400-406 in Discussion). In sum, our data are not very different from that of some previous studies. We are not arguing that receptors do not increase. Instead our point is that the increase is more variable than the increase in MESPC amplitude and thus takes a much bigger sample size to detect. In sum, the difference between the mEPSC data and the receptor data is that the mEPSC data consistently show a ~20-25% increase, whereas the receptor data do not always show an increase and sometimes the increase is only ~10%. Finally, we added two matched culture experiments examining synaptic GluA1 receptor cluster characteristics. GluA1 receptor cluster size decreased in one culture, and increased very modestly in the other (Supplemental Figure 1B), whereas mEPSC amplitude robustly increased (Supplemental Figure 1A; Results, new lines 265-268).

We conclude that these data support the idea that there is another contributor to the TTXinduced increase in quantal size.

Other changes in presentation of GluA2 receptor results: Since the effects on intensity and integral are of lesser magnitude than that on cluster size, we have removed these results from the graphs, although they are presented in Table 1. We have removed Figure 6, the presentation of individual culture results, since these results are now conveyed in Figure 5A-C. We have removed graphs depicting GluA2 receptor cluster size in response to TTX in Rab3A-/- cultures, but these data are still presented in Table 1.

We address other detailed comments below.

Public Reviews:

Reviewer #1 (Public review):

(2) The effects of Rab3A on TTX-induced mini frequency modulation remains unclear, because TTX does not induce a change in mini frequency in the Rab3A+/Ebd control (Fig. 2). The respective conclusions should be revised accordingly (l. 427).

The effects on mini frequency were added for completeness, but given the lack of consistently significant changes with TTX treatment or changes in the KO or Rab3AEbd/Ebd cultures, we have removed comment on these results from the Discussion.

(3) The model is still not supported by the data. In particular, data supporting a negative regulation of Rab3A by APs, Rab3A-dependent release of a tropic factor, or a Rab3Adependent increase in GluA2 abundance are not presented.

We have removed the model from the manuscript.

(4) Data points are not overlapping and appear "quantal" in most box plots. How were the data rounded?

The appearance of quantal variation in cell amplitude means is due to the binning that is part of the creation of the box plot. We have not remade the figures without binning, because the binning provides a visual depiction of the distribution of the data points. We have added the bin sizes to the appropriate figure legends.

Reviewer #2 (Public review):

However, the authors still have not provided further investigation of the mechanisms behind the role of Rab3A in this form of plasticity, and the revision therefore has added little to the significance of the study. Moreover, the experimental design for the investigation of the mismatch between mEPSC amplitude and GluA2 cluster fluorescence remains questionable, making it difficult to draw any credible conclusions from groups of data that not only look similar to the eye but also show no significance statistically.

To our knowledge, no other study has matched measurements of mEPSC amplitude in the same cultures where synaptic receptor levels were assessed. As stated above, we have revised the presentation of GluA2 receptor results, concluding from the lack of significant effects on receptor levels that the mEPSC amplitude increase cannot be fully explained by the receptor data (which is strengthened by addition of two more cultures analyzed for GluA2 immunofluorescence). This is an important addition to the significance of the study.

In summary, this study establishes that neuronal Rab3A plays a role in homeostatic synaptic plasticity, but so do a number of other molecules that have been implicated in homeostatic synaptic plasticity in the past two decades (only will grow with the new techniques such as RNAseq). Without going beyond this finding and demonstrating how exactly Rab3A participates in the induction and/or expression of this form of plasticity, or maybe the potential Rab3A-mediated functional and behavioral defects in vivo, the contribution of the current study to the field is limited. However, given the presynaptic location of Rab3A, this finding could serve as a starting point for researchers interested in pre-postsynaptic cross-talk during homeostatic plasticity in general.

We previously published a review in which we list 19 molecules known at that time to be important for homeostatic synaptic plasticity (see Table 2, Koesters et al., 2024), and they fall into two categories: molecules involved in glutamate receptor expression or trafficking, and signaling molecules. Rab3A is the first synaptic vesicle protein to be implicated in homeostatic plasticity of quantal size. We have added this point to the Discussion, new lines 473-476. By demonstrating that Rab3A is not acting in glia (which release TNF, which regulates receptor expression), and that GluA2 receptor levels do not explain the homeostatic mEPSC increase in our experimental conditions, we have ruled out two major mechanisms.

Reviewer #3 (Public review):

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a frequency effect that is unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. However the change in frequency seems to argue (as the authors do) that some synapses only have CP-AMPARs, while the rest of the synapses have few or none. Another possibility is that there are pre-synaptic NASPM-sensitive receptors that influence release probability. Further, the amplitude data show a strong trend towards smaller amplitude following NASPM treatment (Fig 3B). The p value for both control and TTX neurons was 0.08 - it is very difficult to argue that there is no effect. The decrease on average is larger in the TTX neurons, and some cells show a strong effect. It is possible there is some heterogeneity between neurons on whether GluA1/A2 heteromers or GluA1 homomers are added during HSP. This would impact the weakly supported conclusions about the GluA2 imaging vs mEPSC amplitude data.

We cannot rule out that the NAPSM-induced decrease in mEPSC frequency is due to a loss of presynaptic glutamate receptor enhancement of release probability, and have added this statement to the Results, new lines 202-204. Regarding the p value of 0.08—we are not arguing that NASPM has no effect on mEPSC amplitude, only that it has no effect on the homeostatic increase in amplitude after TTX treatment. An increase in GluA1/A2 heteromers should have been detected in our imaging studies.

Unaddressed issues that would greatly increase the impact of the paper:

(1) Is Rab3A acting pre-synaptically, post-synaptically or both? The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where it is acting (pre or post) would aid substantially in understanding its role. They could use sparse knockdown of Rab3A, or simply mix cultures from KO and WT mice (with appropriate tags/labels). The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. The more support for their suggestion of a pre-synaptic site of control, the better.

We agree that doing co-cultures of Rab3A-/- and Rab3A+/+ neurons is the definitive experiment to determine the locus of action of Rab3A in homeostatic synaptic plasticity. We hope to examine this question in a future manuscript.

(2) Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs (ie the opposite of whatever is happening at excitatory synapses). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling; an effect only at excitatory synapses would argue for a more specific role just at these synapses.

We agree that it would be very interesting to determine if the homeostatic decrease in mIPSCs after activity blockade depends on Rab3A. We hope to address this question in the future.

Recommendations for the authors:

Reviewer #3 (Recommendations for the authors):

Minor points:

The abstract is a bit repetitive in places. Some editing would be advised.

We did not identify anything repetitive in the abstract except the parallel construction referring to the previous findings at the NMJ and current findings in cortical neurons. However, we have eliminated a section in the introduction which went into detail about the receptor imaging results (previous lines 103-110).

Line 77: 'shift toward early awakening' is unclear; do you mean shorter sleep/wake cycle? Other circadian issues? A more complete description is needed.

We have moved the additional detail about the Earlybird mutation’s effect on circadian period from the Results to the Introduction, new lines 77 to 79.

The results section has many passages that seem more like discussion, offering various interpretation and alternatives for the data. While some commentary is appropriate, to justify the next series of experiments and maintain a logical flow, this manuscript has rather a high amount of this. Some editing and shifting material to the discussion might be warranted.

We have reduced the commentary in the Results section.

Line 245: GluA2 homomers are really unlikely, as they won't pass current (unless unedited) and don't often if ever form. But GluA2/A3 heteromers are likely (and detected by their methods).

GluA2 homomers do conduct current, albeit less than heteromers (Swanson et al., 1997; Oh and Derkach, 2005; Coombs et al., 2019). [The Oh and Derkach paper shows a GluA2 homomer current in Supplementary Figure 3]. We have modified the text to acknowledge that the GluA2 receptor imaging will detect heteromers and homomers (Results, new lines 214 to 215).

Line 258: If the number of synaptic pairs analyzed was usually <20, what was the average and range of pairs? This gets into the sampling issue.

We have added the average number of synaptic sites (20.4 ± 6.5) and range (11-38) to the text, Results, new line 229.

Are the stats of the baseline mEPSC amplitude and frequency shifts (WT vs KO on WT feeder layer) given somewhere (lines 398-402)? If not, please add them.

These stats have been added to the text, mEPSC amplitude, (CON, WT on WT, 13.3 ± 0.5 pA; CON, KO on WT, 15.2 ± 1.1 pA, p = 0.23, Kruskal-Wallis test), new lines 325-326 and frequency, (CON, WT on WT, 2.54 ± 0.57 sec-1; CON, KO on WT, 4.46 ± 1.21 sec-1, p = 0.23, Kruskal-Wallis test), new lines, 329-330.

25mM K+ is going to be much more than 'mildly' depolarizing (line 697). Should just skip that word.

‘mildly’ has been removed.

The section on MiniAnalysis seems overly argumentative, and there is no need to discuss flaws in the Wu paper. The important thing (a bit buried at the end of this section) is that the manual mini selection was done blind to condition, which is the normal way of dealing with potential bias. It would be better to limit the methods to describing what was done.

The bulk of the justification of manual analysis has been removed from the text.

The discussion of potential conductance changes (lines 534-6) seems somewhat unwarranted.

Modification of GluA1 phosphorylation in the GluA1/A2 heteromer would not be detected by NASPM (and the NASPM data being a bit inconclusive anyway). Further, auxiliary subunits (like TARPs) can alter conductance of any of the AMPARs. So I don't think they have enough data to exclude such a possibility.

The discussion of contributions of conductance have been removed from the text.

Coombs ID, Soto D, McGee TP, Gold MG, Farrant M, Cull-Candy SG (2019) Homomeric GluA2(R) AMPA receptors can conduct when desensitized. Nat Commun 10:4312.

Oh MC, Derkach VA (2005) Dominant role of the GluR2 subunit in regulation of AMPA receptors by CaMKII. Nat Neurosci 8:853-854.

Swanson GT, Kamboj SK, Cull-Candy SG (1997) Single-channel properties of recombinant AMPA receptors depend on RNA editing, splice variation, and subunit composition. J Neurosci 17:5869.

Version 3

Abstract

Abstract

Following prolonged activity blockade, amplitudes of miniature excitatory postsynaptic currents (mEPSCs) increase, a form of plasticity termed “homeostatic synaptic plasticity.” We previously showed that a presynaptic protein, the small GTPase Rab3A, is required for full expression of the increase in miniature endplate current amplitudes following prolonged blockade of action potential activity at the mouse neuromuscular junction in vivo (Wang et al., 2011), but it is unknown whether this form of Rab3A-dependent homeostatic plasticity shares any characteristics with central synapses. We show here that homeostatic synaptic plasticity of mEPSCs is impaired in mouse cortical neuron cultures prepared from Rab3A-/- and mutant mice expressing a single point mutation of Rab3A, Rab3A Earlybird mice. To determine if Rab3A is involved in the well-established homeostatic increase in postsynaptic AMPA-type receptors (AMPARs), we performed a series of experiments in which electrophysiological recordings of mEPSCs and confocal imaging of synaptic AMPAR immunofluorescence were assessed within the same cultures. We found that the increase in postsynaptic AMPAR levels was more variable than that of mEPSC amplitudes, suggesting other factors may contribute to the homeostatic increase in synaptic strength. Finally, we demonstrate that Rab3A is acting in neurons because only selective loss of Rab3A in neurons, not glia, disrupted the homeostatic increase in mEPSC amplitudes. This is the first demonstration that a protein thought to function presynaptically is required for homeostatic synaptic plasticity of quantal size.

Introduction

One of the most studied phenomena triggered by prolonged activity blockade is the increase in amplitudes of miniature excitatory postsynaptic currents (mEPSCs) in neurons. First demonstrated in cultures of dissociated cortical neurons (Turrigiano et al., 1998) and spinal cord neurons (O’Brien et al., 1998), the compensatory response to prolonged loss of activity has been dubbed “homeostatic synaptic plasticity” (Turrigiano and Nelson, 2004; Pozo and Goda, 2010). These first two studies provided evidence that glutamate receptors were also increased after prolonged inactivity, via an increased response to exogenously applied glutamate (Turrigiano et al., 1998) and increased immunofluorescent labeling of GluA1 receptors at synaptic sites (O’Brien et al., 1998). The accompanying increase in synaptic AMPA receptors has been confirmed with immunofluorescence in multiple studies of hippocampal and cortical cultures treated with TTX, including (Wierenga et al., 2005; Hou et al., 2008; Ibata et al., 2008; Jakawich et al., 2010; Tatavarty et al., 2013; Xu and Pozzo-Miller, 2017; Dubes et al., 2022). It is now well accepted that the homeostatic increase in mEPSC amplitudes in neurons is due to an increase in postsynaptic AMPA receptors.

In our previous work, we found that a TTX cuff applied for 48 hrs around the sciatic nerve in mice led to an increase in the amplitude of the miniature endplate currents (mEPCs) recorded in tibialis anterior muscles. Surprisingly, and unlike at central synapses, we could find no evidence that the acetylcholine receptors levels at the neuromuscular junction (NMJ) were increased in the TTX-cuff treated mice (Wang et al., 2005). This result led us to search for presynaptic molecules that might homeostatically regulate the presynaptic quantum. In previous studies in chromaffin cells, we identified the small GTPase Rab3A, a presynaptic vesicle protein, as a regulator of synaptic vesicle fusion pore opening (Wang et al., 2008), so we examined whether deletion of Rab3A (Rab3A-/-) might prevent homeostatic upregulation of mEPC amplitude at the NMJ. The results were clear: the homeostatic increase in mEPC amplitude induced by the TTX cuff was strongly reduced in the Rab3A-/- mouse, and was completely abolished in the Earlybird mutant (Rab3AEbd/Ebd), which has a single point mutation in Rab3A that causes a shift towards early awakening due to a shorter circadian period, that is more dramatic than that of the Rab3A deletion mouse (Kapfhamer et al., 2002; Wang et al., 2011). These results led us to conclude that the homeostatic increase in mEPC amplitude at the NMJ is a presynaptic phenomenon.

It has remained somewhat of a mystery what the role of Rab3A is in synaptic transmission. The Rab3A-/- mouse has minimal phenotypic abnormalities, with evoked synaptic transmission and mEPSCs essentially normal in hippocampal slices (Geppert et al., 1994). At the Rab3A-/- NMJ, reductions in evoked transmission were detected, but only under conditions of reduced extracellular calcium (Coleman et al., 2007). There are some modest changes in short term plasticity during repetitive stimulus trains, with increased depression observed in response to moderate frequencies in hippocampal slices (Geppert et al., 1994) and at the mammalian NMJ (Coleman and Bykhovskaia, 2009), but increased facilitation in response to high frequency stimulation at the mammalian NMJ (Coleman and Bykhovskaia, 2010). The most dramatic effect of loss of Rab3A is the abolishment of a presynaptic form of long-term potentiation (LTP) at the mossy fiber-CA3 synapse that is induced by 25 Hz stimulation in the presence of NMDA blockers (Weisskopf et al., 1994; Castillo et al., 1997). Our result that prolonged inactivity-induced plasticity of mEPC amplitude at the mammalian NMJ is disrupted with loss of function of Rab3A further cements a role for Rab3A in activity-dependent plasticity of synaptic strength. However, a crucial question that remains is whether the TTX-cuff induced effect on the amplitude of the spontaneous synaptic current at the NMJ shares underlying mechanisms with central synapses. We set out to answer this question for the more well-studied phenomenon of homeostatic plasticity of mEPSCs in dissociated mouse cortical neurons prepared from wild type, Rab3A-/- mice and Rab3aEbd/Ebd mice treated with TTX for 48 hrs. We show here that homeostatic synaptic plasticity of mEPSC amplitude was strongly reduced in cultures from Rab3A-/- mice, and completely abolished in cultures from Rab3AEbd/Ebd mice, supporting a shared mechanism with the neuromuscular junction.

Results

We previously reported that mixed cultures of cortical neurons and glia prepared from postnatal day 0-2 mouse pups responded to a block of action potential-mediated activity by a 48 hr TTX treatment with an increase in mEPSC amplitude (Hanes et al., 2020). Here, we asked the question whether cortical cultures prepared from mice lacking the small GTPase Rab3A, or, expressing a point mutation of Rab3A, Rab3A Earlybird, have an altered homeostatic plasticity response to a prolonged loss of network activity. The Rab3A Earlybird mutation was discovered in a mutagenesis screen in mice for genes involved in the rest-activity cycle (Kapfhamer et al., 2002). In that study, the authors concluded that the Earlybird mutation was likely dominant-negative, as the shift in circadian period was substantially more dramatic than that observed in the Rab3A-/- mice. As described in the Introduction, the effects of loss of Rab3A have been modest, so we included experiments using the Rab3A Earlybird mutant mice to see if there would be a more robust phenotype, and in doing so, strengthen our conclusions based on the Rab3A deletion mice. To obtain Rab3A-/- and Rab3AEbd/Ebdhomozygotes, we established two mouse colonies of heterozygous breeders with cultures prepared from pups derived from a final breeding pair of homozygotes. Although we backcrossed Rab3A+/- with Rab3A+/Ebd for 11 generations, clear differences in mEPSC amplitudes in untreated cultures (see below) and in calcium current amplitudes in adrenal chromaffin cells (unpublished obs.) remained. Therefore, throughout this study we keep the two strains separate and there are two Rab3A+/+ or ‘wild type’ phenotypes.

We found that loss of Rab3A resulted in a greatly impaired homeostatic increase after activity blockade. Example current traces of spontaneously occurring mEPSCs recorded from pyramidal neurons in untreated (CON) 13-14 DIV cortical cultures and sister cultures treated with 500 nM TTX for 48 hr prepared from wild type and Rab3A-/- mice in the Rab3A+/- colony are shown in Figures 1A and B, respectively. Average mEPSC waveforms from the same recordings are shown in Figures 1C and D. The mean mEPSC amplitudes for 30 control and 23 TTX-treated neurons from Rab3A+/+ cultures are displayed in the box and whisker plot in Figure 1E; after activity blockade the average mEPSC amplitude increased from 13.9 ± 0.7 pA to 18.2 ± 0.9 pA. In contrast, in cultures prepared from Rab3A-/- mice, the average mEPSC amplitude was not significantly increased, for 25 untreated cells and 26 TTX-treated cells (Figure 1F, 13.6 ± 0.1 pA vs. 14.3 ± 0.6 pA). We found that TTX treatment also resulted in an increase in mEPSC frequency in cultures prepared from Rab3A+/+ mice, as shown in Figure 1G (CON, 2.26 ± 0.37 sec-1; TTX, 4.62 ± 0.74 sec-1). This TTX-induced frequency increase was strongly reduced in cultures from Rab3A-/- mice, but there were a few outliers with very high frequencies still present after TTX treatment (Figure 1H, CON, 2.74 ± 0.49 Hz; TTX, 3.23 ± 0.93 Hz).

Loss of Rab3A disrupted the TTX-induced increase in amplitudes of mEPSCs recorded in cultured mouse cortical neurons; the increase in frequency was more variable but also appears to be reduced.

(A) Ten second example traces recorded at −60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/- colony. (B) Ten second example traces recorded at −60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A-/- mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A+/+ mice (Rab3A+/- colony; CON, N = 30 cells, 13.9 ± 0.7 pA; TTX, N = 23 cells, 18.2 ± 0.9 pA; from 11 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A-/- mice (CON, N = 25 cells, 13.6 ± 0.7 pA; TTX, N = 26 cells, 14.3 ± 0.6 pA); from 11 cultures. (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 2.26 ± 0.37 sec-1; TTX, 4.62 ± 0.74 sec- 1). (H) Box plots for average mEPSC frequency for same Rab3A-/-cells as in (F) (CON, 2.74 ± 0.49 sec-1; TTX, 3.23 ± 0.93 sec-1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells, bin size 1 pA, 0.5 sec-1; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. For all p-values, * with underline indicates significance with p < 0.05.

We next examined homeostatic plasticity in cultures prepared from wild type mice in the Rab3A+/Ebd colony. Example current traces of mEPSCs, and average mEPSC waveforms, are shown in Figures 2A and C, respectively, for cortical cultures prepared from wild type mice in the Rab3A+/Ebd colony. Treatment with TTX for 48 hr led to a significant increase in the average mEPSC amplitude of 23 TTX-treated cells when compared to 20 untreated cells (Figure 2E, CON, 11.0 ± 0.6 pA; TTX, 15.0 ± 1.3 pA). We note here that while the two wild type strains respond very similarly to activity blockade, the mean mEPSC amplitude in untreated cultures was significantly different in the two wild type strains, 13.9 ± 0.7 pA, wild type from Rab3A+/- colony; 11.0 ± 0.6 pA, wild type from Rab3A+/Ebd colony (p = 0.004, Kruskal-Wallis test).

Normally functioning Rab3A was required for TTX-induced homeostatic plasticity of mEPSC amplitudes in cultured mouse cortical neurons.

(A) Ten second example traces recorded at −60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/Ebd colony. (B) Ten second example traces recorded at −60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3AEbd/Ebd mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3A+/+ mice (Rab3A+/Ebd colony; CON, N = 20 cells, 11.0 ± 0.6 pA; TTX, N = 23 cells, 15.1 ± 1.2 pA; from 6 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3AEbd/Ebd mice (CON, N = 21 cells, 15.1 ± 1.0 pA; TTX, N = 22 cells, 14.6 ± 1.1 pA; from 7 cultures.) (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 1.15 ± 0.19 sec-1; TTX, 2.54 ± 0.55 sec-1). (H) Box plots for average mEPSC frequency for same Rab3AEbd/Ebd cells as in (F) (CON, 1.71 ± 0.41 sec-1; TTX, 3.05 ± 0.80 sec-1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; bin size 1 pA, 0.5 sec-1; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. For all p-values, * with underline indicates significance with p < 0.05.

We found a complete disruption of homeostatic plasticity in cortical cultures prepared from Rab3AEbd/Ebd mice, as can been seen in viewing example mEPSC traces and average mEPSC waveforms (Figures 2B and D, respectively). The lack of TTX effect was confirmed in a comparison of mEPSC amplitude means for 21 untreated and 22 TTX-treated cells (Figure 2F, CON, 15.1 ± 1.0 pA vs. TTX, 14.6 ± 1.1 pA). There was a trend towards increased mEPSC frequency in TTX-treated cultures from this wild type strain, but the difference was not as robust as for the wild type cultures from Rab3A+/- colony, and did not reach statistical significance (Figure 2G, CON, 1.15 ± 0.19 sec-1; TTX, 2.54 ± 0.55 sec-1). This trend was not observed in neurons from cultures prepared from Rab3AEbd/Ebd mice, likely due to an increase in frequency in untreated cells—frequencies remained high in TTX-treated cells (Figure 2H, CON, 1.71 ± 0.41 sec-1; TTX, 3.05 ± 0.80 sec-1). Our results show that the homeostatic increase in mEPSC amplitude after activity blockade is disrupted in both Rab3A-/- and the Rab3AEbd/Ebdcortical neurons, strongly supporting a crucial role of functioning Rab3A in this process.

The disruption in homeostatic plasticity differs for Rab3A-/- and Rab3AEbd/Ebd. In the Rab3A-/- data set, mEPSCs from untreated cultures were indistinguishable from mEPSCs from Rab3A+/+ untreated cultures, demonstrating loss of Rab3A has no impact on basal activity mEPSC amplitudes. In the Rab3AEbd/Ebd data set, mEPSC amplitudes from untreated cultures were significantly larger than those of untreated cultures from wild type mice, as can be seen comparing the CON values in Figures 2E and F (CON, Rab3A+/+, 11.0 ± 0.7 pA, vs. CON, Rab3AEbd/Ebd, 15.1 ± 1.0 pA, p = 0.0027, Kruskal-Wallis test). The increase in mEPSC amplitude in cultures from Rab3AEbd/Ebd mice is consistent with the increase in mEPC amplitude we observed at the Rab3AEbd/EbdNMJ (Wang et al., 2011).

The small GTPase Rab3A is generally thought to function presynaptically to regulate synaptic vesicle trafficking, possibly in an activity-dependent manner (Castillo et al., 1997; Lonart et al., 1998; Leenders et al., 2001; Schluter et al., 2006; Coleman and Bykhovskaia, 2009; Tian et al., 2012). In contrast, homeostatic plasticity of mEPSC amplitude has been attributed to an increase in postsynaptic receptors on the surface of the dendrite (O’Brien et al., 1998; Turrigiano et al., 1998). At the NMJ in vivo, we could find no evidence for an increase in AChRs after TTX block of the sciatic nerve in vivo (Wang et al., 2005). Is Rab3A required for the homeostatic increase in surface AMPA-type glutamate receptors that has been confirmed by multiple studies (see Introduction)?

We explored whether Rab3A might be acting to increase postsynaptic AMPA receptors. In some studies of homeostatic synaptic plasticity it has been found that application of a Ca2+- permeable AMPA receptor-specific inhibitor reversed the homeostatic increase in mEPSC amplitude (Ju et al., 2004; Thiagarajan et al., 2005; Sutton et al., 2006). Figure 3A shows that the TTX-induced increase in mean mEPSC amplitude was nearly identical in a set of 11 CON and 11 TTX-treated cells before and after treatment with 1-Naphthyl acetyl spermine trihydrochloride (NASPM), a synthetic analogue of the Ca2+-permeable AMPA receptor inhibitor philanthotoxin (before NASPM, CON 12.9 ± 3.5 pA; TTX, 17.5 ± 3.1 pA; after NASPM, CON 11.9 ± 2.6 pA; TTX 16.1 ± 3.5 pA). This result indicates that the TTX-induced increase in mEPSC amplitude does not depend on Ca2+-permeable receptors, since the effect of TTX clearly remained when their presence was removed by acute NASPM application. We do not think that there was any technical issue with the NASPM application, because overall, mEPSC amplitudes were reduced modestly in both untreated and TTX-treated cultures (Figure 3B, CON, before NASPM, 12.9 ± 3.5 pA; after NASPM, 11.9 ± 2.6 pA; TTX, before NASPM, 17.5 ± 3.1 pA; after NASPM, 16.1 ± 3.5 pA). Furthermore, NASPM consistently reduced mEPSC frequency (Figure 3C, CON, before NASPM, 1.84 ± 0.55 sec-1; after NASPM, 1.56 ± 0.53 sec-1; TTX, before NASPM, 4.40 ± 3.51 sec-1; after NASPM, 2.68 ± 2.25 sec-1). Similar to the data in Figure 1, frequency is significantly increased after TTX treatment (p = 0.04), but this difference is no longer significant after NASPM application (p = 0.20). One explanation for the effect on frequency is that there are synaptic sites that express only homomers of Ca2+-permeable AMPA receptors (GluA2-lacking AMPARs); mEPSCs from these sites would be expected to be completely blocked by NASPM (see cartoon description in Figure 3D, Left). We cannot rule out an alternative explanation, that presynaptic Ca2+-permeable AMPA receptors normally enhance release probability, and NAPSM prevents this action. The magnitude of the frequency reduction following acute NASPM appears to be greater after TTX treatment, suggesting that loss of activity could promote the establishment of synaptic sites that contain only Ca2+-permeable receptors. However, these new sites do not appear to contribute to the increase in mEPSC amplitude (Figure 3A). The lack of effect of NASPM rules out the contribution of Ca2+-permeable AMPA receptors GluA1, GluA3 and GluA4 (Hollmann et al., 1991; Burnashev et al., 1992; Jonas and Burnashev, 1995), as well as Ca2+-permeable forms of Kainate receptors GluA5 and GluA6 (Koike et al., 1997; Sun et al., 2009) to homeostatic synaptic plasticity.

Homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures treated with TTX for 48 hr was unchanged by acute inhibition of Ca2+ permeable AMPA receptors by NASPM (20 μm).

(A) Box plot comparison of the TTX effect on mEPSC amplitudes in the same pyramidal neurons before and after application of 20 μM NASPM. (N = 11 cells from 3 cultures; Pre-NASPM, CON: 12.9 ± 1.1 pA; TTX: 17.5 ± 0.9 pA; post-NASPM, CON: 11.9 ± 0.8 pA; TTX: 16.1 ± 1.0 pA). (B) Line series plot of mEPSC amplitudes before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NASPM: 12.9 ± 1.1 pA; post-NASPM: 11.9 ± 0.8 pA; TTX, pre-NASPM: 17.5 ± 0.9 pA; post-NASPM: 16.1 ± 0.1 pA. (C) Line series plot of mEPSC frequency before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NAPSM: 1.84 ± 0.55 sec-1; post-NASPM; 1.56 ± 0.53 sec-1; TTX, pre-NASPM: 4.40 ± 1.06 sec-1; post-NASPM, 2.68 ± 0.68 sec-1. (D) A proposed mechanism for why NASPM had a robust effect on frequency without greatly affecting amplitude. A dendrite with spines (Top) is expanded on three postsynaptic sites (Middle) to show possible types of AMPA receptor distributions: Left, at a site comprised only of Ca2+-permeable AMPA receptors (NASPM-sensitive, GluA2-lacking receptors (black)), NASPM would completely inhibit the mEPSC, causing a decrease in frequency to be measured in the overall population; Right, at a site comprised only of Ca2+-impermeable AMPA receptors (NASPM-insensitive, GluA2-containing receptors (green)), NASPM would have no effect on mEPSC amplitude; Middle, at a site comprised of a mix of Ca2+-permeable and Ca2+- impermeable AMPA receptors, NASPM would partially inhibit the mEPSC. Diagram produced in BioRender.com (2024). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; bin size 1 pA; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. Line series plot p values are from student’s paired t test. For all p-values, * with underline indicates significance with p < 0.05.

Having established that homomeric Ca2+-permeable receptors are not contributing to the homeostatic increase in mEPSC amplitude, we turned to immunofluorescence and confocal imaging to assess whether GluA2 receptor expression, which will identify GluA2 homomers and GluA2 heteromers, was increased in our wild type mouse cortical cultures following 48 hr treatment with TTX. Since mEPSCs necessarily report synaptic levels of receptors, we used VGLUT1-immunofluorescence to identify synapses on pyramidal primary apical dendrites labeled with MAP-2 immunofluorescence. We focused on the primary dendrite of pyramidal neurons as a way to reduce variability that might arise from being at widely ranging distances from the cell body, or, from inadvertently sampling dendritic regions arising from inhibitory neurons. In addition, it has been shown that the increase in receptors is most dramatic in this region (Turrigiano et al., 1998). Figure 4 (Top) shows two pyramidal neurons, one from an untreated culture on the left, and one from a TTX-treated culture on the right, prepared from Rab3A+/+ mice. To measure characteristics of synaptically located GluA2 receptor clusters, we zoomed in on the primary dendrite (region in white rectangle). The zoomed regions for single confocal sections of the selected area are shown below and contain pairs of VGLUT1- and GluA2-immunofluorescent clusters indicated by white trapezoidal frames. A dendrite typically required ∼10 confocal sections to be fully captured, and the total number of synaptic pairs for all the sections imaged on a dendrite averaged 20 (20.4 ± 6.5 (SD); range, 11-38) so this is an atypically high number of pairs within a single section; these particular dendrites and sections were selected for illustration purposes. In addition to the synaptic pairs, we observed many GluA2-immunofluorescent clusters not associated with VGLUT1 immunofluorescence (a few are indicated with white arrows), most probably extrasynaptic receptors. There were also GluA2- positive clusters present outside of MAP2-positive dendrites, which may be located on astrocytes (Fan et al., 1999). Finally, GluA2 immunofluorescent clusters close to VGLUT1 immunofluorescence but not located along any apparent MAP2-positive neurites suggests the presence of axon-axonal contacts, although VGLUT1 has also been detected in astrocytes (Ormel et al., 2012). Only sites that contained both VGLUT1 and GluA2 immunofluorescence close to the primary MAP2-postive dendrite or a secondary branch along this primary dendrite were selected for analyses. In the experiment from which these images were collected, we analyzed the distance from the cell body for each synapse. The average distance from the cell body, for dendrites from the untreated cell, was 38.5 ± 2.8 μm; for the TTX-treated cell, it was 42.4 ± 3.2 μm (p = 0.35, Kruskal-Wallis test). Since the greatest distance from the cell body is set by the outer limit of the zoom window, and that window was placed adjacent to the cell body, distances in other experiments would be in this same range.

Identification of synaptic GluA2 receptor immunofluorescence on primary dendrites of pyramidal neurons in high density mouse cortical co-cultures prepared from Rab3A+/+ mice.

(Top) Non-zoomed single confocal sections collected with a 60X oil immersion objective of recognizably pyramidal-shaped neurons, presumed to be excitatory neurons, selected for synaptic GluA2 analysis. A neuron was selected from an untreated coverslip (CON, Left) and a TTX-treated coverslip (TTX, Right) from the same culture prepared from Rab3A+/+ animals (= Rab3A+/+ Culture #2 in Figure 5). The white rectangular boxes indicate 5X zoomed areas shown in the images below. Note that the depth of the single confocal section of the non-zoomed neuron image is not at the same depth as the confocal section in zoomed dendritic image, so some features are not visible in both images. Scale bar, 20 μm. (Bottom) 5X zoom single confocal sections selected for demonstration purposes because they had an unusually high number of identified synaptic pairs along the primary dendrite contained within a single confocal section. Synaptic pairs, highlighted with white trapezoids, were identified based on close proximity of GluA2 (red) and VGLUT1 (white) immunofluorescence, apposed to the MAP2 immunofluorescent primary dendrite (green). Some apparent synaptic pairs are not highlighted with a white trapezoid because there was a different confocal section in which the two immunofluorescent sites were maximally bright. There were GluA2 positive clusters located on the primary dendrites that were not apposed to VGLUT1-positive terminals (4 of these non-synaptic GluA2 clusters are highlighted with white arrows in the MAP2-GluA2 and MAP2-GluA2- VGLUT1 panels). Scale bar, 10 μm.

Variability in the magnitude of the homeostatic response from culture to culture is averaged out in physiological experiments by the pooling of data from many cells across many cultures. To reduce the necessity for many cultures, we chose to pair experiments in the same cultures by recording mEPSCs from one set of coverslips, and processing another set of coverslips from the same culture for immunofluorescence. We also wanted to determine if the homeostatic response of mEPSC amplitudes varied from culture to culture, and if so, whether the GluA2 receptor levels varied in parallel. We completed this matched paradigm of physiology and immunofluorescence for 3 cultures prepared from Rab3A+/+ mice. We first present the results as means for each culture, for control and TTX-treated dishes, for mEPSC data (Figure 5A) and imaging data (Figure 5B). Levels of GluA2 immunofluorescence at synaptic sites were quantified by the size of the GluA2-immunofluorescent receptor cluster, because this showed the greatest response to activity blockade; the average intensity value, and the integral of intensity values across the area of the cluster, are included in a summary of the data in Table 1. Three of three cortical cultures showed an increase in mean mEPSC amplitude after TTX treatment. In contrast, one of the same three cultures showed a decrease in mean GluA2 receptor cluster size. Therefore, the decrease in receptor levels in Culture #3 was due to something other than a lack of homeostatic effect on mEPSCs in that culture. Because there was a clear trend towards increased GluA2 receptor cluster size, but the receptor response appears more variable, we performed immunofluorescence measurements on two additional cortical cultures, both of which showed modest increases after TTX treatment (Figure 5C). We confirmed that there was not an increase in GluA1 receptor levels in two matched cultures that showed robust homeostatic increases in mEPSC amplitudes (Supplementary Figure 1A, culture mean mEPSC amplitudes; Supplementary Figure 1B, culture mean GluA1 receptor cluster sizes).

Comparison of mEPSC amplitudes and GluA2 receptor cluster areas in matched and unmatched mouse cortical cultures prepared from Rab3A+/+ mice and treated with TTX for 48 hr.

(A) Culture averages of mEPSC amplitudes for untreated (CON) and TTX-treated coverslips (TTX) in each of 3 Rab3A+/+ mouse cortical co-cultures. Culture #1, CON, N = 6, 14.2 ± 2.2 pA; TTX, N = 6, 15.9 ± 1.9 pA; Culture #2, CON, N = 7, 13.8 ± 2.4 pA; TTX, N = 8, 17.7 ± 1.8 pA; Culture #3, CON, N = 10, 13.4 ± 0.8 pA; TTX, N = 9, 15.7 ± 1.0 pA. (B) Culture averages of GluA2 receptor cluster size in the same 3 cultures as shown in (A). Culture #1, CON, N = 10, 0.73 ± 0.09 μm2; TTX, N = 9, 0.89 ± 0.09 μm2; Culture #2, CON, N = 9, 0.91 ± 0.12 μm2; TTX, N = 9, 1.42 ± 0.24 μm2; Culture #3, CON, N = 10, 1.05 ± 0.11 μm2; TTX, N = 9, 0.95 ± 0.15 μm2. (C) Two additional culture averages are included that did not have corresponding mEPSC recordings. Culture #4, CON, N = 10, 0.58 ± 0.07 μm2: TTX, N = 10, 0.71 ± 0.04 μm2; Culture 5, CON, N =10, 0.89 ± 0.05 μm2; TTX, N = 10, 0.95 ± 0.08 μm2. (D) Box plots of cell mean mEPSC amplitudes pooled from same 3 cultures as shown in (A). CON, N = 23 cells, 13.7 ± 0.9 pA; TTX, N = 24 cells, 16.4 ± 0.9 pA. (E) Box plots of dendrite mean GluA2 receptor cluster size pooled from same 3 cultures as shown in (A) and (B). CON, N = 29 dendrites, 0.90 ± 0.06 μm2; TTX, N = 28 dendrites, 1.08 ± 0.10 μm2. (F) Box plots of dendrite mean GluA2 receptor cluster size pooled from the same 5 cultures shown in (C). CON, N = 49 dendrites, 0.83 ± 0.04 μm2; TTX, N = 48 dendrites, 0.98 ± 0.07 μm2. Line series, Student’s paired T test; box plots, Kruskal-Wallis test. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10 and 90 percentiles; open circles indicate means of individual cells/dendrites; bin size 1 pA, 0.5 μm2; line, median; dot, mean. For all p-values, * with underline indicates significance with p < 0.05.

We next pooled data from the multiple cultures to examine effects of activity blockade on GluA2 receptors at the level of cell means. Because the average number of synaptic sites identified in the primary dendrites in three cultures was 20.4, we also took 20 mEPSC samples from each cell, the 21st to the 40th mEPSC recorded, to minimize any differences in variability between GluA2 receptors and mEPSCs that might be due to a difference in sampling. In the data pooled from the 3 matched experiments, neurons from Rab3A+/+ cultures showed a significant increase in mean mEPSC amplitudes following activity blockade (Figure 5D; Table 1). This result indicates that the homeostatic response averaged across the 3 cultures was very similar to, but slightly smaller than, that of the data set presented in Figure 1. The mean for size of GluA2 immunofluorescent receptor clusters showed a trend towards higher values after activity blockade (18.1%) that was of similar magnitude to the mEPSC amplitude increase (19.7%), but did not reach statistical significance for 3 cultures (Figure 5E) or 5 cultures (Figure 5F; Table 1). As expected, mean mEPSC amplitude was not increased following activity blockade in the data pooled from a new set of 3 Rab3A-/- cultures (Table 1). For immunofluorescence results from the same Rab3A-/- cultures, no trend towards higher values was apparent for GluA2 receptor cluster characteristics (Table 1). However, given the variability of the GluA2 response in cultures from Rab3A+/+ mice, it would require a much higher number of Rab3A+/+ and Rab3A-/- cultures to make any firm conclusions about the role of Rab3A in the homeostatic response of GluA2 receptors.

Having used VGLUT1 immunofluorescence to mark synaptic sites, we could also ask whether there was a presynaptic effect of activity blockade on the size, intensity or integral of the VGLUT1 signal at sites apposed to the previously examined GluA2 postsynaptic sites. In the pooled data from cultures prepared from Rab3A+/+ mice, we found no change in the size of the VGLUT1-positive regions, and trends towards reduced intensity and integral (Table 2). These data do not support the idea that activity blockade caused an increase in the amount of VGLUT1 per vesicle, although we cannot rule out that a simultaneous reduction in the number of vesicles obscured the effect, since we did not have an independent label for synaptic vesicles. In cultures prepared from Rab3A-/- mice, the size, intensity and integral of the VGLUT1- positive regions did not significantly change (Table 2).

We have demonstrated here for the first time that the synaptic vesicle protein Rab3A influences the homeostatic regulation of quantal amplitude. What are possible ways that Rab3A could exert its effect? It has previously been shown that exogenous addition of TNFα to hippocampal cultures causes an increase in surface expression of GluA1 receptors (although not GluA2 receptors), and that the homeostatic increase in mEPSC amplitudes is abolished in cultures prepared from the TNFα deletion mouse (Stellwagen et al., 2005; Stellwagen and Malenka, 2006). Furthermore, neurons from TNFα+/+ mice plated on glial feeder layers derived from TNFα-/- mice fail to show the increase in mEPSC amplitude after TTX treatment, indicating that the TNFα inducing the receptor increases following activity blockade comes from the glial cells (Stellwagen and Malenka, 2006). It has recently been shown that the source of TNFα is astrocytes rather than microglia (Heir et al., 2024). Rab3A has been detected in astrocytes (Maienschein et al., 1999; Hong et al., 2016), so to determine whether Rab3A is acting via regulating TNFα release from astrocytes, we performed experiments similar to those of Stellwagen and Malenka (2006) (schema illustrated in Figure 6, Left). Under our culturing conditions, the glial feeder layers would be expected to be composed mainly of astrocytes (Heir et al., 2024). We compared the effect of activity blockade on mEPSC amplitudes recorded from cortical neurons from Rab3A+/+ mice plated on Rab3A+/+ glial feeder layers (Figure 6A); neurons from Rab3A+/+ mice plated on Rab3A-/- glia (Figure 6B), and neurons from Rab3A-/- mice plated on Rab3A+/+ glia (Figure 6C). If Rab3A is required for TNFα release from glia, then Rab3A+/+ neurons plated on Rab3A-/- glia should not show a homeostatic increase in mEPSC amplitude after treatment with TTX, and any cultures with Rab3A+/+ glia should have a normal homeostatic response. We found the opposite result: mEPSC amplitudes increased dramatically in cultures where Rab3A was present in neurons (WT on WT, CON, 13.3 ± 0.5 pA, TTX, 16.7 ± 1.2 pA; WT on KO, CON, 13.3 ± 1.0 pA, TTX, 18.8 ± 1.4 pA), but the increase was greatly diminished in cultures where Rab3A was present only in glia (KO on WT, CON, 15.2 ± 1.1 pA, TTX, 16.9 ± 0.7 pA). In the glial feeder layer culture, the loss of Rab3A in the neurons may have caused an increase in mEPSC amplitude at baseline, similar to the Earlybird mutant in dissociated cultures, but this difference was not significant (CON, WT on WT, 13.3 ± 0.5 pA; CON, KO on WT, 15.2 ± 1.1 pA, p = 0.23, Kruskal-Wallis test), so it is also possible this can be attributed to culture to culture variability. Regarding mEPSC frequency, in cultures with Rab3A+/+ neurons, frequency trended upward after activity blockade, whereas in the culture with Rab3A-/- neurons, frequency was already increased (although not significantly, CON, WT on WT, 2.54 ± 0.57 sec-1; CON, KO on WT, 4.46 ± 1.21 sec-1, p = 0.23, Kruskal-Wallis test), and trended downward after activity blockade.

Rab3A in neurons, not astrocytes, was required for full TTX-induced homeostatic plasticity.

(A)-(C), mEPSC amplitude (Middle) and frequency (Right) data from dissociated cortical neurons plated on an astrocyte feeder layer, each prepared separately from the type of mice depicted in the schema (Left): (A) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A+/+ mice. Box plots for mEPSC amplitudes (CON, N = 17 cells, 13.3 ± 0.5 pA; TTX, N = 20 cells, 16.7 ± 1.2 pA) and mEPSC frequency in the same recordings (mean, CON, 2.54 ± 0.57 sec-1; TTX, 3.48 ± 0.64 sec-1); from 4 cultures. (B) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A-/- mice. Box plots for average mEPSC amplitude (CON, N = 11 cells, 13.3 ± 1.0 pA; TTX, N = 11 cells, 18.8 ± 1.4 pA) and average mEPSC frequency in the same recordings (CON, 2.01 ± 0.41 sec-1; TTX, 4.47 ± 1.53 sec-1); from 2 cultures. (C) Neurons from Rab3A-/- neurons plated on astrocytes from Rab3A+/+ mice. Box plots for average mEPSC amplitude (CON, N = 14 cells, 15.2 ± 1.1 pA; TTX, N = 11 cells, 16.9 ± 0.7 pA) and mEPSC frequency in the same recordings (CON, 4.47 ± 1.21 sec-1; TTX, 3.02 ± 0.70 sec-1); from 3 cultures. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. p- values denoted with * and underline indicates significance of p < 0.05.

In summary, in the absence of Rab3A from glia, the homeostatic plasticity of mEPSC amplitude following activity blockade was normal, whereas in the absence of Rab3A in neurons, homeostatic plasticity was greatly diminished. This result makes it highly unlikely that Rab3A is required for the release of TNFα, or another factor from glia, that induces a homeostatic upregulation of postsynaptic receptors and thereby increases mEPSC amplitude following TTX treatment. Neuronal Rab3A appears to mediate the homeostatic increase in mEPSC amplitude following activity blockade.

Discussion

We found that homeostatic synaptic plasticity of mEPSC amplitude in dissociated mixed cultures of mouse cortical neurons and glia behaved remarkably similar to the mouse NMJ in response to loss of Rab3A function: in cultures from Rab3A-/- mice, the increase in mEPSC amplitudes following prolonged network silencing by TTX was strongly diminished, and in cultures from Rab3AEbd/Ebd mice, basal mEPSC amplitude was increased compared to that of wild-type cultures and was not further modified by TTX treatment. These results suggest that normal function of the synaptic vesicle protein Rab3A is required for the homeostatic increase of mEPSC amplitude in cortical cultures. At the NMJ, the effect of Rab3A on homeostatic upregulation of mEPC amplitude was not mediated by upregulation of postsynaptic receptors. In neurons, the increase in mEPSC amplitude following block of activity was not due to increases in Ca2+-permeable GluA1 homomers because acute NASPM application was unable to reverse the TTX-induced increase in amplitude. Imaging of immunofluorescence of GluA2 receptors revealed the response was more variable than that of mEPSC amplitude. The simplest explanation for our findings at both the NMJ and in neurons is that Rab3A regulates a presynaptic contributor to alteration of quantal amplitude during homeostatic synaptic plasticity.

Is the lack of homeostatic plasticity in the Earlybird mutant due to occlusion?

We found that mEPSC amplitude was increased in untreated cultures prepared from Rab3AEbd/Ebd mice, compared to mEPSC amplitude in cultures from wild type mice. If there is a limit to how large an mEPSC can become, it is possible that the mutant Rab3A does not affect homeostatic plasticity, only occludes it. Similar increases in mEPSC amplitude at baseline, combined with a failure to increase further after activity blockade, were noted in cultures prepared from mECP2, AKAP, Homer1a, and Arc/Arg3.1 deletion mice (Shepherd et al., 2006; Hu et al., 2010; Xu and Pozzo-Miller, 2017; Sanderson et al., 2018) as well as in cultures treated with ubiquitin proteasome inhibitors (Jakawich et al., 2010) or microRNA 186-5p (Silva et al., 2019), suggesting the possibility of a generalized phenomenon in which increased mEPSC amplitude is induced when homeostatic regulation is disrupted. Since we did not observe such an increase in mEPSC amplitude at baseline in cultures from Rab3A-/- mice, it remains a possibility that the point mutant may bind to novel partners, causing activities that would not be either facilitated or inhibited by Rab3A. Still, it is a strong coincidence that the novel activity of the mutant Earlybird Rab3A would affect mEPSC amplitude, the same characteristic that is modulated by activity blockade in a Rab3A dependent manner. Although we cannot rule out occlusion, an alternate interpretation is that the presence of the mutant protein mimics the condition of activity blockade.

Homeostatic plasticity of postsynaptic receptors was less consistent than that of mEPSC amplitude

While we found that the effect of TTX treatment on mEPSC amplitude and GluA2 receptor levels were of similar magnitude, the effect on GluA2 receptor levels was not statistically significant for the pooled data sets acquired in the same cultures, nor in a larger data set containing 5 cultures. In light of the extensive literature documenting increases in GluA1 or GluA2 receptors, or both, after activity blockade, the less than convincing increase in GluA2 receptor levels in the current work came as a surprise. The reason for the lack of significance was that the increase in receptors was more variable. We are aware of three possible explanations for greater variability. 1) The number of synaptic sites sampled was smaller for the imaging data set. This was because we limited our analysis to synaptic receptors on the primary dendrite, where the increase in sensitivity to exogenously applied glutamate was shown to be greatest (see Figure 3 in (Turrigiano et al., 1998)). To address this issue, we limited our sample size of mEPSCs in the matched experiments to the same number per cell as the mean sample size per cell for imaging, but there was still a discrepancy in the effect of TTX on mEPSCs vs. receptors. 2) The small sample of only 3 cultures was not enough to reveal the increase in receptors. We attempted to address this by adding 2 more cultures. Although both additional cultures showed increases in GluA2 receptor cluster size, the magnitudes of the increases were not large, and when cells from these cultures were included in the pooled data set, the p value was not substantially improved. The only remaining possibility is: 3) The complexity involved in processing cultures for immunofluorescence introduces variability that obscures the true effect. One can get around technical issues related to immunofluorescence processing by performing Western blots, but then any changes detected could be due to upregulation of extrasynaptic receptors. What remains unclear is why our data were impacted by this but data of other studies were not. Some other studies have used the number of synaptic sites or puncta as the sample size (for example, (Hou et al., 2008; Ibata et al., 2008; Wang et al., 2019)). When we analyzed our data after including as few as 6 randomly selected cluster size measurements per cell, a Kruskal-Wallis test gave highly significant p values for 3 cultures (p = 0.001, n = 174 sites (6 *29 cells)) and 5 cultures (p = 0.005, n = 294 sites (6 *49 cells)). We found only a single study that had comparable sample sizes for receptors and mEPSC amplitudes, and the p value was < 0.01 for mEPSC amplitudes (n = 9 and 10 cells for CON and TTX, respectively), but five times higher (< 0.05) for receptor data (n = 13 and 14 cells for CON and TTX) (Xu and Pozzo-Miller, 2017). This single example is consistent with our suggestion that measurements of changes in receptors after activity blockade are more variable than measurements of mEPSC amplitudes. In sum, our data suggest that TTX treatment of mouse cortical cultures consistently leads to an increase in mEPSC amplitudes, and sometimes leads to an increase in synaptic GluA2 receptor cluster size. Increased receptors likely contribute to increases in mEPSC amplitudes, but our current data, coupled with our previous findings at the mouse neuromuscular junction, support the idea that there are additional sources contributing to the homeostatic increase in quantal size.

Neuronal, not glial, Rab3A is required for homeostatic regulation of quantal amplitude

Astrocytic release of TNFα was shown to mediate the increase in mEPSC amplitudes in prolonged TTX-treated cultures of dissociated hippocampal neurons and cultures of hippocampal slices (Stellwagen and Malenka, 2006; Heir et al., 2024), but we found that loss of Rab3A in glia did not disrupt the increase in mEPSC amplitude after activity blockade. Further, we previously showed that the homeostatic increase in NMJ mEPC amplitude was completely normal in the absence of TNFα (Wang et al., 2011). This leaves neuronal Rab3A as the potential controller of homeostatic regulation of quantal amplitude.

Is Rab3A acting in the presynaptic cell to cause homeostatic increase in mEPSC amplitude?

Our current results support the surprising possibility that the presynaptic vesicle protein Rab3A either regulates postsynaptic AMPARs or regulates a presynaptic contributor to homeostatic synaptic plasticity. To our knowledge, there is no evidence localizing Rab3A to postsynaptic dendritic sites. Other presynaptic molecules, such as SNARE proteins, synaptotagmins, and NSF, have been identified in the postsynaptic compartment (Lledo et al., 1998; Nishimune et al., 1998; Osten et al., 1998; Song et al., 1998; Araki et al., 2010; Kennedy et al., 2010; Suh et al., 2010; Jurado et al., 2013; Hussain and Davanger, 2015; Gu et al., 2016; Wu et al., 2017). A recent report implicates Rab3A in the localization of plasma membrane proteins, including the EGF receptor, in rafts in HEK cells and Jurkat-T cells, which indicates Rab3A can be found in other specialized locations besides the presynaptic nerve terminal (Diaz-Rohrer et al., 2023). Without additional experiments targeting Rab3A deletion to the pre- or post-synaptic compartment, we cannot rule out that Rab3A is regulating homeostatic plasticity in mouse cortical cultures by acting postsynaptically on receptor trafficking. However, given our previous finding of a lack of any increase in AChR after TTX treatment the NMJ (Wang et al., 2011), the simplest interpretation is that in these two situations, Rab3A is required for an increase in the presynaptic quantum rather than an increase in postsynaptic receptors.

The presynaptic quantum, or the amount of transmitter released during a single fusion event, is an important contributor to quantal size. The amount of transmitter released during vesicle fusion can be affected by the kinetics of the fusion pore opening (Chang et al., 2017). We previously used amperometric measurements in mouse adrenal chromaffin cells to show loss of Rab3A increased the occurrence of very small amplitude fusion pore feet (Wang et al., 2008). Paired with our finding that there is an increase in the occurrence of very slow-rising, abnormally shaped mEPCs at the neuromuscular junction of Rab3A-/- mice, our data suggest that small synaptic vesicles may have a fusion pore step under certain circumstances (see also (Chiang et al., 2018)). If activity blockade causes an increase in mEPSC amplitude due to a more rapid opening of a fusion pore, or a larger fusion pore conductance, it is possible that Rab3A is required for this modulation.

Another way to increase the presynaptic quantum is to increase levels of the transmitter transporter (Daniels et al., 2004; De Gois et al., 2005; Wilson et al., 2005; Hartman et al., 2006). However, we found that immunofluorescence for the glutamate transporter, VGLUT1, was not increased after activity blockade under our experimental conditions. A similar lack of increase in VGLUT1 levels, relative to synapsin, was previously observed in hippocampal cultures treated with NBQX, an AMPA receptor blocker (see supplemental data, (Wilson et al., 2005)). Taken together with our findings, it appears that there is not strong evidence that the amount of VGLUT1 at synapses is a contributing factor to the increase in mEPSC amplitude in neuronal cultures after activity blockade with TTX.

Finally, the amount of transmitter released by a vesicle could be increased due to larger vesicle size. Larger vesicle diameter has been observed following activity disruption in multiple paradigms, including ocular deprivation, cortical ablation, and immobilization (Kawana et al., 1971; Lloret and Saavedra, 1975; Cheresharov et al., 1978). Loss of Rab3 family members is associated with increased vesicle size (Riedel et al., 2002; Schonn et al., 2010). Finally, an increase in volume has been shown to increase miniature endplate junctional current amplitude at the Drosophila NMJ (Karunanithi et al., 2002). Rab3A could be required for an activity-dependent increase in vesicle size, but the increase in diameter needed for a 25% increase in transmitter, and therefore mEPSC amplitude, would be difficult to detect.

The number of molecules shown to be required for homeostatic synaptic plasticity of quantal size in neuronal cultures may have surpassed ∼20 (see (Koesters et al., 2024) for a recent summary), but fall into two main categories: molecules involved in glutamate receptor expression or trafficking, and signaling molecules. To our knowledge, this is the first evidence that a presynaptic protein is participating in this process. Homeostatic synaptic plasticity is becoming increasingly implicated in essential normal functions such as sleep (Tononi and Cirelli, 2014; Diering et al., 2017; Torrado Pacheco et al., 2021), and pathological conditions such as epilepsy, neuropsychiatric disorders, Huntington’s, and alcohol use disorder (Trasande and Ramirez, 2007; Fernandes and Carvalho, 2016; Wang et al., 2017; Lovinger and Abrahao, 2018; Smith-Dijak et al., 2019; Lignani et al., 2020; Suzuki et al., 2021; Kavalali and Monteggia, 2023). Therefore, our identification of a novel Rab3A-dependent regulatory pathway in homeostatic synaptic plasticity may have important therapeutic implications. Our findings also further cement the importance of Rab3A in activity-dependent modulation of synaptic strength.

Materials and methods

Animals

To determine the role of Rab3A in homeostatic plasticity in mouse cortical cultures, we employed two distinct genetic mouse strains with altered Rab3A function. Rab3A deletion mice were generated as follows: Rab3A+/- heterozygous mice were bred and genotyped as previously described (Kapfhamer et al., 2002; Wang et al., 2008), and maintained as a colony using multiple heterozygous breeding pairs. Homozygous Rab3A+/+ or Rab3A-/- mouse pups were obtained by breeding Rab3A+/+ pairs and Rab3A-/- pairs, respectively, so that the pups were the first-generation progeny of a homozygous mating. This protocol minimizes the tendency of multiple generations of Rab3A+/+ and Rab3A-/- homozygous breedings to produce progeny that are more and more genetically distinct. Rab3AEbd/Ebdmice were identified in an EU-mutagenesis screen of C57BL/6J mice, and after a cross to C3H/HeJ, were backcrossed for 3 generations to C57BL/6J (Kapfhamer et al., 2002). Rab3A+/Ebd heterozygous mice were bred at Wright State University and genotyped in a two-step procedure: 1. a PCR reaction (forward primer: TGA CTC CTT CAC TCC AGC CT; reverse primer: TGC ACT GCA TTA AAT GAC TCC T) followed by 2. a restriction digest with enzyme Bsp1286I (New England Biolabs) that distinguishes the Earlybird mutant by its different base-pair products. Rab3A+/- mice were backcrossed with Rab3A+/+ mice from the Earlybird heterozygous colony for 11 generations in an attempt to establish a single wild type strain, but differences in mEPSC amplitude and adrenal chromaffin cell calcium currents persisted, likely due to genes that are close to the Rab3A site, resulting in two wild type strains: 1. Rab3A+/+ from the Rab3A+/- colony, and 2. Rab3A+/+ from the Rab3A+/Ebd colony.

Primary Culture of Mouse Cortical Neurons

Primary dissociated cultures of mixed neuronal and glia populations were prepared as previously described (Hanes et al., 2020). Briefly, postnatal day 0-2 (P0-P2) Rab3A+/+, Rab3A-/- or Rab3AEbd/Ebd neonates were euthanized by rapid decapitation, as approved by the Wright State University Institutional Animal Care and Use Committee, and brains were quickly removed. Each culture was prepared from the cortices harvested from two animals; neonates were not sexed. Cortices were collected in chilled Neurobasal-A media (Gibco) with osmolarity adjusted to 270 mOsm and supplemented with 40 U/ml DNAse I (ThermoFisher Scientific). The tissues were digested with papain (Worthington Biochemical) at 20 U/ml at 37°C for 20 minutes followed by trituration with a sterile, fire-polished Pasteur pipette, then filtered through a 100 μm cell strainer, and centrifuged at 1100 rpm for 2 minutes. After discarding the supernatant, the pellet was resuspended in room temperature Neurobasal-A media (270 mOsm), supplemented with 5% fetal bovine serum for glia growth, and 2% B-27 supplement to promote neuronal growth (Gibco), L-glutamine, and gentamicin (ThermoFisher Scientific). Neurons were counted and plated at 0.15 * 106 cells/coverslip onto 12 mm coverslips pre-coated with poly-L-lysine (BioCoat, Corning). This plating density produces a “high density” culture characterized by a complex mesh of neuronal processes criss-crossing the field of view, completely filling the space between cell bodies (see Figure 4). We found in preliminary experiments that an increase in mEPSC amplitude in mouse cortical cultures was inconsistent when cell density was low, likely due to lower levels of baseline activity, although it is a limitation of this study that we did not directly measure activity levels in cultures prepared from Rab3A+/+, Rab3A-/- or Rab3AEbd/Ebd mice, and therefore do not know if differences in results are due to differences in activity levels. While unlikely for Rab3A+/+ and Rab3A-/- studies, given the mEPSC amplitude in untreated cultures was relatively small, it remains possible that in Rab3AEbd/Ebd studies, the larger mEPSC amplitude in untreated cultures was due to a loss of activity in these cultures. The culture media for the first day (0 DIV) was the same as the above Neurobasal-A media supplemented with FBS, B-27, L-glutamine, and gentamicin, and was switched after 24 hours (1 DIV) to media consisting of Neurobasal-A (270 mOsm), 2% B-27, and L-glutamine without FBS to avoid its toxic effects on neuronal viability and health (Stellwagen and Malenka, 2006). Half of the media was changed twice weekly and experiments were performed at 13-14 DIV. Two days prior to experiments, tetrodotoxin (TTX) (500 nM; Tocris), a potent Na+ channel blocker, was added to some cultures to chronically silence all network activity and induce homeostatic synaptic plasticity mechanisms, while untreated sister cultures served as controls. Cultures prepared from mutant mice were compared with cultures from wild-type mice from their respective colonies. Note that the cultures comprising the Rab3A+/+ data here are a subset of the data previously published in Hanes et al., 2020. This smaller data set was restricted to the time period over which cultures were also prepared from Rab3A-/- mice.

Preparation of Glial Feeder Layers

Glial feeder layers were prepared from the cortices of P0-P2 Rab3A+/+ or Rab3A-/- mouse pups as described previously (Stellwagen and Malenka, 2006). Briefly, cortices were dissected and cells were dissociated as described above. Cell suspensions of mixed neuronal and glial populations were plated onto glass coverslips pre-coated with poly-L-lysine in Dulbecco’s Modified Eagle Media (ThermoFisher Scientific) supplemented with 5% FBS (to promote glial proliferation and to kill neurons), L-glutamine, and gentamicin, and maintained in an incubator at 37°C, 5% CO2; cultures were maintained in this manner for up to 1 month to generate purely glial cultures (all neurons typically died off by 7 DIV). Culture media was replaced after 24 hours, and subsequent media changes were made twice weekly, replacing half of the culture media with fresh media. Feeder layers were not used for neuronal seeding until all native neurons were gone and glia approached 100% confluency (visually inspected).

Plating of Neurons on Glial Feeder Layers

Cortical neurons were obtained as described above. The cell pellet obtained was resuspended in Neurobasal-A (osmolarity adjusted to 270 mOsm) containing B27 (2%, to promote neuronal growth), L-glutamine, and 5-fluorodeoxyuridine (FdU, a mitotic inhibitor; Sigma). Addition of FdU was used to prevent glial proliferation and contamination of the feeder layer with new glia, promoting only neuronal growth on the feeder layers (FdU-containing media was used for the maintenance of these cultures and all subsequent media changes). Glial culture media was removed from the feeder layer cultures, and the neuronal cell suspension was plated onto the glial feeder cultures. The culture strategy used to distinguish the relative roles of neuronal and glial Rab3A is outlined in Figure 8, Left. At 1 DIV, all of the culture media was removed and replaced with fresh Neurobasal-A media containing FdU described above, and half of the media was replaced twice per week for all subsequent media changes. Cultures were maintained in a 37°C, 5% CO2 incubator for 13-14 DIV.

Whole-Cell Voltage Clamp to Record mEPSCs

At 13-14 DIV, mEPSCs from TTX-treated and untreated sister cultures of Rab3A+/+ or Rab3A-/- neurons from the Rab3A+/- colony, or Rab3A+/+ or Rab3AEbd/Ebd neurons from the Rab3A+/Ebd colony, were recorded via whole-cell voltage clamp to assess the role of Rab3A in homeostatic synaptic plasticity. Recordings were taken from pyramidal neurons, which were identified visually by a prominent apical dendrite; images were taken of all cells recorded from. Cells were continuously perfused with a solution consisting of (in mM): NaCl (115), KCl (5), CaCl2 (2.5), MgCl2 (1.3), dextrose (23), sucrose (26), HEPES (4.2), pH = 7.2 (Stellwagen and Malenka, 2006). On the day of recording, the osmolarity of the media from the cultures was measured (normally 285 – 295 mOsm) and the perfusate osmolarity was adjusted to match the culture osmolarity to protect against osmotic shock to the neurons. To isolate glutamatergic mEPSCs, TTX (500 nM) and picrotoxin (50 μM), were included in the perfusion solution to block action potentials and GABAergic currents, respectively. The NMDA receptor antagonist, APV, was not included in the perfusion solution because in pilot studies, all mEPSCs were blocked by 10 µM CNQX and 50 μm picrotoxin, demonstrating no APV-sensitive mEPSCs were present (data not shown). Previous studies have described NMDA mEPSCs in neuronal cultures, but recordings were performed in 0 Mg2+ (Watt et al., 2000; Sutton et al., 2006). In our recordings, the presence of extracellular Mg2+ (1.3 mM) and TTX would mean that the majority of NMDA receptors are blocked by extracellular Mg2+ at resting Vm. Patch electrodes (3.5 – 5 MΩ) were filled with an internal solution containing (in mM): K-gluconate (130), NaCl (10), EGTA (1), CaCl2 (0.132), MgCl2 (2), HEPES (10), pH = 7.2. Osmolarity was adjusted to 10 mOsm less than the perfusion solution osmolarity, which we have found improves success rate of achieving low access resistance recordings. Neurons were clamped at a voltage of −60 mV using an Axopatch 200B patch-clamp (Axon Instruments), recorded from for 2 – 5 minutes, and data were collected with Clampex 10.0/10.2 (Axon Instruments).

NASPM application

The antagonist of Ca2+-permeable AMPA receptors, N-naphthyl acetylspermine (NASPM, 20 μM; Tocris), was applied during recordings in a subset of experiments. NASPM is a synthetic analog of Joro Spider Toxin (JSTX) (Koike et al., 1997), and chemicals in this family block Ca2+ permeable glutamate receptors (Blaschke et al., 1993; Herlitze et al., 1993; Iino et al., 1996). The presence of the edited GluA2 subunit disrupts both the Ca2+ permeability and the sensitivity to JSTX and related substances (Blaschke et al., 1993; Bochet et al., 1994; Jonas and Burnashev, 1995). Therefore, 20 μM NASPM is expected to completely block AMPA receptors containing GluA1, GluA3 or GluA4 subunits (Hollmann et al., 1991; Burnashev et al., 1992), or the Kainate receptors containing GluA5 or GluA6 (Koike et al., 1997; Sun et al., 2009), but be ineffective against any heteromer or homomer containing GluA2 subunits. Because the effect of the spider toxins and their analogues are use-dependent (Herlitze et al., 1993; Iino et al., 1996; Koike et al., 1997), NASPM was applied with a depolarizing high K+ solution (25 mM KCl, 95 mM NaCl) which we expected to trigger release of glutamate and opening of glutamate-activated receptors, allowing entry of NASPM into the pore. Baseline recordings were performed for 2 minutes in our standard perfusate, followed by perfusion of NASPM + High K+ solution for 45 seconds, and then perfusion of a NASPM only solution for 5 minutes, after which recording was recommenced for 5 minutes (because we found in pilot experiments that frequency was reduced following NASPM application).

Analysis of mEPSCs

Miniature excitatory postsynaptic currents (mEPSCs) were manually selected using Mini Analysis software (Synaptosoft), a standard program used for mEPSC event detection and analysis. Records were filtered at 2 kHz using a low-pass Butterworth filter prior to selection. The program detection threshold was set at 3 pA but the smallest mEPSC selected was 4.01 pA. A fully manual detection process means that the same criterion (“does this look like an mEPSC?”) is applied to all events, not just the false positives or the false negatives, which prevents the bias from being primarily at one end or the other of the range of mEPSC amplitudes. It is important to note that when performing the MiniAnalysis process, the researcher did not know whether a record was from an untreated cell or a TTX-treated cell.

Immunofluorescence, microscopy, and data analysis

Primary cultures of mouse cortical neurons were grown for 13-14 DIV. Antibodies to GluA2 (mouse ab against N-terminal, EMD Millipore) were added directly to live cultures at 1:40 dilution, and incubated at 37 °C in a CO2 incubator for 45 minutes. Cultures were rinsed 3 times with PBS/5% donkey serum before being fixed with 4% paraformaldehyde. After 3 rinses in PBS/5% donkey serum, cultures were incubated in CY3-labeled donkey-anti-mouse secondary antibodies for 1 hour at room temperature, rinsed in PBS/5% donkey serum, permeabilized with 0.2% saponin, and incubated in chick anti-MAP2 (1:2500, AbCAM) and rabbit anti-VGLUT1 (1:4000, Synaptic Systems) for 1 hour at room temperature in PBS/5% donkey serum. After rinsing with PBS/5% donkey serum, coverslips were incubated with 488-anti chick and CY5-anti rabbit secondary antibodies for 1 hour at room temperature, rinsed, blotted to remove excess liquid, and inverted on a drop of Vectashield (Vector Labs). Coverslips were sealed with nail polish and stored at 4 °C for < 1 week before imaging. All secondary antibodies were from Jackson Immunoresearch and were used at 1:225 dilution.

Coverslips were viewed on a Fluoview FV1000 laser scanning confocal microscope with a 60x oil immersion, 1.35 NA objective. Once a pyramidal neuron was identified and a single confocal section of the cell acquired (see Top, Figure 4), Fluoview 2.1 software was used to zoom in on the primary dendrite (5X) and a series of confocal sections were taken every 0.5 μm. Images were analyzed offline with ImagePro 6 (Cybernetics). The composite image was used to locate synaptic sites containing both VGLUT1 and GluA2 immunoreactivity in close apposition to each other and to the primary dendrite or a secondary branch identified with MAP2 immunoreactivity. An area of interest (AOI) was manually drawn around the GluA2 cluster in the confocal section in which it was the brightest. The AOIs for a dendrite were saved in a single file; the AOI number and the confocal section it was associated with were noted for later retrieval. For quantification, AOIs were loaded, an individual AOI was called up on the appropriate section, and the count/measurement tool used to apply a threshold. Multiple thresholds were used, 400-550, but we found that within a given experiment, the magnitude of the TTX effect was not altered by different thresholds, therefore, all data presented here use a threshold of 450. Pixels within the cluster that were above the threshold were automatically outlined, and size, average intensity, and integral of the outlined region saved to Origin2020 (OriginLab) file for calculating mean values. For analyzing VGLUT1 sites, previous AOIs were examined, and if not directly over the VGLUT1 site, moved to encompass it (this happened when GluA2 and VGLUT1 were side by side rather than overlapping). The threshold was 200 for all VGLUT1 data except Rab3A+/+ Culture #1, where threshold was 400. There was no threshold that worked for both that culture and the other cultures, with too much of the background regions being included for Culture #1 with a threshold of 200, and too little of the synaptic site area being included for the other Cultures with a threshold of 400.

Statistics

In our previous publication (Hanes et al., 2020), we described the characteristics of homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures using multiple mEPSC amplitude quantiles from each cell to create data sets with thousands of samples. We compared the cumulative distribution functions for mEPSC amplitudes in untreated vs. TTX- treated cultures, as well as sorting the values from smallest to largest, and plotted TTX values vs. control values in the rank order plot first described in (Turrigiano et al., 1998) to demonstrate synaptic scaling. Finally, we plotted the ratio of TTX values/control values vs. control values to show that the ratio, which represents the effect of TTX, increased with increasing control values, which we termed “divergent scaling.” Here, we wanted to avoid the inflation of sample size caused by pooling multiple mEPSCs per cell in order to statistically compare the effects of TTX in cultures of wild type mice, Rab3A-/- mice and Rab3AEbd/Ebd mutant mice. Therefore, we computed the mean mEPSC amplitude for each cell and compared means from untreated and TTX-treated cultures with the non-parametric Kruskal-Wallis test, with N = the number of cells. The overall means are presented ± SEM in the text, and p values are displayed above each plot. The same process was carried out for the characteristics of synaptic GluA2 receptor clusters and VGLUT1 sites in imaging experiments, with N = the number of dendrites (one dendrite was sampled for each cell). Although we quantified three characteristics of the GluA2 clusters and VGLUT1 sites, size, average intensity, and integral, we have not corrected p values for multiple comparisons. Means are presented as box plots + data, with the box at 25 and 75 percentiles, whiskers at 10 and 90 percentiles, a dot indicating the mean, and a line indicating the median. In Figure 3, mEPSC amplitudes were measured within the same cell before and after NASPM application, so a paired t test was used to compare the pre- and post-NASPM mEPSC amplitudes. In Figure 5A-C the mean mEPSC amplitude for untreated and TTX-treated coverslips in the same culture are compared with a paired t test. Statistics were computed and plots were created using Origin2020 (OriginLab).

eLife Assessment

This study presents valuable findings on the role of the small GTPase Rab3A in homeostatic plasticity. While the study demonstrates that Rab3A is required for homeostatic scaling, the evidence supporting the model put forward by the authors is incomplete. The work will be of interest to researchers in the field of synaptic transmission and plasticity.

Reviewer #1 (Public review):

Koesters and colleagues investigated the role of the small GTPase Rab3A in homeostatic scaling of miniature synaptic transmission in primary mouse cortical cultures using electrophysiology and immunohistochemistry. The major finding is that TTX incubation for 48 hours does not induce an increase in the amplitude of excitatory synaptic miniature events in neuronal cortical cultures derived from Rab3A KO and Rab3A Earlybird mutant mice. NASPM application had comparable effects on mEPSC amplitude in control and after TTX, implying that Ca2+-permeable glutamate receptors are unlikely modulated during synaptic scaling. Immunohistochemical analysis revealed no significant changes in GluA2 puncta size, intensity, and integral after TTX treatment in control and Rab3A KO cultures. Finally, they provide evidence that loss of Rab3A in neurons, but not astrocytes, blocks homeostatic scaling. Based on these data, the authors propose a model in which neuronal Rab3A is required for homeostatic scaling of synaptic transmission, potentially through GluA2-independent mechanisms.

The major finding - impaired homeostatic up-scaling after TTX treatment in Rab3A KO and Rab3 earlybird mutant neurons - is supported by data of high quality. However, the paper falls short of providing any evidence or direction regarding potential mechanisms. The data on GluA2 modulation after TTX incubation are likely statistically underpowered, and do not allow drawing solid conclusions, such as GluA2-independent mechanisms of up-scaling.

The study should be of interest to the field because it implicates a presynaptic molecule in homeostatic scaling, which is generally thought to involve postsynaptic neurotransmitter receptor modulation. However, it remains unclear how Rab3A participates in homeostatic plasticity.

Major (remaining) point:

(1) Direct quantitative comparison between electrophysiology and GluA2 imaging data is complicated by many factors, such as different signal-to-noise ratios. Hence, comparing the variability of the increase in mini amplitude vs. GluA2 fluorescence area is not valid. Thus, I recommend removing the sentence "We found that the increase in postsynaptic AMPAR levels was more variable than that of mEPSC amplitudes, suggesting other factors may contribute to the homeostatic increase in synaptic strength." from the abstract.

Similarly, the data do not directly support the conclusion of GluA2-independent mechanisms of homeostatic scaling. Statements like "We conclude that these data support the idea that there is another contributor to the TTX- induced increase in quantal size." should be thus revised or removed.

Reviewer #2 (Public review):

I thank the authors for their efforts in the revision. In general, I believe the main conclusion that Rab3A is required for TTX-induced homeostatic synaptic plasticity is well-supported by the data presented, and this is an important addition to the repertoire of molecular players involved in homeostatic compensations. I also acknowledge that the authors are more cautious in making conclusions based on the current evidence, and the structure and logic have been much improved.

The only major concern I have still falls on the interpretation of the mismatch between GluA2 cluster size and mEPSC amplitude. The authors argue that they are only trying to say that changes in the cluster size are more variable than those in the mEPSC amplitude, and they provide multiple explanations for this mismatch. It seems incongruous to state that the simplest explanation is a presynaptic factor when you have all these alternative factors that very likely have contributed to the results. Further, the authors speculate in the discussion that Rab3A does not regulate postsynaptic GluA2 but instead regulates a presynaptic contributor. Do the authors mean that, in their model, the mEPSC amplitude increases can be attributed to two factors- postsynaptic GluA2 regulation and a presynaptic contribution (which is regulated by Rab3A)? If so, and Rab3A does not affect GluA2 whatsoever, shouldn't we see GluA2 increase even in the absence of Rab3A? The data in Table 1 seems to indicate otherwise.

I also question the way the data are presented in Figure 5. The authors first compare 3 cultures and then 5 cultures altogether, if these experiments are all aimed to answer the same research question, then they should be pooled together. Interestingly, the additional two cultures both show increases in GluA2 clusters, which makes the decrease in culture #3 even more perplexing, for which the authors comment in line 261 that this is due to other factors. Shouldn't this be an indicator that something unusual has happened in this culture? Data in this figure is sufficient to support that GluA2 increases are variable across cultures, which hardly adds anything new to the paper or to the field. The authors further cite a study with comparable sample sizes, which shows a similar mismatch based on p values (Xu and Pozzo-Miller 2007), yet the effect sizes in this study actually match quite well (both ~160%). P values cannot be used to show whether two effects match, but effect sizes can. Therefore, the statement in lines 411-413 "... consistently leads to an increase in mEPSC amplitudes, and sometimes leads to an increase in synaptic GluA2 receptor cluster size" is not very convincing, and can hardly be used to support "the idea that there are additional sources contributing to the homeostatic increase in quantal size".

I would suggest simply showing mEPSC and immunostaining data from all cultures in this experiment as additional evidence for homeostatic synaptic plasticity in WT cultures, and leave out the argument for "mismatch". The presynaptic location of Rab3A is sufficient to speculate a presynaptic regulation of this form of homeostatic compensation.

Minor concerns:

(1) Line 214, I see the authors cite literature to argue that GluA2 can form homomers and can conduct currents. While GluA2 subunits edited at the Q/R site (they are in nature) can form homomers with very low efficiency in exogenous systems such as HEK293 cells (as done in the cited studies), it's unlikely for this to happen in neurons (they can hardly traffick to synapses if possible at all).

(2) Lines 221-222, the authors may have misinterpreted the results in Turrigiano 1998. This study does not show that the increase in receptors is most dramatic in the apical dendrite, in fact, this is the only region they have tested. The results in Figures 3b-c show that the effect size is independent of the distance from soma.

(3) Lines 309-310 (and other places mentioning TNFa), the addition of TNFa to this experiment seems out of place. The authors have not performed any experiment to validate the presence/absence of TNFa in their system (citing only 1 study from another lab is insufficient). Although it's convincing that glia Rab3A is not required for homeostatic plasticity here, the data does not suggest Rab3A's role (or the lack of) for TNFa in this process.

Reviewer #3 (Public review):

This manuscript presents a number of interesting findings that have the potential to increase our understanding of the mechanism underlying homeostatic synaptic plasticity (HSP). The data broadly support that Rab3A plays a role in HSP, although the site and mechanism of action remain uncertain.

The authors clearly demonstrate that Rab3A plays a role in HSP at excitatory synapses, with substantially less plasticity occurring in the Rab3A KO neurons. There is also no apparent HSP in the Earlybird Rab3A mutation, although baseline synaptic strength is already elevated. In this context, it is unclear if the plasticity is absent, already induced by this mutation, or just occluded by a ceiling effect due to the synapses already being strengthened. Occlusion may also occur in the mixed cultures when Rab3A is missing from neurons but not astrocytes. The authors do appropriately discuss these options. The authors have solid data showing that Rab3A is unlikely to be active in astrocytes, Finally, they attempt to study the linkage between changes in synaptic strength and AMPA receptor trafficking during HSP, and conclude that trafficking may not be solely responsible for the changes in synaptic strength during HSP.

Strengths:

This work adds another player into the mechanisms underlying an important form of synaptic plasticity. The plasticity is likely only reduced, suggesting Rab3A is only partially required and perhaps multiple mechanisms contribute. The authors speculate about some possible novel mechanisms, including whether Rab3A is active pre-synaptically to regulate quantal amplitude.

As Rab3A is primarily known as a pre-synaptic molecule, this possibility is intriguing. However, it is based on the partial dissociation of AMPAR trafficking and synaptic response and lacks strong support. On average, they saw a similar magnitude of change in mEPSC amplitude and GluA2 cluster area and integral, but the GluA2 data was not significant due to higher variability. It is difficult to determine if this is due to biology or methodology - the imaging method involves assessing puncta pairs (GluA2/VGlut1) clearly associated with a MAP2 labeled dendrite. This is a small subset of synapses, with usually less than 20 synapses per neuron analyzed, which would be expected to be more variable than mEPSC recordings averaged across several hundred events. However, when they reduce the mEPSC number of events to similar numbers as the imaging, the mESPC amplitudes are still less variable than the imaging data. The reason for this remains unclear. The pool of sampled synapses is still different between the methods and recent data has shown that synapses have variable responses during HSP. Further, there could be variability in the subunit composition of newly inserted AMPARs, and only assessing GluA2 could mask this (see below). It is intriguing that pre-synaptic changes might contribute to HSP, especially given the likely localization of Rab3A. But it remains difficult to distinguish if the apparent difference in imaging and electrophysiology is a methodological issue rather than a biological one. Stronger data, especially positive data on changes in release, will be necessary to conclude that pre-synaptic factors are required for HSP, beyond the established changes in post-synaptic receptor trafficking.

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a strong frequency effect that is unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. But the change in frequency seems to argue (as the authors do) that some synapses only have CP-AMPARs, while the rest of the synapses have few or none. Another possibility is that there are pre-synaptic NASPM-sensitive receptors that influence release probability. Further, the amplitude data show a strong trend towards smaller amplitude following NASPM treatment (Fig 3B). The p value for both control and TTX neurons was 0.08 - it is very difficult to argue that there is no effect. The decrease on average is larger in the TTX neurons, and some cells show a strong effect. It is possible there is some heterogeneity between neurons on whether GluA1/A2 heteromers or GluA1 homomers are added during HSP. This would impact the conclusions about the GluA2 imaging as compared to the mEPSC amplitude data.

To understand the role of Rab3A in HSP will require addressing two main issues:

(1) Is Rab3A acting pre-synaptically, post-synaptically or both? The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where it is acting (pre or post) would aid substantially in understanding its role. The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. More concrete support for the authors' suggestion of a pre-synaptic site of control would be helpful.

(2) Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs or a decrease in GABA release (ie the opposite of whatever is happening at excitatory synapses). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling; an effect only at excitatory synapses would argue for a more specific role just at those synapses.

Author response:

The following is the authors’ response to the previous reviews

General Response to Reviewers:

We thank the Reviewers for their comments, which continue to substantially improve the quality and clarity of the manuscript, and therefore help us to strengthen its message while acknowledging alternative explanations.

All three reviewers raised the concern that we have not proven that Rab3A is acting on a presynaptic mechanism to increase mEPSC amplitude after TTX treatment of mouse cortical cultures. The reviewers’ main point is that we have not shown a lack of upregulation of postsynaptic receptors in mouse cortical cultures. We want to stress that we agree that postsynaptic receptors are upregulated after activity block in neuronal cultures. However, the reviewers are not acknowledging that we have previously presented strong evidence at the mammalian NMJ that there is no increase in AChR after activity blockade, and therefore the requirement for Rab3A in the homeostatic increase in quantal amplitude points to a presynaptic contribution. We agree that we should restrict our firmest conclusions to the data in the current study, but in the Discussion we are proposing interpretations. We have added the following new text:

“The impetus for our current study was two previous studies in which we examined homeostatic regulation of quantal amplitude at the NMJ. An advantage of studying the NMJ is that synaptic ACh receptors are easily identified with fluorescently labeled alpha-bungarotoxin, which allows for very accurate quantification of postsynaptic receptor density. We were able to detect a known change due to mixing 2 colors of alpha-BTX to within 1% (Wang et al., 2005). Using this model synapse, we showed that there was no increase in synaptic AChRs after TTX treatment, whereas miniature endplate current increased 35% (Wang et al., 2005). We further showed that the presynaptic protein Rab3A was necessary for full upregulation of mEPC amplitude (Wang et al., 2011). These data strongly suggested Rab3A contributed to homeostatic upregulation of quantal amplitude via a presynaptic mechanism. With the current study showing that Rab3A is required for the homeostatic increase in mEPSC amplitude in cortical cultures, one interpretation is that in both situations, Rab3A is required for an increase in the presynaptic quantum.”

The point we are making is that the current manuscript is an extension of that work and interpretation of our findings regarding the variability of upregulation of postsynaptic receptors in our mouse cortical cultures further supports the idea that there is a Rab3Adependent presynaptic contribution to homeostatic increases in quantal amplitude.

Public Reviews:

Reviewer #1 (Public review):

Koesters and colleagues investigated the role of the small GTPase Rab3A in homeostatic scaling of miniature synaptic transmission in primary mouse cortical cultures using electrophysiology and immunohistochemistry. The major finding is that TTX incubation for 48 hours does not induce an increase in the amplitude of excitatory synaptic miniature events in neuronal cortical cultures derived from Rab3A KO and Rab3A Earlybird mutant mice. NASPM application had comparable effects on mEPSC amplitude in control and after TTX, implying that Ca2+-permeable glutamate receptors are unlikely modulated during synaptic scaling. Immunohistochemical analysis revealed no significant changes in GluA2 puncta size, intensity, and integral after TTX treatment in control and Rab3A KO cultures. Finally, they provide evidence that loss of Rab3A in neurons, but not astrocytes, blocks homeostatic scaling. Based on these data, the authors propose a model in which neuronal Rab3A is required for homeostatic scaling of synaptic transmission, potentially through GluA2-independent mechanisms.

The major finding - impaired homeostatic up-scaling after TTX treatment in Rab3A KO and Rab3 earlybird mutant neurons - is supported by data of high quality. However, the paper falls short of providing any evidence or direction regarding potential mechanisms. The data on GluA2 modulation after TTX incubation are likely statistically underpowered, and do not allow drawing solid conclusions, such as GluA2-independent mechanisms of up-scaling.

The study should be of interest to the field because it implicates a presynaptic molecule in homeostatic scaling, which is generally thought to involve postsynaptic neurotransmitter receptor modulation. However, it remains unclear how Rab3A participates in homeostatic plasticity.

Major (remaining) point:

(1) Direct quantitative comparison between electrophysiology and GluA2 imaging data is complicated by many factors, such as different signal-to-noise ratios. Hence, comparing the variability of the increase in mini amplitude vs. GluA2 fluorescence area is not valid. Thus, I recommend removing the sentence "We found that the increase in postsynaptic AMPAR levels was more variable than that of mEPSC amplitudes, suggesting other factors may contribute to the homeostatic increase in synaptic strength." from the abstract.

We have not removed the statement, but altered it to soften the conclusion. It now reads, “We found that the increase in postsynaptic AMPAR levels in wild type cultures was more variable than that of mEPSC amplitudes, which might be explained by a presynaptic contribution, but we cannot rule out variability in the measurement.”.

Similarly, the data do not directly support the conclusion of GluA2-independent mechanisms of homeostatic scaling. Statements like "We conclude that these data support the idea that there is another contributor to the TTX- induced increase in quantal size." should be thus revised or removed.

This particular statement is in the previous response to reviewers only, we deleted the sentence that starts, “The simplest explanation Rab3A regulates a presynaptic contributor….”. and “Imaging of immunofluorescence more variable…”. We deleted “ our data suggest….consistently leads to an increase in mEPSC amplitude and sometimes leads to….” We added “…the lack of a robust increase in receptor levels leaves open the possibility that there is a presynaptic contributor to quantal size in mouse cortical cultures. However, the variability could arise from technical factors associated with the immunofluorescence method, and the mechanism of Rab3A-dependent plasticity could be presynaptic for the NMJ and postsynaptic for cortical neurons.”

Reviewer #2 (Public review):

I thank the authors for their efforts in the revision. In general, I believe the main conclusion that Rab3A is required for TTX-induced homeostatic synaptic plasticity is wellsupported by the data presented, and this is an important addition to the repertoire of molecular players involved in homeostatic compensations. I also acknowledge that the authors are more cautious in making conclusions based on the current evidence, and the structure and logic have been much improved.

The only major concern I have still falls on the interpretation of the mismatch between GluA2 cluster size and mEPSC amplitude. The authors argue that they are only trying to say that changes in the cluster size are more variable than those in the mEPSC amplitude, and they provide multiple explanations for this mismatch. It seems incongruous to state that the simplest explanation is a presynaptic factor when you have all these alternative factors that very likely have contributed to the results. Further, the authors speculate in the discussion that Rab3A does not regulate postsynaptic GluA2 but instead regulates a presynaptic contributor. Do the authors mean that, in their model, the mEPSC amplitude increases can be attributed to two factors- postsynaptic GluA2 regulation and a presynaptic contribution (which is regulated by Rab3A)? If so, and Rab3A does not affect GluA2 whatsoever, shouldn't we see GluA2 increase even in the absence of Rab3A? The data in Table 1 seems to indicate otherwise.

The main body of this comment is addressed in the General Response to Reviewers. In addition, we deleted text “current data, coupled with our previous findings at the mouse neuromuscular junction, support the idea that there are additional sources contributing to the homeostatic increase in quantal size.” We added new text, so the sentence now reads: “Increased receptors likely contribute to increases in mESPC amplitudes in mouse cortical cultures, but because we do not have a significant increase in GluA2 receptors in our experiments, it is impossible to conclude that the increase is lacking in cultures from Rab3A-/- neurons.”

I also question the way the data are presented in Figure 5. The authors first compare 3 cultures and then 5 cultures altogether, if these experiments are all aimed to answer the same research question, then they should be pooled together. Interestingly, the additional two cultures both show increases in GluA2 clusters, which makes the decrease in culture #3 even more perplexing, for which the authors comment in line 261 that this is due to other factors. Shouldn't this be an indicator that something unusual has happened in this culture?

Data in this figure is sufficient to support that GluA2 increases are variable across cultures, which hardly adds anything new to the paper or to the field.

A major goal of performing the immunofluorescence measurements in the same cultures for which we had electrophysiological results was to address the common impression that the homeostatic effect itself is highly variable, as the reviewer notes in the comment “…GluA2 increases are variable across cultures…” Presumably, if GluA2 increases are the mechanism of the mEPSC amplitude increases, then variable GluA2 increases should correlate with variable mEPSC amplitude increases, but that is not what we observed. We are left with the explanation that the immunofluorescence method itself is very variable. We have added the point to the Discussion, which reads, “the variability could arise from technical factors associated with the immunofluorescence method, and the mechanism of Rab3A-dependent homeostatic plasticity could be presynaptic for the NMJ and postsynaptic for cortical neurons.”

Finally, the implication of “Shouldn’t this be an indicator that something unusual has happened in this culture?” if it is not due to culture to culture variability in the homeostatic response itself, is that there was a technical problem with accurately measuring receptor levels. We have no reason to suspect anything was amiss in this set of coverslips (the values for controls and for TTX-treated were not outside the range of values in other experiments). In any of the coverslips, there may be variability in the amount of primary anti-GluA2 antibody, as this was added directly to the culture rather than prepared as a diluted solution and added to all the coverslips. But to remove this one experiment because it did not give the expected result is to allow bias to direct our data selection.

The authors further cite a study with comparable sample sizes, which shows a similar mismatch based on p values (Xu and Pozzo-Miller 2007), yet the effect sizes in this study actually match quite well (both ~160%). P values cannot be used to show whether two effects match, but effect sizes can. Therefore, the statement in lines 411-413 "... consistently leads to an increase in mEPSC amplitudes, and sometimes leads to an increase in synaptic GluA2 receptor cluster size" is not very convincing, and can hardly be used to support "the idea that there are additional sources contributing to the homeostatic increase in quantal size.”

We have the same situation; our effect sizes match (19.7% increase for mEPSC amplitude; 18.1% increase for GluA2 receptor cluster size, see Table 1), but in our case, the p value for receptors does not reach statistical significance. Our point here is that there is published evidence that the variability in receptor measurements is greater than the variability in electrophysiological measurements. But we have softened this point, removing the sentences containing “…consistently leads and sometimes...” and “……additional sources contributing…”.

I would suggest simply showing mEPSC and immunostaining data from all cultures in this experiment as additional evidence for homeostatic synaptic plasticity in WT cultures, and leave out the argument for "mismatch". The presynaptic location of Rab3A is sufficient to speculate a presynaptic regulation of this form of homeostatic compensation.

We have removed all uses of the word “mismatch,” but feel the presentation of the 3 matched experiments, 23-24 cells (Figure 5A, D), and the additional 2 experiments for a total of 5 cultures, 48-49 cells (Figure 5C, F), is important in order to demonstrate that the lack of statistically significant receptor response is due neither to a variable homeostatic response in the mEPSC amplitudes, nor to a small number of cultures.

Minor concerns:

(1) Line 214, I see the authors cite literature to argue that GluA2 can form homomers and can conduct currents. While GluA2 subunits edited at the Q/R site (they are in nature) can form homomers with very low efficiency in exogenous systems such as HEK293 cells (as done in the cited studies), it's unlikely for this to happen in neurons (they can hardly traffic to synapses if possible at all).

We were unable to identify a key reference that characterized GluA2 homomers vs. heteromers in native cortical neurons, but we have rewritten the section in the manuscript to acknowledge the low conductance of homomers:

“…to assess whether GluA2 receptor expression, which will identify GluA2 homomers and GluA2 heteromers (the former unlikely to contribute to mEPSCs given their low conductance relative to heteromers (Swanson et al., 1997; Mansour et al., 2001)…”

(2) Lines 221-222, the authors may have misinterpreted the results in Turrigiano 1998. This study does not show that the increase in receptors is most dramatic in the apical dendrite, in fact, this is the only region they have tested. The results in Figures 3b-c show that the effect size is independent of the distance from soma.

Figure 3 in Turrigiano et al., shows that the increase in glutamate responsiveness is higher at the cell body than along the primary dendrite. We have revised our description to indicate that an increase in responsiveness on the primary dendrite has been demonstrated in Turrigiano et al. 1998.

“We focused on the primary dendrite of pyramidal neurons as a way to reduce variability that might arise from being at widely ranging distances from the cell body, or, from inadvertently sampling dendritic regions arising from inhibitory neurons. In addition, it has been shown that there is a clear increase in response to glutamate in this region (Turrigiano et al., 1998).”

“…synaptic receptors on the primary dendrite, where a clear increase in sensitivity to exogenously applied glutamate was demonstrated (see Figure 3 in (Turrigiano et al., 1998)).

(3) Lines 309-310 (and other places mentioning TNFa), the addition of TNFa to this experiment seems out of place. The authors have not performed any experiment to validate the presence/absence of TNFa in their system (citing only 1 study from another lab is insufficient). Although it's convincing that glia Rab3A is not required for homeostatic plasticity here, the data does not suggest Rab3A's role (or the lack of) for TNFa in this process.

We have modified the paragraph in the Discussion that addresses the glial results, to describe more clearly the data that supported an astrocytic TNF-alpha mechanism: “TNF-alpha accumulates after activity blockade, and directly applied to neuronal cultures, can cause an increase in GluA1 receptors, providing a potential mechanism by which activity blockade leads to the homeostatic upregulation of postsynaptic receptors (Beattie et al., 2002; Stellwagen et al., 2005; Stellwagen and Malenka, 2006).”

We have also acknowledged that we cannot rule out TNF-alpha coming from neurons in the cortical cultures: “…suggesting the possibility that neuronal Rab3A can act via a non-TNF-alpha mechanism to contribute to homeostatic regulation of quantal amplitude, although we have not ruled out a neuronal Rab3A-mediated TNF-alpha pathway in cortical cultures.”

Reviewer #3 (Public review):

This manuscript presents a number of interesting findings that have the potential to increase our understanding of the mechanism underlying homeostatic synaptic plasticity (HSP). The data broadly support that Rab3A plays a role in HSP, although the site and mechanism of action remain uncertain.

The authors clearly demonstrate that Rab3A plays a role in HSP at excitatory synapses, with substantially less plasticity occurring in the Rab3A KO neurons. There is also no apparent HSP in the Earlybird Rab3A mutation, although baseline synaptic strength is already elevated. In this context, it is unclear if the plasticity is absent, already induced by this mutation, or just occluded by a ceiling effect due to the synapses already being strengthened. Occlusion may also occur in the mixed cultures when Rab3A is missing from neurons but not astrocytes. The authors do appropriately discuss these options. The authors have solid data showing that Rab3A is unlikely to be active in astrocytes, Finally, they attempt to study the linkage between changes in synaptic strength and AMPA receptor trafficking during HSP, and conclude that trafficking may not be solely responsible for the changes in synaptic strength during HSP.

Strengths:

This work adds another player into the mechanisms underlying an important form of synaptic plasticity. The plasticity is likely only reduced, suggesting Rab3A is only partially required and perhaps multiple mechanisms contribute. The authors speculate about some possible novel mechanisms, including whether Rab3A is active pre-synaptically to regulate quantal amplitude.

As Rab3A is primarily known as a pre-synaptic molecule, this possibility is intriguing. However, it is based on the partial dissociation of AMPAR trafficking and synaptic response and lacks strong support. On average, they saw a similar magnitude of change in mEPSC amplitude and GluA2 cluster area and integral, but the GluA2 data was not significant due to higher variability. It is difficult to determine if this is due to biology or methodology - the imaging method involves assessing puncta pairs (GluA2/VGlut1) clearly associated with a MAP2 labeled dendrite. This is a small subset of synapses, with usually less than 20 synapses per neuron analyzed, which would be expected to be more variable than mEPSC recordings averaged across several hundred events. However, when they reduce the mEPSC number of events to similar numbers as the imaging, the mESPC amplitudes are still less variable than the imaging data. The reason for this remains unclear. The pool of sampled synapses is still different between the methods and recent data has shown that synapses have variable responses during HSP. Further, there could be variability in the subunit composition of newly inserted AMPARs, and only assessing GluA2 could mask this (see below). It is intriguing that pre-synaptic changes might contribute to HSP, especially given the likely localization of Rab3A. But it remains difficult to distinguish if the apparent difference in imaging and electrophysiology is a methodological issue rather than a biological one. Stronger data, especially positive data on changes in release, will be necessary to conclude that pre-synaptic factors are required for HSP, beyond the established changes in post-synaptic receptor trafficking.

Regarding the concern that the lack of increase in receptors is due to a technical issue, please see General Response to Reviewers, above. We have also softened our conclusions throughout, acknowledging we cannot rule out a technical issue.

Other questions arise from the NASPM experiments, used to justify looking at GluA2 (and not GluA1) in the immunostaining. First, there is a strong frequency effect that is unclear in origin. One would expect NASPM to merely block some fraction of the post-synaptic current, and not affect pre-synaptic release or block whole synapses. But the change in frequency seems to argue (as the authors do) that some synapses only have CP-AMPARs, while the rest of the synapses have few or none. Another possibility is that there are pre-synaptic NASPM-sensitive receptors that influence release probability. Further, the amplitude data show a strong trend towards smaller amplitude following NASPM treatment (Fig 3B). The p value for both control and TTX neurons was 0.08 - it is very difficult to argue that there is no effect. The decrease on average is larger in the TTX neurons, and some cells show a strong effect. It is possible there is some heterogeneity between neurons on whether GluA1/A2 heteromers or GluA1 homomers are added during HSP. This would impact the conclusions about the GluA2 imaging as compared to the mEPSC amplitude data.

The key finding in Figure 3 is that NASPM did not eliminate the statistically significant increase in mEPSC amplitude after TTX treatment (Fig 3A). Whether or not NASPM sensitive receptors contribute to mESPC amplitude is a separate question (Fig 3B). We are open to the possibility that NASPM reduces mEPSC amplitude in both control and TTX treated cells (p = 0.08 for both), but that does not change our conclusion that NASPM has no effect on the TTX-induced increase in mEPSC amplitude. The mechanism underlying the decrease in mEPSC frequency following NASPM is interesting, but does not alter our conclusions regarding the role of Rab3A in homeostatic synaptic plasticity of mEPSC amplitude. In addition, the Reviewer does not acknowledge the Supplemental Figure #1, which shows a similar lack of correspondence between homeostatic increases in mEPSC amplitude and GluA1 receptors in two cultures where matched data were obtained. Therefore, we do not think our lack of a robust increase in receptors can be explained by our failing to look at the relevant receptor.

To understand the role of Rab3A in HSP will require addressing two main issues:

(1) Is Rab3A acting pre-synaptically, post-synaptically or both? The authors provide good evidence that Rab3A is acting within neurons and not astrocytes. But where it is acting (pre or post) would aid substantially in understanding its role. The general view in the field has been that HSP is regulated post-synaptically via regulation of AMPAR trafficking, and considerable evidence supports this view. More concrete support for the authors' suggestion of a pre-synaptic site of control would be helpful.

We agree that definitive evidence for a presynaptic role of Rab3A in homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures requires demonstrating that loss of Rab3A in postsynaptic neurons does not disrupt the plasticity, whereas loss in presynaptic neurons does. Without these data, we can only speculate that the Rab3A-dependence of homeostatic plasticity of quantal size in cortical neurons may be similar to that of the neuromuscular junction, where it cannot be receptors. We have added to the Discussion that the mechanism of Rab3A regulation of homeostatic plasticity of quantal amplitude could different between cortical neurons and the neuromuscular junction (lines 448-450 in markup,). Establishing a way to co-culture Rab3A-/- and Rab3A+/+ neurons in ratios that would allow us to record from a Rab3A-/- neuron that has mainly Rab3A+/+ inputs (or vice versa) is not impossible, but requires either transfection or transgenic expression with markers that identify the relevant genotype, and will be the subject of future experiments.

(2): Rab3A is also found at inhibitory synapses. It would be very informative to know if HSP at inhibitory synapses is similarly affected. This is particularly relevant as at inhibitory synapses, one expects a removal of GABARs or a decrease in GABA release (ie the opposite of whatever is happening at excitatory synapses). If both processes are regulated by Rab3A, this might suggest a role for this protein more upstream in the signaling; an effect only at excitatory synapses would argue for a more specific role just at those synapses.

We agree with the Reviewer, that it is important to determine the generality of Rab3A function in homeostatic plasticity. Establishing the homeostatic effect on mIPSCs and then examining them in Rab3A-/- cultures is a large undertaking and will be the subject of future experiments.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor (remaining) points:

(1) The figure referenced in the first response to the reviewers (Figure 5G) does not exist.

We meant Figure 5F, which has been corrected in the current response.

(2) I recommend showing the data without binning (despite some overlap).

The box plot in Origin will not allow not binning, but we can make the bin size so small that for all intents and purposes, there is close to 1 sample in each bin. When we do this, the majority of data are overlapped in a straight vertical line. Previously described concerns were regarding the gaps in the data, but it should be noted that these are cell means and we are not depicting the distributions of mEPSC amplitudes within a recording or across multiple recordings.

(3) Please auto-scale all axes from 0 (e.g., Fig 1E, F).

We have rescaled all mEPSC amplitude axes in box plots to go from 0 (Figures 1, 2 and 6).

(4) Typo in Figure legend 3: "NASPM (20 um)" => uM

Fixed.

Reviewer #2 (Recommendations for the authors):

(1) Line 140, frequencies are reported in Hz while other places are in sec-1, while these are essentially the same, they should be kept consistent in writing.

All mEPSC frequencies have been changed to sec-1, except we have left “Hz” for repetitive stimulation and filtering.

(2) Paragraph starting from line 163 (as well as other places where multiple groups are compared, such as the occlusion discussion), the authors assessed whether there was a change in baseline between WT and mutant group by doing pairwise tests, this is not the right test. A two-way ANOVA, or at least a multivariant test would be more appropriate.

We have performed a two-way ANOVA, with genotype as one factor, and treatment as the other factor. The p values in Figures 1 and 2 have been revised to reflect p values from the post-hoc Tukey test on the specific interactions (for each particular genotype, TTX vs CON effects). The difference in the two WT strains, untreated, was not significant in the Post-Hoc Tukey test, and we have revised the text. The difference between the untreated WT from the Rab3A+/Ebd colony and the untreated Rab3AEbd/Ebd mutant was still significant in the Post-Hoc Tukey test, and this has replaced the Kruskal-Wallis test. The two-way ANOVA was also applied to the neuron-glia experiments and p values in Figure 6 adjusted accordingly.

(3) Relevant to the second point under minor concerns, I suggest this sentence be removed, as reducing variability and avoiding inhibitory projects are reasons good enough to restrict the analysis to the apical dendrites.

We have revised the description of the Turrigiano et al., 1998 finding from their Figure 3 and feel it still strengthens the justification for choosing to analyze only synapses on the apical dendrite.

Reviewer #3 (Recommendations for the authors):

Minor points:

The comments on lines 256-7 could seem misleading - the NASPM results wouldn't rule out contribution of those other subunits, only non-GluA2 containing combinations of those subunits. I would suggest revising this statement. Also, NASPM does likely have an effect, just not one that changes much with TTX treatment.

At new line 213 (markup) we have added the modifier “homomeric” to clarify our point that the lack of NASPM effect on the increase in mEPSC amplitude after TTX indicates that the increase is not due to more homomeric Ca2+-permeable receptors. We have always stated that NASPM reduces mEPSC amplitude, but it is in both control and treated cultures.

Strong conclusions based on a single culture (lines 314-5) seem unwarranted.

We have softened this statement with a “suggesting that” substituted for the previous “Therefore,” but stand by our point that the mEPSC amplitude data support a homeostatic effect of TTX in Culture #3, so the lack of increase in GluA2 cluster size needs an explanation other than variability in the homeostatic effect itself.

Saying (line 554) something is 'the only remaining possibility' also seems unwarranted.

We have softened this statement to read, “A remaining possibility…”.

Beattie EC, Stellwagen D, Morishita W, Bresnahan JC, Ha BK, Von Zastrow M, Beattie MS, Malenka RC (2002) Control of synaptic strength by glial TNFalpha. Science 295:2282-2285.

Mansour M, Nagarajan N, Nehring RB, Clements JD, Rosenmund C (2001) Heteromeric AMPA receptors assemble with a preferred subunit stoichiometry and spatial arrangement. Neuron 32:841-853. Stellwagen D, Malenka RC (2006) Synaptic scaling mediated by glial TNF-alpha. Nature 440:1054-1059.

Stellwagen D, Beattie EC, Seo JY, Malenka RC (2005) Differential regulation of AMPA receptor and GABA receptor trafficking by tumor necrosis factor-alpha. J Neurosci 25:3219-3228.

Swanson GT, Kamboj SK, Cull-Candy SG (1997) Single-channel properties of recombinant AMPA receptors depend on RNA editing, splice variation, and subunit composition. J Neurosci 17:5869.

Turrigiano GG, Leslie KR, Desai NS, Rutherford LC, Nelson SB (1998) Activity-dependent scaling of quantal amplitude in neocortical neurons. Nature 391:892-896.

Wang X, Wang Q, Yang S, Bucan M, Rich MM, Engisch KL (2011) Impaired activity-dependent plasticity of quantal amplitude at the neuromuscular junction of Rab3A deletion and Rab3A earlybird mutant mice. J Neurosci 31:3580-3588.

Wang X, Li Y, Engisch KL, Nakanishi ST, Dodson SE, Miller GW, Cope TC, Pinter MJ, Rich MM (2005) Activity-dependent presynaptic regulation of quantal size at the mammalian neuromuscular junction in vivo. J Neurosci 25:343-351.

Version 4

Abstract

Abstract

Following prolonged activity blockade, amplitudes of miniature excitatory postsynaptic currents (mEPSCs) increase, a form of plasticity termed “homeostatic synaptic plasticity.” We previously showed that a presynaptic protein, the small GTPase Rab3A, is required for full expression of the increase in miniature endplate current amplitudes following prolonged blockade of action potential activity at the mouse neuromuscular junction in vivo, where an increase in postsynaptic receptors does not contribute (Wang et al., 2005; Wang et al., 2011). It is unknown whether this form of Rab3A-dependent homeostatic plasticity at the neuromuscular junction shares any characteristics with central synapses. We show here that homeostatic synaptic plasticity of mEPSCs is impaired in mouse cortical neuron cultures prepared from Rab3A-/- and mutant mice expressing a single point mutation of Rab3A, Rab3A Earlybird mice. To determine if Rab3A is involved in the well-established homeostatic increase in postsynaptic AMPA-type receptors (AMPARs), we performed a series of experiments in which electrophysiological recordings of mEPSCs and confocal imaging of synaptic AMPAR immunofluorescence were assessed within the same cultures. We found that the increase in postsynaptic AMPAR levels in wild type cultures was more variable than that of mEPSC amplitudes, which might be explained by a presynaptic contribution, but we cannot rule out variability in the measurement. Finally, we demonstrate that Rab3A is acting in neurons because only selective loss of Rab3A in neurons, not glia, disrupted the homeostatic increase in mEPSC amplitudes. This is the first demonstration that a protein thought to function presynaptically is required for homeostatic synaptic plasticity of quantal size in central neurons.

Introduction

One of the most studied phenomena triggered by prolonged activity blockade is the increase in amplitudes of miniature excitatory postsynaptic currents (mEPSCs) in neurons. First demonstrated in cultures of dissociated cortical neurons (Turrigiano et al., 1998) and spinal cord neurons (O’Brien et al., 1998), the compensatory response to prolonged loss of activity has been dubbed “homeostatic synaptic plasticity” (Turrigiano and Nelson, 2004; Pozo and Goda, 2010). These first two studies provided evidence that glutamate receptors were also increased after prolonged inactivity, via an increased response to exogenously applied glutamate (Turrigiano et al., 1998) and increased immunofluorescent labeling of GluA1 receptors at synaptic sites (O’Brien et al., 1998). The accompanying increase in synaptic AMPA receptors has been confirmed with immunofluorescence in multiple studies of hippocampal and cortical cultures treated with TTX, including (Wierenga et al., 2005; Hou et al., 2008; Ibata et al., 2008; Jakawich et al., 2010; Tatavarty et al., 2013; Xu and Pozzo-Miller, 2017; Dubes et al., 2022). It is now well accepted that the homeostatic increase in mEPSC amplitudes in neurons is due to an increase in postsynaptic AMPA receptors.

In our previous work, we found that a TTX cuff applied for 48 hrs around the sciatic nerve in mice led to an increase in the amplitude of the miniature endplate currents (mEPCs) recorded in tibialis anterior muscles. Surprisingly, and unlike at central synapses, we could find no evidence that the acetylcholine receptors levels at the neuromuscular junction (NMJ) were increased in the TTX-cuff treated mice (Wang et al., 2005). This result led us to search for presynaptic molecules that might homeostatically regulate the presynaptic quantum. In previous studies in chromaffin cells, we identified the small GTPase Rab3A, a presynaptic vesicle protein, as a regulator of synaptic vesicle fusion pore opening (Wang et al., 2008), so we examined whether deletion of Rab3A (Rab3A-/-) might prevent homeostatic upregulation of mEPC amplitude at the NMJ. The results were clear: the homeostatic increase in mEPC amplitude induced by the TTX cuff was strongly reduced in the Rab3A-/- mouse, and was completely abolished in the Earlybird mutant (Rab3AEbd/Ebd), which has a single point mutation in Rab3A that causes a shift towards early awakening due to a shorter circadian period, that is more dramatic than that of the Rab3A deletion mouse (Kapfhamer et al., 2002; Wang et al., 2011). These results led us to conclude that the homeostatic increase in mEPC amplitude at the NMJ is a presynaptic phenomenon.

It has remained somewhat of a mystery what the role of Rab3A is in synaptic transmission. The Rab3A-/- mouse has minimal phenotypic abnormalities, with evoked synaptic transmission and mEPSCs essentially normal in hippocampal slices (Geppert et al., 1994). At the Rab3A-/- NMJ, reductions in evoked transmission were detected, but only under conditions of reduced extracellular calcium (Coleman et al., 2007). There are some modest changes in short term plasticity during repetitive stimulus trains, with increased depression observed in response to moderate frequencies in hippocampal slices (Geppert et al., 1994) and at the mammalian NMJ (Coleman and Bykhovskaia, 2009), but increased facilitation in response to high frequency stimulation at the mammalian NMJ (Coleman and Bykhovskaia, 2010). The most dramatic effect of loss of Rab3A is the abolishment of a presynaptic form of long-term potentiation (LTP) at the mossy fiber-CA3 synapse that is induced by 25 Hz stimulation in the presence of NMDA blockers (Weisskopf et al., 1994; Castillo et al., 1997). Our result that prolonged inactivity-induced plasticity of mEPC amplitude at the mammalian NMJ is disrupted with loss of function of Rab3A further cements a role for Rab3A in activity-dependent plasticity of synaptic strength. However, a crucial question that remains is whether the TTX-cuff induced effect on the amplitude of the spontaneous synaptic current at the NMJ shares underlying mechanisms with central synapses. We set out to answer this question for the more well-studied phenomenon of homeostatic plasticity of mEPSCs in dissociated mouse cortical neurons prepared from wild type, Rab3A-/- mice and Rab3aEbd/Ebd mice treated with TTX for 48 hrs. We show here that homeostatic synaptic plasticity of mEPSC amplitude was strongly reduced in cultures from Rab3A-/- mice, and completely abolished in cultures from Rab3AEbd/Ebd mice, supporting a shared mechanism with the neuromuscular junction.

Results

We previously reported that mixed cultures of cortical neurons and glia prepared from postnatal day 0-2 mouse pups responded to a block of action potential-mediated activity by a 48 hr TTX treatment with an increase in mEPSC amplitude (Hanes et al., 2020). Here, we asked the question whether cortical cultures prepared from mice lacking the small GTPase Rab3A, or, expressing a point mutation of Rab3A, Rab3A Earlybird, have an altered homeostatic plasticity response to a prolonged loss of network activity. The Rab3A Earlybird mutation was discovered in a mutagenesis screen in mice for genes involved in the rest-activity cycle (Kapfhamer et al., 2002). In that study, the authors concluded that the Earlybird mutation was likely dominant-negative, as the shift in circadian period was substantially more dramatic than that observed in the Rab3A-/- mice. As described in the Introduction, the effects of loss of Rab3A have been modest, so we included experiments using the Rab3A Earlybird mutant mice to see if there would be a more robust phenotype, and in doing so, strengthen our conclusions based on the Rab3A deletion mice. To obtain Rab3A-/- and Rab3AEbd/Ebd homozygotes, we established two mouse colonies of heterozygous breeders with cultures prepared from pups derived from a final breeding pair of homozygotes. Although we backcrossed Rab3A+/- with Rab3A+/Ebd for 11 generations, clear differences in mEPSC amplitudes in untreated cortical cultures for the two wild type strains (see below) and in calcium current amplitudes in adrenal chromaffin cells for the two wild type strains (unpublished obs.) remained. Therefore, throughout this study there are two Rab3A+/+ or ‘wild type’ phenotypes.

We found that loss of Rab3A greatly impaired the homeostatic increase after activity blockade. Example current traces of spontaneously occurring mEPSCs recorded from pyramidal neurons in untreated (CON) 13-14 DIV cortical cultures and sister cultures treated with 500 nM TTX for 48 hr prepared from wild type and Rab3A-/- mice in the Rab3A+/- colony are shown in Figures 1A and B, respectively. Average mEPSC waveforms from the same recordings are shown in Figures 1C and D. The mean mEPSC amplitudes for 30 control and 23 TTX-treated neurons from Rab3A+/+ cultures are displayed in the box and whisker plot in Figure 1E; after activity blockade the average mEPSC amplitude increased from 13.9 ± 0.7 pA to 18.2 ± 0.9 pA. In contrast, in cultures prepared from Rab3A-/- mice, the average mEPSC amplitude was not significantly increased, for 25 untreated cells and 26 TTX-treated cells (Figure 1F, 13.6 ± 0.1 pA vs. 14.3 ± 0.6 pA). We found that TTX treatment also resulted in an increase in mEPSC frequency in cultures prepared from Rab3A+/+ mice, as shown in Figure 1G (CON, 2.26 ± 0.37 sec-1; TTX, 4.62 ± 0.74 sec-1). This TTX-induced frequency increase was strongly reduced in cultures from Rab3A-/- mice, but there were a few outliers with very high frequencies still present after TTX treatment (Figure 1H, CON, 2.74 ± 0.49 sec-1; TTX, 3.23 ± 0.93 sec-1).

Loss of Rab3A disrupted the TTX-induced increase in amplitudes of mEPSCs recorded in cultured mouse cortical neurons; the increase in frequency was more variable but also appears to be reduced.

(A) Ten second example traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/- colony. (B) Ten second example traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A-/- mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A+/+ mice (Rab3A+/- colony; CON, N = 30 cells, 13.9 ± 0.7 pA; TTX, N = 23 cells, 18.2 ± 0.9 pA; from 11 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells in cultures prepared from Rab3A-/- mice (CON, N = 25 cells, 13.6 ± 0.7 pA; TTX, N = 26 cells, 14.3 ± 0.6 pA); from 11 cultures. (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 2.26 ± 0.37 sec-1; TTX, 4.62 ± 0.74 sec- 1). (H) Box plots for average mEPSC frequency for same Rab3A-/-cells as in (F) (CON, 2.74 ± 0.49 sec-1; TTX, 3.23 ± 0.93 sec-1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells, bin size 0.1 pA, 0.5 sec-1; line, median; dot, mean. p values (shown on the graphs) are from Tukey’s post-hoc test following a two-way ANOVA. For all p-values, * with underline indicates significance with p < 0.05.

We next examined homeostatic plasticity in cultures prepared from wild type mice in the Rab3A+/Ebd colony. Example current traces of mEPSCs, and average mEPSC waveforms, are shown in Figures 2A and C, respectively, for cortical cultures prepared from wild type mice in the Rab3A+/Ebd colony. Treatment with TTX for 48 hr led to a significant increase in the average mEPSC amplitude of 23 TTX-treated cells when compared to 20 untreated cells (Figure 2E, CON, 11.0 ± 0.6 pA; TTX, 15.1 ± 1.2 pA). We note here that while the two wild type strains respond very similarly to activity blockade, the mean mEPSC amplitude in untreated cultures was different between the two strains, 13.9 ± 0.7 pA, wild type from Rab3A+/- colony; 11.0 ± 0.6 pA, wild type from Rab3A+/Ebd colony, but this difference did not reach statistical significance in a Tukey means comparison after a two-way ANOVA (p = 0.27).

Normally functioning Rab3A was required for TTX-induced homeostatic plasticity of mEPSC amplitudes in cultured mouse cortical neurons.

(A) Ten second example traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3A+/+ mice from the Rab3A+/Ebd colony. (B) Ten second example traces recorded at -60 mV in pyramidal cortical neurons from an untreated (“CON”) and TTX-treated (“TTX”) neuron in cultures prepared from Rab3AEbd/Ebd mice. (C & D) Average traces for the recordings shown in A and B, respectively. (E) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3A+/+ mice (Rab3A+/Ebd colony; CON, N = 20 cells, 11.0 ± 0.6 pA; TTX, N = 23 cells, 15.1 ± 1.2 pA; from 6 cultures). (F) Box plots for average mEPSC amplitudes from untreated cells and TTX-treated cells from cultures prepared from Rab3AEbd/Ebd mice (CON, N = 21 cells, 15.1 ± 1.0 pA; TTX, N = 22 cells, 14.6 ± 1.1 pA; from 7 cultures.) (G) Box plots for average mEPSC frequency for same Rab3A+/+ cells as in (E) (CON, 1.15 ± 0.19 sec-1; TTX, 2.54 ± 0.55 sec-1). (H) Box plots for average mEPSC frequency for same Rab3AEbd/Ebd cells as in (F) (CON, 1.71 ± 0.41 sec-1; TTX, 3.05 ± 0.80 sec-1). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; bin size 0.1 pA, 0.5 sec-1; line, median; dot, mean. p values (shown on the graphs) are from a Tukey’s post-hoc test following a two-way ANOVA. For all p-values, * with underline indicates significance with p < 0.05.

We found a complete disruption of homeostatic plasticity in cortical cultures prepared from Rab3AEbd/Ebd mice, as can been seen in viewing example mEPSC traces and average mEPSC waveforms (Figures 2B and D, respectively). The lack of TTX effect was confirmed in a comparison of mEPSC amplitude means for 21 untreated and 22 TTX-treated cells (Figure 2F, CON, 15.1 ± 1.0 pA vs. TTX, 14.6 ± 1.1 pA). There was a trend towards increased mEPSC frequency in TTX-treated cultures from this wild type strain, but the difference was not as robust as for the wild type cultures from Rab3A+/- colony, and did not reach statistical significance (Figure 2G, CON, 1.15 ± 0.19 sec-1; TTX, 2.54 ± 0.55 sec-1). This trend was not observed in neurons from cultures prepared from Rab3AEbd/Ebd mice, likely due to an increase in frequency in untreated cells—frequencies remained high in TTX-treated cells (Figure 2H, CON, 1.71 ± 0.41 sec-1; TTX, 3.05 ± 0.80 sec-1). Our results show that the homeostatic increase in mEPSC amplitude after activity blockade is disrupted in both Rab3A-/- and the Rab3AEbd/Ebdcortical neurons, strongly supporting a crucial role of functioning Rab3A in this process.

The disruption in homeostatic plasticity differs for Rab3A-/- and Rab3AEbd/Ebd. In the Rab3A-/- data set, mEPSCs from untreated cultures were indistinguishable from mEPSCs from Rab3A+/+ untreated cultures, demonstrating loss of Rab3A has no impact on basal activity mEPSC amplitudes. In the Rab3AEbd/Ebddata set, mEPSC amplitudes from untreated cultures were significantly larger than those of untreated cultures from wild type mice, as can be seen comparing the CON values in Figures 2E and F (CON, Rab3A+/+, 11.0 ± 0.7 pA, vs. CON, Rab3AEbd/Ebd, 15.1 ± 1.0 pA, p = 0.04, Tukey means comparison after two-way ANOVA). The increase in mEPSC amplitude in cultures from Rab3AEbd/Ebd mice is consistent with the increase in mEPC amplitude we observed at the Rab3AEbd/Ebd NMJ (Wang et al., 2011).

The small GTPase Rab3A is generally thought to function presynaptically to regulate synaptic vesicle trafficking, possibly in an activity-dependent manner (Castillo et al., 1997; Lonart et al., 1998; Leenders et al., 2001; Schluter et al., 2006; Coleman and Bykhovskaia, 2009; Tian et al., 2012). In contrast, homeostatic plasticity of mEPSC amplitude has been attributed to an increase in postsynaptic receptors on the surface of the dendrite (O’Brien et al., 1998; Turrigiano et al., 1998). At the NMJ in vivo, we could find no evidence for an increase in AChRs after TTX block of the sciatic nerve in vivo (Wang et al., 2005). Is Rab3A required for the homeostatic increase in surface AMPA-type glutamate receptors that has been confirmed by multiple studies (see Introduction)?

In some studies of homeostatic synaptic plasticity it has been found that application of a Ca2+-permeable AMPA receptor-specific inhibitor reversed the homeostatic increase in mEPSC amplitude (Ju et al., 2004; Thiagarajan et al., 2005; Sutton et al., 2006). Figure 3A shows that the TTX-induced increase in mean mEPSC amplitude was nearly identical in a set of 11 CON and 11 TTX-treated cells before and after treatment with 1-Naphthyl acetyl spermine trihydrochloride (NASPM), a synthetic analogue of the Ca2+-permeable AMPA receptor inhibitor philanthotoxin (before NASPM, CON 12.9 ± 3.5 pA; TTX, 17.5 ± 3.1 pA; after NASPM, CON 11.9 ± 2.6 pA; TTX 16.1 ± 3.5 pA). This result indicates that the TTX-induced increase in mEPSC amplitude does not depend on Ca2+-permeable receptors, since the effect of TTX clearly remained when their presence was removed by acute NASPM application. We do not think that there was any technical issue with the NASPM application, because overall, mEPSC amplitudes were reduced modestly in both untreated and TTX-treated cultures (Figure 3B, CON, before NASPM, 12.9 ± 3.5 pA; after NASPM, 11.9 ± 2.6 pA; TTX, before NASPM, 17.5 ± 3.1 pA; after NASPM, 16.1 ± 3.5 pA). Furthermore, NASPM consistently reduced mEPSC frequency (Figure 3C, CON, before NASPM, 1.84 ± 0.55 sec-1; after NASPM, 1.56 ± 0.53 sec-1; TTX, before NASPM, 4.40 ± 3.51 sec-1; after NASPM, 2.68 ± 2.25 sec-1). Similar to the data in Figure 1, frequency is significantly increased after TTX treatment (p = 0.04), but this difference is no longer significant after NASPM application (p = 0.20). One explanation for the effect on frequency is that there are synaptic sites that express only homomers of Ca2+-permeable AMPA receptors (GluA2-lacking AMPARs); mEPSCs from these sites would be expected to be completely blocked by NASPM (see cartoon description in Figure 3D, Left). We cannot rule out an alternative explanation, that presynaptic Ca2+-permeable AMPA receptors normally enhance release probability, and NAPSM prevents this action. The magnitude of the frequency reduction following acute NASPM appears to be greater after TTX treatment, suggesting that loss of activity could promote the establishment of synaptic sites that contain only Ca2+-permeable receptors. However, these new sites do not appear to contribute to the increase in mEPSC amplitude (Figure 3A). The lack of effect of NASPM rules out the contribution of homomeric Ca2+-permeable AMPA receptors GluA1, GluA3 and GluA4 (Hollmann et al., 1991; Burnashev et al., 1992; Jonas and Burnashev, 1995), as well as Ca2+-permeable forms of Kainate receptors GluA5 and GluA6 (Koike et al., 1997; Sun et al., 2009) to homeostatic synaptic plasticity.

Homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures treated with TTX for 48 hr was unchanged by acute inhibition of Ca2+ permeable AMPA receptors by NASPM (20 μm).

(A) Box plot comparison of the TTX effect on mEPSC amplitudes in the same pyramidal neurons before and after application of 20 μM NASPM. (N = 11 cells from 3 cultures; Pre-NASPM, CON: 12.9 ± 1.1 pA; TTX: 17.5 ± 0.9 pA; post-NASPM, CON: 11.9 ± 0.8 pA; TTX: 16.1 ± 1.0 pA). (B) Line series plot of mEPSC amplitudes before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NASPM: 12.9 ± 1.1 pA; post-NASPM: 11.9 ± 0.8 pA; TTX, pre-NASPM: 17.5 ± 0.9 pA; post-NASPM: 16.1 ± 0.1 pA. (C) Line series plot of mEPSC frequency before and after acute perfusion with 20 μM NASPM for untreated and TTX-treated pyramidal neurons; same cells as in (A); CON, pre-NAPSM: 1.84 ± 0.55 sec-1; post-NASPM; 1.56 ± 0.53 sec-1; TTX, pre-NASPM: 4.40 ± 1.06 sec-1; post-NASPM, 2.68 ± 0.68 sec-1. (D) A proposed mechanism for why NASPM had a robust effect on frequency without greatly affecting amplitude. A dendrite with spines (Top) is expanded on three postsynaptic sites (Middle) to show possible types of AMPA receptor distributions: Left, at a site comprised only of Ca2+-permeable AMPA receptors (NASPM-sensitive, GluA2-lacking receptors (black)), NASPM would completely inhibit the mEPSC, causing a decrease in frequency to be measured in the overall population; Right, at a site comprised only of Ca2+-impermeable AMPA receptors (NASPM-insensitive, GluA2-containing receptors (green)), NASPM would have no effect on mEPSC amplitude; Middle, at a site comprised of a mix of Ca2+-permeable and Ca2+-impermeable AMPA receptors, NASPM would partially inhibit the mEPSC. Diagram produced in BioRender.com (2024). Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles; open circles represent means from individual cells; bin size 0.1 pA; line, median; dot, mean. p values (shown on the graphs) are from Kruskal-Wallis test. Line series plot p values are from student’s paired t test. For all p-values, * with underline indicates significance with p < 0.05.

Having established that homomeric Ca2+-permeable receptors are not contributing to the homeostatic increase in mEPSC amplitude, we turned to immunofluorescence and confocal imaging to assess whether GluA2 receptor expression, which will identify GluA2 homomers and GluA2 heteromers (the former unlikely to contribute to mEPSCs given their low conductance relative to heteromers (Swanson et al., 1997; Mansour et al., 2001), was increased in our wild type mouse cortical cultures following 48 hr treatment with TTX. Since mEPSCs necessarily report synaptic levels of receptors, we used VGLUT1-immunofluorescence to identify synapses on pyramidal primary apical dendrites labeled with MAP-2 immunofluorescence. We focused on the primary dendrite of pyramidal neurons as a way to reduce variability that might arise from being at widely ranging distances from the cell body, or, from inadvertently sampling dendritic regions arising from inhibitory neurons. In addition, it has been shown that there is a clear increase in response to glutamate in this region (Turrigiano et al., 1998). Figure 4 (Top) shows two pyramidal neurons, one from an untreated culture on the left, and one from a TTX-treated culture on the right, prepared from Rab3A+/+ mice. To measure characteristics of synaptically located GluA2 receptor clusters, we zoomed in on the primary dendrite (region in white rectangle). The zoomed regions for single confocal sections of the selected area are shown below and contain pairs of VGLUT1- and GluA2-immunofluorescent clusters indicated by white trapezoidal frames. A dendrite typically required ∼10 confocal sections to be fully captured, and the total number of synaptic pairs for all the sections imaged on a dendrite averaged 20 (20.4 ± 6.5 (SD); range, 11-38) so this is an atypically high number of pairs within a single section; these particular dendrites and sections were selected for illustration purposes. In addition to the synaptic pairs, we observed many GluA2-immunofluorescent clusters not associated with VGLUT1 immunofluorescence (a few are indicated with white arrows), most probably extrasynaptic receptors. There were also GluA2-positive clusters present outside of MAP2-positive dendrites, which may be located on astrocytes (Fan et al., 1999). Finally, GluA2 immunofluorescent clusters close to VGLUT1 immunofluorescence but not located along any apparent MAP2-positive neurites suggests the presence of axon-axonal contacts, although VGLUT1 has also been detected in astrocytes (Ormel et al., 2012). Only sites that contained both VGLUT1 and GluA2 immunofluorescence close to the primary MAP2-postive dendrite or a secondary branch along this primary dendrite were selected for analyses. In the experiment from which these images were collected, we analyzed the distance from the cell body for each synapse. The average distance from the cell body, for dendrites from the untreated cell, was 38.5 ± 2.8 μm; for the TTX-treated cell, it was 42.4 ± 3.2 μm (p = 0.35, Kruskal-Wallis test). Since the greatest distance from the cell body is set by the outer limit of the zoom window, and that window was placed adjacent to the cell body, distances in other experiments would be in this same range.

Identification of synaptic GluA2 receptor immunofluorescence on primary dendrites of pyramidal neurons in high density mouse cortical co-cultures prepared from Rab3A+/+ mice.

(Top) Non-zoomed single confocal sections collected with a 60X oil immersion objective of recognizably pyramidal-shaped neurons, presumed to be excitatory neurons, selected for synaptic GluA2 analysis. A neuron was selected from an untreated coverslip (CON, Left) and a TTX-treated coverslip (TTX, Right) from the same culture prepared from Rab3A+/+ animals (= Rab3A+/+ Culture #2 in Figure 5). The white rectangular boxes indicate 5X zoomed areas shown in the images below. Note that the depth of the single confocal section of the non-zoomed neuron image is not at the same depth as the confocal section in zoomed dendritic image, so some features are not visible in both images. Scale bar, 20 μm. (Bottom) 5X zoom single confocal sections selected for demonstration purposes because they had an unusually high number of identified synaptic pairs along the primary dendrite contained within a single confocal section. Synaptic pairs, highlighted with white trapezoids, were identified based on close proximity of GluA2 (red) and VGLUT1 (white) immunofluorescence, apposed to the MAP2 immunofluorescent primary dendrite (green). Some apparent synaptic pairs are not highlighted with a white trapezoid because there was a different confocal section in which the two immunofluorescent sites were maximally bright. There were GluA2 positive clusters located on the primary dendrites that were not apposed to VGLUT1-positive terminals (4 of these non-synaptic GluA2 clusters are highlighted with white arrows in the MAP2-GluA2 and MAP2-GluA2-VGLUT1 panels). Scale bar, 10 μm.

Variability in the magnitude of the homeostatic response from culture to culture is averaged out in physiological experiments by the pooling of data from many cells across many cultures. To reduce the necessity for many cultures, we chose to pair experiments in the same cultures by recording mEPSCs from one set of coverslips, and processing another set of coverslips from the same culture for immunofluorescence. We also wanted to determine if the homeostatic response of mEPSC amplitudes varied from culture to culture, and if so, whether the GluA2 receptor levels varied in parallel. We completed this matched paradigm of physiology and immunofluorescence for 3 cultures prepared from Rab3A+/+ mice. We first present the results as means for each culture, for control and TTX-treated dishes, for mEPSC data (Figure 5A) and imaging data (Figure 5B). Levels of GluA2 immunofluorescence at synaptic sites were quantified by the size of the GluA2-immunofluorescent receptor cluster, because this showed the greatest response to activity blockade; the average intensity value, and the integral of intensity values across the area of the cluster, are included in a summary of the data in Table 1. Three of three cortical cultures showed an increase in mean mEPSC amplitude after TTX treatment. In contrast, one of the same three cultures showed a decrease in mean GluA2 receptor cluster size, suggesting that the decrease in receptor levels in Culture #3 was not due to a lack of homeostatic effect on mEPSC amplitudes in that culture. Because there was a clear trend towards increased GluA2 receptor cluster size, but the receptor response appears more variable, we performed immunofluorescence measurements on two additional cortical cultures, both of which showed modest increases after TTX treatment (Figure 5C). We confirmed that there was also not a consistent increase in GluA1 receptor levels in two matched cultures that showed robust homeostatic increases in mEPSC amplitudes (Supplementary Figure 1A, culture mean mEPSC amplitudes; Supplementary Figure 1B, culture mean GluA1 receptor cluster sizes).

Comparison of mEPSC amplitudes and GluA2 receptor cluster areas in matched and unmatched mouse cortical cultures prepared from Rab3A+/+ mice and treated with TTX for 48 hr.

(A) Culture averages of mEPSC amplitudes for untreated (CON) and TTX-treated coverslips (TTX) in each of 3 Rab3A+/+ mouse cortical co-cultures. Culture #1, CON, N = 6, 14.2 ± 2.2 pA; TTX, N = 6, 15.9 ± 1.9 pA; Culture #2, CON, N = 7, 13.8 ± 2.4 pA; TTX, N = 8, 17.7 ± 1.8 pA; Culture #3, CON, N = 10, 13.4 ± 0.8 pA; TTX, N = 9, 15.7 ± 1.0 pA. (B) Culture averages of GluA2 receptor cluster size in the same 3 cultures as shown in (A). Culture #1, CON, N = 10, 0.73 ± 0.09 μm2; TTX, N = 9, 0.89 ± 0.09 μm2; Culture #2, CON, N = 9, 0.91 ± 0.12 μm2; TTX, N = 9, 1.42 ± 0.24 μm2; Culture #3, CON, N = 10, 1.05 ± 0.11 μm2; TTX, N = 9, 0.95 ± 0.15 μm2. (C) Two additional culture averages are included that did not have corresponding mEPSC recordings. Culture #4, CON, N = 10, 0.58 ± 0.07 μm2: TTX, N = 10, 0.71 ± 0.04 μm2; Culture 5, CON, N =10, 0.89 ± 0.05 μm2; TTX, N = 10, 0.95 ± 0.08 μm2. (D) Box plots of cell mean mEPSC amplitudes pooled from same 3 cultures as shown in (A). CON, N = 23 cells, 13.7 ± 0.9 pA; TTX, N = 24 cells, 16.4 ± 0.9 pA. (E) Box plots of dendrite mean GluA2 receptor cluster size pooled from same 3 cultures as shown in (A) and (B). CON, N = 29 dendrites, 0.90 ± 0.06 μm2; TTX, N = 28 dendrites, 1.08 ± 0.10 μm2. (F) Box plots of dendrite mean GluA2 receptor cluster size pooled from the same 5 cultures shown in (C). CON, N = 49 dendrites, 0.83 ± 0.04 μm2; TTX, N = 48 dendrites, 0.98 ± 0.07 μm2. Line series, Student’s paired T test; box plots, Kruskal-Wallis test. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10 and 90 percentiles; open circles indicate means of individual cells/dendrites; bin size 0.1 pA, 0.05 μm2; line, median; dot, mean. For all p-values, * with underline indicates significance with p < 0.05.

We next pooled data from the multiple cultures to examine effects of activity blockade on GluA2 receptors at the level of cell means. Because the average number of synaptic sites identified in the primary dendrites in three cultures was 20.4, we also took 20 mEPSC samples from each cell, the 21st to the 40th mEPSC recorded, to minimize any differences in variability between GluA2 receptors and mEPSCs that might be due to a difference in sampling. In the data pooled from the 3 matched experiments, neurons from Rab3A+/+ cultures showed a significant increase in mean mEPSC amplitudes following activity blockade (Figure 5D; Table 1). This result indicates that the homeostatic response averaged across the 3 cultures was very similar to, but slightly smaller than, that of the data set presented in Figure 1. The mean for size of GluA2 immunofluorescent receptor clusters showed a trend towards higher values after activity blockade (18.1%) that was of similar magnitude to the mEPSC amplitude increase (19.7%), but did not reach statistical significance for 3 cultures (Figure 5E) or 5 cultures (Figure 5F; Table 1). As expected, mean mEPSC amplitude was not increased following activity blockade in the data pooled from a new set of 3 Rab3A-/- cultures (Table 1). For immunofluorescence results from the same Rab3A-/- cultures, no trend towards higher values was apparent for GluA2 receptor cluster characteristics (Table 1). However, given the variability of the GluA2 response in cultures from Rab3A+/+ mice, it would require a much higher number of Rab3A+/+ and Rab3A-/- cultures to make any firm conclusions about the role of Rab3A in the homeostatic response of GluA2 receptors.

Having used VGLUT1 immunofluorescence to mark synaptic sites, we could also ask whether there was a presynaptic effect of activity blockade on the size, intensity or integral of the VGLUT1 signal at sites apposed to the previously examined GluA2 postsynaptic sites. In the pooled data from cultures prepared from Rab3A+/+ mice, we found no change in the size of the VGLUT1-positive regions, and trends towards reduced intensity and integral (Table 2). These data do not support the idea that activity blockade caused an increase in the amount of VGLUT1 per vesicle, although we cannot rule out that a simultaneous reduction in the number of vesicles obscured the effect, since we did not have an independent label for synaptic vesicles. In cultures prepared from Rab3A-/- mice, the size, intensity and integral of the VGLUT1-positive regions did not significantly change (Table 2).

We have demonstrated here for the first time that the synaptic vesicle protein Rab3A influences the homeostatic regulation of quantal amplitude. What are possible ways that Rab3A could exert its effect? It has previously been shown that exogenous addition of TNFα to hippocampal cultures causes an increase in surface expression of GluA1 receptors (although not GluA2 receptors), and that the homeostatic increase in mEPSC amplitudes is abolished in cultures prepared from the TNFα deletion mouse (Stellwagen et al., 2005; Stellwagen and Malenka, 2006). Furthermore, neurons from TNFα+/+ mice plated on glial feeder layers derived from TNFα-/- mice fail to show the increase in mEPSC amplitude after TTX treatment, indicating that the TNFα inducing the receptor increases following activity blockade comes from the glial cells (Stellwagen and Malenka, 2006). It has recently been shown that the source of TNFα is astrocytes rather than microglia (Heir et al., 2024).

Rab3A has been detected in astrocytes (Maienschein et al., 1999; Hong et al., 2016), so to determine whether Rab3A is acting via regulating TNFα release from astrocytes, we performed experiments similar to those of Stellwagen and Malenka (2006) (schema illustrated in Figure 6, Left). Under our culturing conditions, the glial feeder layers would be expected to be composed mainly of astrocytes (Heir et al., 2024). We compared the effect of activity blockade on mEPSC amplitudes recorded from cortical neurons from Rab3A+/+ mice plated on Rab3A+/+ glial feeder layers (Figure 6A); neurons from Rab3A+/+ mice plated on Rab3A-/- glia (Figure 6B), and neurons from Rab3A-/- mice plated on Rab3A+/+ glia (Figure 6C). If Rab3A is required for TNFα release from glia, then Rab3A+/+ neurons plated on Rab3A-/- glia should not show a homeostatic increase in mEPSC amplitude after treatment with TTX, and any cultures with Rab3A+/+ glia should have a normal homeostatic response. We found the opposite result: mEPSC amplitudes increased in cultures where Rab3A was present in neurons (WT on WT, CON, 13.3 ± 0.5 pA, TTX, 16.7 ± 1.2 pA; WT on KO, CON, 13.3 ± 1.0 pA, TTX, 18.8 ± 1.4 pA), but the increase was greatly diminished in cultures where Rab3A was present only in glia (KO on WT, CON, 15.2 ± 1.1 pA, TTX, 16.9 ± 0.7 pA). In the glial feeder layer culture, the loss of Rab3A in the neurons may have caused an increase in mEPSC amplitude at baseline, similar to the Earlybird mutant in dissociated cultures, but this difference was not significant (CON, WT on WT, 13.3 ± 0.5 pA; CON, KO on WT, 15.2 ± 1.1 pA, p = 0.78, Tukey’s post-hoc test after two-way ANOVA). Regarding mEPSC frequency, in cultures with Rab3A+/+ neurons, frequency trended upward after activity blockade, whereas in the culture with Rab3A-/- neurons, frequency was already increased (although not significantly, CON, WT on WT, 2.54 ± 0.57 sec-1; CON, KO on WT, 4.46 ± 1.21 sec-1, p = 0.59, Tukey’s), and trended downward after activity blockade.

Rab3A in neurons, not astrocytes, was required for full TTX-induced homeostatic plasticity.

(A)-(C), mEPSC amplitude (Middle) and frequency (Right) data from dissociated cortical neurons plated on an astrocyte feeder layer, each prepared separately from the type of mice depicted in the schema (Left): (A) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A+/+ mice. Box plots for mEPSC amplitudes (CON, N = 17 cells, 13.3 ± 0.5 pA; TTX, N = 20 cells, 16.7 ± 1.2 pA) and mEPSC frequency in the same recordings (mean, CON, 2.54 ± 0.57 sec-1; TTX, 3.48 ± 0.64 sec-1); from 4 cultures. (B) Neurons from Rab3A+/+ mice plated on astrocytes from Rab3A-/- mice. Box plots for average mEPSC amplitude (CON, N = 11 cells, 13.3 ± 1.0 pA; TTX, N = 11 cells, 18.8 ± 1.4 pA) and average mEPSC frequency in the same recordings (CON, 2.01 ± 0.41 sec-1; TTX, 4.47 ± 1.53 sec-1); from 2 cultures. (C) Neurons from Rab3A-/- neurons plated on astrocytes from Rab3A+/+ mice. Box plots for average mEPSC amplitude (CON, N = 14 cells, 15.2 ± 1.1 pA; TTX, N = 11 cells, 16.9 ± 0.7 pA) and mEPSC frequency in the same recordings (CON, 4.47 ± 1.21 sec-1; TTX, 3.02 ± 0.70 sec-1); from 3 cultures. Box plot parameters: box ends, 25th and 75th percentiles; whiskers, 10th and 90th percentiles, bin size 0.1 pA; open circles represent means from individual cells; line, median; dot, mean. p values (shown on the graphs) are from Tukey’s post-hoc test following a two-way ANOVA. p-values denoted with * and underline indicates significance of p < 0.05.

In summary, in the absence of Rab3A from glia, the homeostatic plasticity of mEPSC amplitude following activity blockade was normal, whereas in the absence of Rab3A in neurons, homeostatic plasticity was greatly diminished. This result makes it highly unlikely that Rab3A is required for the release of TNFα, or another factor from glia, that induces a homeostatic upregulation of postsynaptic receptors and thereby increases mEPSC amplitude following TTX treatment. Neuronal Rab3A appears to mediate the homeostatic increase in mEPSC amplitude following activity blockade.

Discussion

We found that homeostatic synaptic plasticity of mEPSC amplitude in dissociated mixed cultures of mouse cortical neurons and glia behaved remarkably similar to the mouse NMJ in response to loss of Rab3A function: in cultures from Rab3A-/- mice, the increase in mEPSC amplitudes following prolonged network silencing by TTX was strongly diminished, and in cultures from Rab3AEbd/Ebd mice, basal mEPSC amplitude was increased compared to that of wild-type cultures and was not further modified by TTX treatment. These results suggest that normal function of the synaptic vesicle protein Rab3A is required for the homeostatic increase of mEPSC amplitude in cortical cultures.

Is Rab3A acting in the presynaptic cell to cause homeostatic increase in mEPSC amplitude?

The impetus for our current study was two previous studies in which we examined homeostatic regulation of quantal amplitude at the NMJ. An advantage of studying the NMJ is that synaptic ACh receptors are easily identified with fluorescently labeled α-bungarotoxin, which allows for very accurate quantification of postsynaptic receptor density. We were able to detect a known change due to mixing 2 colors of α-bungarotoxin to within 1% (Wang et al., 2005). Using this model synapse, we showed that there was no increase in synaptic AChRs after TTX treatment, whereas miniature endplate current increased 35% (Wang et al., 2005). We further showed that the presynaptic protein Rab3A was necessary for full upregulation of mEPC amplitude (Wang et al., 2011). These data strongly suggested Rab3A contributed to homeostatic upregulation of quantal amplitude via a presynaptic mechanism. With the current study showing that Rab3A is required for the homeostatic increase in mEPSC amplitude in cortical neuron cultures, one interpretation is that in both situations, Rab3A is required for an increase in the presynaptic quantum.

The presynaptic quantum, or the amount of transmitter released during a single fusion event, is an important contributor to quantal size. The amount of transmitter released during vesicle fusion can be affected by the kinetics of the fusion pore opening (Chang et al., 2017). We previously used amperometric measurements in mouse adrenal chromaffin cells to show loss of Rab3A increased the occurrence of very small amplitude fusion pore feet (Wang et al., 2008). Paired with our finding that there is an increase in the occurrence of very slow-rising, abnormally shaped mEPCs at the neuromuscular junction of Rab3A-/- mice, our data suggest that small synaptic vesicles may have a fusion pore step under certain circumstances (see also (Chiang et al., 2018)). If activity blockade causes an increase in mEPSC amplitude due to a more rapid opening of a fusion pore, or a larger fusion pore conductance, it is possible that Rab3A is required for this modulation.

Another way to increase the presynaptic quantum is to increase levels of the transmitter transporter (Daniels et al., 2004; De Gois et al., 2005; Wilson et al., 2005; Hartman et al., 2006). However, we found that immunofluorescence for the glutamate transporter, VGLUT1, was not increased after activity blockade under our experimental conditions. A similar lack of increase in VGLUT1 levels, relative to synapsin, was previously observed in hippocampal cultures treated with NBQX, an AMPA receptor blocker (see supplemental data, (Wilson et al., 2005)). Taken together with our findings, it appears that there is not strong evidence that the amount of VGLUT1 at synapses is a contributing factor to the increase in mEPSC amplitude in neuronal cultures after activity blockade with TTX.

Finally, the amount of transmitter released by a vesicle could be increased due to larger vesicle size. Larger vesicle diameter has been observed following activity disruption in multiple paradigms, including ocular deprivation, cortical ablation, and immobilization (Kawana et al., 1971; Lloret and Saavedra, 1975; Cheresharov et al., 1978). Loss of Rab3 family members is associated with increased vesicle size (Riedel et al., 2002; Schonn et al., 2010). An increase in volume has been shown to increase miniature endplate junctional current amplitude at the Drosophila NMJ (Karunanithi et al., 2002). Rab3A could be required for an activity-dependent increase in vesicle size, but the increase in diameter needed for a 25% increase in transmitter, and therefore mEPSC amplitude (∼3%), would be difficult to detect.

Is Rab3A acting on postsynaptic receptors to cause increase in mEPSC amplitude?

In neurons, the increase in mEPSC amplitude following block of activity was not due to increases in Ca2+-permeable GluA1 homomers because acute NASPM application was unable to reverse the TTX-induced increase in amplitude. While we found that the effect of TTX treatment on mEPSC amplitude and GluA2 receptor levels were of similar magnitude, the effect on GluA2 receptor levels was not statistically significant for the pooled data sets acquired in the same 3 cultures, nor in a larger data set containing an additional 2 cultures. In light of the extensive literature documenting increases in GluA1 or GluA2 receptors, or both, after activity blockade, the less than convincing increase in GluA2 receptor levels in the current work came as a surprise. The reason for the lack of significance was that the increase in receptors was more variable. We are aware of three possible explanations for greater variability. 1) The number of synaptic sites sampled was smaller for the imaging data set. This was because we limited our analysis to synaptic receptors on the primary dendrite, where a clear increase in sensitivity to exogenously applied glutamate was demonstrated (see Figure 3 in (Turrigiano et al., 1998)). To address this issue, we limited our sample size of mEPSCs in the matched experiments to the same number per cell as the mean sample size per cell for imaging, but there was still a discrepancy in the effect of TTX on mEPSCs vs. receptors. 2) The small sample of only 3 cultures was not enough to reveal the increase in receptors. We attempted to address this by adding 2 more cultures. Although both additional cultures showed small increases in GluA2 receptor cluster size, when cells from these cultures were included in the pooled data set, the p value was not substantially improved. A remaining possibility is: 3) The complexity involved in processing cultures for immunofluorescence introduces variability that obscures the true effect. One can get around technical issues related to immunofluorescence processing by performing Western blots, but then any changes detected could be due to upregulation of extrasynaptic receptors.

What remains unclear is why our data were impacted by the variability inherent in immunofluorescence measurements, but data of other studies were not. Some other studies have used the number of synaptic sites or puncta as the sample size (for example, (Hou et al., 2008; Ibata et al., 2008; Wang et al., 2019)). When we analyzed our data after including as few as 6 randomly selected cluster size measurements per cell, a Kruskal-Wallis test gave highly significant p values for 3 cultures (p = 0.001, n = 174 sites (6 *29 cells)) and 5 cultures (p = 0.005, n = 294 sites (6 *49 cells)). We found only a single study that had comparable sample sizes for receptors and mEPSC amplitudes, and the p value was < 0.01 for mEPSC amplitudes (n = 9 and 10 cells for CON and TTX, respectively), but five times higher (< 0.05) for receptor data (n = 13 and 14 cells for CON and TTX) (Xu and Pozzo-Miller, 2017). This single example is consistent with our suggestion that measurements of changes in receptors after activity blockade are more variable than measurements of mEPSC amplitudes. In sum, the lack of a robust increase in receptor levels leaves open the possibility that there is a presynaptic contributor to quantal size in mouse cortical cultures. However, the variability could arise from technical factors associated with the immunofluorescence method, and the mechanism of Rab3A-dependent homeostatic plasticity could be presynaptic for the NMJ and postsynaptic for cortical neurons.

Increased receptors likely contribute to increases in mEPSC amplitudes, but because we do not have a significant increase in GluA2 receptors in our experiments, it is impossible to conclude that the increase is lacking in cultures from Rab3A-/- neurons. Furthermore, without additional experiments targeting Rab3A deletion to the pre-or post-synaptic compartment, we cannot rule out that Rab3A is regulating homeostatic plasticity in mouse cortical cultures by acting postsynaptically on receptor trafficking. To our knowledge, there is no evidence localizing Rab3A to postsynaptic dendritic sites, although other presynaptic molecules, such as SNARE proteins, synaptotagmins, and NSF, have been identified in the postsynaptic compartment (Lledo et al., 1998; Nishimune et al., 1998; Osten et al., 1998; Song et al., 1998; Araki et al., 2010; Kennedy et al., 2010; Suh et al., 2010; Jurado et al., 2013; Hussain and Davanger, 2015; Gu et al., 2016; Wu et al., 2017). A recent report implicates Rab3A in the localization of plasma membrane proteins, including the EGF receptor, in rafts in HEK cells and Jurkat-T cells, which indicates Rab3A can be found in other specialized locations besides the presynaptic nerve terminal (Diaz-Rohrer et al., 2023).

Neuronal, not glial, Rab3A is required for homeostatic regulation of quantal amplitude

Astrocytic release of TNFα was shown to mediate the increase in mEPSC amplitudes in prolonged TTX-treated cultures of dissociated hippocampal neurons and cultures of hippocampal slices (Stellwagen and Malenka, 2006; Heir et al., 2024). TNF-α accumulates after activity blockade, and directly applied to neuronal cultures, can cause an increase in GluA1 receptors, providing a potential mechanism by which activity blockade leads to the homeostatic upregulation of postsynaptic receptors (Beattie et al., 2002; Stellwagen et al., 2005; Stellwagen and Malenka, 2006). However, we found that loss of Rab3A in glia did not disrupt the increase in mEPSC amplitude after activity blockade, so it is unlikely that the Rab3A-dependent homeostatic increase in mEPSC amplitude is via the astrocytic-TNF-α pathway. Interestingly, we previously showed that the homeostatic increase in NMJ mEPC amplitude was completely normal in the absence of TNF-α (Wang et al., 2011), suggesting the possibility that neuronal Rab3A can act via a non-TNF-α mechanism to contribute to homeostatic regulation of quantal amplitude, although we have not ruled out a neuronal Rab3A-mediated TNF-α pathway in cortical cultures.

Is the lack of homeostatic plasticity in the Earlybird mutant due to occlusion?

We found that mEPSC amplitude was increased in untreated cultures prepared from Rab3AEbd/Ebd mice, compared to mEPSC amplitude in cultures from wild type mice. If there is a limit to how large an mEPSC can become, it is possible that the mutant Rab3A does not affect homeostatic plasticity, only occludes it. Similar increases in mEPSC amplitude at baseline, combined with a failure to increase further after activity blockade, were noted in cultures prepared from mECP2, AKAP, Homer1a, and Arc/Arg3.1 deletion mice (Shepherd et al., 2006; Hu et al., 2010; Xu and Pozzo-Miller, 2017; Sanderson et al., 2018) as well as in cultures treated with ubiquitin proteasome inhibitors (Jakawich et al., 2010) or microRNA 186-5p (Silva et al., 2019), suggesting the possibility of a generalized phenomenon in which increased mEPSC amplitude is induced when homeostatic regulation is disrupted. Since we did not observe such an increase in mEPSC amplitude at baseline in cultures from Rab3A-/- mice, it remains a possibility that the point mutant may bind to novel partners, causing activities that would not be either facilitated or inhibited by Rab3A. Still, it is a strong coincidence that the novel activity of the mutant Earlybird Rab3A would affect mEPSC amplitude, the same characteristic that is modulated by activity blockade in a Rab3A dependent manner. Although we cannot rule out occlusion, an alternate interpretation is that the presence of the mutant protein mimics the condition of activity blockade.

Rab3A is the first presynaptic function protein required for full homeostatic plasticity of mEPSC amplitude

The number of molecules shown to be required for homeostatic synaptic plasticity of quantal size in neuronal cultures may have surpassed ∼20 (see (Koesters et al., 2024) for a recent summary), but fall into two main categories: molecules involved in glutamate receptor expression or trafficking, and signaling molecules. To our knowledge, this is the first evidence that a presynaptic protein is participating in this process. Homeostatic synaptic plasticity is becoming increasingly implicated in essential normal functions such as sleep (Tononi and Cirelli, 2014; Diering et al., 2017; Torrado Pacheco et al., 2021), and pathological conditions such as epilepsy, neuropsychiatric disorders, Huntington’s, and alcohol use disorder (Trasande and Ramirez, 2007; Fernandes and Carvalho, 2016; Wang et al., 2017; Lovinger and Abrahao, 2018; Smith-Dijak et al., 2019; Lignani et al., 2020; Suzuki et al., 2021; Kavalali and Monteggia, 2023). Therefore, our identification of a novel Rab3A-dependent regulatory pathway in homeostatic synaptic plasticity may have important therapeutic implications. Our findings also further cement the importance of Rab3A in activity-dependent modulation of synaptic strength.

Materials and Methods

Animals

To determine the role of Rab3A in homeostatic plasticity in mouse cortical cultures, we employed two distinct genetic mouse strains with altered Rab3A function. Rab3A deletion mice were generated as follows: Rab3A+/- heterozygous mice were bred and genotyped as previously described (Kapfhamer et al., 2002; Wang et al., 2008), and maintained as a colony using multiple heterozygous breeding pairs. Homozygous Rab3A+/+ or Rab3A-/- mouse pups were obtained by breeding Rab3A+/+ pairs and Rab3A-/- pairs, respectively, so that the pups were the first-generation progeny of a homozygous mating. This protocol minimizes the tendency of multiple generations of Rab3A+/+ and Rab3A-/- homozygous breedings to produce progeny that are more and more genetically distinct. Rab3AEbd/Ebdmice were identified in an EU-mutagenesis screen of C57BL/6J mice, and after a cross to C3H/HeJ, were backcrossed for 3 generations to C57BL/6J (Kapfhamer et al., 2002). Rab3A+/Ebd heterozygous mice were bred at Wright State University and genotyped in a two-step procedure: 1. a PCR reaction (forward primer: TGA CTC CTT CAC TCC AGC CT; reverse primer: TGC ACT GCA TTA AAT GAC TCC T) followed by 2. a restriction digest with enzyme Bsp1286I (New England Biolabs) that distinguishes the Earlybird mutant by its different base-pair products. Rab3A+/- mice were backcrossed with Rab3A+/+ mice from the Earlybird heterozygous colony for 11 generations in an attempt to establish a single wild type strain, but differences in mEPSC amplitude and adrenal chromaffin cell calcium currents persisted, likely due to genes that are close to the Rab3A site, resulting in two wild type strains: 1. Rab3A+/+ from the Rab3A+/- colony, and 2. Rab3A+/+ from the Rab3A+/Ebd colony.

Primary Culture of Mouse Cortical Neurons

Primary dissociated cultures of mixed neuronal and glia populations were prepared as previously described (Hanes et al., 2020). Briefly, postnatal day 0-2 (P0-P2) Rab3A+/+, Rab3A-/- or Rab3AEbd/Ebd neonates were euthanized by rapid decapitation, as approved by the Wright State University Institutional Animal Care and Use Committee, and brains were quickly removed. Each culture was prepared from the cortices harvested from two animals; neonates were not sexed. Cortices were collected in chilled Neurobasal-A media (Gibco) with osmolarity adjusted to 270 mOsm and supplemented with 40 U/ml DNAse I (ThermoFisher Scientific). The tissues were digested with papain (Worthington Biochemical) at 20 U/ml at 37°C for 20 minutes followed by trituration with a sterile, fire-polished Pasteur pipette, then filtered through a 100 μm cell strainer, and centrifuged at 1100 rpm for 2 minutes. After discarding the supernatant, the pellet was resuspended in room temperature Neurobasal-A media (270 mOsm), supplemented with 5% fetal bovine serum for glia growth, and 2% B-27 supplement to promote neuronal growth (Gibco), L-glutamine, and gentamicin (ThermoFisher Scientific). Neurons were counted and plated at 0.15 * 106 cells/coverslip onto 12 mm coverslips pre-coated with poly-L-lysine (BioCoat, Corning). This plating density produces a “high density” culture characterized by a complex mesh of neuronal processes criss-crossing the field of view, completely filling the space between cell bodies (see Figure 4). We found in preliminary experiments that an increase in mEPSC amplitude in mouse cortical cultures was inconsistent when cell density was low, likely due to lower levels of baseline activity, although it is a limitation of this study that we did not directly measure activity levels in cultures prepared from Rab3A+/+, Rab3A-/- or Rab3AEbd/Ebd mice, and therefore do not know if differences in results are due to differences in activity levels. While unlikely for Rab3A+/+ and Rab3A-/- studies, given the mEPSC amplitude in untreated cultures was relatively small, it remains possible that in Rab3AEbd/Ebd studies, the larger mEPSC amplitude in untreated cultures was due to a loss of activity in these cultures. The culture media for the first day (0 DIV) was the same as the above Neurobasal-A media supplemented with FBS, B-27, L-glutamine, and gentamicin, and was switched after 24 hours (1 DIV) to media consisting of Neurobasal-A (270 mOsm), 2% B-27, and L-glutamine without FBS to avoid its toxic effects on neuronal viability and health (Stellwagen and Malenka, 2006). Half of the media was changed twice weekly and experiments were performed at 13-14 DIV. Two days prior to experiments, tetrodotoxin (TTX) (500 nM; Tocris), a potent Na+ channel blocker, was added to some cultures to chronically silence all network activity and induce homeostatic synaptic plasticity mechanisms, while untreated sister cultures served as controls. Cultures prepared from mutant mice were compared with cultures from wild-type mice from their respective colonies. Note that the cultures comprising the Rab3A+/+ data here are a subset of the data previously published in Hanes et al., 2020. This smaller data set was restricted to the time period over which cultures were also prepared from Rab3A-/- mice.

Preparation of Glial Feeder Layers

Glial feeder layers were prepared from the cortices of P0-P2 Rab3A+/+ or Rab3A-/- mouse pups as described previously (Stellwagen and Malenka, 2006). Briefly, cortices were dissected and cells were dissociated as described above. Cell suspensions of mixed neuronal and glial populations were plated onto glass coverslips pre-coated with poly-L-lysine in Dulbecco’s Modified Eagle Media (ThermoFisher Scientific) supplemented with 5% FBS (to promote glial proliferation and to kill neurons), L-glutamine, and gentamicin, and maintained in an incubator at 37°C, 5% CO2; cultures were maintained in this manner for up to 1 month to generate purely glial cultures (all neurons typically died off by 7 DIV). Culture media was replaced after 24 hours, and subsequent media changes were made twice weekly, replacing half of the culture media with fresh media. Feeder layers were not used for neuronal seeding until all native neurons were gone and glia approached 100% confluency (visually inspected).

Plating of Neurons on Glial Feeder Layers

Cortical neurons were obtained as described above. The cell pellet obtained was resuspended in Neurobasal-A (osmolarity adjusted to 270 mOsm) containing B27 (2%, to promote neuronal growth), L-glutamine, and 5-fluorodeoxyuridine (FdU, a mitotic inhibitor; Sigma). Addition of FdU was used to prevent glial proliferation and contamination of the feeder layer with new glia, promoting only neuronal growth on the feeder layers (FdU-containing media was used for the maintenance of these cultures and all subsequent media changes). Glial culture media was removed from the feeder layer cultures, and the neuronal cell suspension was plated onto the glial feeder cultures. The culture strategy used to distinguish the relative roles of neuronal and glial Rab3A is outlined in Figure 8, Left. At 1 DIV, all of the culture media was removed and replaced with fresh Neurobasal-A media containing FdU described above, and half of the media was replaced twice per week for all subsequent media changes. Cultures were maintained in a 37°C, 5% CO2 incubator for 13-14 DIV.

Whole-Cell Voltage Clamp to Record mEPSCs

At 13-14 DIV, mEPSCs from TTX-treated and untreated sister cultures of Rab3A+/+ or Rab3A-/- neurons from the Rab3A+/- colony, or Rab3A+/+ or Rab3AEbd/Ebd neurons from the Rab3A+/Ebd colony, were recorded via whole-cell voltage clamp to assess the role of Rab3A in homeostatic synaptic plasticity. Recordings were taken from pyramidal neurons, which were identified visually by a prominent apical dendrite; images were taken of all cells recorded from. Cells were continuously perfused with a solution consisting of (in mM): NaCl (115), KCl (5), CaCl2 (2.5), MgCl2 (1.3), dextrose (23), sucrose (26), HEPES (4.2), pH = 7.2 (Stellwagen and Malenka, 2006). On the day of recording, the osmolarity of the media from the cultures was measured (normally 285 – 295 mOsm) and the perfusate osmolarity was adjusted to match the culture osmolarity to protect against osmotic shock to the neurons. To isolate glutamatergic mEPSCs, TTX (500 nM) and picrotoxin (50 μM), were included in the perfusion solution to block action potentials and GABAergic currents, respectively. The NMDA receptor antagonist, APV, was not included in the perfusion solution because in pilot studies, all mEPSCs were blocked by 10 µM CNQX and 50 μm picrotoxin, demonstrating no APV-sensitive mEPSCs were present (data not shown). Previous studies have described NMDA mEPSCs in neuronal cultures, but recordings were performed in 0 Mg2+ (Watt et al., 2000; Sutton et al., 2006). In our recordings, the presence of extracellular Mg2+ (1.3 mM) and TTX would mean that the majority of NMDA receptors are blocked by extracellular Mg2+ at resting Vm. Patch electrodes (3.5 – 5 MΩ) were filled with an internal solution containing (in mM): K-gluconate (130), NaCl (10), EGTA (1), CaCl2 (0.132), MgCl2 (2), HEPES (10), pH = 7.2. Osmolarity was adjusted to 10 mOsm less than the perfusion solution osmolarity, which we have found improves success rate of achieving low access resistance recordings. Neurons were clamped at a voltage of -60 mV using an Axopatch 200B patch-clamp (Axon Instruments), recorded from for 2 – 5 minutes, and data were collected with Clampex 10.0/10.2 (Axon Instruments).

NASPM application

The antagonist of Ca2+-permeable AMPA receptors, N-naphthyl acetylspermine (NASPM, 20 μM; Tocris), was applied during recordings in a subset of experiments. NASPM is a synthetic analog of Joro Spider Toxin (JSTX) (Koike et al., 1997), and chemicals in this family block Ca2+ permeable glutamate receptors (Blaschke et al., 1993; Herlitze et al., 1993; Iino et al., 1996). The presence of the edited GluA2 subunit disrupts both the Ca2+ permeability and the sensitivity to JSTX and related substances (Blaschke et al., 1993; Bochet et al., 1994; Jonas and Burnashev, 1995). Therefore, 20 μM NASPM is expected to completely block AMPA receptors containing GluA1, GluA3 or GluA4 subunits (Hollmann et al., 1991; Burnashev et al., 1992), or the Kainate receptors containing GluA5 or GluA6 (Koike et al., 1997; Sun et al., 2009), but be ineffective against any heteromer or homomer containing GluA2 subunits. Because the effect of the spider toxins and their analogues are use-dependent (Herlitze et al., 1993; Iino et al., 1996; Koike et al., 1997), NASPM was applied with a depolarizing high K+ solution (25 mM KCl, 95 mM NaCl) which we expected to trigger release of glutamate and opening of glutamate-activated receptors, allowing entry of NASPM into the pore. Baseline recordings were performed for 2 minutes in our standard perfusate, followed by perfusion of NASPM + High K+ solution for 45 seconds, and then perfusion of a NASPM only solution for 5 minutes, after which recording was recommenced for 5 minutes (because we found in pilot experiments that frequency was reduced following NASPM application).

Analysis of mEPSCs

Miniature excitatory postsynaptic currents (mEPSCs) were manually selected using Mini Analysis software (Synaptosoft), a standard program used for mEPSC event detection and analysis. Records were filtered at 2 kHz using a low-pass Butterworth filter prior to selection. The program detection threshold was set at 3 pA but the smallest mEPSC selected was 4.01 pA. A fully manual detection process means that the same criterion (“does this look like an mEPSC?”) is applied to all events, not just the false positives or the false negatives, which prevents the bias from being primarily at one end or the other of the range of mEPSC amplitudes. It is important to note that when performing the MiniAnalysis process, the researcher did not know whether a record was from an untreated cell or a TTX-treated cell.

Immunofluorescence, microscopy, and data analysis

Primary cultures of mouse cortical neurons were grown for 13-14 DIV. Antibodies to GluA2 (mouse ab against N-terminal, EMD Millipore) were added directly to live cultures at 1:40 dilution, and incubated at 37 °C in a CO2 incubator for 45 minutes. Cultures were rinsed 3 times with PBS/5% donkey serum before being fixed with 4% paraformaldehyde. After 3 rinses in PBS/5% donkey serum, cultures were incubated in CY3-labeled donkey-anti-mouse secondary antibodies for 1 hour at room temperature, rinsed in PBS/5% donkey serum, permeabilized with 0.2% saponin, and incubated in chick anti-MAP2 (1:2500, AbCAM) and rabbit anti-VGLUT1 (1:4000, Synaptic Systems) for 1 hour at room temperature in PBS/5% donkey serum. After rinsing with PBS/5% donkey serum, coverslips were incubated with 488-anti chick and CY5-anti rabbit secondary antibodies for 1 hour at room temperature, rinsed, blotted to remove excess liquid, and inverted on a drop of Vectashield (Vector Labs). Coverslips were sealed with nail polish and stored at 4 °C for < 1 week before imaging. All secondary antibodies were from Jackson Immunoresearch and were used at 1:225 dilution.

Coverslips were viewed on a Fluoview FV1000 laser scanning confocal microscope with a 60x oil immersion, 1.35 NA objective. Once a pyramidal neuron was identified and a single confocal section of the cell acquired (see Top, Figure 4), Fluoview 2.1 software was used to zoom in on the primary dendrite (5X) and a series of confocal sections were taken every 0.5 μm. Images were analyzed offline with ImagePro 6 (Cybernetics). The composite image was used to locate synaptic sites containing both VGLUT1 and GluA2 immunoreactivity in close apposition to each other and to the primary dendrite or a secondary branch identified with MAP2 immunoreactivity. An area of interest (AOI) was manually drawn around the GluA2 cluster in the confocal section in which it was the brightest. The AOIs for a dendrite were saved in a single file; the AOI number and the confocal section it was associated with were noted for later retrieval. For quantification, AOIs were loaded, an individual AOI was called up on the appropriate section, and the count/measurement tool used to apply a threshold. Multiple thresholds were used, 400-550, but we found that within a given experiment, the magnitude of the TTX effect was not altered by different thresholds, therefore, all data presented here use a threshold of 450. Pixels within the cluster that were above the threshold were automatically outlined, and size, average intensity, and integral of the outlined region saved to Origin2020 (OriginLab) file for calculating mean values. For analyzing VGLUT1 sites, previous AOIs were examined, and if not directly over the VGLUT1 site, moved to encompass it (this happened when GluA2 and VGLUT1 were side by side rather than overlapping). The threshold was 200 for all VGLUT1 data except Rab3A+/+ Culture #1, where threshold was 400. There was no threshold that worked for both that culture and the other cultures, with too much of the background regions being included for Culture #1 with a threshold of 200, and too little of the synaptic site area being included for the other Cultures with a threshold of 400.

Statistics

In our previous publication (Hanes et al., 2020), we described the characteristics of homeostatic plasticity of mEPSC amplitudes in mouse cortical cultures using multiple mEPSC amplitude quantiles from each cell to create data sets with thousands of samples. We compared the cumulative distribution functions for mEPSC amplitudes in untreated vs. TTX-treated cultures, as well as sorting the values from smallest to largest, and plotted TTX values vs. control values in the rank order plot first described in (Turrigiano et al., 1998) to demonstrate synaptic scaling. Finally, we plotted the ratio of TTX values/control values vs. control values to show that the ratio, which represents the effect of TTX, increased with increasing control values, which we termed “divergent scaling.” Here, we wanted to avoid the inflation of sample size caused by pooling multiple mEPSCs per cell. To statistically compare the effects of TTX in cultures of wild type mice, Rab3A-/- mice and Rab3AEbd/Ebd mutant mice, or, WT-WT, WT-KO and KO-WT neuron-glial cultures, we performed a two-way ANOVA, followed by a Tukey test, and used the p value for each specific situation (for example, CON vs. TTX for Rab3A+/+; CON vs. TTX for Rab3A-/-, and so on. The overall means are presented ± SEM in the text, and p values are displayed above each plot, with N = the number of cells. For the characteristics of synaptic GluA2 receptor clusters and VGLUT1 sites in imaging experiments, with N = the number of dendrites (one dendrite was sampled for each cell), and statistical comparisons were made with the non-parametric Kruskal-Wallis test. Although we quantified three characteristics of the GluA2 clusters and VGLUT1 sites, size, average intensity, and integral, we have not corrected p values for multiple comparisons. Means are presented as box plots + data, with the box at 25 and 75 percentiles, whiskers at 10 and 90 percentiles, a dot indicating the mean, and a line indicating the median. In Figure 3, mEPSC amplitudes were measured within the same cell before and after NASPM application, so a paired t test was used to compare the pre- and post-NASPM mEPSC amplitudes. In Figure 5A-C the mean mEPSC amplitude for untreated and TTX-treated coverslips in the same culture are compared with a paired t test. Statistics were computed and plots were created using Origin2020 (OriginLab).